### 2024-08-15 Test1.Rmd
First substantive commit, I believe.  Initial mortality analysis.

---
title: "Test1"
author: "Jeremiah Groom"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

To find the mean and variance of proportions where the number of values for each proportion are not necessarily equal, you can follow these steps:

### Mean of Proportions

1. **Collect Data**: Suppose you have multiple groups, each with a different number of observations. Let $n_i$ be the number of observations in the $i$-th group, and let $p_i$ be the proportion of successes in the $i$-th group.

2. **Total Observations and Successes**: Calculate the total number of observations $N$ and the total number of successes $S$ across all groups.
   $$
   N = \sum_{i=1}^{k} n_i
   $$
   $$
   S = \sum_{i=1}^{k} (p_i \times n_i)
   $$

3. **Overall Proportion (Mean of Proportions)**: The mean of the proportions is the overall proportion of successes across all groups.
   $$
   \hat{p} = \frac{S}{N} = \frac{\sum_{i=1}^{k} (p_i \times n_i)}{\sum_{i=1}^{k} n_i}
   $$

### Variance of Proportions

1. **Calculate Individual Variances**: For each group, the variance of the proportion $p_i$ is:
   $$
   \text{Var}(p_i) = \frac{p_i (1 - p_i)}{n_i}
   $$

2. **Weighted Variance**: Since the groups have different sizes, you need to weight the variances by the group sizes. The overall variance of the proportions can be calculated as follows:

   $$
   \text{Var}(\hat{p}) = \frac{1}{N} \left( \sum_{i=1}^{k} n_i \times \text{Var}(p_i) \right)
   $$
   where $\text{Var}(p_i) = \frac{p_i (1 - p_i)}{n_i}$.

3. **Plugging in Individual Variances**: Substitute the individual variances into the formula.
   $$
   \text{Var}(\hat{p}) = \frac{1}{N} \left( \sum_{i=1}^{k} n_i \times \frac{p_i (1 - p_i)}{n_i} \right)
   $$
   $$
   \text{Var}(\hat{p}) = \frac{1}{N} \left( \sum_{i=1}^{k} p_i (1 - p_i) \right)
   $$

### Summary

- **Mean of Proportions**: 
  $$
  \hat{p} = \frac{\sum_{i=1}^{k} (p_i \times n_i)}{\sum_{i=1}^{k} n_i}
  $$

- **Variance of Proportions**:
  $$
  \text{Var}(\hat{p}) = \frac{1}{\sum_{i=1}^{k} n_i} \left( \sum_{i=1}^{k} p_i (1 - p_i) \right)
  $$

By following these steps, you can accurately calculate the mean and variance of proportions, even when the number of values for each proportion are not necessarily equal.

### 2024-08-15 Analysis Code/Data_Prep.R

@@ -1,0 +1,176 @@
+
+
+source("Global.R")
+
+
+##### ---- Code ---- #####
+
+# DATA PREP
+
+
+# Climate variables and reduction to a single table
+tmp <- read_csv(paste0(DATA.LOC, "tmp.20.10.csv")) %>% dplyr::select(-INVYR) %>% mutate(delt.temp = post.temp - pre.temp) 
+precip <- read_csv(paste0(DATA.LOC, "precip.20.10.csv")) %>% dplyr::select(-INVYR) %>% mutate(delt.precip = post.precip - pre.precip)
+vpdmin <- read_csv(paste0(DATA.LOC, "vpdmin.20.10.csv")) %>% dplyr::select(-INVYR) %>% mutate(delt.vpdmin = post.vpdmin - pre.vpdmin)
+vpdmax <- read_csv(paste0(DATA.LOC, "vpdmax.20.10.csv")) %>% dplyr::select(-INVYR) %>% mutate(delt.vpdmax = post.vpdmax - pre.vpdmax)
+
+
+# clim.dat = all climate data
+clim.dat <- list(tmp, precip, vpdmin, vpdmax) %>% reduce(left_join, by = c("STATECD", "PLOT_FIADB")) %>% 
+  mutate(State_Plot = paste0(PLOT_FIADB, STATECD),
+         State_Plot = as.numeric(State_Plot)) %>% 
+  dplyr::select(-STATECD, -PLOT_FIADB)
+
+
+
+## Cleaned tree data, removing trees beyond 24 feet (only subplot data)
+Tree1 <- readr::read_csv(unzip(paste0(DATA.LOC, "Cleaned_Trees_2019v2.zip"), "Cleaned_Trees_2019v2.csv")) %>% filter(DIST <= 24, SDN != 3)
+   # SDN = 1 is alive-alive, SDN = 2 = alive-dead.  We are removing all ingrowth.
+
+# Deleting extracted csv file (~125 MB)
+if (file.exists("Cleaned_Trees_2019v2.csv") == TRUE) {
+  file.remove("Cleaned_Trees_2019v2.csv")
+}
+
+
+PlotDat <- orig[, 1:5]
+dat.cent.start <- tibble(var.deltvar = factor(QUANT.LEVELS, level = QUANT.LEVELS))
+
+
+
+## Function to determine the number of trees that died in a plot and their climate quantiles.
[…]

### 2024-08-15 Analysis Code/DeadTree_Analysis.R

@@ -1,0 +1,219 @@
+source("Global.R")
+source(paste0(CODE.LOC, "Functions.R"))
+
+# Obtaining the data to work with: 
+# Opening the zipped RDS file, output from Data_Prep.R
+dieout.dat <- readr::read_rds(unzip(paste0(DATA.LOC, "dieout_dat.zip"), paste0(DATA.LOC,"dieout_dat.rds")))
+
+# Deleting written RDS file (~150 MB)
+if (file.exists(paste0(DATA.LOC, "dieout_dat.rds")) == TRUE) {
+  file.remove(paste0(DATA.LOC, "dieout_dat.rds"))
+}
+
+spp.id <- dieout.dat$spp.id
+spp.list <- dieout.dat$spp.list
+ado_vals <- dieout.dat$ado_vals
+all_vals <- dieout.dat$all_vals
+quant.array <- dieout.dat$quant.array
+quant.matrix <- dieout.dat$quant.matrix
+centroid_array <- dieout.dat$centroid.array
+centroid.array <- dieout.dat$centroid.array
+quant.n <- dieout.dat$quant.n
+quant.lims <- dieout.dat$quant.lims
+
+
+
+
+# Questions about the plot data
+# 1) Are there intensification plots
+#sum(orig$intensification)  # Yes, 6327
+# 2) How many forested plots?
+#length(orig$propfor[orig$propfor > 0])  # 18672 
+# 3) Do our trees exclusively occur in forested plots?
+#orig2 <- orig %>% mutate(State_Plot2 = as.numeric(paste0(PLOT_FIADB, STATECD))) %>%
+#  left_join(clim.dat, by = c("State_Plot2" = "State_Plot"))
+#orig2$trees <- apply(orig2[, 10:59], 1, sum)
+#summary(orig2$propfor[orig2$trees > 0])
+#orig2[orig2$propfor == 0 & orig2$trees > 0,] # 79 plots
+# It is evident that the proportion of forest/non-forest/missed areas of plots is not necessarily constant between the two visits. 
+# 4) We have climate data for 37,612 plots and plot data for 38656 plots.  Are we missing climate data for any plots with our species of interest?
+#dim(clim.dat) ; dim(orig)   # 37612 13 , 38656 59
[…]

### 2024-08-15 Analysis Code/Functions.R

@@ -1,0 +1,92 @@
+
+##### ---- Functions ---- #####
+
+
+## Functions used, organized by script that uses them ##
+
+
+# Find the estimates for mu_hat (the ratio estimator), Y_hat (numerator, thing we're interested in), and 
+#    X_Hat (denominator, density of things like total trees of a species). See Equation 6 and related.
+mean.q.fcn <- function(dat_tot, resp, spp.sel1) {     #dat_tot = data frame of species total for a quantile (denominator). resp = response data frame (numerator) for quantile.
+  
+  Zt <- Yv <- R <- 0 #  rep(0,length(sppname))    # Mean difference in response for a given species: Zt = Mean value for total trees, 
+  #Yv = mean value for/of trees of interest (mean # that died), R = ratio of the two.
+  #for (i in 1:length(sppname)) {                         # i indexes species
+  col.spp <- grep(paste0("nd.", spp.sel1),  colnames(dat_tot))
+  col.name <- paste0("nd.", spp.sel1)
+  
+  Zt_i <- Yv_i <-  rep(0, strat.num)  # Weighted values for each strata
+  for (h in 1:strat.num) {                  # h indexes strata
+    n_h <- length(dat_tot$STRATUM[dat_tot$STRATUM == strat2$STRATUM[h]] )    # Number of plots in stratum h with species of interest
+    Zt_i[h] <- strat2$W_h[h] * sum(get(col.name, dat_tot)[dat_tot$STRATUM == strat2$STRATUM[h]]) / n_h     
+    Yv_i[h] <- strat2$W_h[h] * sum(get(col.name, resp)[resp$STRATUM == strat2$STRATUM[h]]) / n_h     
+  }
+  Zt <- sum(Zt_i) 
+  Yv <- sum(Yv_i) 
+  R <- Yv / Zt
+  
+  means <- list(Zt = Zt, Yv = Yv, R = R)
+  return(means)
+}
+
+# Helper functions for ratio.SE.fcn: 
+# Function that finds variance or covariance for individual strata in the function ratio.SE.fcn
[…]

### 2024-08-15 Global.R

@@ -1,0 +1,53 @@
+## Global conditions for all scripts to use ##
+
+
+##### ---- Libraries ---- #####
+library(tidyverse)
+library(readr)
+library(viridis)
+library(cowplot)
+library(patchwork)
+library(readxl)
+
+
+
+##### ---- Constants ---- #####
+DATA.LOC <- "Data/"
+CODE.LOC <- "Analysis Code/"
+RESULTS.LOC <- "Results/"
+
+### Select spp for analysis
+#SEL.SPP <- "X202" # Douglas fir, 8600 + 
+#SEL.SPP <- "X98" # Sitka spruce
+#SEL.SPP <- "X108" # Lodgepole pine
+#SEL.SPP <- "X312" # Bigleaf maple
+SEL.SPP <- "X127" # Gray pine
+VAR1 <- "pre.vpdmin"
+VAR.DELT <- "delt.vpdmin"
+QUANT.PROBS <- c(0.25, 0.75)
+QUANT.LEVELS <- c("LiHc", "MiHc", "HiHc", "LiMc", "MiMc", "HiMc", "LiLc", "MiLc", "HiLc")
+N.PLOT.LIM <- 10  # Limit to the minimum number of plots for which an estimate will be calculated
+
+
+
+## Stratum info - not sure we need this. 'orig' already has W_h, # of strata = length(unique(W_h))
+strat <- read_csv(paste0(DATA.LOC, "strat_info052120.csv"), show_col_types = FALSE) %>%
+  mutate(W_h = P1POINTCNT/p1pntcnt_eu)             # W_h is the stratum weight
+strat2 <- strat %>% select(STRATUM, P1POINTCNT, W_h)  # Reducing the number of columns to those we need 
+strat.num <- dim(strat2)[1]
+
+
+n_quant <- length(QUANT.LEVELS)
+
+# From the previous analysis (Groom and Monleon 2023)
+orig <- read_csv(paste0(DATA.LOC, "Occ_OriginalVisit2.csv")) %>%
+  dplyr::select(STATECD, PLOT_FIADB, State_Plot, STRATUM, W_h, propfor, propnonfor, propwater, propmiss, intensification, starts_with("X"))
+
+# Number of possible species, indexes, names list
+# Data frame of plots and species considered, and plot weights
+
[…]

### 2025-07-25 Manuscript.qmd
Only push for Est_Redo branch.  Revised flow of estimation so that furrr worked on iterations first.  Allows for manipulation of bootstrap estimates.

---
title: "Tree growth, mortality, and climatic stress in west-coast states, USA"
author:
  - name: "Jeremiah D. Groom"
    affiliations:
      - name: "Groom Analytics LLC"
        address: "1975 SE Crystal Lake Dr., Unit 173"
        city: "Corvallis"
        state: "OR"
        postal-code: "97333"
    orcid: "0000-0002-0110-1036"
    email: "jeremy@groomanalytics.com"
  - name: "Bryce Frank"
    affiliations:
      - name: "USDA Forest Service"
        department: "Pacific Northwest Field Station"
        address: "3625 93rd Ave SW"
        city: "Olympia"
        state: "WA"
        postal-code: "98512"
abstract: |
  Your abstract goes here. This should be a concise summary of your research question, methods, key findings, and implications. Keep it within journal word limits (typically 150-300 words).
keywords: [keyword1, keyword2, keyword3, keyword4]
date: today
format:
  docx:
    reference-doc: template_use2.docx 
    fig-width: 6
    fig-height: 4
    fig-dpi: 300
    toc: false
    number-sections: true
    highlight-style: github
execute:
  echo: false
  warning: false
  message: false
  cache: true
---

```{r setup, include = FALSE}

# Set global chunk options
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 6,
  fig.height = 4,
  dpi = 300
)
```

```{r libraries}

# Load required R libraries
library(tidyverse)    # Data manipulation and visualization
library(here)         # File path management
library(knitr)        # Document generation
library(kableExtra)   # Enhanced table formatting
library(ggplot2)      # Advanced plotting
library(dplyr)        # Data manipulation
library(readr)        # Data import
library(scales)       # Scale functions for visualization
library(patchwork)    # Combining plots
library(flextable)    # Professional tables for Word output

```

```{r constants}

# Define any custom functions
format_p_value <- function(p) {
  case_when(
    p < 0.001 ~ "< 0.001",
    p < 0.01 ~ sprintf("= %.3f", p),
    TRUE ~ sprintf("= %.2f", p)
  )
}
```

```{r load-data}
source("Global.R")

spp.names.use <- spp.names %>% dplyr::select()

```

```{r table-creation}
# State mortality
s.m.1 <-

```

# Introduction

Your introduction should provide background context, review relevant literature, identify gaps in current knowledge, and clearly state your research objectives and hypotheses.

## Background and Rationale

Provide the scientific background and rationale for your study. Cite relevant literature using the format [@author2023] for in-text citations.

## Study Objectives

Clearly state your primary and secondary objectives or research questions.

# Methods

## Study Design

Describe your study design, including the overall approach and methodology.

## Participants

Describe your study population, inclusion/exclusion criteria, and recruitment methods.

## Data Collection

Detail your data collection procedures, instruments used, and any quality control measures.

## Statistical Analysis

Describe your statistical analysis plan, including software used, significance levels, and specific tests performed.

# Results

Describe your main findings here, referencing

```{r statistical-tests}
```

The analysis revealed a significant difference between groups , with a mean difference of  units.

# Discussion

## Main Findings

Summarize your key findings and their interpretation in the context of existing literature.

## Limitations

Acknowledge study limitations and their potential impact on your findings.

## Implications and Future Directions

Discuss the implications of your findings for theory, practice, and future research.

# Conclusion

Provide a concise summary of your main findings and their significance.

# Acknowledgments

Acknowledge funding sources, collaborators, and others who contributed to the work.

# References

::: {#refs}
:::

# Supplementary Material

```{r supplementary-table}
```

```{r supplementary-figure}

```

### 2025-07-25 Analysis Code/Data_Prep_Analysis.R

@@ -3,24 +3,27 @@
 ####### ----------------------------------------------------------#######
 
 
-# The purpose of this code is to allow the user to have control over the desired
-#  analysis pathway.  There are six different analysis pathways, three each 
-#  for tree mortality and basal area growth.  
-#  1) Analysis of four climate variables in 9 climate variable/change variable
-#     quadrants.
-#  2) Analysis by size class. The user selects the DBH breakpoint and the 
-#    TREE file is separated into two files. Each is analyzed for the 9 quadrants.
-#       - The diameter at the first visit is used for filtering
-#  3) Analysis by site class (SITECLCD, 7).  The 9 climate quadrants are each 
-#   subdivided into 7 site classes.  This is a change in plot interpretation,
-#    not TREE file.  There become effectively 63 quadrant conditions.  
+# The purpose of this code is to allow the user to explore the evolving data sets.
+#  There are two analysis branches, one for growth and the other for mortality .
 #
-#   The data preparation is combined with the analysis code because of the 
-#   variety of analysis types.  The data preparation code does not separately
-#   create output files for analysis code to operate upon because the file size
-#   becomes unwieldy. This process will keep the Git repo size substantially
-#   smaller.
+#  Users set many output conditions in Global.R
 #
+#   In step 1 the conditions for the analysis are loaded.  Step 2 involves either
+#   reconstructing the data from FIA SQL tables or loading existing data sets
+#   and avoiding that time commitment.  Climate data and lat/longitude data are
+#   loaded as well. Step 3 prepares the data for analysis and determines data 
+#   conditions.  Step 4 is the analysis of data and generation of result tables.
+#   Results are generated by species across the 3-state range, by state, and by
+#   climate variable values (the initial-visit climate variable value and the 
[…]

### 2025-07-25 Analysis Code/Functions.R

@@ -426,32 +426,29 @@ fire.frac.table.fcn <- function(tablename, tableloc, treedat, parseddat) {
 
 
 
+
+# Determine where to save results (data and plots)
+save.loc.fcn <- function(i){
+  paste0(RESULTS.LOC, "Quantile_Only/", 
+         switch(i, "1" = "Growth", "2" = "Mort"), 
+         "_figs_", CLIM.VAR.USE, "/")
+}
+
+
+
+
+
+
+
+
 ### ------ Mortality and growth estimation -------  ####
 
-#  This function is nested within domain.est.spp.fcn, which is in turn nested within bs.fcn and finally d_mort.grow.fcn.  It is called in DeadTree_Analysis
-# Find the estimates for mu_hat (the ratio estimator), Y_hat (numerator, thing we're interested in), and 
+#  This function is nested within calculate_domain_estimates.fcn, which is in turn nested within generate_bootstrap_array.fcn.  
+#  This function finds the estimates for mu_hat (the ratio estimator), Y_hat (numerator, thing we're interested in), and 
 #    X_Hat (denominator, density of things like total trees of a species). See Equation 6 and related.
-mean.q.fcn <- function(dat_tot, dat_vals, spp.sel1) {     #dat_tot = data frame of species total for a domain (denominator). dat_vals = response data frame (numerator) for domain
-  
-  Zt <- Yv <- R <- 0 #  rep(0,length(sppname))    # Mean difference in response for a given species: Zt = Mean value for total trees, 
-  #Yv = mean value for/of trees of interest (mean # that died), R = ratio of the two.
-  #for (i in 1:length(sppname)) {                         # i indexes species
-  col.spp <- grep(paste0("nd.", spp.sel1), colnames(dat_tot))
+mean.q.fcn <- function(dat_tot, dat_vals, spp.sel1) {
   col.name <- paste0("nd.", spp.sel1)
   
-  # Old slow code.  For understanding the data.table usage.
-  #  Zt_i <- Yv_i <- rep(0, length(strata) )  # Weighted values for each strata
-  #  for (h in 1:strat.num) {                  # h indexes strata
-  #    w_h <- unique(dat_tot$w[dat_tot$stratum == strata[h]])
[…]

### 2025-07-25 Analysis Code/Nplot_Climvar_anaysis.R

@@ -82,6 +82,10 @@ ggplot(data = map.dat.1, aes(x = LON, y = LAT)) +
 cwd.plt.startcwd <- cwd.map.fcn('pre_mean', title.txt = "FIA plot initial CWD values (10-yr mean prior to first visit)", lab.use = "Initial Climatic\nWater Deficit")
 cwd.plt.deltcwd <- cwd.map.fcn('difference', title.txt = "FIA plot 10-year difference in CWD values", lab.use = "Climatic Water\nDeficit Change")
 
+cwd.maps <- plot_grid(cwd.plt.startcwd, cwd.plt.deltcwd, nrow = 1)
+ggsave("CWD_map.png", cwd.maps, device = "png")
+
+
 #Elevation and change in Climatic Water Deficit
 ggplot(map.dat.1, aes(ELEV, difference)) + 
   geom_bin2d(bins = 70) +
@@ -109,4 +113,77 @@ data.use %>% filter(SPCD == spp.x) %>%
 plts.yr <- map(spp.consid, plts.yr.fcn)
 plts.yr.all <- reduce(plts.yr, left_join, by = "INVYR")
 # Yes, aside from 2011 (which is a spill-over year) there is balance.  Roughly the same number of plots are sampled per year.  
+
+
+
+
+
+# Investigating CWD over time
+# Starting with summer_means from Climate_Dat_Preparation
+
+
+summer_means %>% filter(is.na(mean) == FALSE)
+
+ggplot(summer_means %>% filter(puid == "10094_41_57"), aes(year, mean)) + geom_line()
+
+#ggplot(summer_means, aes(year, mean)) + geom_hex() + theme_bw() + geom_smooth(method = "lm")
+
+# Trying all 47k plots with geom_path and alpha transparency
+tic()
+g_alpha <- ggplot(summer_means, aes(year, mean, group = puid)) + geom_path(alpha = 0.01)
+ggsave("g_alpha.png", g_alpha, dev = "png")
+toc()
+
+
+# Let's sample 1k plots...
+samp_puid <- sample(unique(summer_means$puid), 1000, replace = FALSE) 
+s2 <- summer_means %>% filter(puid %in% samp_puid)
+
+g_alpha2 <- ggplot(s2, aes(year, mean, group = puid)) + geom_path(alpha = 0.03) + theme_bw()
+ggsave("g_alpha_1k.png", g_alpha2, dev = "png")
+
+
[…]

### 2025-07-25 Analysis Code/Overall_Mort_Est.R

@@ -1,55 +1,82 @@
-### Code for finding the overall mortality for each species ###
+### Code for finding the state level and overall growth/mortality for each species ###
 ## It doesn't matter what the climate variable is as the deepest level = state.###
 
+for(k in 1:2){ # 1 = growth, 2 = mortality
+  
+  g.m.labels <- c("Growth", "Mortality")
+  
+  
+  if(k == 1) {
+    t.all <- map(as.numeric(gsub("X", "", SEL.SPP)), clim.growth.resp.fcn, clim.var = CLIM.VAR.USE, treedat.sel = treedat.use, clim.dat = climate.use)
+    a.all <- parse.tree.clim.fcn(t.all, clim.var = CLIM.VAR.USE, analysis.type = "grow", resp.dat = "growth.val", tot.dat = "growth.n.trees", selected.spp = SEL.SPP, clim.dat = climate.use)
+  } else { 
+    t.all <- map(as.numeric(gsub("X", "", SEL.SPP)), clim.mort.resp.fcn, clim.var = CLIM.VAR.USE, treedat.sel = treedat.use, clim.dat = climate.use)
+    a.all <- parse.tree.clim.fcn(t.all, clim.var = CLIM.VAR.USE, analysis.type = "mort", resp.dat = "died.out", tot.dat = "all.trees", selected.spp = SEL.SPP, clim.dat = climate.use)
+  }
+  
+  
+  
+  # Now simplifying names
+  vals_dat <- a.all$vals_dat
+  all_dat <- a.all$all_dat
+  domain.array <- a.all$domain.array
+  state.array <- a.all$state.array
+  state.n <- a.all$state.n
+  all.array <- a.all$all.array
+  all.n <- a.all$all.n
+  centroid.array <- a.all$centroid.array
+  domain.n <- a.all$domain.n 
+  quant.lims <- a.all$quant.lims
+  quant.lims.delta <- a.all$quant.lims.delt
+  
+  
+  strata <- unique(vals_dat$stratum)
+  strata.num <- vals_dat %>% dplyr::select(stratum, puid) %>%
+    ungroup() %>% 
+    mutate(row.id = row_number())
+  
+  
+  
+  # Setting up parallel computing
+  plan(multisession, workers = n.cores) 
+  
+  # State-level mortality estimates.
+  state.list <- unique(vals_dat$STATECD) # 6 = California, 41 = Oregon, 53 = Washington
+  state.domain.index <- 1:length(state.list)
+  
+  ## Generating the bootstrap values: 
[…]

### 2025-07-25 Analysis Code/Temp_NewEstCode.R

@@ -1,0 +1,80 @@
+
+
+
+# Finding domain summaries
+domain.summaries <- domain.index %>% 
+  map(\(d) domain.sum.fcn(bootstrap_results, d)) %>%
+  do.call(rbind, .) %>%
+  arrange(Species, Domain)
+
+saveRDS(list(bootstrap_results = bootstrap_results, domain.summaries = domain.summaries), file = paste0(save.loc, "Domain_Analysis_Output.RDS"))
+
+
+
+ # - BREAK: all domain types included and saved above, only 6 quantiles continued below.
+
+
+
+
+
+## -- Figure code 
+
+
+
+# Read the data files
+#dvs_data <- read.csv("DvS.csv")
+#d_lmh_data <- read.csv("D_LMH.csv")
+#s_lmh_data <- read.csv("S_LMH.csv")
+
+
+
+diff.panel.fcn <- function(diff.dat, remove.y, fig.title, lab.right) {
+
+ggplot(diff.dat, aes(y = as.numeric(species_label) + y.offset)) + 
+  geom_vline(xintercept = 0, linewidth = 0.5) + 
+  geom_segment(aes(x = LCI.95, xend = UCI.95, 
+                   y = as.numeric(species_label) + y.offset, 
+                   yend = as.numeric(species_label) + y.offset,
+                   color = Domain), 
+               linewidth = 0.8) + 
+  geom_point(aes(x = Means, y = as.numeric(species_label) + y.offset,
+                 fill = ifelse(significant, "black", "white"),
+                 color = Domain),
+             size = 2, shape = 21, stroke = 0.8) +
+  #scale_y_continuous(breaks = 1:length(levels(DvS2$species_label)),
+  #                    labels = levels(DvS2$species_label)) +
+  scale_y_continuous(breaks = 1:length(levels(diff.dat$species_label)),
+                     labels = levels(diff.dat$species_label)) +
+  scale_color_viridis(discrete = TRUE, name = "Domain", option = "B", begin = 0.2, end = 0.6) +
+  scale_fill_identity() + 
+  labs(x = NULL, y = NULL, title = fig.title) +
+  theme_bw() + 
+  theme(axis.text.y = element_text(size = 10, face = "italic"),
+        legend.position = "inside",
+        legend.position.inside = c(if(lab.right) 0.8 else 0.2, 0.95),
[…]

### 2025-07-25 Global.R

@@ -23,7 +23,7 @@ library(tictoc)  # For development, timing routines.
 # Code controls:
 CALC.FIREPROP <- TRUE   # Set this to TRUE if wanting to calculate or recalculate
                           # the proportion of dead trees killed by fire by quantile.
-RUN.STATES <- FALSE # Should ANALYSIS.PATHWAY 1 find mortality estimates by state?
+RUN.STATES <- TRUE # Should ANALYSIS.PATHWAY 1 find mortality estimates by state?
 
 # File locations:
 DATA.LOC <- "Data/"

### 2025-08-09 Manuscript.Rmd
Changes to RMarkdown and associated files.  Including Manuscript_information.R, which assimilates data and figures for the manuscript.

---
title: "Tree growth, mortality, and climatic stress in west-coast states, USA"
author:
  - name: "Jeremiah D. Groom"
    affiliations:
      - name: "Groom Analytics LLC"
        address: "1975 SE Crystal Lake Dr., Unit 173"
        city: "Corvallis"
        state: "OR"
        postal-code: "97333"
    orcid: "0000-0002-0110-1036"
    email: "jeremy@groomanalytics.com"
  - name: "Bryce Frank"
    affiliations:
      - name: "USDA Forest Service"
        department: "Pacific Northwest Field Station"
        address: "3625 93rd Ave SW"
        city: "Olympia"
        state: "WA"
        postal-code: "98512"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  officedown::rdocx_document:
    reference_docx: "template_use2.docx"

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

library(kableExtra)   # Enhanced table formatting
library(scales)       # Scale functions for visualization
library(flextable)    # Professional tables for Word output
library(officedown)
library(officer)
library(rvg)
#devtools::install_github("adletaw/captioner")
library(captioner) # For figure captions.

```

```{r other_settings}
fp <- fp_par(
  text.align = "center", 
  padding.bottom = 20, padding.top = 120, 
  border.bottom = fp_border())

ft <- fp_text(shading.color='#EFEFEF', bold = TRUE)
```

```{r constants}

source("Global.R")
source(paste0(CODE.LOC, "Functions.R"))

# Define any custom functions
format_p_value <- function(p) {
  case_when(
    p < 0.001 ~ "< 0.001",
    p < 0.01 ~ sprintf("= %.3f", p),
    TRUE ~ sprintf("= %.2f", p)
  )
}

set_flextable_defaults(
  font.family = "Times New Roman",
  font.size = 11,
  theme_fun = "theme_vanilla"
)

centeredP <- fp_par(text.align = "center")

```

```{r load-data}
# Some general analysis info + plots
analysis.stats <- read_rds(paste0(RESULTS.OTHER, "analysis.stats.RDS"))

# Species scientific names
spp.names.use <- analysis.stats$gym.angio.dat

# Growth data for species overall and states
grow.state <- read_csv(paste0(RESULTS1.LOC, "Growth_figs_", CLIM.VAR.USE, "/State_Growth_", CLIM.VAR.USE, ".csv"))
grow.all <- read_csv(paste0(RESULTS1.LOC, "Growth_figs_", CLIM.VAR.USE, "/All_Growth_", CLIM.VAR.USE, ".csv"))

# Growth data for species overall and states
mort.state <- read_csv(paste0(RESULTS1.LOC, "Mort_figs_", CLIM.VAR.USE, "/State_Mortality_", CLIM.VAR.USE, ".csv"))
mort.all <- read_csv(paste0(RESULTS1.LOC, "Mort_figs_", CLIM.VAR.USE, "/All_Mortality_", CLIM.VAR.USE, ".csv"))

```

```{r table-creation}
# State growth
s.g.dat <- state.table.fcn(grow.state, grow.all, 1)

growth.ft <- flextable(s.g.dat) %>%
  fontsize(size = 9, part = "body") %>%
  fontsize(size = 10, part = "header") %>%
  line_spacing(i = NULL, j = NULL, space = 0.5, part = "body") %>% # Control space between lines. 
  theme_zebra(
    odd_header = "#FFFFFF",   # White background for all
    even_header = "#FFFFFF",
    odd_body = "#FFFFFF",     
    even_body = "#FFFFFF"     
  ) %>%
  set_table_properties(layout = "autofit", width = 0)

max.grow.spp <- maxmin.fcn(grow.all, "Means", "LCI.95", "UCI.95", extreme = "max", cm2 = TRUE)
min.grow.spp <- maxmin.fcn(grow.all, "Means", "LCI.95", "UCI.95", extreme = "min", cm2 = TRUE)

# State mortality
s.m.dat <- state.table.fcn(mort.state, mort.all, 2)

mort.ft <- flextable(s.m.dat) %>%
  fontsize(size = 9, part = "body") %>%
  fontsize(size = 10, part = "header") %>%
  line_spacing(i = NULL, j = NULL, space = 0.5, part = "body") %>% # Control space between lines. 
  theme_zebra(
    odd_header = "#FFFFFF",   # White background for all
    even_header = "#FFFFFF",  
    odd_body = "#FFFFFF",     
    even_body = "#FFFFFF"     
  ) %>%
  set_table_properties(layout = "autofit", width = 0)

max.mort.spp <- maxmin.fcn(mort.all, "Means", "LCI.95", "UCI.95", extreme = "max", cm2 = FALSE, digits = 2)
min.mort.spp <- maxmin.fcn(mort.all, "Means", "LCI.95", "UCI.95", extreme = "min", cm2 = FALSE, digits = 2)

```

```{r set_captions}

# 
fig_nums <- captioner::captioner(prefix = "Figure")
tab_nums <- captioner::captioner(prefix = "Table")

f.ref <- function(x) {
 stringr::str_extract(fig_nums(x), "[^:]*")
}

t.ref <- function(x)  {
  stringr::str_extract(tab_nums(x), "[^:]*")
}

tab.tbl_growth <- tab_nums(name = "tbl_growth", 
                        caption = "Decadal growth estimates by state and overall for analysis species. Mean estimates with confidence intervals and plot number are provided. Estimates are not made for states with fewer than 10 plots.")

tab.tbl_mort <- tab_nums(name = "tbl_mort", 
                        caption = "Decadal mortality estimates by state and overall for analysis species. Mean estimates with confidence intervals and plot number are provided. Estimates are not made for states with fewer than 10 plots.")

fig.cwdmap_cap <- fig_nums(name = "fig_cwdmap", 
                        caption = "Fuzzed FIA plot locations displayed with initial visit climatic water deficit (CWD) values (A) and CWD change between visits (C) for the U.S. states, north to south, of Washington, Oregon, and California (see inset map B for the location of these states within the USA).")

fig.domains <- fig_nums(name = "fig_domains",
                        caption = "Decadal (A) growth and (B) mortality domain estimates for *Abies lasiocarpa*.  The six domains are defined as FIA plots that fall [A]bove or [B]elow the CWD change threshold with [L]ow, [M]edium, or [H]igh initial CWD values (see text for details). E.g., 'AL' is Above CWD change threshold, Low initial CWD value. The distribution of plot CWD change and initial values (C) is presented as well, with the blue horizontal line at a change in CWD of 5 mm representing the change threshold and the green vertical lines representing the quantile boundaries between Low, Medium, and High initial CWD values." )

fig.diff_grow <- fig_nums(name = "fig_diffgrow", 
                          caption = paste0("Differences in mean growth rates (cm\u00B2/decade) between domains by species.  Domain abbreviations are described in the legend for ",
                                           f.ref("fig_domains"), 
                                           ". See each column legend for the key of domain differences.  Mean domain differences (circles) with 95% CIs (bars) that do not intersect zero are filled."))

fig.diff_mort <- fig_nums(name = "fig_diffmort", 
                          caption = paste0("Differences in mean decadal mortality rates between domains by species.  Domain abbreviations are described in the legend for ",
                                           f.ref("fig_domains"), 
                                           ". See each column legend for the key of domain differences.  Mean domain differences (circles) with 95% CIs (bars) that do not intersect zero are filled."))

```

# Abstract

# Introduction

# Methods

# Results

Our study tracked the fate of `r comma(analysis.stats$tree.num)` trees from `r length(sel.spp)` in `r comma(analysis.stats$plot.num)` plots that were visited twice by FIA crews.  Of the `r length(sel.spp)` species, `r analysis.stats$n.gym.spp` are gymnosperms.  Overall growth rates for species varied between `r min.grow.spp$mean` (*`r min.grow.spp$spp`*) and `r max.grow.spp$mean` (*`r max.grow.spp$spp`*, `r t.ref("tbl_growth")`) cm$^2$/decade.  Decadal mean mortality rates varied between `r min.mort.spp$mean` (*`r min.mort.spp$spp`*) and `r max.mort.spp$mean` (*`r max.mort.spp$spp`*, `r t.ref("tbl_mort")`). Growth and mortality rates differed among states (Tables `r tab_nums("tbl_growth", display = "num")` and `r tab_nums("tbl_mort", display = "num")`).

Forested plot 10-year differences in CWD do not visually coincide with initial CWD values (`r f.ref("fig_cwdmap")`) although most plots increased in their CWD values to some degree; `r analysis.stats$cwd.lt0.pct`% of plots decreased in mean CWD values between visits.  Individual species quantiles for CWD / delta CWD. DO I want to include some sort of quantile-thing for the range of delta-CWD for species?  Then cite the dashboard.

# Discussion

The initial CWD values coincide fairly well with mean annual precipitation [NOAA Climate.gov](https://www.climate.gov/news-features/featured-images/new-maps-annual-average-temperature-and-precipitation-us-climate).

# Conclusion

# Acknowledgments

# References

::: {#refs}
:::

<!---BLOCK_LANDSCAPE_START--->

```{r tbl-growth, tab.cap=tab.tbl_growth}

#tab_num <- run_autonum(seq_id = "Table", pre_label = "Table ", post_label = ": ", bkm = "tbl-growth")
#block_caption(label = "Decadal growth estimates by state and overall for analysis species. Mean estimates with confidence intervals and plot number are provided. Estimates are not made for states with fewer than 10 plots.", style = "caption", autonum = tab_num)

growth.ft

```

```{r tbl-mort, tab.cap = tab.tbl_mort}
#tab_num <- run_autonum(seq_id = "Table", pre_label = "Table ", post_label = ": ", bkm = "tbl-mortality")
#block_caption(label = "Decadal mortality estimates by state and overall for analysis species. Mean estimates with confidence intervals and plot number are provided. Estimates are not made for states with fewer than 10 #plots.", style = "caption", autonum = tab_num)

mort.ft

```

<!---BLOCK_LANDSCAPE_STOP--->

```{r mapfig, fig.height = 7,  fig.dpi=300, fig.cap=fig.cwdmap_cap, fig.cap.pre="", fig.cap.sep=""}

# Output from Manuscript_information.R
knitr::include_graphics(paste0(RESULTS.OTHER, "cwd_maps_figure.png"))

```

```{r domainsfig, fig.width = 6,  fig.dpi=300, fig.cap=fig.domains, fig.cap.pre="", fig.cap.sep=""}

# Output from Manuscript_information.R
knitr::include_graphics(paste0(RESULTS.OTHER, "Example_spp_figure.png"))

```

```{r diff_grow_fig, fig.height = 6, width = 6, fig.dpi=300, fig.cap=fig.diff_grow, fig.cap.pre="", fig.cap.sep=""}

# Output from Manuscript_information.R
knitr::include_graphics(paste0(RESULTS1.LOC, "/Growth_figs_cwd/Growth_Panel_Plot.png"))

```

```{r diff_mort_fig, fig.height = 6, width = 6, fig.dpi=300, fig.cap=fig.diff_mort, fig.cap.pre="", fig.cap.sep=""}

# Output from Manuscript_information.R
knitr::include_graphics(paste0(RESULTS1.LOC, "/Mort_figs_cwd/Mort_Panel_Plot.png"))

```

### 2025-08-09 Analysis Code/Data_Prep_Analysis.R

@@ -31,7 +31,12 @@
 source("Global.R")
 source(paste0(CODE.LOC, "Functions.R"))
 
-
+library(furrr)
+library(parallel)
+library(RSQLite) # For obtaining SQLite FIA databases
+library(tictoc)  # For development, timing routines.
+library(data.table)
+library(abind) # for combining matrices into arrays
 #### 2) Base data prep --------------------------------------------------------
 # These are the base data sets that will be transformed by the analysis types.
 
@@ -119,7 +124,7 @@ PlotDat <- left_join(PlotDat, plotN3, by = c("stratum"))
 # Now just checking to see how Wh and n_h/N compare.
 #  plt.use <- PlotDat %>% group_by(stratum, w) %>%
 #  summarize(n.plts = n()) %>%
-#  left_join(plotN3, by = "stratum") %>%
+#  left_join(plotN3, by = "Stratum") %>%
 #  mutate(w.plts = n_h.plts / tot.n) 
 
 #  It looks like the fit between Wh and n_h/N is pretty tightly linear.  
@@ -135,10 +140,10 @@ PlotDat <- left_join(PlotDat, plotN3, by = c("stratum"))
 ## First the data are summarized by species and climate variable (clim.mort.resp.fcn, 
 #   clim.growth.resp.fcn).  Then the data are combined into arrays for 
 #    analysis (parse.tree.clim.fcn).
-tree.mort.dat <- map(sel.spp, clim.mort.resp.fcn, clim.var = CLIM.VAR.USE, treedat.sel = treedat.use, clim.dat = climate.use) 
+tree.mort.dat <- purrr::map(sel.spp, clim.mort.resp.fcn, clim.var = CLIM.VAR.USE, treedat.sel = treedat.use, clim.dat = climate.use) 
 
 # Create lists of tree growth and number of trees that grew in each plot
-tree.grow.dat <- map(sel.spp, clim.growth.resp.fcn, clim.var = CLIM.VAR.USE, treedat.sel = treedat.use, clim.dat = climate.use) 
+tree.grow.dat <- purrr::map(sel.spp, clim.growth.resp.fcn, clim.var = CLIM.VAR.USE, treedat.sel = treedat.use, clim.dat = climate.use) 
 
 
[…]

### 2025-08-09 Analysis Code/Functions.R

@@ -8,10 +8,11 @@
 domain.divide.fcn <- function(domain.level, spcd2, tree.plots2, PlotDat) {
   domain.name <- paste0(domain.level, ".",  spcd2)
   domain.died.out <- left_join(PlotDat, tree.plots2 %>% 
-                                dplyr::select(puid, var.deltvar) %>% # var.deltvar defined in clim.tree.resp.fcn, "LiHc" etc.
-                                filter(var.deltvar == domain.level), by = "puid") %>%
+                                 dplyr::select(puid, var.deltvar) %>% # var.deltvar defined in clim.tree.resp.fcn, "LiHc" etc.
+                                 filter(var.deltvar == domain.level), by = "puid") %>%
     mutate(var.deltvar = ifelse(is.na(var.deltvar), 0, 1)) %>%
     rename(!!domain.name := var.deltvar) 
+  return(domain.died.out)
 }
 
 
@@ -58,7 +59,7 @@ clim.mort.resp.fcn <- function(spcd, clim.var, treedat.sel, clim.dat) {
         mutate(., var.quants = ifelse(!!sym(var1) > quant.lims[2], 1, ifelse(!!sym(var1) < quant.lims[1], -1, 0)),
                deltvar.quants = ifelse(!!sym(var.delt) > quant.lims.delta, 1, 0),
                vq = case_match(var.quants, -1 ~ "L", 0 ~ "M", 1 ~ "H"),
-               dvq = case_match(deltvar.quants, 0 ~ "S", 1 ~ "D"),
+               dvq = case_match(deltvar.quants, 0 ~ "B", 1 ~ "A"),
                var.deltvar = factor(paste0(dvq, vq), levels = DOMAIN.LEVELS))
       } else {
         values$errorMessage <- "Error: Expecting 6 or 9 estimation domains. Check Global.R:DOMAIN.LEVELS and Functions.R:clim.mort.resp.fcn."
@@ -91,12 +92,14 @@ clim.mort.resp.fcn <- function(spcd, clim.var, treedat.sel, clim.dat) {
   Tree3 <- left_join(Tree2, tree.plots2, by = "puid") %>%
     group_by(puid) %>% #, var.deltvar, get(var1), get(var.delt)) %>%  # This will assign the names `get(var1)` and `get(var.delt)`, which are cleaned up below
     reframe(n.trees = sum(ntree.rep),
[…]

### 2025-08-09 Analysis Code/Manuscript_information.R

@@ -1,0 +1,303 @@
+## This code provides inforamtion summaries for the manuscript and generates 
+#   non-estimate figures and (maybe) tables.
+#   The file Data_Prep_Analysis.R sources this code.  
+#   The file Manuscript.Rmd will open and use its outputs.
+
+
+### Q: How many trees did we track? How many plots?
+tree.num <- dim(tree.plt.data %>% filter(SPCD %in% sel.spp))[1]
+plot.num <- dim(PlotDat)[1]
+
+## Number of Gymnosperms?  Which ones?
+gym.angio.dat <- spp.names %>% filter(SPCD %in% sel.spp) %>%
+  mutate(sci_name = paste(GENUS, SPECIES),
+         SpeciesCode = paste0("X", SPCD)) %>%
+  dplyr::select(SpeciesCode, SPCD, COMMON_NAME, SPECIES_SYMBOL, SFTWD_HRDWD, sci_name)
+n.gym.spp <- length(gym.angio.dat$SPCD[gym.angio.dat$SFTWD_HRDWD == "S"])
+
+
+### Q: What percentage of change in CWD is negative & positive?
+# Quantiles of data below different thresholds
+n.cwd <- nrow(climate.use)
+n.zero <- nrow(climate.use %>% filter(difference < 0))
+n.min.3 <- nrow(climate.use %>% filter(difference < -3))
+
+cwd.lt.0.pct <- round(n.zero/n.cwd, 3) * 100 #Percent CWD diff < 0  =  8%
+cwd.lt.min3.pct <- n.min.3/n.cwd #Percent CWD diff < -3  = 0.6% (This is useful for the mapping)
+
+
+
+analysis.stats <- list(tree.num = tree.num,
+                       plot.num = plot.num,
+                       cwd.lt0.pct = cwd.lt.0.pct,
+                       gym.angio.dat = gym.angio.dat,
+                       n.gym.spp = n.gym.spp) 
+
+write_rds(analysis.stats, paste0(RESULTS.OTHER, "analysis.stats.RDS"))
+
+
+
+
+##### Figure: Example growth, mortality and CWD scatterplot -----------------------
+domain.figs <- list()
+for(k in 1:2){
+  plt.dat <- readRDS(paste0(save.loc.fcn(k), "Domain_Analysis_Output.RDS"))
+  plt.dat2 <- plt.dat$domain.summaries
+  
+  sppnum.to.plot <- "X81"
[…]

### 2025-08-09 Analysis Code/Nplot_Climvar_anaysis.R

@@ -3,7 +3,8 @@
 # --> Run Data_Prep_Analysis.R up through data.use (around line 91)
 
 ### -- Number of plots per species -- ###
-
+## This bit of code is for determining which species to investigate based on plot number.  
+#  In this case, 400 plots is used as a breakpoint.
 n.spp.plts <- data.use %>% select(SPCD, puid) %>%
   distinct() %>%
   group_by(SPCD) %>%

### 2025-08-09 Analysis Code/Overall_Mort_Est.R

@@ -56,7 +56,7 @@ for(k in 1:2){ # 1 = growth, 2 = mortality
   
   # Finding and saving domain summaries
   state.domain.summaries <- state.domain.index %>% 
-    map(\(d) domain.sum.fcn(state.bootstrap_results, d)) %>%
+    map(\(d) domain.sum.fcn(state.bootstrap_results, d, domain_n =state.n)) %>%
     do.call(rbind, .) %>%
     mutate(State = case_match(Domain, 1 ~ "CA", 2 ~ "WA", 3 ~ "OR")) %>%
     arrange(Species, Domain)
@@ -74,7 +74,7 @@ for(k in 1:2){ # 1 = growth, 2 = mortality
     n_iter = BS.N
   )
   
-  all.species.summaries <- domain.sum.fcn(all.bootstrap_results, 1)
+  all.species.summaries <- domain.sum.fcn(all.bootstrap_results, 1, domain_n =all.n)
   
   # Writing values for species across states to another file
   write_csv(all.species.summaries, file = paste0(save.loc.fcn(k), "All_", g.m.labels[k], "_", CLIM.VAR.USE, ".csv"))

### 2025-08-09 Analysis Code/Temp_NewEstCode.R

@@ -1,80 +1,0 @@
-
-
-
-# Finding domain summaries
-domain.summaries <- domain.index %>% 
-  map(\(d) domain.sum.fcn(bootstrap_results, d)) %>%
-  do.call(rbind, .) %>%
-  arrange(Species, Domain)
-
-saveRDS(list(bootstrap_results = bootstrap_results, domain.summaries = domain.summaries), file = paste0(save.loc, "Domain_Analysis_Output.RDS"))
-
-
-
- # - BREAK: all domain types included and saved above, only 6 quantiles continued below.
-
-
-
-
-
-## -- Figure code 
-
-
-
-# Read the data files
-#dvs_data <- read.csv("DvS.csv")
-#d_lmh_data <- read.csv("D_LMH.csv")
-#s_lmh_data <- read.csv("S_LMH.csv")
-
-
-
-diff.panel.fcn <- function(diff.dat, remove.y, fig.title, lab.right) {
-
-ggplot(diff.dat, aes(y = as.numeric(species_label) + y.offset)) + 
-  geom_vline(xintercept = 0, linewidth = 0.5) + 
-  geom_segment(aes(x = LCI.95, xend = UCI.95, 
-                   y = as.numeric(species_label) + y.offset, 
-                   yend = as.numeric(species_label) + y.offset,
-                   color = Domain), 
-               linewidth = 0.8) + 
-  geom_point(aes(x = Means, y = as.numeric(species_label) + y.offset,
-                 fill = ifelse(significant, "black", "white"),
-                 color = Domain),
-             size = 2, shape = 21, stroke = 0.8) +
-  #scale_y_continuous(breaks = 1:length(levels(DvS2$species_label)),
-  #                    labels = levels(DvS2$species_label)) +
-  scale_y_continuous(breaks = 1:length(levels(diff.dat$species_label)),
-                     labels = levels(diff.dat$species_label)) +
-  scale_color_viridis(discrete = TRUE, name = "Domain", option = "B", begin = 0.2, end = 0.6) +
-  scale_fill_identity() + 
-  labs(x = NULL, y = NULL, title = fig.title) +
-  theme_bw() + 
-  theme(axis.text.y = element_text(size = 10, face = "italic"),
-        legend.position = "inside",
-        legend.position.inside = c(if(lab.right) 0.8 else 0.2, 0.95),
[…]

### 2025-08-09 Global.R

@@ -2,19 +2,19 @@
 
 
 ##### ---- Libraries ---- #####
-library(tidyverse)
 library(readr)
 library(viridis)
 library(cowplot)
 library(patchwork)
 library(readxl)
-library(data.table)
-library(abind) # for combining matrices into arrays
-
+library(tidyverse)
+library(here)         # File path management
 library(furrr)
 library(parallel)
-library(RSQLite) # For obtaining SQLite FIA databases
-library(tictoc)  # For development, timing routines.
+# Generation of maps
+library(sf)
+library(ggspatial)
+library(maps)
 
 
 
@@ -30,8 +30,7 @@ DATA.LOC <- "Data/"
 CODE.LOC <- "Analysis Code/"
 RESULTS.LOC <- "Results/"
 RESULTS1.LOC <- "Results/Quantile_Only/"  # For ANALYSIS.PATHWAY 1
-RESULTS2.LOC <- "Results/Size_Class/"  # For ANALYSIS.PATHWAY 2
-RESULTS3.LOC <- "Results/Site_Class/"  # For ANALYSIS.PATHWAY 3
+RESULTS.OTHER <- "Results/Other_Results/"  # For summaries, maps, etc.
 
 
 
@@ -68,7 +67,7 @@ INIT.CLIM.BREAKS <- c(75, 125) # if USE.QUANT.PROBS are FALSE, these are the dom
 if(THREE.CHANGE.CATEGORIES) {
 DOMAIN.LEVELS <- c("LiHc", "MiHc", "HiHc", "LiMc", "MiMc", "HiMc", "LiLc", "MiLc", "HiLc")
 } else {
-DOMAIN.LEVELS <- c("DL", "DM", "DH", "SL", "SM", "SH")
+DOMAIN.LEVELS <- c("AL", "AM", "AH", "BL", "BM", "BH") # Above/Below threshold and Low/Medium/High
 }
 N.PLOT.LIM <- 10  # Limit to the minimum number of plots for which an estimate will be calculated
 BS.N <- 200    # Bootstrap iteration number
@@ -97,7 +96,8 @@ orig <- read_csv(paste0(DATA.LOC, "Occ_OriginalVisit2.csv")) %>%
 spp.list <- as.numeric(gsub("X", "", names(orig)[(ncol(orig) - (length(grep("X", names(orig))) - 1)):ncol(orig)]))
[…]

### 2025-08-21 Manuscript.Rmd
Added Supplemental Material.Rmd which includes tables summarizing CWD stats for all tree species. Debugged many R code files, added Methods to Manuscript.Rmd

@@
 
+# Percent of plots sampled with REMPER = 9, 10, and 11 years (rounded)
+pct.remeas <- as.vector(analysis.stats$pct.remeas)
+
 # Species scientific names
@@
 
-# Growth data for species overall and states
+# Growth data for species overall, states, differences
 grow.state <- read_csv(paste0(RESULTS1.LOC, "Growth_figs_", CLIM.VAR.USE, "/State_Growth_", CLIM.VAR.USE, ".csv"))
 grow.all <- read_csv(paste0(RESULTS1.LOC, "Growth_figs_", CLIM.VAR.USE, "/All_Growth_", CLIM.VAR.USE, ".csv"))
+grow.diff <- read_rds(paste0(save.loc.fcn(1), "Processed_6Domain_Data.RDS"))
 
-# Growth data for species overall and states
+# Mortality data for species overall and states
 mort.state <- read_csv(paste0(RESULTS1.LOC, "Mort_figs_", CLIM.VAR.USE, "/State_Mortality_", CLIM.VAR.USE, ".csv"))
 mort.all <- read_csv(paste0(RESULTS1.LOC, "Mort_figs_", CLIM.VAR.USE, "/All_Mortality_", CLIM.VAR.USE, ".csv"))
+mort.diff <- read_rds(paste0(save.loc.fcn(2), "Processed_6Domain_Data.RDS"))
+
+# Significant findings, Growth
+grow.n <- nrow(grow.diff$AvB2)
+
+grow.AvB.sig <- length(grow.diff$AvB2$significant[grow.diff$AvB2$significant == TRUE])
+grow.AvB.signeg <- length(grow.diff$AvB2$significant[grow.diff$AvB2$significant == TRUE & grow.diff$AvB2$Means < 0])
+grow.A.sig <- length(grow.diff$A_LMH2$significant[grow.diff$A_LMH2$significant == TRUE])
+grow.A.signeg <- length(grow.diff$A_LMH2$significant[grow.diff$A_LMH2$significant == TRUE & grow.diff$A_LMH2$Means < 0])
+grow.B.sig <- length(grow.diff$B_LMH2$significant[grow.diff$B_LMH2$significant == TRUE])
+grow.B.signeg <- length(grow.diff$B_LMH2$significant[grow.diff$B_LMH2$significant == TRUE & grow.diff$B_LMH2$Means < 0])
+
+# Significant findings, Mort
+mort.AvB.sig <- length(mort.diff$AvB2$significant[mort.diff$AvB2$significant == TRUE])
+mort.AvB.signeg <- length(mort.diff$AvB2$significant[mort.diff$AvB2$significant == TRUE & mort.diff$AvB2$Means < 0])
+mort.A.sig <- length(mort.diff$A_LMH2$significant[mort.diff$A_LMH2$significant == TRUE])
+mort.A.signeg <- length(mort.diff$A_LMH2$significant[mort.diff$A_LMH2$significant == TRUE & mort.diff$A_LMH2$Means < 0])
+mort.B.sig <- length(mort.diff$B_LMH2$significant[mort.diff$B_LMH2$significant == TRUE])
+mort.B.signeg <- length(mort.diff$B_LMH2$significant[mort.diff$B_LMH2$significant == TRUE & mort.diff$B_LMH2$Means < 0])
 
@@
   fontsize(size = 10, part = "header") %>%
+    italic(j = 1, part = "body") %>%
   line_spacing(i = NULL, j = NULL, space = 0.5, part = "body") %>% # Control space between lines. 
@@
   fontsize(size = 10, part = "header") %>%
+  italic(j = 1, part = "body") %>%
   line_spacing(i = NULL, j = NULL, space = 0.5, part = "body") %>% # Control space between lines. 
@@
 min.mort.spp <- maxmin.fcn(mort.all, "Means", "LCI.95", "UCI.95", extreme = "min", cm2 = FALSE, digits = 2)
+
+# CWD amounts
+min.mean.init <- analysis.stats$init.cwd.stats$Mean[analysis.stats$init.cwd.min.mean]
+min.mean.init.spp <- analysis.stats$init.cwd.stats$Species[analysis.stats$init.cwd.min.mean]
+max.mean.init <- analysis.stats$init.cwd.stats$Mean[analysis.stats$init.cwd.max.mean]
+max.mean.init.spp <- analysis.stats$init.cwd.stats$Species[analysis.stats$init.cwd.max.mean]
+
+max.q95.diff <- analysis.stats$diff.cwd.stats$q95[analysis.stats$diff.cwd.95max]
+max.q95.diff.spp <- analysis.stats$diff.cwd.stats$Species[analysis.stats$diff.cwd.95max]
+max.range.diff <- analysis.stats$diff.cwd.90range.max
+max.q95.diff.spp <- analysis.stats$diff.cwd.stats$Species[analysis.stats$diff.cwd.95max]
 
@@
                                            f.ref("fig_domains"), 
-                                           ". See each column legend for the key of domain differences.  Mean domain differences (circles) with 95% CIs (bars) that do not intersect zero are filled."))
+                                           ".
+See each column legend for the key of domain differences.
+Mean domain differences (circles) with 95% CIs (bars) that do not intersect zero are filled.
+Species above the dashed line are coniferous; those below are deciduous."))
 
@@
 
+## FIA data
+
+We used U.S.
+National Forest Inventory datasets for our analyses of forestland in California, Oregon, and Washington, USA.
+Forestland is defined as an area of 4,050 m$^2$ that is at least 10 percent stocked or potentially stocked with tree species, excluding urban and agricultural land uses (Reams et al., 2005).
+Additionally, as our study relied on plot revisit data, we only examined plots that were considered forestland during both visits.
+
+FIA plots consist of four points where all trees with stem diameters greater than or equal to 12.7 cm (for most species measured at 1.37 m above the ground) were tallied in a 7.32 m radius circular subplot (total area 672.5 m$^2$) with additional trees sampled in an annulus around each subplot (macroplot) around each subplot, with a total radius of 18 m from the center point and a total area of 1012 m$^2$.
+We did not include data from FIA microplots.
+Bechtold and Scott (2005) provide plot design details and measurement protocols.
+
+Since 2000, FIA plots have been selected via a spatially balanced probability sample (McRoberts 2005, Reams et al., 2005) with a target plot density of one ground plot every 24 km$^2$.
+Some regions and ownerships intensify the plot density; we excluded these additional plots to simplify stratification estimation, relying on [601] FIA-defined strata and using the density of FIA Phase 1 sampling points within and across strata to determine strata weights.
+
+Plots in the three states are sampled every 10 years with a target of 1/10 of plots sampled in a panel every year (Thompson 2015).
+Our analysis depends on examining the fate and growth of trees from two sequential visits (referred to as the first and second plot visit), ideally from a complete sample panel remeasure.
+We obtained our FIA data from the USDA FIA DataMart web portal (https://research.fs.usda.gov/products/dataandtools/fia-datamart; downloaded May 5, 2025).
+They included plot data sampled from 2001 to 2023.
+Our total sample size was [12500] forested plots (Figure 1).
+
+Individual trees within plots were retained in the analysis if they were alive at the initial visit and met the stem diameter criterion during that visit (i.e., plot ingrowth was not included).
+The growth analysis additionally restricted trees considered to those that were alive at both visits.
+The mortality analysis excluded trees killed through silvicultural practices.
+Both analyses excluded trees that shrank in diameter, physically moved due to landslides or other event, were erroneously tallied at either visit, were missed due to data collection procedural changes, or existed at any time in a nonsampled area.
+We used only FIA plots sampled specifically for tree mortality and growth.
+
+## Climate data
+
+We obtained plot-level CWD data from ClimateNA (Wang et al.
+2016) for June, July, and August (summer months) between 1980 and 2023 (Bryce: I just averaged the CWD values for these months.
+It wasn’t available seasonally, was it?
+Also, can you provide more info on how CWD is calculated?
+I know it is PET-AET, but I’m not sure how to interpret a monthly/seasonal value).
+Latitude and longitude were based on FIA-fuzzed coordinates (citation?
+Bryce – should we tighten this up by running the analysis on non-fuzzed coordinates?).
+We found the 20-year mean of summer CWD values for each plot prior to and including the first visit.
+We refer this this value as the “initial CWD value”.
+We then found the mean of summer CWD values between visits and including the second visit year.
+The difference between this second value and initial CWD value is the “CWD change value” (maybe ‘difference’ really is a better, less loaded term).
+Both the initial CWD and CWD change values are mapped in Figure 1.
+
+## Analysis
+Estimation approach: Following Section 4.3.4 in Scott et al.
+(2005) we obtained point estimates of species growth and mortality using a ratio estimators (Eqn 4.16).
+For mortality, the numerator is the sum of stratum-weighted means of trees per plot of a given species that died in plots.
+The number of sampled trees in each plot is expanded by the area sampled (subplot or macroplot area) adjusted by the proportion of the plot area available for sampling and the number of years between visits.
+The denominator is similar but is calculated as the sum of stratum-weighted means of all trees per plot of a given species.
+
+The numerator of species-specific growth estimates is the sum of stratum-weighted means of the increase in basal area (cm$^2$) per plot while the denominator is the same as was used for the mortality estimate.
+This produces a point estimate for a species’ mean growth.
+Mortality and growth values were standardized to a decadal period.
+
+The analysis finds species growth and mortality estimates for six domains based on CWD values (Figure 2C).
+The initial CWD values are divided into three quantiles: values below the first quartile (Low), between the first and third quartiles (Medium), and above the third quartile (High).
+These initial CWD value divisions allowed for species-specific assessments.
+For all species we defined an arbitrary boundary of 5 mm to define plots with a greater change in CWD (Above [the boundary]) and those with less CWD change (Below).
+This was done because species did not differ greatly in CWD change (Table S2).
+The category Below includes CWD changes that are negative (i.e., mean CWD was larger prior to the initial visit than between visits).
+We also report species mortality and growth estimates for all plots across and within Oregon, Washington, and California irrespective of CWD values.
+We limited the number of species examined to those with more than 400 plots.
+We report domain estimates for all CWD-based domains with 10 or more plots.
+
+We obtained empirical means and 95% confidence intervals via bootstrap.
+For every species we sampled plots with replacement from within each of the strata and obtained point estimates.
+For each domain we calculated a mean and upper and lower 95% CIs with the 2.5% and 97.5% quantiles from 1000 sampling iterations.
+This approach allows for nonsymmetrical CIs around a mean and prevents CIs from including negative values.
+We estimated differences between pairs of CWD domains by subtracting within-iteration domain point estimates.
+[permutation analysis]
+
 # Results
@@
 Our study tracked the fate of `r comma(analysis.stats$tree.num)` trees from `r length(sel.spp)` in `r comma(analysis.stats$plot.num)` plots that were visited twice by FIA crews.
+The two visits for `r pct.remeas[2]`% of plots were 10 years apart(`r pct.remeas[1]`% were 9 years, `r pct.remeas[3]`% were 11 years).
 Of the `r length(sel.spp)` species, `r analysis.stats$n.gym.spp` are gymnosperms.
@@
 
-Forested plot 10-year differences in CWD do not visually coincide with initial CWD values (`r f.ref("fig_cwdmap")`) although most plots increased in their CWD values to some degree; `r analysis.stats$cwd.lt0.pct`% of plots decreased in mean CWD values between visits.
-Individual species quantiles for CWD / delta CWD.
-DO I want to include some sort of quantile-thing for the range of delta-CWD for species?
-Then cite the dashboard.
+Forested plot 10-year differences in CWD do not visually coincide with initial CWD values (Figure `r fig_nums("fig_cwdmap", display = "num")`); however, `r analysis.stats$cwd.gt0.pct`% of plots increased in their CWD values to some degree during that time period.
+Tree species exhibited a variety of initial-visit CWD ranges, with means ranging from `r min.mean.init` (*`r min.mean.init.spp`*) to `r max.mean.init` (*`r max.mean.init.spp`*; Table S1).
+The CWD difference tended to be comparatively small, with a maximum 95th percentile value of `r max.q95.diff` (*`r max.q95.diff.spp`*, Table S2).
+The distribution of individual species plot domain values for CWD and change in CWD can be explored on the dashboard().
+
+The confidence intervals for 10-year mortality and growth rate domains did not necessarily follow the same patterns nor did all confidence intervals for a species' mortality or growth estimates overlap (e.g., Figures `r fig_nums("fig_domains", display = "num")`A and `r fig_nums("fig_domains", display = "num")`B).
+The relationship between initial CWD values and plot-level change in CWD were not reliably linear (e.g., Figure `r fig_nums("fig_domains", display = "num")`C) and varied by species (see dashboard).
+
+We examined differences in 10-year growth estimates among domains (Figure `r fig_nums("fig_diffgrow", display = "num")`).
+The column "Above vs.
+Below Threshold" compares the domains above the CWD change threshold with those below it (e.g., for *Abies lasiocarpa*, subtracting the estimates for columns BL, BM, and BH from AL, AM, AH, respectively; Figure `r fig_nums("fig_domains", display = "num")`A).
+There were `r grow.AvB.sig` estimates out of `r grow.n` which had 95% CIs that did not overlap with zero; `r grow.AvB.signeg` of these were negative.
+Negative estimates indicate that the growth rate was greater for a species in plots below the change estimates.
+
+The second column of Figure `r fig_nums("fig_diffgrow", display = "num")` shows the differences in growth estimates AH - AL, AH - AM, and AM - AL for each species; effectively, the top row of domains shown in Figure `r fig_nums("fig_domains", display = "num")`A.
+`r grow.A.sig` comparisons had 95% CIs that did not include 0, of which `r grow.A.signeg` were negative.
+Negative values indicate that a lower-valued initial CWD domain had a greater growth rate than a higher-valued domain.
+The third column displays the same domain differences but for the bottom row of domains shown in `r fig_nums("fig_domains", display = "num")`A.
+`r grow.B.sig` comparisons had 95% CIs that did not include 0, of which `r grow.B.signeg` were negative.
+Like the second column, negative differences indicate lower initial-value CWD domains had higher growth than higher domains.
+
+We examined mortality estimates in a similar fashion.
+The first column in Figure `r fig_nums("fig_diffmort", display = "num")` has `r mort.AvB.sig` comparisons with 95% CIs that do not overlap zero, `r mort.AvB.sig - mort.AvB.signeg` of which are positive, indicating mortality is greater for those species in one or more of the above-threshold CWD change domains.
+The middle column had `r mort.A.sig` "significant" comparisons, `r mort.A.signeg` of which were negative (for the above-threshold change CWD domains, there was more mortality for lower initial-value CWD domains).
+The right column had `r mort.B.sig` "significant" comparisons, `r mort.B.signeg` of which were negative (for the below-threshold change CWD domains, there was more mortality for lower initial-value CWD domains).
 
@@
 ::: {#refs}
+Wang, T., Hamann, A. Spittlehouse, D.L. and Carroll, C. 2016. Locally downscaled and spatially customizable climate data for historical and future periods for North America. PLoS One 11: e0156720
+
 :::
@@
 
-#tab_num <- run_autonum(seq_id = "Table", pre_label = "Table ", post_label = ": ", bkm = "tbl-growth")
-#block_caption(label = "Decadal growth estimates by state and overall for analysis species.
-Mean estimates with confidence intervals and plot number are provided.
-Estimates are not made for states with fewer than 10 plots.", style = "caption", autonum = tab_num)
-
 growth.ft
@@
 ```{r tbl-mort, tab.cap = tab.tbl_mort}
-#tab_num <- run_autonum(seq_id = "Table", pre_label = "Table ", post_label = ": ", bkm = "tbl-mortality")
-#block_caption(label = "Decadal mortality estimates by state and overall for analysis species.
-Mean estimates with confidence intervals and plot number are provided.
-Estimates are not made for states with fewer than 10 #plots.", style = "caption", autonum = tab_num)
 

### 2025-08-21 Manuscript.qmd
Added Supplemental Material.Rmd which includes tables summarizing CWD stats for all tree species. Debugged many R code files, added Methods to Manuscript.Rmd

@@
     reference-doc: template_use2.docx 
-    fig-width: 6
-    fig-height: 4
-    fig-dpi: 300
+    prefer-html: false
     toc: false
@@
 
-# Load required R libraries
-library(tidyverse)    # Data manipulation and visualization
-library(here)         # File path management
-library(knitr)        # Document generation
 library(kableExtra)   # Enhanced table formatting
-library(ggplot2)      # Advanced plotting
-library(dplyr)        # Data manipulation
-library(readr)        # Data import
 library(scales)       # Scale functions for visualization
-library(patchwork)    # Combining plots
 library(flextable)    # Professional tables for Word output
+#library(officer)
 
@@
 ```{r constants}
+
+source("Global.R")
 
@@
 }
+
+library(flextable)
+set_flextable_defaults(
+  font.family = "Times",
+  font.size = 11,
+  theme_fun = "theme_vanilla"
+)
+
 ```
@@
 ```{r load-data}
-source("Global.R")
 
-spp.names.use <- spp.names %>% dplyr::select()
+# Species scientific names
+spp.names.use <- spp.names %>% dplyr::select(SPCD, GENUS, SPECIES) %>%
+  mutate(SpeciesCode = paste0("X", SPCD), 
+         SciName = paste(GENUS, SPECIES)) %>%
+  dplyr::select(SpeciesCode, SciName)
+
+# Growth data for species overall and states
+grow.state <- read_csv(paste0(RESULTS1.LOC, "Growth_figs_", CLIM.VAR.USE, "/State_Growth_", CLIM.VAR.USE, ".csv"))
+grow.all <- read_csv(paste0(RESULTS1.LOC, "Growth_figs_", CLIM.VAR.USE, "/All_Growth_", CLIM.VAR.USE, ".csv"))
+
+#class(grow.state)
+#class(grow.all)
+```
+
+```{r table-creation}
+# State growth
+s.g.1 <- dplyr::left_join(spp.names.use, grow.state, by = c("SpeciesCode" = "Species")) %>%
+  dplyr::mutate_if(is.numeric, round, 2) %>% 
+  dplyr::mutate(results = case_when(is.na(Median) ~ "", 
+                              !is.na(Median) ~ paste0(Means, " (", LCI.95, ",", UCI.95, "; ", n.plts, ")"))) %>%
+  dplyr::select(SciName, State, results) %>%
+  tidyr::pivot_wider(names_from = State, values_from = results) %>%
+  dplyr::select(-`NA`) %>%
+  dplyr::rename("California" = "CA", "Washington" = "WA", "Oregon" = "OR")
+
+s.g.2 <- dplyr::left_join(spp.names.use, grow.all, by = c("SpeciesCode" = "Species")) %>%
+  dplyr::mutate_if(is.numeric, round, 2) %>% 
+  dplyr::mutate(results = case_when(is.na(Median) ~ "", 
+                              !is.na(Median) ~ paste0(Means, " (", LCI.95, ",", UCI.95, "; ", n.plts, ")"))) %>%
+  dplyr::select(SciName, results) %>%
+ dplyr::rename("All" = "results")
+
+s.g.3 <- dplyr::left_join(s.g.1, s.g.2, by = "SciName") %>% 
+  tidyr::replace_na(list(California = "", Washington = "", Oregon = "")) %>%
+  dplyr::filter(All != "")
+
+growth.ft <- flextable(s.g.3) %>%
+  fontsize(size = 9, part = "body") %>%
+  line_spacing(i = NULL, j = NULL, space = 0.2, part = "body") %>% # Control space between lines. 
+  theme_zebra(
+    odd_header = "#FFFFFF",   # White background for header
+    even_header = "#FFFFFF",  # White background for header
+    odd_body = "#F9F9F9",     # Light gray for odd rows
+    even_body = "#FFFFFF"     # White for even rows
+  ) %>%
+  autofit() %>%
+  set_table_properties(layout = "autofit", width = 1)
+
+# State mortality
+#s.m.1 <-
 
@@
 
-```{r table-creation}
-# State mortality
-s.m.1 <-
+\newpage
+<!---BLOCK_LANDSCAPE_START--->
+```{r tbl-growth}
+#| tbl-cap: "Growth estimates by state and overall for analysis species.
+Mean estimates with confidence intervals and plot number are provided.
+Estimates are not made for states with fewer than 10 plots."
+growth.ft  
+```
 
-```
+<!---BLOCK_LANDSCAPE_STOP--->
 

### 2025-08-21 Supplemental Material.Rmd
Added Supplemental Material.Rmd which includes tables summarizing CWD stats for all tree species. Debugged many R code files, added Methods to Manuscript.Rmd

---
title: "Supplemental material for Tree growth, mortality, and climatic stress in west-coast states, USA"
date: "`r Sys.Date()`"
author: "Jeremiah D. Groom"
output: 
  officedown::rdocx_document:
    mapstyles:
      Normal: ['First Paragraph']
    reference_docx: "template_use2.docx"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)

library(officedown)
library(officer)
library(kableExtra)   # Enhanced table formatting
library(scales)       # Scale functions for visualization
library(flextable)    # Professional tables for Word output
library(rvg)
library(captioner) # For figure captions.

fp <- fp_par(
  text.align = "center", 
  padding.bottom = 20, padding.top = 120, 
  border.bottom = fp_border())

ft <- fp_text(shading.color='#EFEFEF', bold = TRUE)

set_flextable_defaults(
  font.family = "Times New Roman",
  font.size = 11,
  theme_fun = "theme_vanilla"
)

```

```{r load_data}

source('Global.R')
# Some general analysis info + plots
analysis.stats <- read_rds(paste0(RESULTS.OTHER, "analysis.stats.RDS"))

```

```{r cwd_summary_tables}
cwd.init <- analysis.stats$init.cwd.stats 
cwd.diff <- analysis.stats$diff.cwd.stats

init.ft <- flextable(cwd.init) %>%
  fontsize(size = 10, part = "body") %>%
  fontsize(size = 11, part = "header") %>%
  italic(j = 1, part = "body") %>%
  line_spacing(i = NULL, j = NULL, space = 0.5, part = "body") %>% # Control space between lines. 
    set_table_properties(layout = "autofit", width = 0) %>%
  border_remove() %>%
  hline_top(part = "all") %>%
  hline_bottom(part = "all")

diff.ft <- flextable(cwd.diff) %>%
  fontsize(size = 10, part = "body") %>%
  fontsize(size = 11, part = "header") %>%
  italic(j = 1, part = "body") %>%
  line_spacing(i = NULL, j = NULL, space = 0.5, part = "body") %>% # Control space between lines. 
    set_table_properties(layout = "autofit", width = 0) %>%
  border_remove() %>%
  hline_top(part = "all") %>%
  hline_bottom(part = "all")

```

```{r table-legends}
tab_nums <- captioner::captioner(prefix = "Table S")

t.ref <- function(x)  {
  stringr::str_extract(tab_nums(x), "[^:]*")
}

tbl_cwd_init <- tab_nums(name = "tbl_cwdinit", 
                        caption = "Summary statistics for initial-visit plot CWD values by species. The number of species' plots (n), minimum, mean, median, and maximum CWD values are provided, along with the 5%, 25%, 75%, and 95% quantiles (q).")

tbl_cwd_diff <- tab_nums(name = "tbl_cwddiff", 
                        caption = "Summary statistics for the difference in plot CWD values between visits by species. The number of species' plots (n), minimum, mean, median, and maximum CWD values are provided, along with the 5%, 25%, 75%, and 95% quantiles (q).")

```

\newpage

<!---BLOCK_LANDSCAPE_START--->

```{r table-cwd-init, tab.cap=tbl_cwd_init}

init.ft

```

\newpage

```{r table-cwd-diff, tab.cap=tbl_cwd_diff}

diff.ft

```

<!---BLOCK_LANDSCAPE_STOP--->

### 2025-08-21 Analysis Code/Data_Prep_Analysis.R

@@ -194,6 +194,7 @@ if(RUN.STATES == TRUE) {
 tic() 
 # For ANALYSIS.PATHWAY == 1, one climate variable, and 16 dedicated cores, the analysis takes about 8 minutes.
 # With 6 domains and 23 species, 16 dedicated cores, 3.8 minutes
+# With 6 domains, 1000 iterations, 23 spp, 16 cores, 15 min.
 
 for(k in 1:length(ANALYSIS.TYPE)) {  # 1 = grow, 2 = mortality
   
@@ -224,12 +225,7 @@ for(k in 1:length(ANALYSIS.TYPE)) {  # 1 = grow, 2 = mortality
     quant.lims <- mort.grow.dat$quant.lims
     quant.lims.delta <- mort.grow.dat$quant.lims.delt
     
-    # The quant.spp is a data frame of the change quantiles.  This is for 
-    #  reporting purposes (sent over to Manuscript_information.R)
-    quants.spp <- quant.lims %>%
-        map_dfr(~ data.frame(q25 = .["25%"], q75 = .["75%"]), .id = "species")
-    
-    
+
     ## First, adjusting the species list
     #spp.id <- paste0("X", spp.list)     # Can use SEL.SPP
     spp.list <- as.numeric(gsub("X", "", SEL.SPP))
@@ -252,8 +248,10 @@ for(k in 1:length(ANALYSIS.TYPE)) {  # 1 = grow, 2 = mortality
     ## Quantile estimates for mortality or growth, across select species.
     domain.index <- 1:n_domain
     
-    
+
     ## Generating the bootstrap values: 
+   # 200 iterations = 104.99 s, or 1.75 min, or 29.2 hours
+   # 1000 iterations = 426.21 s, or 7.1 min, or 118.3 hrs or 5 days....
     bootstrap_results <- generate_bootstrap_array.fcn(
       vals.dat = vals_dat,
       all.dat = all_dat,
@@ -263,14 +261,14 @@ for(k in 1:length(ANALYSIS.TYPE)) {  # 1 = grow, 2 = mortality
       n_iter = BS.N
     )
 
-    
     # Finding and saving domain summaries
     domain.summaries <- domain.index %>% 
       purrr::map(\(d) domain.sum.fcn(bootstrap_results, d, domain_n = domain.n)) %>%
       do.call(rbind, .) %>%
       arrange(Species, Domain)
     
[…]

### 2025-08-21 Analysis Code/FIA_SQL_compile.R

@@ -16,7 +16,7 @@ db.extract.fcn <- function(state, db.loc) {
            !is.na(PREV_STATUS_CD),   # Remove entries where no previous STATUSCD
            PREV_STATUS_CD != 2,     # Remove entries where tree previously found dead (dead --> dead)
            STATUSCD != 0,           # Remove entries where tree is not presently in sample
-           AGENTCD2 != 80,
+           AGENTCD2 != 80,          # Silvicultural practices
            PREVDIA >= 5,             # Remove microplot trees that became ingrowth.
            DIA2 == 0 | DIA2 >= 5     # Keep DIA = NA in case the tree burned and died, no DIA.  Remove the small trees.
     )
@@ -45,40 +45,40 @@ db.extract.fcn <- function(state, db.loc) {
 }
 
 
-all.states <- map(statelist, db.extract.fcn, db.loc = SQL.LOC)  # Appling the function to all three states
+all.states <- purrr::map(statelist, db.extract.fcn, db.loc = SQL.LOC)  # Appling the function to all three states
 
-extract.plots <- map(all.states , ~.[["plot"]])   # Binding by rows all of the plot data across states
+extract.plots <- purrr::map(all.states , ~.[["plot"]])   # Binding by rows all of the plot data across states
 plot_vals <- bind_rows(extract.plots)
 
-extract.trees <- map(all.states , ~.[["tree"]]) 
+extract.trees <- purrr::map(all.states , ~.[["tree"]]) 
 tree_vals <- bind_rows(extract.trees)
 
-extract.cond <- map(all.states , ~.[["cond"]]) 
+extract.cond <- purrr::map(all.states , ~.[["cond"]]) 
 cond_vals <- bind_rows(extract.cond)
 
-extract.sccm <- map(all.states , ~.[["sccm"]]) 
+extract.sccm <- purrr::map(all.states , ~.[["sccm"]]) 
 sccm_vals <- bind_rows(extract.sccm)
 
-extract.peu <- map(all.states , ~.[["peu"]]) 
+extract.peu <- purrr::map(all.states , ~.[["peu"]]) 
 peu_vals <- bind_rows(extract.peu)
 
-extract.pev <- map(all.states , ~.[["pev"]]) 
+extract.pev <- purrr::map(all.states , ~.[["pev"]]) 
[…]

### 2025-08-21 Analysis Code/Functions.R

@@ -465,7 +465,8 @@ mean.q.fcn <- function(dat_tot, dat_vals, spp.sel1) {
   
   # Old slow code.  For understanding the data.table usage.
   #  Zt_i <- Yv_i <- rep(0, length(strata) )  # Weighted values for each strata
-  #  for (h in 1:strat.num) {                  # h indexes strata
+  #strat.num <- length(unique(strata))  
+  #for (h in 1:strat.num) {                  # h indexes strata
   #    w_h <- unique(dat_tot$w[dat_tot$stratum == strata[h]])
   #    n_h <- unique(dat_tot$n_h.plts[dat_tot$stratum == strata[h]])    # Number of plots in stratum h 
   #    Yv_i[h] <- w_h * sum(get(col.name, dat_vals)[dat_vals$stratum == strata[h]]) / n_h 
@@ -500,14 +501,20 @@ calculate_domain_estimates.fcn <- function(samp, domain_data, domain_idx, select
       if (n_q < N.PLOT.LIM) {
         return(NA_real_)
       } else {
-        return(mean.q.fcn(
+       # return(mean.q.fcn(
+        mean.q.fcn(
           dat_tot = d.all_use[samp$row.id, ],
           dat_vals = d.vals_use[samp$row.id, ],
           spp.sel1 = spp.sel
-        )$R)
+        )$R
       }
     })
   
+
+ # df <- rbind(estimates, estimates.yv, estimates.zt) %>% data.frame()
+#  colnames(df) <- SEL.SPP
+ # write.csv(df, "Oregon.csv")
+
   return(estimates)
 }
 
@@ -578,6 +585,8 @@ generate_bootstrap_array.fcn <- function(vals.dat, all.dat, domain.array, domain
   
   return(bootstrap_array2)
 }
+
+
 
 
 
@@ -658,6 +667,9 @@ cm2.fcn <- function(k, results.table){
 ## This plot provides the 
 diff.panel.fcn <- function(diff.dat, remove.y, fig.title, lab.right) {
   
+  # Kludgy way to get a conifer/deciduous break.  X312 = bigleaf maple.
+  decid.break <- length(SEL.SPP) - which(SEL.SPP == "X312") + 1.5
+  
   ggplot(diff.dat, aes(y = as.numeric(species_label) + y.offset)) + 
[…]

### 2025-08-21 Analysis Code/Manuscript_information.R

@@ -8,6 +8,11 @@
 tree.num <- dim(tree.plt.data %>% filter(SPCD %in% sel.spp))[1]
 plot.num <- dim(PlotDat)[1]
 
+## Q: What percentage of plots with 9, 10, and 11 yr remeasure?
+remeas <- tree.plt.data %>% dplyr::select(puid, REMPER) %>% distinct() %>% mutate(REMPER2 = round(REMPER, digits = 0))
+pct.remeas <- round(100 * table(remeas$REMPER2)/nrow(remeas), digits = 1)[4:6]
+
+
 ## Number of Gymnosperms?  Which ones?
 gym.angio.dat <- spp.names %>% filter(SPCD %in% sel.spp) %>%
   mutate(sci_name = paste(GENUS, SPECIES),
@@ -23,15 +28,50 @@ n.zero <- nrow(climate.use %>% filter(difference < 0))
 n.min.3 <- nrow(climate.use %>% filter(difference < -3))
 
 cwd.lt.0.pct <- round(n.zero/n.cwd, 3) * 100 #Percent CWD diff < 0  =  8%
+cwd.gt.0.pct <- round(1 - n.zero/n.cwd, 3) * 100 # Percent CWD diff > 0 (91.6%)
 cwd.lt.min3.pct <- n.min.3/n.cwd #Percent CWD diff < -3  = 0.6% (This is useful for the mapping)
 
 
 
+mort.out <- readRDS(paste0(save.loc.fcn(2), "Domain_Analysis_Output.RDS"))
+
+init.cwd.stats <- summ.spp.fcn(mort.out$all_dat, "pre_mean", gym.angio.dat) 
+diff.cwd.stats <- summ.spp.fcn(mort.out$all_dat, "difference", gym.angio.dat)
+
+init.cwd.min.mean <- which(init.cwd.stats$Mean == min(init.cwd.stats$Mean))
+init.cwd.max.mean <- which(init.cwd.stats$Mean == max(init.cwd.stats$Mean))
+init.range = init.cwd.stats$q95 - init.cwd.stats$q05
+init.cwd.90range.min <- which(init.range == min(init.range))
+init.cwd.90range.max <- which(init.range == max(init.range))
+
+
+diff.cwd.min.mean <- which(diff.cwd.stats$Mean == min(diff.cwd.stats$Mean))
+diff.cwd.max.mean <- which(diff.cwd.stats$Mean == max(diff.cwd.stats$Mean))
[…]

### 2025-08-21 Analysis Code/Overall_Mort_Est.R

@@ -7,10 +7,10 @@ for(k in 1:2){ # 1 = growth, 2 = mortality
   
   
   if(k == 1) {
-    t.all <- map(as.numeric(gsub("X", "", SEL.SPP)), clim.growth.resp.fcn, clim.var = CLIM.VAR.USE, treedat.sel = treedat.use, clim.dat = climate.use)
+    t.all <- purrr::map(as.numeric(gsub("X", "", SEL.SPP)), clim.growth.resp.fcn, clim.var = CLIM.VAR.USE, treedat.sel = treedat.use, clim.dat = climate.use)
     a.all <- parse.tree.clim.fcn(t.all, clim.var = CLIM.VAR.USE, analysis.type = "grow", resp.dat = "growth.val", tot.dat = "growth.n.trees", selected.spp = SEL.SPP, clim.dat = climate.use)
   } else { 
-    t.all <- map(as.numeric(gsub("X", "", SEL.SPP)), clim.mort.resp.fcn, clim.var = CLIM.VAR.USE, treedat.sel = treedat.use, clim.dat = climate.use)
+    t.all <- purrr::map(as.numeric(gsub("X", "", SEL.SPP)), clim.mort.resp.fcn, clim.var = CLIM.VAR.USE, treedat.sel = treedat.use, clim.dat = climate.use)
     a.all <- parse.tree.clim.fcn(t.all, clim.var = CLIM.VAR.USE, analysis.type = "mort", resp.dat = "died.out", tot.dat = "all.trees", selected.spp = SEL.SPP, clim.dat = climate.use)
   }
   
@@ -56,7 +56,7 @@ for(k in 1:2){ # 1 = growth, 2 = mortality
   
   # Finding and saving domain summaries
   state.domain.summaries <- state.domain.index %>% 
-    map(\(d) domain.sum.fcn(state.bootstrap_results, d, domain_n =state.n)) %>%
+    purrr::map(\(d) domain.sum.fcn(state.bootstrap_results, d, domain_n =state.n)) %>%
     do.call(rbind, .) %>%
     mutate(State = case_match(Domain, 1 ~ "CA", 2 ~ "WA", 3 ~ "OR")) %>%
     arrange(Species, Domain)

### 2025-08-21 Global.R

@@ -23,7 +23,9 @@ library(maps)
 # Code controls:
 CALC.FIREPROP <- TRUE   # Set this to TRUE if wanting to calculate or recalculate
                           # the proportion of dead trees killed by fire by quantile.
-RUN.STATES <- TRUE # Should ANALYSIS.PATHWAY 1 find mortality estimates by state?
+RUN.STATES <- TRUE # Should Data_Prep_Analysis.R find mortality estimates by state and overall?
+
+RUN.SUMMARY <- TRUE # Should Data_Prep_Analysis.R calculate 
 
 # File locations:
 DATA.LOC <- "Data/"

### 2025-08-21 Manuscript.Rmd
Fix to mortality calcs, changed bootstrap to sample from the entire pool of plots, not within each stratum.

@@
 
-FIA plots consist of four points where all trees with stem diameters greater than or equal to 12.7 cm (for most species measured at 1.37 m above the ground) were tallied in a 7.32 m radius circular subplot (total area 672.5 m$^2$) with additional trees sampled in an annulus around each subplot (macroplot) around each subplot, with a total radius of 18 m from the center point and a total area of 1012 m$^2$.
+FIA plots consist of four points where all trees with stem diameters greater than or equal to 12.7 cm (for most species measured at 1.37 m above the ground) were tallied in a 7.32 m radius circular subplot (total area 672.5 m$^2$) with additional trees sampled in an anulus around each subplot (macroplot) around each subplot, with a total radius of 18 m from the center point and a total area of 1012 m$^2$.
 We did not include data from FIA microplots.
@@
 Since 2000, FIA plots have been selected via a spatially balanced probability sample (McRoberts 2005, Reams et al., 2005) with a target plot density of one ground plot every 24 km$^2$.
-Some regions and ownerships intensify the plot density; we excluded these additional plots to simplify stratification estimation, relying on [601] FIA-defined strata and using the density of FIA Phase 1 sampling points within and across strata to determine strata weights.
+Some regions and ownerships intensify the plot density; we excluded these additional plots to simplify stratification estimation, relying on FIA-defined strata and using the density of FIA Phase 1 sampling points within and across strata to determine strata weights.
 
@@
 They included plot data sampled from 2001 to 2023.
-Our total sample size was [12500] forested plots (Figure 1).
 
@@
 
-Our study tracked the fate of `r comma(analysis.stats$tree.num)` trees from `r length(sel.spp)` in `r comma(analysis.stats$plot.num)` plots that were visited twice by FIA crews.
-The two visits for `r pct.remeas[2]`% of plots were 10 years apart(`r pct.remeas[1]`% were 9 years, `r pct.remeas[3]`% were 11 years).
+Our study tracked the fate of `r comma(analysis.stats$tree.num)` trees from `r length(sel.spp)` species in `r comma(analysis.stats$plot.num)` plots that were visited twice by FIA crews.
+The two visits for `r pct.remeas[2]`% of plots were 10 years apart (`r pct.remeas[1]`% were 9 years, `r pct.remeas[3]`% were 11 years).
+Plots were distributed in `r analysis.stats$strat.num` strata.
 Of the `r length(sel.spp)` species, `r analysis.stats$n.gym.spp` are gymnosperms.
-Overall growth rates for species varied between `r min.grow.spp$mean` (*`r min.grow.spp$spp`*) and `r max.grow.spp$mean` (*`r max.grow.spp$spp`*, `r t.ref("tbl_growth")`) cm$^2$/decade.
+Mean overall growth rates for species varied between `r min.grow.spp$mean` (*`r min.grow.spp$spp`*) and `r max.grow.spp$mean` (*`r max.grow.spp$spp`*, `r t.ref("tbl_growth")`) cm$^2$/decade.
 Decadal mean mortality rates varied between `r min.mort.spp$mean` (*`r min.mort.spp$spp`*) and `r max.mort.spp$mean` (*`r max.mort.spp$spp`*, `r t.ref("tbl_mort")`).

### 2025-08-21 Analysis Code/Data_Prep.R

@@ -1,223 +1,0 @@
-
-
-source("Global.R")
-source(paste0(CODE.LOC, "Functions.R"))
-
-
-##### ---- Code ---- #####
-
-# DATA PREP
-
-# Climate variables and reduction to a single table
-tmp <- read_csv(paste0(DATA.LOC, "tmp.20.10.csv")) %>% dplyr::select(-INVYR) %>% mutate(delt.temp = post.temp - pre.temp) 
-precip <- read_csv(paste0(DATA.LOC, "precip.20.10.csv")) %>% dplyr::select(-INVYR) %>% mutate(delt.precip = post.precip - pre.precip)
-vpdmin <- read_csv(paste0(DATA.LOC, "vpdmin.20.10.csv")) %>% dplyr::select(-INVYR) %>% mutate(delt.vpdmin = post.vpdmin - pre.vpdmin)
-vpdmax <- read_csv(paste0(DATA.LOC, "vpdmax.20.10.csv")) %>% dplyr::select(-INVYR) %>% mutate(delt.vpdmax = post.vpdmax - pre.vpdmax)
-
-
-# clim.dat = all climate data
-clim.dat <- list(tmp, precip, vpdmin, vpdmax) %>% reduce(left_join, by = c("STATECD", "PLOT_FIADB")) %>% 
-  mutate(State_Plot = paste0(PLOT_FIADB, STATECD),
-         State_Plot = as.numeric(State_Plot)) %>% 
-  dplyr::select(-STATECD, -PLOT_FIADB)
-
-
-
-##### Tree and plot info  ####
-## Some info (harvest, fire deaths for trees) was absent in the range-shift analysis.  We need to introduce it here. 
-##  The following code relies on downloaded and zipped SQLite FIA data from Oregon, Washington, and CA.  See the Global.R file:
-#    the zip files should be placed in the same folder, defined by SQL.LOC.  The following code unpacks, reduces, combines, and saves
-#   the PLOT, TREE, and COND files for each state as a zipped RDS file.  
-if (file.exists(paste0(DATA.LOC, "Addl_PlotTreeInfo.zip")) == FALSE) {
-  source(paste0(CODE.LOC, "FIA_SQL_compile.R"))
-}
-
-fia.tables <- read_rds(unzip(paste0(DATA.LOC, "Addl_PlotTreeInfo.zip"), "Addl_PlotTreeInfo.rds")) # This unzips the folder in the parent directory.
[…]

### 2025-08-21 Analysis Code/Data_Prep_Analysis.R

@@ -180,11 +180,12 @@ death.prop <- read_csv(paste0(RESULTS1.LOC, "Mort_figs_", CLIM.VAR.USE, "/Fire_P
 
 # If desired, the analysis will obtain mortality estimates for the selected species
 #  by state and overall. These values can be used to evaluate species for inclusion in the
-#  main analysis
+#  main analysis.  200 iterations, 16 cores = 4 minutes
+tic()
 if(RUN.STATES == TRUE) {
   source(paste0(CODE.LOC, "Overall_Mort_Est.R"))
 }
-
+toc()
 
 
 
@@ -358,15 +359,15 @@ if(n_domain == 6) {
     ggsave(paste0(save.loc.fcn(k), filename.use, "Panel_Plot.png"), plot = diff.plt2, device = "png", width = 10, height = 10, units = 'in')
   }
 }
-domain.index %>% 
-  purrr::map(\(d) domain.sum.fcn(bootstrap_results, d, domain_n = domain.n))
 
-  # Analysis Pathway 1: no subgroups of size or site class
+
+
+# Analysis Pathway 1: no subgroups of size or site class
 for(k in 1:2){ # 1 = growth, 2 = mortality
   plt.dat <- readRDS(paste0(save.loc.fcn(k), "Domain_Analysis_Output.RDS"))
   plt.dat2 <- plt.dat$domain.summaries
   
-
+  
   mort.grow.dat <- get(paste0("arrays.", ANALYSIS.TYPE[k])) 
   
   domain.matrix <- mort.grow.dat$domain.matrix
@@ -375,10 +376,10 @@ for(k in 1:2){ # 1 = growth, 2 = mortality
   quant.lims.delta <- mort.grow.dat$quant.lims.delt
   
   
-    
-    # Plotting paired plots of mortality/growth by quantile and a scatterplot of plot distribution by quantiles.
-    SEL.SPP %>% purrr::map(\(s) pair.plts.fcn(sppnum.to.plot = s, use.dat = plt.dat2, domain.matrix = domain.matrix,
-                 quant.lims = quant.lims, domain.n = domain.n, k = k))
+  
+  # Plotting paired plots of mortality/growth by quantile and a scatterplot of plot distribution by quantiles.
+  SEL.SPP %>% purrr::map(\(s) pair.plts.fcn(sppnum.to.plot = s, use.dat = plt.dat2, domain.matrix = domain.matrix,
+                                            quant.lims = quant.lims, domain.n = domain.n, k = k))
 }
 
 

### 2025-08-21 Analysis Code/Functions.R

@@ -28,7 +28,7 @@ clim.mort.resp.fcn <- function(spcd, clim.var, treedat.sel, clim.dat) {
   # For alive/dead tree assessments
   Tree2 <- treedat.sel %>% 
     filter(SPCD == spcd) %>%
-    dplyr::select(puid, REMPER, TREE, STATUSCD, ntree.rep) %>% distinct()
+    dplyr::select(puid, REMPER, SUBP, TREE, STATUSCD, ntree.rep)
   
   tree.plots <- clim.dat %>% filter(puid %in% Tree2$puid) %>% 
     dplyr::select(puid, all_of(var1), all_of(var.delt))
@@ -47,8 +47,7 @@ clim.mort.resp.fcn <- function(spcd, clim.var, treedat.sel, clim.dat) {
   }
   
   # Applying quantiles to a particular species
-  if(ANALYSIS.PATHWAY != 3) {
-    tree.plots2 <- tree.plots %>% dplyr::select(puid, all_of(c(var1, var.delt))) %>% 
+  tree.plots2 <- tree.plots %>% dplyr::select(puid, all_of(c(var1, var.delt))) %>% 
       {if (n_domain == 9) {
         mutate(., var.quants = ifelse(!!sym(var1) > quant.lims[2], 1, ifelse(!!sym(var1) < quant.lims[1], -1, 0)),
                deltvar.quants = ifelse(!!sym(var.delt) > quant.lims.delta[2], 1, ifelse(!!sym(var.delt) < quant.lims.delta[1], -1, 0)),
@@ -61,11 +60,9 @@ clim.mort.resp.fcn <- function(spcd, clim.var, treedat.sel, clim.dat) {
                vq = case_match(var.quants, -1 ~ "L", 0 ~ "M", 1 ~ "H"),
                dvq = case_match(deltvar.quants, 0 ~ "B", 1 ~ "A"),
                var.deltvar = factor(paste0(dvq, vq), levels = DOMAIN.LEVELS))
-      } else {
-        values$errorMessage <- "Error: Expecting 6 or 9 estimation domains. Check Global.R:DOMAIN.LEVELS and Functions.R:clim.mort.resp.fcn."
-        return(.)
       }}
-    
+        
+    # Not currently used, but available    
     # For centroid calcluation
     dat.cent <- tree.plots2 %>% group_by(var.deltvar) %>% # Quartile centers, good for finding centroids in each quadrant
[…]

### 2025-08-21 Analysis Code/Manuscript_information.R

@@ -7,6 +7,7 @@
 ### Q: How many trees did we track? How many plots?
 tree.num <- dim(tree.plt.data %>% filter(SPCD %in% sel.spp))[1]
 plot.num <- dim(PlotDat)[1]
+strat.num <- length(unique(PlotDat$stratum))
 
 ## Q: What percentage of plots with 9, 10, and 11 yr remeasure?
 remeas <- tree.plt.data %>% dplyr::select(puid, REMPER) %>% distinct() %>% mutate(REMPER2 = round(REMPER, digits = 0))
@@ -55,6 +56,7 @@ diff.cwd.95max <- which(diff.cwd.stats$q95 == max(diff.cwd.stats$q95))
 
 analysis.stats <- list(tree.num = tree.num,
                        plot.num = plot.num,
+                       strat.num = strat.num,
                        pct.remeas = pct.remeas,
                        cwd.lt0.pct = cwd.lt.0.pct,
                        cwd.gt0.pct = cwd.gt.0.pct,
@@ -111,8 +113,8 @@ for(k in 1:2){
   
   text.size <- 9
   
-  p1 <- domain.grid.plt.fcn(c("DL", "DM", "DH"), 1:3) 
-  p2 <- domain.grid.plt.fcn(c("SL", "SM", "SH"), 4:6) 
+  p1 <- domain.grid.plt.fcn(c("DL", "DM", "DH"), 1:3, use.dat2, qt.max, q.g.p.labs) 
+  p2 <- domain.grid.plt.fcn(c("SL", "SM", "SH"), 4:6, use.dat2, qt.max, q.g.p.labs) 
   #p_all <- plot_grid(p1, p2, ncol = 1)
   list1 <- list(p1 = p1, p2 = p2)
   fig.letter <- switch(k, "A", "B")

### 2025-09-09 Manuscript.Rmd
Created a third table for the manuscript summarizing the two difference figures findings.

@@
 # Significant findings, Growth
-grow.n <- nrow(grow.diff$AvB2)
+grow.n.AvB <- nrow(grow.diff$AvB2)
+grow.n.A <- nrow(grow.diff$A_LMH2)
+grow.n.B <- nrow(grow.diff$B_LMH2)
 
@@
 # Significant findings, Mort
+mort.n.AvB <- nrow(mort.diff$AvB2)
+mort.n.A <- nrow(mort.diff$A_LMH2)
+mort.n.B <- nrow(mort.diff$B_LMH2)
+
 mort.AvB.sig <- length(mort.diff$AvB2$significant[mort.diff$AvB2$significant == TRUE])
@@
 mort.B.signeg <- length(mort.diff$B_LMH2$significant[mort.diff$B_LMH2$significant == TRUE & mort.diff$B_LMH2$Means < 0])
+
+# Summary of significant Growth and Mortality findings for table creation.
+datasets.grow <- list(grow.diff$AvB2, grow.diff$A_LMH2, grow.diff$B_LMH2)
+datasets.mort <- list(mort.diff$AvB2, mort.diff$A_LMH2, mort.diff$B_LMH2)
+datatypes <- list("Above vs. Below", "Above", "Below")
+n.comp.grow <- list(grow.n.AvB, grow.n.A, grow.n.B)
+n.comp.mort <- list(mort.n.AvB, mort.n.A, mort.n.B)
+
+growth.summ <- pmap(list(datasets.grow, datatypes, n.comp.grow), summ.fig.fcn) %>% bind_rows()
+mortality.summ <- pmap(list(datasets.mort, datatypes, n.comp.mort), summ.fig.fcn) %>% bind_rows()
 
@@
 
+# Adding a section identifier to each table
+growth.summ2 <- growth.summ %>%
+  mutate(Type = "Growth", .before = 1)
+
+mortality.summ2 <- mortality.summ %>%
+  mutate(Type = "Mortality", .before = 1)
+
+# Combine the tables
+grow.mort.summ <- bind_rows(growth.summ2, mortality.summ2)
+
+# Create flextable
+g.m.summ.ft <- flextable(grow.mort.summ) %>%
+  merge_v(j = "Type") %>%  # Merge cells in Section column
+  valign(j = "Type", valign = "top") %>%  # Top align section labels
+  flextable::align(j = "Type", align = "left") %>%   # Left align section labels
+  line_spacing(i = NULL, j = NULL, space = 0.5, part = "body") %>% # Control space between lines. 
+  set_table_properties(layout = "autofit", width = 0) %>%
+  theme_zebra(
+    odd_header = "#FFFFFF",   # White background for all
+    even_header = "#FFFFFF",
+    odd_body = "#FFFFFF",     
+    even_body = "#FFFFFF"     
+  ) %>%
+    flextable::align(j = 3:7, align = "center", part = "all")
+
 ```
@@
 # 
-fig_nums <- captioner::captioner(prefix = "Figure")
+fig_nums <- captioner::captioner(prefix = "")
 tab_nums <- captioner::captioner(prefix = "Table")
@@
 f.ref <- function(x) {
- stringr::str_extract(fig_nums(x), "[^:]*")
+ caption_text <- fig_nums(x)
+  # Remove any leading numbers and extract just "Figure X"
+  stringr::str_extract(caption_text, "\\d+")
 }
@@
 tab.tbl_growth <- tab_nums(name = "tbl_growth", 
-                        caption = "Decadal growth estimates by state and overall for analysis species.
+                        caption = "Decadal growth estimates by state and overall for analysis species (cm\u00B2/decade).
 Mean estimates with confidence intervals and plot number are provided.
@@
 fig.domains <- fig_nums(name = "fig_domains",
-                        caption = "Decadal (A) growth and (B) mortality domain estimates for *Abies lasiocarpa*.
+                        caption = "Decadal (A) growth and (B) mortality domain estimates for *Calocedrus decurrens*.
 The six domains are defined as FIA plots that fall [A]bove or [B]elow the CWD change threshold with [L]ow, [M]edium, or [H]igh initial CWD values (see text for details).
@@
 Species above the dashed line are coniferous; those below are deciduous."))
+
+tab.summ_grow_mort <- tab_nums(name = "tbl_summ_grow_mort",
+                               caption = paste0("Summary of number of comparisons with 95% CIs that do not overlap with zero ([T]otal, [P]ositive-, and [N]egative-valued estimates) for Figures ", fig_nums("fig_diffgrow", display = "num"), " and ", fig_nums("fig_diffmort", display = "num"), " by analysis (T)ype (Growth or Mortality) and type of (C)omparison.
+The column Domains lists the domains compared; All is the sum of values within a Comparison.
+See text for other domain labels.
+'n' is the total number of differences examined within a Comparison."))
 
@@
 We obtained empirical means and 95% confidence intervals via bootstrap.
-For every species we sampled plots with replacement from within each of the strata and obtained point estimates.
+For every species we sampled plots with replacement and obtained point estimates.
 For each domain we calculated a mean and upper and lower 95% CIs with the 2.5% and 97.5% quantiles from 1000 sampling iterations.
@@
 
-We examined differences in 10-year growth estimates among domains (Figure `r fig_nums("fig_diffgrow", display = "num")`).
+We examined differences in 10-year growth estimates (Figure `r fig_nums("fig_diffgrow", display = "num")`) and mortality estimates (Figure `r fig_nums("fig_diffmort", display = "num")`) among domains.
 The column "Above vs.
 Below Threshold" compares the domains above the CWD change threshold with those below it (e.g., for *Abies lasiocarpa*, subtracting the estimates for columns BL, BM, and BH from AL, AM, AH, respectively; Figure `r fig_nums("fig_domains", display = "num")`A).
-There were `r grow.AvB.sig` estimates out of `r grow.n` which had 95% CIs that did not overlap with zero; `r grow.AvB.signeg` of these were negative.
-Negative estimates indicate that the growth rate was greater for a species in plots below the change estimates.
+Negative estimates for the "Above vs.
+Below Threshold" in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")`  indicate that the 10-year growth or mortality rate was greater for a species in plots below the change estimates, respectively.
+`r t.ref("tbl_summ_grow_mort")` summarizes the number of domain differences in each of the columns in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")` where the 95% CI does not include zero as well as the number of comparisons per figure column.
 
-The second column of Figure `r fig_nums("fig_diffgrow", display = "num")` shows the differences in growth estimates AH - AL, AH - AM, and AM - AL for each species; effectively, the top row of domains shown in Figure `r fig_nums("fig_domains", display = "num")`A.
-`r grow.A.sig` comparisons had 95% CIs that did not include 0, of which `r grow.A.signeg` were negative.
-Negative values indicate that a lower-valued initial CWD domain had a greater growth rate than a higher-valued domain.
-The third column displays the same domain differences but for the bottom row of domains shown in `r fig_nums("fig_domains", display = "num")`A.
-`r grow.B.sig` comparisons had 95% CIs that did not include 0, of which `r grow.B.signeg` were negative.
-Like the second column, negative differences indicate lower initial-value CWD domains had higher growth than higher domains.
+Growth domain comparisons (`r fig_nums("fig_diffgrow")`, `r t.ref("tbl_summ_grow_mort")`) did not appear to be produce substantially more "significant" negative vs.
+positive estimates, and all three columns of `r fig_nums("fig_diffgrow")` had similar numbers of "significant" comparisons.
+Growth may have been susceptible (negatively or positively) to a change in CWD and differed by initial CWD values, and the responses are species-specific.
+*Thuja plicata* exhibited greater growth in the Above and Below Medium domains than in the Lower or Higher domains.
+*Abies amablis* had lower growth in the Above and Below High domains than the others, where initial-visit CWD was the greatest.
+*Picea englelmannii* had lower growth in the Low domain than the High domain, indicating that more growth was occurring where trees were more water stressed.
 
-We examined mortality estimates in a similar fashion.
-The first column in Figure `r fig_nums("fig_diffmort", display = "num")` has `r mort.AvB.sig` comparisons with 95% CIs that do not overlap zero, `r mort.AvB.sig - mort.AvB.signeg` of which are positive, indicating mortality is greater for those species in one or more of the above-threshold CWD change domains.
-The middle column had `r mort.A.sig` "significant" comparisons, `r mort.A.signeg` of which were negative (for the above-threshold change CWD domains, there was more mortality for lower initial-value CWD domains).
-The right column had `r mort.B.sig` "significant" comparisons, `r mort.B.signeg` of which were negative (for the below-threshold change CWD domains, there was more mortality for lower initial-value CWD domains).
+Mortality domain comparisons are similar in certain regards to the growth comparisons: results for the three types of domain comparisons did not appear to be predominantly positive or negative (`r fig_nums("fig_diffmort")`, `r t.ref("tbl_summ_grow_mort")`).
+There were fewer mortality estimates with "significant" findings than growth.
+There were fewer differences for "Above vs.
+Below Threshold" comparisons (`r t.ref("tbl_summ_grow_mort")`) than for the comparisons Above or Below the threshold, indicating that mortality may not be as strongly associated with the change in CWD as by the starting value.
+*Abies concolor* had greater mortality at the Above Low and High categories than the Above Middle.
+*Calocedrus decurrens* had a higher mortality rate at the High Above domains than the other Above domains.
+*Picea engelmannii* was the opposite; mortality was lowest at the Above High domain compared to other domains.
+*Psuedotsuga menziesii* exhibited greater mortality at its Below High domain than other Below domains.
+The two *Quercus* species did not exhibit strong patterns to their mortality.
 
@@
 
-```{r mapfig, fig.height = 7,  fig.dpi=300, fig.cap=fig.cwdmap_cap, fig.cap.pre="", fig.cap.sep=""}
+```{r mapfig, fig.height = 7,  fig.dpi=300, fig.cap=fig.cwdmap_cap, fig.cap.sep=""}
 
@@
 
-```{r domainsfig, fig.width = 6,  fig.dpi=300, fig.cap=fig.domains, fig.cap.pre="", fig.cap.sep=""}
+```{r domainsfig, fig.width = 6,  fig.dpi=300, fig.cap=fig.domains, fig.cap.sep=""}
 
@@
 
-```{r diff_grow_fig, fig.height = 6, width = 6, fig.dpi=300, fig.cap=fig.diff_grow, fig.cap.pre="", fig.cap.sep=""}
+```{r diff_grow_fig, fig.height = 6, width = 6, fig.dpi=300, fig.cap=fig.diff_grow, fig.cap.sep=""}
 
@@
 
-```{r diff_mort_fig, fig.height = 6, width = 6, fig.dpi=300, fig.cap=fig.diff_mort, fig.cap.pre="", fig.cap.sep=""}
+```{r diff_mort_fig, fig.height = 6, width = 6, fig.dpi=300, fig.cap=fig.diff_mort, fig.cap.sep=""}
 
@@
 ```
+
+\newpage
+
+```{r tbl-summ-grow-mort, tab.cap = tab.summ_grow_mort}
+
+g.m.summ.ft
+
+```

### 2025-09-09 Analysis Code/Functions.R

@@ -765,8 +765,8 @@ domain.dist.plt.fcn <- function(plot.spp, domain.matrix, var1, var.delt, quant.l
   plot.vals.plt <- ggplot(q_plot_spp, aes(get(var1), get(var.delt), color = factor(Domain))) + 
     geom_point() + 
     #stat_ellipse(type = "norm", level = 0.95, col = "orange", lwd = 2) +
-    geom_hline(yintercept = quant.lims.delt.plt[1], col = "blue") +
-    geom_hline(yintercept = quant.lims.delt.plt[2], col = "blue") +
+    geom_hline(yintercept = quant.lims.delt.plt[1], col = "blue", linewidth = 1.5, linetype = 2) +
+    geom_hline(yintercept = quant.lims.delt.plt[2], col = "blue", linewidth = 1.5, linetype = 2) +
     geom_vline(xintercept = quant.lims.plt[1], col = "green") +
     geom_vline(xintercept = quant.lims.plt[2], col = "green") +
     geom_hline(yintercept = 0, col = "black") +
@@ -1001,6 +1001,31 @@ maxmin.fcn <- function(table1, mean_col, lower_ci_col, upper_ci_col,
   return(maxmin_output)
 }
 
+## Summary of domain difference figures.  A pmap() call will 
+#   operate on a list of datasets. Datatype = type of comparison,
+#   n.comp is the number of total comparisons in a figure column (sig and non-sig).
+summ.fig.fcn <- function(dataset, datatype, n.comp) {
+  
+  x.a <- dataset %>% 
+    mutate(np = ifelse(Means < 0, "Negative", "Positive"),
+           pos.sig = ifelse(significant == TRUE & np == "Positive", 1, 0)) 
+  
+  x.a1 <- x.a %>% 
+    group_by(Domain) %>%
+    reframe(Total = length(significant[significant == TRUE]),
+            Positive = sum(pos.sig),
+            Negative = Total - Positive)
+  
+  x.a2 <- x.a1 %>% 
+    mutate(across(-Domain, \(x) sum(x)),
+           Domain = "All") %>% 
+    distinct() %>%
+    bind_rows(x.a1) %>%
+    mutate(Comparison = datatype,
+           n = ifelse(Domain == "All", n.comp, "")) %>%
+    relocate(Comparison) 
+}
+
 
 
[…]

### 2025-09-09 Analysis Code/Manuscript_information.R

@@ -36,8 +36,8 @@ cwd.lt.min3.pct <- n.min.3/n.cwd #Percent CWD diff < -3  = 0.6% (This is useful 
 
 mort.out <- readRDS(paste0(save.loc.fcn(2), "Domain_Analysis_Output.RDS"))
 
-init.cwd.stats <- summ.spp.fcn(mort.out$all_dat, "pre_mean", gym.angio.dat) 
-diff.cwd.stats <- summ.spp.fcn(mort.out$all_dat, "difference", gym.angio.dat)
+init.cwd.stats <- summ.spp.fcn(data.all = mort.out$all_dat, "pre_mean", gym.angio.dat) 
+diff.cwd.stats <- summ.spp.fcn(data.all = mort.out$all_dat, "difference", gym.angio.dat)
 
 init.cwd.min.mean <- which(init.cwd.stats$Mean == min(init.cwd.stats$Mean))
 init.cwd.max.mean <- which(init.cwd.stats$Mean == max(init.cwd.stats$Mean))
@@ -137,17 +137,34 @@ for(k in 1:2){
     q_plot_spp <- domain.matrix %>% dplyr::select(all_of(sppnum.to.plot), all_of(var1), all_of(var.delt)) %>%
       filter(get(sppnum.to.plot) > 0) 
     pts.max.y <- ceiling(layer_scales(plot.vals.plt)$y$range$range[2] * 10) / 10 # Extracts max Y extent of ggplot
+    pts.min.y <- ceiling(layer_scales(plot.vals.plt)$y$range$range[1] * 10) / 10 # Extracts max Y extent of ggplot
     
     #ceiling(max(use.dat2$UCI.95, na.rm = TRUE) * 10) / 10 # For consistent y-axis heights
     
     
     range.x <- range(q_plot_spp$pre_mean, na.rm = TRUE)
     pts.min.x <- min(q_plot_spp$pre_mean) + 0.065 * (range.x[2] - range.x[1]) 
+    pts.max.x <- max(q_plot_spp$pre_mean)
+    
+    # Locations for domain text in figure:
+    q.lim.n <- which(names(quant.lims) == sppnum.to.plot)
+    q.lim.plot <- quant.lims[[q.lim.n]]
+    # Low/med/high positions
+    l.txt.loc <- q.lim.plot[1] - (q.lim.plot[1] - pts.min.x)/2
+    m.txt.loc <- q.lim.plot[1] + (q.lim.plot[2] - q.lim.plot[1])/2
[…]

### 2025-09-10 Global.R
Reran code with 1000 iterations

@@ -72,7 +72,7 @@ DOMAIN.LEVELS <- c("LiHc", "MiHc", "HiHc", "LiMc", "MiMc", "HiMc", "LiLc", "MiLc
 DOMAIN.LEVELS <- c("AL", "AM", "AH", "BL", "BM", "BH") # Above/Below threshold and Low/Medium/High
 }
 N.PLOT.LIM <- 10  # Limit to the minimum number of plots for which an estimate will be calculated
-BS.N <- 200    # Bootstrap iteration number
+BS.N <- 1000    # Bootstrap iteration number
  
 
 # Species number as number

### 2025-09-11 Analysis Code/Data_Prep_Analysis.R
First push with edits/files that will assist with producing the Shiny app.  I expect the app.R file may change dramatically in future revisions.

@@ -151,6 +151,18 @@ arrays.mort <- parse.tree.clim.fcn(tree.mort.dat, clim.var = CLIM.VAR.USE, analy
 
 arrays.grow <- parse.tree.clim.fcn(tree.grow.dat, clim.var = CLIM.VAR.USE, analysis.type = "grow", resp.dat = "growth.val", tot.dat = "growth.n.trees", selected.spp = SEL.SPP, clim.dat = climate.use)
 
+if(file.exists(paste0(DATA.LOC, "analysis.arrays.RDS")) == FALSE){
+  write_rds(list(arrays.mort = arrays.mort, arrays.grow = arrays.grow), file = paste0(DATA.LOC, "analysis.arrays.RDS"))
+  zip(zipfile = file.path(DATA.LOC, "analysis.arrays.zip"), files = file.path(DATA.LOC, "analysis.arrays.RDS"),
+      flags = '-r9Xj' )  # This weird bit keeps the parent directory from being included in the zip folder
+  
+  # Deleting written CSV file (~200 MB), cleaning up after the data zip process. 
+  if (file.exists(file.path(DATA.LOC, "analysis.arrays.RDS")) == TRUE) {
+    file.remove(file.path(DATA.LOC, "analysis.arrays.RDS"))
+  }
+  
+}
+
 
 #### 4) Analysis and plotting  --------------------------------------------------------
 
@@ -379,7 +391,7 @@ for(k in 1:2){ # 1 = growth, 2 = mortality
   
   # Plotting paired plots of mortality/growth by quantile and a scatterplot of plot distribution by quantiles.
   SEL.SPP %>% purrr::map(\(s) pair.plts.fcn(sppnum.to.plot = s, use.dat = plt.dat2, domain.matrix = domain.matrix,
-                                            quant.lims = quant.lims, domain.n = domain.n, k = k))
+                                            quant.lims = quant.lims, domain.n = domain.n, k = k, SHINYAPP.IN.USE = FALSE))
 }
 
 

### 2025-09-11 Analysis Code/Functions.R

@@ -718,7 +718,7 @@ diff.panel.fcn <- function(diff.dat, remove.y, fig.title, lab.right) {
 
 ### Plot of domain estimates for Growth or Mortality ###
 ## Used in pair.plts.fcn below.
-domain.grid.plt.fcn <- function(domains, domain.index, use.dat2, qt.max, q.g.p.labs) {
+domain.grid.plt.fcn <- function(domains, domain.index, use.dat2, qt.max, q.g.p.labs, ylabs) {
   ggplot(data = use.dat2 %>% filter(Domain %in% domain.index), aes(factor(Domain), Means, fill = factor(Domain))) + 
     geom_col() + 
     geom_errorbar(aes(ymax = UCI.95, ymin = LCI.95), width = 0.1) + 
@@ -836,7 +836,7 @@ domain.map.fcn <- function(domain.matrix, spp.num, n.plots.used, virid.use) {
 
 ## Function for plotting the mortality numbers by quantile and the distribution of plots in quantiles
 pair.plts.fcn <- function(sppnum.to.plot, use.dat, domain.matrix,
-                          quant.lims, domain.n, k){
+                          quant.lims, domain.n, k, SHINYAPP.IN.USE){
   
   use.dat2 <- use.dat %>% filter(Species == sppnum.to.plot)
   
@@ -844,8 +844,8 @@ pair.plts.fcn <- function(sppnum.to.plot, use.dat, domain.matrix,
   plot.domain.dat <- use.dat %>% 
     filter(Species == sppnum.to.plot) %>%
     left_join(domain.level.table, by = c("Domain" = "q.num"))
-  
-  # Details for plotting
+
+    # Details for plotting
   virid.use <- viridis_pal(option = "H", begin = 0.1, end = 0.9)(n_domain)  # Get colors for plotting
   
   qt.max <- ceiling(max(use.dat2$UCI.95, na.rm = TRUE) * 10) / 10 # For consistent y-axis heights
@@ -859,13 +859,13 @@ pair.plts.fcn <- function(sppnum.to.plot, use.dat, domain.matrix,
   
   # These will be arranged in a plot below
   if(n_domain == 9){
[…]

### 2025-09-11 Analysis Code/Manuscript_information.R

@@ -113,8 +113,8 @@ for(k in 1:2){
   
   text.size <- 9
   
-  p1 <- domain.grid.plt.fcn(c("DL", "DM", "DH"), 1:3, use.dat2, qt.max, q.g.p.labs) 
-  p2 <- domain.grid.plt.fcn(c("SL", "SM", "SH"), 4:6, use.dat2, qt.max, q.g.p.labs) 
+  p1 <- domain.grid.plt.fcn(c("DL", "DM", "DH"), 1:3, use.dat2, qt.max, q.g.p.labs, ylabs) 
+  p2 <- domain.grid.plt.fcn(c("SL", "SM", "SH"), 4:6, use.dat2, qt.max, q.g.p.labs, ylabs) 
   #p_all <- plot_grid(p1, p2, ncol = 1)
   list1 <- list(p1 = p1, p2 = p2)
   fig.letter <- switch(k, "A", "B")

### 2025-09-11 app.R

@@ -1,0 +1,305 @@
+# Forest Analysis Shiny Application
+# Load required libraries
+library(shiny)
+library(shinydashboard)
+library(DT)
+library(plotly)
+library(readxl)
+library(here)
+
+# Source required files
+source("Global.R")
+source(paste0(CODE.LOC, "Functions.R"))
+
+# Load species names data
+spp_names_full <- readxl::read_xlsx(paste0(DATA.LOC, "FullSppNames.xlsx"))
+
+# Details for plotting
+virid.use <- viridis_pal(option = "H", begin = 0.1, end = 0.9)(n_domain)  # Get colors for plotting
+
+# Create species choice list with readable labels
+species_choices <- spp_names_full %>%
+  filter(SPCD %in% as.numeric(gsub("X", "", SEL.SPP))) %>%
+  mutate(
+    species_code = paste0("X", SPCD),
+    display_name = paste0(GENUS, " ", SPECIES, " (", COMMON_NAME, ")")
+  ) %>%
+  arrange(GENUS, SPECIES) %>%
+  pull(display_name, name = species_code)
+
+# Load analysis arrays once at startup
+load_analysis_data <- function() {
+  # Load the main analysis arrays
+  if (file.exists(paste0(DATA.LOC, "analysis.arrays.zip"))) {
+    temp_file <- unzip(paste0(DATA.LOC, "analysis.arrays.zip"), "analysis.arrays.RDS")
+    analysis_data <- readRDS(temp_file)
+    file.remove(temp_file)  # Clean up
+    return(analysis_data)
+  } else {
+    stop("analysis.arrays.zip file not found in Data/ directory")
+  }
+}
+
+# Pre-load the analysis data
+analysis_arrays <- load_analysis_data()
+
+# UI
+ui <- dashboardPage(
+  dashboardHeader(title = "Forest Climate Response Analysis"),
+  
+  dashboardSidebar(
+    sidebarMenu(
+      menuItem("Analysis", tabName = "analysis", icon = icon("tree"))
+    )
+  ),
+  
+  dashboardBody(
+    tags$head(
+      tags$style(HTML("
+        .content-wrapper, .right-side {
+          background-color: #f4f4f4;
+        }
+        .box {
+          border-radius: 5px;
+        }
+        .selectize-input {
+          font-size: 14px;
+        }
+      "))
+    ),
+    
+    tabItems(
+      tabItem(tabName = "analysis",
+              fluidRow(
+                box(
[…]

### 2025-09-12 Manuscript.Rmd
This push contains a few changes for the manuscript and the working, published v1 of the app.

@@
 Growth may have been susceptible (negatively or positively) to a change in CWD and differed by initial CWD values, and the responses are species-specific.
-*Thuja plicata* exhibited greater growth in the Above and Below Medium domains than in the Lower or Higher domains.
-*Abies amablis* had lower growth in the Above and Below High domains than the others, where initial-visit CWD was the greatest.
-*Picea englelmannii* had lower growth in the Low domain than the High domain, indicating that more growth was occurring where trees were more water stressed.
+*Abies amabilis*, *Abies grandis*, *Calocedrus decurrens*, *Pinus ponderosa*, *Pseudotsuga menziesii*, *Tsuga mertensiana*, *Acer macrophyllum*, and *Alnus rubra* exhibited greater growth in the Low domains than the Medium and/or High domains.
+Except for *Alnus rubra*, these results predominantly occurred in both the Below domain (2) or the Above and Below domains (5).
+*Thuja plicata* had higher growth in the Above and Below Medium domains than either High or Low domains.
+*Abies concolor*, *Larix occidentalis*, *Picea engelmannii*, *Pinus contorta*, *Pinus lambertiana*, *Arbutus menziesii*, and *Quercus kelloggii* had greater growth in the more water stressed High domains than the Medium and/or Low domains.
+Except for *Larix occidentalis*, these results predominantly occurred in both the Above domain (3) or the Above and Below domains (3).
+Other species exhibited no 'significant' findings or displayed a mixture of results.
 
@@
 Below Threshold" comparisons (`r t.ref("tbl_summ_grow_mort")`) than for the comparisons Above or Below the threshold, indicating that mortality may not be as strongly associated with the change in CWD as by the starting value.
+*Larix occidentalis*, *Picea engelmannii*,	*Pinus contorta*,	*Pinus monticola*, *Thuja plicata*,	and *Arbutus menziesii* had greater mortality in the lower water deficit Low domains than Medium and/or High domains.
+These results occurred in Above (3) or both Above and Below (3) domains.
+*Abies lasiocarpa*,	*Calocedrus decurrens*,	*Pinus lambertiana*, *Pinus ponderosa*,	*Pseudotsuga menziesii*,	*Tsuga heterophylla*,	*Tsuga mertensiana*,	*Acer macrophyllum*,	and *Notolithocarpus densiflorus* had more mortality in the High domains than the Medium and/or Low domains.
+A(1111) both() B(11111)
+
 *Abies concolor* had greater mortality at the Above Low and High categories than the Above Middle.

### 2025-09-12 Analysis Code/Data_Prep_Analysis.R

@@ -391,7 +391,8 @@ for(k in 1:2){ # 1 = growth, 2 = mortality
   
   # Plotting paired plots of mortality/growth by quantile and a scatterplot of plot distribution by quantiles.
   SEL.SPP %>% purrr::map(\(s) pair.plts.fcn(sppnum.to.plot = s, use.dat = plt.dat2, domain.matrix = domain.matrix,
-                                            quant.lims = quant.lims, domain.n = domain.n, k = k, SHINYAPP.IN.USE = FALSE))
+                                            quant.lims = quant.lims, domain.n = domain.n, k = k, 
+                                            SHINYAPP.IN.USE = FALSE, SHINYAPP.FONTSIZE = SHINYAPP.FONTSIZE)) # fontsize irrelevant here
 }
 
 

### 2025-09-12 Analysis Code/Functions.R

@@ -718,8 +718,9 @@ diff.panel.fcn <- function(diff.dat, remove.y, fig.title, lab.right) {
 
 ### Plot of domain estimates for Growth or Mortality ###
 ## Used in pair.plts.fcn below.
-domain.grid.plt.fcn <- function(domains, domain.index, use.dat2, qt.max, q.g.p.labs, ylabs) {
-  ggplot(data = use.dat2 %>% filter(Domain %in% domain.index), aes(factor(Domain), Means, fill = factor(Domain))) + 
+domain.grid.plt.fcn <- function(domains, domain.index, use.dat2, qt.max, 
+                                q.g.p.labs, ylabs, text.size) {
+  g1 <- ggplot(data = use.dat2 %>% filter(Domain %in% domain.index), aes(factor(Domain), Means, fill = factor(Domain))) + 
     geom_col() + 
     geom_errorbar(aes(ymax = UCI.95, ymin = LCI.95), width = 0.1) + 
     scale_fill_manual(values = virid.use[domain.index]) + 
@@ -728,7 +729,8 @@ domain.grid.plt.fcn <- function(domains, domain.index, use.dat2, qt.max, q.g.p.l
     theme_bw() +
     theme(legend.position = "none") + 
     labs(x = NULL, y = ylabs) +
-    theme(text = element_text(size = 7))
+    theme(text = element_text(size = text.size))
+  return(g1)
 }
 
 
@@ -740,7 +742,7 @@ domain.grid.plt.fcn <- function(domains, domain.index, use.dat2, qt.max, q.g.p.l
 # for each species, var1, var.delt = carried through from input for pairs.plts.fcn,
 # quant.lims, domain.n = quantile boundaries and number of plots, established in parse.tree.clim.fcn,
 # size.trees = one of "small diameter ", "large diameter ", and (for pathways 1 and 3) "".  
-domain.dist.plt.fcn <- function(plot.spp, domain.matrix, var1, var.delt, quant.lims, n.plots.used, size.trees) {
+domain.dist.plt.fcn <- function(plot.spp, domain.matrix, var1, var.delt, quant.lims, n.plots.used, size.trees, text.size) {
   
   q_plot_spp <- domain.matrix %>% dplyr::select(all_of(plot.spp), all_of(var1), all_of(var.delt)) %>%
     filter(get(plot.spp) > 0) 
[…]

### 2025-09-12 Global.R

@@ -73,7 +73,11 @@ DOMAIN.LEVELS <- c("AL", "AM", "AH", "BL", "BM", "BH") # Above/Below threshold a
 }
 N.PLOT.LIM <- 10  # Limit to the minimum number of plots for which an estimate will be calculated
 BS.N <- 1000    # Bootstrap iteration number
- 
+
+
+## Shiny app settings
+SHINYAPP.IN.USE <- FALSE
+SHINY.FONTSIZE <- 14
 
 # Species number as number
 sel.spp <- as.numeric(gsub("X", "", SEL.SPP))

### 2025-09-12 app.R

@@ -1,171 +1,75 @@
-# Forest Analysis Shiny Application
-# Load required libraries
+# Tree Growth and Mortality Analysis Shiny App
+
 library(shiny)
 library(shinydashboard)
 library(DT)
 library(plotly)
-library(readxl)
-library(here)
 
 # Source required files
 source("Global.R")
 source(paste0(CODE.LOC, "Functions.R"))
 
-# Load species names data
-spp_names_full <- readxl::read_xlsx(paste0(DATA.LOC, "FullSppNames.xlsx"))
-
-# Details for plotting
-virid.use <- viridis_pal(option = "H", begin = 0.1, end = 0.9)(n_domain)  # Get colors for plotting
-
-# Create species choice list with readable labels
-species_choices <- spp_names_full %>%
-  filter(SPCD %in% as.numeric(gsub("X", "", SEL.SPP))) %>%
-  mutate(
-    species_code = paste0("X", SPCD),
-    display_name = paste0(GENUS, " ", SPECIES, " (", COMMON_NAME, ")")
-  ) %>%
-  arrange(GENUS, SPECIES) %>%
-  pull(display_name, name = species_code)
-
-# Load analysis arrays once at startup
-load_analysis_data <- function() {
-  # Load the main analysis arrays
-  if (file.exists(paste0(DATA.LOC, "analysis.arrays.zip"))) {
-    temp_file <- unzip(paste0(DATA.LOC, "analysis.arrays.zip"), "analysis.arrays.RDS")
-    analysis_data <- readRDS(temp_file)
-    file.remove(temp_file)  # Clean up
-    return(analysis_data)
-  } else {
-    stop("analysis.arrays.zip file not found in Data/ directory")
-  }
-}
-
-# Pre-load the analysis data
-analysis_arrays <- load_analysis_data()
-
 # UI
-ui <- dashboardPage(
-  dashboardHeader(title = "Forest Climate Response Analysis"),
+ui <- fluidPage(
+  tags$head(  # Container for logo
+    tags$style(HTML("
+      .logo-container {
+        position: absolute;
+        top: 10px;
+        right: 20px;
+        z-index: 1000;
+      }
+      .logo-container img {
+        transition: opacity 0.3s ease;
+      }
+      .logo-container img:hover {
+        opacity: 0.8;
+      }
+    "))
+  ),
   
-  dashboardSidebar(
-    sidebarMenu(
-      menuItem("Analysis", tabName = "analysis", icon = icon("tree"))
[…]

### 2025-09-20 Manuscript.Rmd
I updated the Manuscript.RMD file and created the first iteration of the permutation test (Permutation_Test.R)

@@
 We estimated differences between pairs of CWD domains by subtracting within-iteration domain point estimates.
+
+We
+
 [permutation analysis]
@@
 The CWD difference tended to be comparatively small, with a maximum 95th percentile value of `r max.q95.diff` (*`r max.q95.diff.spp`*, Table S2).
-The distribution of individual species plot domain values for CWD and change in CWD can be explored on the dashboard().
+The distribution of individual species plot domain values for CWD and change in CWD can be explored on the dashboard (https://tinyurl.com/FIAgrowmort).
 
@@
 These results occurred in Above (3) or both Above and Below (3) domains.
-*Abies lasiocarpa*,	*Calocedrus decurrens*,	*Pinus lambertiana*, *Pinus ponderosa*,	*Pseudotsuga menziesii*,	*Tsuga heterophylla*,	*Tsuga mertensiana*,	*Acer macrophyllum*,	and *Notolithocarpus densiflorus* had more mortality in the High domains than the Medium and/or Low domains.
-A(1111) both() B(11111)
+Mortality rates for *Abies lasiocarpa*,	*Calocedrus decurrens*,	*Pinus lambertiana*, *Pinus ponderosa*,	*Pseudotsuga menziesii*,	*Tsuga heterophylla*,	*Tsuga mertensiana*,	*Acer macrophyllum*,	and *Notolithocarpus densiflorus* were greater in the High domains than the Medium and/or Low domains.
+These species were approximately evenly split between differences being present in the Above (4) or Below (5) domains, but none with 'significant' findings in both the Above and Below domains.
 
-*Abies concolor* had greater mortality at the Above Low and High categories than the Above Middle.
-*Calocedrus decurrens* had a higher mortality rate at the High Above domains than the other Above domains.
-*Picea engelmannii* was the opposite; mortality was lowest at the Above High domain compared to other domains.
-*Psuedotsuga menziesii* exhibited greater mortality at its Below High domain than other Below domains.
-The two *Quercus* species did not exhibit strong patterns to their mortality.
+Mortality: 
+Picea engelmannii: Highest upper mortality = low. 
+Abies lasiocarpa: Lowest mortaity in Low. 
+- delta CWD lower for PIEN than for PILA.  Mean of High Above = 7.5 vs. 9ish.  
+- Similar patterns of delta CWD vs. CWD
+- ABLA = Least mortality in Low, in OLY penninsula, NW Cascades of WA.  High mortality overall (0.22)
+- PIEN = most mort in NE WA. High mortality overall (0.28)
+- PILA = most mort in mid/southern Sierras.  Seems like high mortality (0.20)
+
+ABLA and PIEN are conspecifics / codominants.
+Perret et al.
+2023 estimated rangewide mortality of ABLA of 25% +/-2.7, mort concentrated in eastern OR cascades (and ID and CO).
+83% of mort due to fire or bio disturbance.
+Conspecifics mostly affected by beetles, ABLA by fire.
+PIEN = higher survival and longevity (used to be; they estimate around a high of 1% annual mortality for fir, below 0.5% annually in spruce, PIEN), ABLA = faster growing, more prolific regeneration (Andrus 2018).
+We found that PIEN had higher mortality in the Above Low and Medium domains.
+We note that the Above domains represents a proportionally large increase for Low and Medium, with CWD means of approximately 50 and 80 mm, respectively.
+The same is true for ABLA, however, where the mean for Low is approximately 40 mm yet PIEN has less mortality in the Low domains and PIEN has more mortality in the Above Low domain.
+The difference may be geographical: ABLA low-mortality plots are primarily on the Olympic Peninsula of Washington and the northwestern Cascades, where rainfall is abundant.
+PIEN is generally absent from these locations.
+However, PIEN has the least mortality in the Above domain in the Blue Mountains of Oregon and Washington while ABLA does not show a similar finding for the same geographical region (their AH domains largely overlap and exist only in that mountain range).
+PIEN mortality in N&NE WA may have been driven by spruce beetles (WADNR 2023)
+
+Pinus contorta: Greatest mort in Low, least in High
+Pinus ponderosa: Greatest mort in High
+- In southern Sierra I see that they co-occur but have opposite mortality signals.
+Also a scattering in the Blue Mtns of opposite signals domains.
+Different fire strategies; Keely et al.
+2011 may be useful.
+
+Growth: 
+PIEN: greater growth in High, above and below.  So, greater growth where least mortality.  
+ABLA: not really any difference in growth by domain.  Supposed to be faster growing than PIEN, but not seeing it.
+
+Lowest growth rate is for PIEN BL, at 15 in2.
+Highest growth for ABLA is 12.5 in2.
+ABLA is supposed to be a faster grower (height) but maybe not DIA?
+Diameter growth may be slow (trees 10-20" may be 150-200 yrs; Alexander et al.
+1990).
+On other hand, PIEN appears to have the capability of achieving greater diameters relatively sooner than ABLA (Parish et al.
+1999).
+PIEN growth in the High domains may be related to more growing days relative to other domains.
+
+Communities:
+Pinus lambertiana PILA
+Pinus monticola PIMO
+(Pinus ponderosa) PIPO
+(Pinus contorta) PICO
+
+These two are interesting because the lowest mortality domain for PIMO (Above, High) coincides geographically in the southern Sierra with the highest mortality domain for PILA (and for much of the highest-mortality domain for PIPO).
+PILA has the highest growth in AH whereas there isn't much of a growth story for PIMA.
+PIPO and PICO have great range overlap.
+PICO has the highest mortality in Low (NE WA), particularly AL and lowest mortality in BH (Sierras, OR Cascades).
+Note that the mortality rate for PICO in WA was the highest of any species in any state, and more than twice as high as the mortality rate for WA and 5 times higher than CA.
+PICO growth is lowest in AL.
+PIPO has the highest growth in Low.
+PIPO in NE WA has much less acreage of pine beetle damage in 2023 than PICO, not really a factor for PIMO.
+
+What is happening in the southern Sierras?  N/NE wA?
+
+Pseudotsuga menziesii PSME
+So much data.
+Over 4500 plots.
+Except for High, where mortality of Above and Below was approximately equivalent, we see that mortality was high in AL and AM.
+BL, BM had the lowest mortality.
+These coincided with moister plots.
+Growth was highest for Below except for BH.
+PSME mortality appears to be drought related - see Cline et al.
+2025.
+-Tree mortality increased 2020-2023, prolongued drought and warming.  Soil moisture, temperature, and VPD key predictors of old tree (125-500yr) mortality. 
+-strongest signal at mid-elevation west of Cascade crest, OR
+-Strongest for suppressed/intermediate, non-dominant or codominant individuals, smaller diameter (competition exacerbated the effect of climate variables on smaller-tree mortality at drier sites). 
+-insect mortality more prevalent following drought. 
+- lots of relevant citations
+- Western hemlock (Tsuga heterophylla) in the S/I cohort also died more frequently.  ACMA also did at one site.  
+- PNW heat dome of 2021 caused 11% mortality of C/D trees at Falls Creek site
+
+Acer macrophyllum ACMA
+We found that mortality was higher for BH than BL, BM.
+Maybe more interesting, growth was higher for Low, then Medium, then High irrespective of Above, Below.
+We cannot explain outright the high mortality for ACMA in California.
+However, Betzen et al.
+(2021) found widespread evidence in Washington of ACMA decline (crown dieback, thinning, discoloration, reduced leaf size) and reduced growth and mortality associated with higher temperatures and water limitations.
+These findings are technically in alignment with the overall pattern in growth and mortality we observed, but does not specifically address the very high mortality rate of CA relative to OR and WA.
 
@@
 ::: {#refs}
+
+Betzen, J.J., Ramsey, A., Omdal, D., Ettl, G.J. and Tobin, P.C., 2021. Bigleaf maple, Acer macrophyllum Pursh, decline in western Washington, USA. Forest Ecology and Management, 501, p.119681.
+
 Wang, T., Hamann, A. Spittlehouse, D.L. and Carroll, C. 2016. Locally downscaled and spatially customizable climate data for historical and future periods for North America. PLoS One 11: e0156720
+
+Alexander, R.R., Shearer, R.C. and Shepperd, W.D., 1990. Abies lasiocarpa (Hook.) Nutt. subalpine fir. Silvics of North America, 1, pp.60-70.
+
+Washington State Department of Natural Resources (DNR) Forest Reliance Division.
+2024.
+Forest health highlights in Washington: A summary of insect, disease, and other disturbance conditions affecting Washington’s forests.
+
+Washington State Department of Natural Resources (WADNR).
+2024.
+Forest Health Highlights in Washington – 2024.
+Washington State Department of Natural Resources, Forest Resilience Division, Olympia, WA.
+https://dnr.wa.gov/sites/default/files/2025-03/rp_2023_fhh_report_final_wr.pdf, accessed 9/18/2025
 

### 2025-09-20 Analysis Code/Data_Prep_Analysis.R

@@ -151,8 +151,8 @@ arrays.mort <- parse.tree.clim.fcn(tree.mort.dat, clim.var = CLIM.VAR.USE, analy
 
 arrays.grow <- parse.tree.clim.fcn(tree.grow.dat, clim.var = CLIM.VAR.USE, analysis.type = "grow", resp.dat = "growth.val", tot.dat = "growth.n.trees", selected.spp = SEL.SPP, clim.dat = climate.use)
 
-if(file.exists(paste0(DATA.LOC, "analysis.arrays.RDS")) == FALSE){
-  write_rds(list(arrays.mort = arrays.mort, arrays.grow = arrays.grow), file = paste0(DATA.LOC, "analysis.arrays.RDS"))
+if(file.exists(paste0(DATA.LOC, "analysis.arrays.zip")) == FALSE){
+  write_rds(list(arrays.mort = arrays.mort, arrays.grow = arrays.grow, PlotDat = PlotDat), file = paste0(DATA.LOC, "analysis.arrays.RDS"))
   zip(zipfile = file.path(DATA.LOC, "analysis.arrays.zip"), files = file.path(DATA.LOC, "analysis.arrays.RDS"),
       flags = '-r9Xj' )  # This weird bit keeps the parent directory from being included in the zip folder
   
@@ -208,6 +208,7 @@ tic()
 # For ANALYSIS.PATHWAY == 1, one climate variable, and 16 dedicated cores, the analysis takes about 8 minutes.
 # With 6 domains and 23 species, 16 dedicated cores, 3.8 minutes
 # With 6 domains, 1000 iterations, 23 spp, 16 cores, 15 min.
+# With 6 domains, 1000 iterations, 23 spp, 20 cores, 16 min.
 
 for(k in 1:length(ANALYSIS.TYPE)) {  # 1 = grow, 2 = mortality
   

### 2025-09-20 Analysis Code/Functions.R

@@ -48,29 +48,29 @@ clim.mort.resp.fcn <- function(spcd, clim.var, treedat.sel, clim.dat) {
   
   # Applying quantiles to a particular species
   tree.plots2 <- tree.plots %>% dplyr::select(puid, all_of(c(var1, var.delt))) %>% 
-      {if (n_domain == 9) {
-        mutate(., var.quants = ifelse(!!sym(var1) > quant.lims[2], 1, ifelse(!!sym(var1) < quant.lims[1], -1, 0)),
-               deltvar.quants = ifelse(!!sym(var.delt) > quant.lims.delta[2], 1, ifelse(!!sym(var.delt) < quant.lims.delta[1], -1, 0)),
-               vq = case_match(var.quants, -1 ~ "Li", 0 ~ "Mi", 1 ~ "Hi"),
-               dvq = case_match(deltvar.quants, -1 ~ "Lc", 0 ~ "Mc", 1 ~ "Hc"),
-               var.deltvar = factor(paste0(vq, dvq), levels = DOMAIN.LEVELS)) 
-      } else if (n_domain == 6) {
-        mutate(., var.quants = ifelse(!!sym(var1) > quant.lims[2], 1, ifelse(!!sym(var1) < quant.lims[1], -1, 0)),
-               deltvar.quants = ifelse(!!sym(var.delt) > quant.lims.delta, 1, 0),
-               vq = case_match(var.quants, -1 ~ "L", 0 ~ "M", 1 ~ "H"),
-               dvq = case_match(deltvar.quants, 0 ~ "B", 1 ~ "A"),
-               var.deltvar = factor(paste0(dvq, vq), levels = DOMAIN.LEVELS))
-      }}
-        
-    # Not currently used, but available    
-    # For centroid calcluation
-    dat.cent <- tree.plots2 %>% group_by(var.deltvar) %>% # Quartile centers, good for finding centroids in each quadrant
-      reframe(end.x = mean(get(var1)),        # Referred to as "end" because of distance measurements from the main centroid to each of the outer (end) centroids.
-              end.y = mean(get(var.delt))) %>% 
-      right_join(domain.level.table %>% select(-q.num), by = "var.deltvar") %>%
-      arrange(var.deltvar)
-    
-   
+    {if (n_domain == 9) {
+      mutate(., var.quants = ifelse(!!sym(var1) > quant.lims[2], 1, ifelse(!!sym(var1) < quant.lims[1], -1, 0)),
[…]

### 2025-09-20 Analysis Code/Permutation_Test.R

@@ -1,0 +1,246 @@
+## Monte Carlo permutation analysis ##
+
+source("Global.R")
+source(paste0(CODE.LOC, "Functions.R"))
+
+library(furrr)
+library(parallel)
+library(tictoc)  # For development, timing routines.
+library(data.table)
+library(abind) # for combining matrices into arrays
+library(simctest) # Gandy 2009 permutation reduction approach
+
+## Constants
+
+# Number of maximum MC permutation iterations:
+PERM.ITER.N <- 10
+
+
+
+
+## Loading data to be used.  Assuming analysis.arrays.RDS exist. If not, Data_Prep_Analysis will need to be run.
+if(file.exists(paste0(DATA.LOC, "analysis.arrays.zip")) == TRUE){
+  arrays.use <- read_rds(unzip(paste0(DATA.LOC, "analysis.arrays.zip"), "analysis.arrays.RDS"))
+  if(file.exists("analysis.arrays.RDS")) {
+    file.remove("analysis.arrays.RDS") # Too big for GitHub
+  }  
+} else {
+  print("File analysis.arrays.zip not found")
+}
+
+
+# Number of MC permutation iterations:
+arrays.grow <- arrays.use$arrays.grow
+arrays.mort <- arrays.use$arrays.mort
+PlotDat <- arrays.use$PlotDat
+# BS.N <- 100
+
+
+# Finding the analysis results to pull out the minimum number of significant findings
+#  per contrast. This is used in the Gandy simctest() function at the end.
+contrast.min <- rep(NA, 6)
+for(k in 1:2) {
+plt.dat <- readRDS(paste0(save.loc.fcn(k), "Processed_6Domain_Data.RDS"))
+
+sig.AvB2 <- length(plt.dat$AvB2$Species[plt.dat$AvB2$significant == TRUE])
+sig.A_LMH2 <- length(plt.dat$A_LMH2$Species[plt.dat$A_LMH2$significant == TRUE])
+sig.B_LMH2 <- length(plt.dat$B_LMH2$Species[plt.dat$B_LMH2$significant == TRUE])
+
+contrast.min[(k-1) * 3 + 1:3] <- c(sig.AvB2, sig.A_LMH2, sig.B_LMH2)
+}
+contrast.min.val <- min(contrast.min)                # Minimum number of significant contrasts
+contrast.min.position <- which(contrast.min == min(contrast.min)) # Position of the minimum number (id which contrast)
+
+
+# Need climate names for files and axes.
[…]

### 2025-09-20 Analysis Code/Data_Prep_Analysis.R
Small fix to Data_Prep_Analysis.R and Overall_Mort_est.R

@@ -272,7 +272,9 @@ for(k in 1:length(ANALYSIS.TYPE)) {  # 1 = grow, 2 = mortality
       domain.array = domain.array,
       domain.n = domain.n,
       selected.spp = SEL.SPP,
-      n_iter = BS.N
+      n_iter = BS.N, 
+      strata.num = strata.num,
+      PlotDat = PlotDat
     )
 
     # Finding and saving domain summaries

### 2025-09-20 Analysis Code/Overall_Mort_Est.R

@@ -51,7 +51,9 @@ for(k in 1:2){ # 1 = growth, 2 = mortality
     domain.array = state.array,
     domain.n = state.n,
     selected.spp = SEL.SPP,
-    n_iter = BS.N
+    n_iter = BS.N, 
+    strata.num = strata.num,
+    PlotDat = PlotDat
   )
   
   # Finding and saving domain summaries
@@ -71,7 +73,9 @@ for(k in 1:2){ # 1 = growth, 2 = mortality
     domain.array = all.array,
     domain.n = all.n,
     selected.spp = SEL.SPP,
-    n_iter = BS.N
+    n_iter = BS.N, 
+    strata.num = strata.num,
+    PlotDat = PlotDat
   )
   
   all.species.summaries <- domain.sum.fcn(all.bootstrap_results, 1, domain_n =all.n)

### 2025-09-25 Analysis Code/Data_Prep_Analysis.R
The generate_bootstrap_array.fcn() has been optimized to run much faster.  The Permutation_Test.R, Functions.R, and Data_Prep_Analysis.R have been modified to align with changes to this function.

@@ -36,6 +36,7 @@ library(parallel)
 library(RSQLite) # For obtaining SQLite FIA databases
 library(tictoc)  # For development, timing routines.
 library(data.table)
+library(matrixStats) # For fast iteration processing
 library(abind) # for combining matrices into arrays
 #### 2) Base data prep --------------------------------------------------------
 # These are the base data sets that will be transformed by the analysis types.
@@ -265,7 +266,11 @@ for(k in 1:length(ANALYSIS.TYPE)) {  # 1 = grow, 2 = mortality
 
     ## Generating the bootstrap values: 
    # 200 iterations = 104.99 s, or 1.75 min, or 29.2 hours
-   # 1000 iterations = 426.21 s, or 7.1 min, or 118.3 hrs or 5 days....
+   # 1000 iterations, 10 cores, batches of 100: 50 s
+  # 1000 iterations, 5 cores, batches of 200:  41 s # This appears best
+  # 1000 iterations, 4 cores, batches of 250: 44 s
+  # 1000 iterations, 1 core, batch of 1000: 110 s  
+    #tic()
     bootstrap_results <- generate_bootstrap_array.fcn(
       vals.dat = vals_dat,
       all.dat = all_dat,
@@ -276,7 +281,7 @@ for(k in 1:length(ANALYSIS.TYPE)) {  # 1 = grow, 2 = mortality
       strata.num = strata.num,
       PlotDat = PlotDat
     )
-
+  #toc()
     # Finding and saving domain summaries
     domain.summaries <- domain.index %>% 
       purrr::map(\(d) domain.sum.fcn(bootstrap_results, d, domain_n = domain.n)) %>%

### 2025-09-25 Analysis Code/Functions.R

@@ -469,39 +469,49 @@ mean.q.fcn <- function(dat_tot, dat_vals, spp.sel1) {
 
 
 # Calculate estimates for all species in a given domain using map
+# NOTE: this code sidesteps mean.q.fcn entirely in favor of vectorization.
 calculate_domain_estimates.fcn <- function(samp, domain_data, domain_idx, selected.spp, domain.n) {
   
-  d.all_use <- domain_data[[domain_idx]]$all
-  d.vals_use <- domain_data[[domain_idx]]$vals
+  d.all_use <- domain_data[[domain_idx]]$all[samp$row.id, ]
+  d.vals_use <- domain_data[[domain_idx]]$vals[samp$row.id, ]
   
   # Convert species names to numeric for domain.n lookup
   spp.use.numeric <- as.numeric(gsub("X", "", selected.spp))
+  spp_cols <- paste0("nd.", spp.use.numeric)
   
-  # Use map to calculate estimates for all species
-  estimates <- spp.use.numeric %>% 
-    map_dbl(\(spp.sel) {
-      # Check if we have enough plots for this species in this domain. Otherwise returns NA.
-      spp.id1 <- paste0("X", spp.sel)
-      n_q <- get(spp.id1, domain.n)[domain_idx]
-      
-      if (n_q < N.PLOT.LIM) {
-        return(NA_real_)
-      } else {
-        # return(mean.q.fcn(
-        mean.q.fcn(
-          dat_tot = d.all_use[samp$row.id, ],
-          dat_vals = d.vals_use[samp$row.id, ],
-          spp.sel1 = spp.sel
-        )$R
-      }
-    })
+  # Single data.table conversion and join
+  dt_all <- as.data.table(d.all_use)
+  dt_vals <- as.data.table(d.vals_use)
   
   
-  # df <- rbind(estimates, estimates.yv, estimates.zt) %>% data.frame()
-  #  colnames(df) <- SEL.SPP
-  # write.csv(df, "Oregon.csv")
+  # More efficient: calculate everything in one data.table operation
+  # First, calculate stratum-level summaries for both datasets (weights, # plots)
+  all_stratum <- dt_all[, c(
+    lapply(.SD, sum),
+    list(w_h = w[1], n_h = n_h.plts[1])
+  ), by = stratum, .SDcols = spp_cols]
   
-  return(estimates)
+  # Adding values by stratum for each species
+  vals_stratum <- dt_vals[, lapply(.SD, sum), by = stratum, .SDcols = spp_cols]
+  
[…]

### 2025-09-25 Analysis Code/Permutation_Test.R

@@ -1,19 +1,20 @@
 ## Monte Carlo permutation analysis ##
 
-source("Global.R")
-source(paste0(CODE.LOC, "Functions.R"))
-
 library(furrr)
 library(parallel)
 library(tictoc)  # For development, timing routines.
 library(data.table)
+library(matrixStats) # For fast iteration processing
 library(abind) # for combining matrices into arrays
 library(simctest) # Gandy 2009 permutation reduction approach
+
+source("Global.R")
+source(paste0(CODE.LOC, "Functions.R"))
 
 ## Constants
 
 # Number of maximum MC permutation iterations:
-PERM.ITER.N <- 10
+PERM.ITER.N <- 3
 
 
 
@@ -73,9 +74,11 @@ perm.output <- list(perm.grow = perm.grow, perm.mort = perm.mort)
 
 
 ## Here's the main permutation function.  
-perm.test.fcn <- function() {
-  
-  gandy.out <- c(rep(NA, 6))
+perm.test.fcn <- function(arrays.grow, arrays.mort, PlotDat) {
+  gandy.out <- numeric(6)
+  sig.tests.out <- matrix(NA, nrow = ncol(arrays.grow$domain.n) * 3, ncol = 13 )
+  colnames(sig.tests.out) <- c("Species", "G.AvB.domain", "G.AvB.sig", "G.A_LMH.domain", "G.A_LMH.sig", "G.B_LMH.domain", "G.B_LMH.sig",
+                            "M.AvB.domain", "M.AvB.sig", "M.A_LMH.domain", "M.A_LMH.sig", "M.B_LMH.domain", "M.B_LMH.sig")
   for(k in 1:length(ANALYSIS.TYPE)) {  # 1 = grow, 2 = mortality
     
     # Obtaining the data to work with: Grabbing the list objects from above
@@ -104,14 +107,13 @@ perm.test.fcn <- function() {
     
     # Strata for bootstrap resampling procedure
     strata <- unique(vals_dat$stratum)
-    strata.num <- vals_dat %>% dplyr::select(stratum, puid) %>%
+    strata.num <- vals_dat %>% dplyr::select(stratum, puid, w, n_h.plts) %>%
       ungroup() %>% 
       mutate(row.id = row_number())
     
     ## THIS IS WHERE THE DOMAIN ASSIGNMENTS ARE SAMPLED FOR THE PERMUTATION:
     domain.samp.array <- samp.domain.array.fcn(domain.matrix)
[…]

### 2025-09-25 Global.R

@@ -73,6 +73,8 @@ DOMAIN.LEVELS <- c("AL", "AM", "AH", "BL", "BM", "BH") # Above/Below threshold a
 }
 N.PLOT.LIM <- 10  # Limit to the minimum number of plots for which an estimate will be calculated
 BS.N <- 1000    # Bootstrap iteration number
+BATCH.SIZE <- 200 # Parallel computing batch number of iterations to process per core
+N.WORKERS <- 5
 
 
 ## Shiny app settings
@@ -106,8 +108,8 @@ spp.names <- readxl::read_xlsx(paste0(DATA.LOC, "FullSppNames.xlsx")) %>% dplyr:
 
 
 ## Items for parallel computing
-n.cores <- round(detectCores() * 0.75, 0) # Number of cores, package "parallel".  Using 75% of the available cores, rounded.
-
+#n.cores <- round(detectCores() * 0.75, 0) # Number of cores, package "parallel".  Using 75% of the available cores, rounded.
+n.cores <- N.WORKERS
 
    # State layers
 states <- map_data("state")   # Loaded with ggplot2

### 2025-09-26 Manuscript.Rmd
Some updates to the permutation analysis and the manuscript

@@
 
-What is happening in the southern Sierras?  N/NE wA?
+What is happening in the southern Sierras?
+Southern Sierras: Paper by Restaino et al.
+2019 examined mortality in Sierras.
+2012-2016 CA drought most extreme in past 1000 yrs (citations) - hot drought (high temps, low precip).
+Killed 129m trees in CA, primarily on west slope of Central & S Sierras (citation = USDA 2017).
+In the areas of the greatest mortality, Fettig et al.
+(2019) found tree mortality rates of around 50% with PIPO exhibiting almost 90% mortality rates.
+Mortality was hightest for PIPO at the lowest elevations and attributable to western pine beetles, and concentrated in the largest diameter trees.
+PILA was the second highest at 48%, and killed by Mt pine beetle.
+ABCO and CADE exhibited 26% and 23% mortality.
+Only one Quercus died.
+Restaino et al.
+2019 also found PIPO to experience high mortality followed by PILA, ABCO, and CADE.
+QUKE had 1% mortality (191 trees total, presumably 1 or 2 died?).
+Large PIPO were most likely to die, esp in dense stands and dry areas.
+For PILA mortality positively affected by stand BA size.
+CADE survived much better in wetter sites.
+ABCO mortality greater at dry sites, particularly for larger stands.
+In our study we had no evidence of low mortality rates for QUKE - about 20%.
+We had some evidence of high mortality for ABCO in the Sierras, although not overwhelming.
+CADE definitely had the highest mortality in the southern Sierras.
+PIMO appears to be doing great...why?
+
+N/NE wA?  
+McNellis et al.
+(2015) used FIA plot data to obtain mortality estimates but did not obtain high values for N/NE WA.
+Stanke (2021) shows a low forest stability index for PICO in that region.
+ABLA has a moderately negative index in that region.
+PIEN roughly aligns.
+PIPO has agreeably low mortality for that region.
+PSME is high relative to coastal trees, although the differences are not great (maybe why Stanke thinks they're doing fine).
+THPL is doing pretty well in NE WA; note that there is a call of alarm for the species (Andrus et al.
+2023), even in that region.
 

### 2025-09-26 Analysis Code/Functions.R

@@ -1039,12 +1039,12 @@ state.table.fcn <- function(state.dat, all.dat, sig.places) {
                   n.plts = as.integer(n.plts),
                   results = case_when(
                     is.na(Median) ~ "", 
-                    grepl("NA", paste0(Means, " (", LCI.95, ",", UCI.95, "; ", n.plts, ")")) ~ "",
-                    !is.na(Median) ~ paste0(Means, " (", LCI.95, ",", UCI.95, "; ", n.plts, ")")
+                    grepl("NA", paste0(Means, " (", LCI.95, ", ", UCI.95, "; ", n.plts, ")")) ~ "",
+                    !is.na(Median) ~ paste0(Means, " (", LCI.95, ", ", UCI.95, "; ", n.plts, ")")
                   ))%>%
     dplyr::mutate(results = case_when(is.na(Median) ~ "", 
                                       Median == "NA" ~ "",
-                                      !is.na(Median) ~ paste0(Means, " (", LCI.95, ",", UCI.95, "; ", n.plts, ")"))) %>%
+                                      !is.na(Median) ~ paste0(Means, " (", LCI.95, ", ", UCI.95, "; ", n.plts, ")"))) %>%
     dplyr::select(sci_name, State, results) %>%
     tidyr::pivot_wider(names_from = State, values_from = results) %>%
     dplyr::rename("California" = "CA", "Washington" = "WA", "Oregon" = "OR") %>%
@@ -1056,8 +1056,8 @@ state.table.fcn <- function(state.dat, all.dat, sig.places) {
                   n.plts = as.integer(n.plts),
                   results = case_when(
                     is.na(Median) ~ "", 
-                    grepl("NA", paste0(Means, " (", LCI.95, ",", UCI.95, "; ", n.plts, ")")) ~ "",
-                    !is.na(Median) ~ paste0(Means, " (", LCI.95, ",", UCI.95, "; ", n.plts, ")")
+                    grepl("NA", paste0(Means, " (", LCI.95, ", ", UCI.95, "; ", n.plts, ")")) ~ "",
+                    !is.na(Median) ~ paste0(Means, " (", LCI.95, ", ", UCI.95, "; ", n.plts, ")")
                   )) %>%
     dplyr::select(sci_name, results) %>%
     dplyr::rename("All" = "results")

### 2025-09-26 Analysis Code/Permutation_Test.R

@@ -240,7 +240,7 @@ plan(list(
 #print(plan('list'))
 
 tic()
-x <- furrr::future_map(1:n.output, function(i) {
+x <- furrr::future_map(1:total.output, function(i) {
   J <-i 
   y <- perm.test.fcn(arrays.grow = arrays.grow,
                      arrays.mort = arrays.mort, 
@@ -252,6 +252,44 @@ toc()
 
 
 plan(sequential)  # This closes parallel workers
+
+
+
+## Code for opening previously-run permutation tests
+x <- read_rds(paste0(DATA.LOC, "perm_test_out_092625.RDS"))
+
+gandy_table <- data.frame(do.call(rbind, lapply(x, function(xi) xi$gandy.out)))
+
+n.sto <- length(x)
+sto.rows <- nrow(x[[1]]$sig.tests.out)
+sto.cols <- ncol(x[[1]]$sig.tests.out)
+
+sig_tests_table <- map_dfr(seq_along(x), ~ {
+  as.data.frame(x[[.x]]$sig.tests.out) %>%
+    mutate(id = .x)
+})
+
+
+sig_array <- array(dim = c(sto.rows, sto.cols, n.sto))
+
+# Fill each slice with the corresponding matrix
+for(i in 1:n.sto) {
+  sig_array[, , i] <- x[[i]]$sig.tests.out
+}
+
+
+odd_cols <- seq(3, 13, by = 2)  # Creates c(3, 5, 7, 9, 11, 13)
+
+sig_array2 <- sig_array[, odd_cols, ]
+  
+sig_array2 <- apply(sig_array2, c(1, 2, 3), as.numeric)
+
+sig_array_prop <- apply(sig_array2, c(1, 2), function(x) mean(x == 1, na.rm = TRUE))
+
+# Convert to data frame
+proportions_df <- as.data.frame(proportions)
+colnames(proportions_df) <- paste0("col_", odd_cols)
+
 
 #tic()
 #perm_results <- 1:PERM.ITER.N %>% 

### 2025-09-30 Analysis Code/Functions.R
Changed the permutation analysis to output tables of instances where significant permutation contrasts have CIs at least as far from zero as test contrasts.

@@ -681,6 +681,45 @@ diff.fig.prep.fcn <- function(diff.table, diff.levels){
 
 ### MC Permutation analysis functions --------------------------------------
 
+sig.dist.fcn <- function(data.set){
+  sig.data.set <- data.set %>% mutate(
+    sig.dist = case_when(
+      LCI.95 > 0 & Means > 0 ~ LCI.95,
+      UCI.95 < 0 & Means < 0 ~ UCI.95,
+      LCI.95 < 0 & Means > 0 ~ 0,
+      UCI.95 > 0 & Means < 0 ~ 0,
+      .default = NA)
+  ) %>% dplyr::select(Domain, sig.dist)
+  return(sig.data.set)
+}    
+
+
+# Function for finding the distance from zero a CI is if significant.
+sig.dist.fcn2 <- function(data.set, k.level){
+  domain.col <- paste0("Domain", k.level)
+  sig.col <- paste0("sig.dist", k.level)
+  gt.col <- paste0("abs.gt", k.level)
+  sig.data.set <- data.set %>% mutate(
+    sig.dist = case_when(
+      LCI.95 > 0 & Means > 0 ~ LCI.95,
+      UCI.95 < 0 & Means < 0 ~ UCI.95,
+      LCI.95 < 0 & Means > 0 ~ 0,
+      UCI.95 > 0 & Means < 0 ~ 0,
+      .default = NA)
+    ) %>% left_join(sig.dist %>% dplyr::select(Species, order.c, all_of(domain.col), all_of(sig.col)), by = c("Species", "Domain" = domain.col)) %>%
+    arrange(order.c) %>%
+    mutate(
+      sig.temp = get(sig.col),
+      abs.gt = ifelse(abs(sig.dist) >= abs(sig.temp) & sig.temp != 0, 1, 0)) %>%
+    dplyr::select(Species, order.c, Domain, abs.gt) %>%
+    rename("{domain.col}" := "Domain",
+           "{gt.col}" := "abs.gt")
+    return(sig.data.set)
+}
+
+
+
+
 # Function to randomly permute non-zero values within a column
 permute_nonzero <- function(x) {
   nonzero_idx <- which(x != 0) # Row numbers where values != 0

### 2025-09-30 Analysis Code/Permutation_Test.R

@@ -2,6 +2,7 @@
 
 library(furrr)
 library(parallel)
+library(future) # Involved in parallel computing
 library(tictoc)  # For development, timing routines.
 library(data.table)
 library(matrixStats) # For fast iteration processing
@@ -10,13 +11,6 @@ library(simctest) # Gandy 2009 permutation reduction approach
 
 source("Global.R")
 source(paste0(CODE.LOC, "Functions.R"))
-
-## Constants
-
-# Number of maximum MC permutation iterations:
-PERM.ITER.N <- 3
-
-
 
 
 ## Loading data to be used.  Assuming analysis.arrays.RDS exist. If not, Data_Prep_Analysis will need to be run.
@@ -38,19 +32,44 @@ PlotDat <- arrays.use$PlotDat
 
 
 # Finding the analysis results to pull out the minimum number of significant findings
-#  per contrast. This is used in the Gandy simctest() function at the end.
+#  per contrast and the distance of the CI from zero.
+sig.dist <- NULL
 contrast.min <- rep(NA, 6)
 for(k in 1:2) {
 plt.dat <- readRDS(paste0(save.loc.fcn(k), "Processed_6Domain_Data.RDS"))
 
+# Counting the number of sig findings by column for k (growth, mortality)
 sig.AvB2 <- length(plt.dat$AvB2$Species[plt.dat$AvB2$significant == TRUE])
 sig.A_LMH2 <- length(plt.dat$A_LMH2$Species[plt.dat$A_LMH2$significant == TRUE])
 sig.B_LMH2 <- length(plt.dat$B_LMH2$Species[plt.dat$B_LMH2$significant == TRUE])
 
 contrast.min[(k-1) * 3 + 1:3] <- c(sig.AvB2, sig.A_LMH2, sig.B_LMH2)
+
+# Finding the CI distance from zero by column for k (growth, mortality)
+sig.dist.AvB <- sig.dist.fcn(plt.dat$AvB)
+sig.dist.A_LMH <- sig.dist.fcn(plt.dat$A_LMH)
+sig.dist.B_LMH <- sig.dist.fcn(plt.dat$B_LMH)
+
+sig.dist.k <- data.frame(sig.dist.AvB, sig.dist.A_LMH, sig.dist.B_LMH) %>%
+  mutate(order.c = seq_along(1:n())) %>%
+  relocate(order.c)
+if(is.null(sig.dist)){
[…]

### 2025-10-02 Manuscript.Rmd
Corrected errors in the app and manuscript tables and figures - growth values were in in2, not cm2.  Corrected version of manuscript included in push.

@@
 # State growth
-s.g.dat <- state.table.fcn(grow.state, grow.all, 1)
+grow.state.cm <- cm2.fcn(1, grow.state)
+grow.all.cm <- cm2.fcn(1, grow.all)
+s.g.dat <- state.table.fcn(grow.state.cm, grow.all.cm, 1)
 
@@
 
-Growth domain comparisons (`r fig_nums("fig_diffgrow")`, `r t.ref("tbl_summ_grow_mort")`) did not appear to be produce substantially more "significant" negative vs.
-positive estimates, and all three columns of `r fig_nums("fig_diffgrow")` had similar numbers of "significant" comparisons.
+Growth domain comparisons (Figure `r fig_nums("fig_diffgrow", display = "num")`, `r t.ref("tbl_summ_grow_mort")`) did not appear to be produce substantially more "significant" negative vs.
+positive estimates, and all three columns of Figure `r fig_nums("fig_diffgrow", display = "num")` had similar numbers of "significant" comparisons.
 Growth may have been susceptible (negatively or positively) to a change in CWD and differed by initial CWD values, and the responses are species-specific.
@@
 
-Mortality domain comparisons are similar in certain regards to the growth comparisons: results for the three types of domain comparisons did not appear to be predominantly positive or negative (`r fig_nums("fig_diffmort")`, `r t.ref("tbl_summ_grow_mort")`).
+Mortality domain comparisons are similar in certain regards to the growth comparisons: results for the three types of domain comparisons did not appear to be predominantly positive or negative (Figure `r fig_nums("fig_diffmort", display = "num")`, `r t.ref("tbl_summ_grow_mort")`).
 There were fewer mortality estimates with "significant" findings than growth.
@@
 These species were approximately evenly split between differences being present in the Above (4) or Below (5) domains, but none with 'significant' findings in both the Above and Below domains.
+
+(Section on geography?
+Growth highest for Abies grandis in BL, which coincides with coast range for OR, WA, CA plus Olympics.
+CA really only has trees in BL; these probably are the ones that grew the most.
+Then, mention some mortality patterns observed below, especially N/NE WA and Sierra Nevadas)
+
+(Thoughts about permutation analysis): The number of significant findings per contrast was unexpected by chance except for Mortality, Above vs.
+Below.
+We examined the likelihood of each of the significant comparisons arising by chance, adjusted by the number of significant findings (120).
+Results were considered significant if their permutation p-value was below 0.00041 (0.05/120).
+We might need to run 10,000 permutations.
 
@@
 PICO has the highest mortality in Low (NE WA), particularly AL and lowest mortality in BH (Sierras, OR Cascades).
-Note that the mortality rate for PICO in WA was the highest of any species in any state, and more than twice as high as the mortality rate for WA and 5 times higher than CA.
+Note that the mortality rate for PICO in WA was the highest of any species in any state, and more than twice as high as the mortality rate for OR and 5 times higher than CA.
 PICO growth is lowest in AL.

### 2025-10-02 Analysis Code/Functions.R

@@ -985,7 +985,7 @@ domain.map.fcn <- function(domain.matrix, spp.num, n.plots.used, virid.use, poin
 
 ## Function for plotting the mortality numbers by quantile and the distribution of plots in quantiles
 pair.plts.fcn <- function(sppnum.to.plot, use.dat, domain.matrix,
-                          quant.lims, domain.n, k, SHINYAPP.IN.USE, SHINY.FONTSIZE){
+                          quant.lims, domain.n, k, SHINYAPP.IN.USE, SHINY.FONTSIZE, cm = TRUE){
   
   use.dat2 <- use.dat %>% filter(Species == sppnum.to.plot)
   
@@ -1003,7 +1003,13 @@ pair.plts.fcn <- function(sppnum.to.plot, use.dat, domain.matrix,
   # Create a grid of plots to match bivariate plot
   q.g.p.labs <- paste0(DOMAIN.LEVELS, ", n = ", as.numeric(use.dat2$n.plts)) # Labels for the domain grid plot
   
-  ylabs <- if (k == 1) "Mean Growth (inches^2)" else "Mean Annual Mortality Rate"
+  ylabs <- if (k == 1 & cm == FALSE) {
+    "Mean Decadal Growth (inches^2)"
+  } else if (k == 1 & cm == TRUE) {
+    "Mean Decadal Growth (cm^2)"
+  } else {
+    "Mean Decadal Mortality Rate"
+  }
   
   text_size <- if(SHINYAPP.IN.USE == TRUE) SHINY.FONTSIZE else 7
   point_size <- if(SHINYAPP.IN.USE == TRUE) 1 else 0.5

### 2025-10-02 Analysis Code/Manuscript_information.R

@@ -1,4 +1,4 @@
-## This code provides inforamtion summaries for the manuscript and generates 
+## This code provides information summaries for the manuscript and generates 
 #   non-estimate figures and (maybe) tables.
 #   The file Data_Prep_Analysis.R sources this code.  
 #   The file Manuscript.Rmd will open and use its outputs.
@@ -86,6 +86,17 @@ for(k in 1:2){
   plt.dat <- readRDS(paste0(save.loc.fcn(k), "Domain_Analysis_Output.RDS"))
   plt.dat2 <- plt.dat$domain.summaries
   
+  # Need climate names for files and axes.
+  clim.names <- read_csv(paste0(DATA.LOC, "ClimateNames.csv"), show_col_types = FALSE) %>%
+    filter(filename == CLIM.VAR.USE)
+  
+  var1 <- clim.names$values[grep("pre", clim.names$values)]
+  var.delt <- clim.names$values[grep("d", clim.names$values)]
+  var.label <- clim.names$label[clim.names$values == var1]
+  #var.filename <- clim.names$filename[clim.names$values == var1]
+  var.delt.label <- clim.names$label[clim.names$values == var.delt]
+  
+  
   sppnum.to.plot <- "X81"
   mort.grow.dat <- get(paste0("arrays.", ANALYSIS.TYPE[k])) 
   
@@ -96,6 +107,7 @@ for(k in 1:2){
   
   
   use.dat2 <- plt.dat2 %>% filter(Species == sppnum.to.plot)
+  use.dat2 <- cm2.fcn(k, use.dat2)  # Transforming growth from inches2 to cm2
   
   plot.domain.dat <- plt.dat2 %>% 
     filter(Species == sppnum.to.plot) %>%
@@ -113,8 +125,8 @@ for(k in 1:2){
   
   text.size <- 9
   
-  p1 <- domain.grid.plt.fcn(c("DL", "DM", "DH"), 1:3, use.dat2, qt.max, q.g.p.labs, ylabs) 
-  p2 <- domain.grid.plt.fcn(c("SL", "SM", "SH"), 4:6, use.dat2, qt.max, q.g.p.labs, ylabs) 
+  p1 <- domain.grid.plt.fcn(c("DL", "DM", "DH"), 1:3, use.dat2, qt.max, q.g.p.labs, ylabs, text.size) 
+  p2 <- domain.grid.plt.fcn(c("SL", "SM", "SH"), 4:6, use.dat2, qt.max, q.g.p.labs, ylabs, text.size) 
[…]

### 2025-10-02 app.R

@@ -184,7 +184,8 @@ server <- function(input, output, session) {
         plt_dat <- analysis_data$mort_output
       }
       
-      plt_dat2 <- plt_dat$domain.summaries
+#      browser()
+      plt_dat2 <- cm2.fcn(k, plt_dat$domain.summaries) # Transforming growth data to cm2
       domain.matrix <- mort_grow_dat$domain.matrix
       domain.n <- mort_grow_dat$domain.n
       quant.lims <- mort_grow_dat$quant.lims
@@ -193,7 +194,7 @@ server <- function(input, output, session) {
       virid.use <<- viridis_pal(option = "H", begin = 0.1, end = 0.9)(n_domain)  # Get colors for plotting
       
       
-    # browser()
+     #browser()
       # Generate the plot using pair.plts.fcn with SHINYAPP.IN.USE = TRUE
       plot_result <- pair.plts.fcn(
         sppnum.to.plot = input$species, 

### 2025-10-06 Manuscript.Rmd
The permutation analysis may be in its final form.  The figures 3 and 4 were altered to show MC permutation p-value symbols, and tables were added to the supplemental material to display the actual permuatation p-values.

@@
 This approach allows for nonsymmetrical CIs around a mean and prevents CIs from including negative values.
-We estimated differences between pairs of CWD domains by subtracting within-iteration domain point estimates.
+We estimated contrasts between pairs of CWD domains by subtracting within-iteration domain point estimates.
 
-We
+Our analysis approach results in many comparisons.
+We conducted a Monte Carlo (MC) permutation analysis to provide species-level false positive rates.
+The MC permutation analysis had 1000 iterations; for each iteration the bootstrap sampling was conducted 1000 times.
+The domain assignments for each species was sampled without replacement.
+The distance of contrast CI bounds from zero were compared against test CI bounds: The permutation recorded a null significant value if the absolute values of the permutation iteration CI bounds were at least as far from zero as observed bounds were.
+The false detection rate for a
 
@@
 
-```{r diff_grow_fig, fig.height = 6, width = 6, fig.dpi=300, fig.cap=fig.diff_grow, fig.cap.sep=""}
+```{r diff_grow_fig, fig.height = 6.5, width = 6.5, fig.dpi=300, fig.cap=fig.diff_grow, fig.cap.sep=""}
 
@@
 
-```{r diff_mort_fig, fig.height = 6, width = 6, fig.dpi=300, fig.cap=fig.diff_mort, fig.cap.sep=""}
+```{r diff_mort_fig, fig.height = 6.5, width = 6.5, fig.dpi=300, fig.cap=fig.diff_mort, fig.cap.sep=""}
 

### 2025-10-06 Supplemental Material.Rmd
The permutation analysis may be in its final form.  The figures 3 and 4 were altered to show MC permutation p-value symbols, and tables were added to the supplemental material to display the actual permuatation p-values.

@@
 source('Global.R')
+source(paste0(CODE.LOC, "Functions.R"))
 # Some general analysis info + plots
@@
 
+```{r domain_pvals}
+
+psig.dat2 <- analysis.stats$psig.dat2
+psig.g <- psig.dat2[, 1:7]
+psig.m <- psig.dat2[, c(1, 8:13)]
+
+psig.ft.g <- psig.ft.fcn(psig.g)
+psig.ft.m <- psig.ft.fcn(psig.m)
+
+```
+
 ```{r table-legends}
@@
 The number of species' plots (n), minimum, mean, median, and maximum CWD values are provided, along with the 5%, 25%, 75%, and 95% quantiles (q).")
+
+tbl_psig_growth <- tab_nums(name = "tbl_psiggrowth", 
+                            caption = "Monte Carlo permutation p-values for tree species growth domains.
+Values correspond to Manuscript Figure 3 values.
+Values of zero represented no permutation results with confidence intervals equal to or greater than those observed in the data.")
+
+tbl_psig_mort <- tab_nums(name = "tbl_psigmort", 
+                            caption = "Monte Carlo permutation p-values for tree species mortality domains.
+Values correspond to Manuscript Figure 4 values.
+Values of zero represented no permutation results with confidence intervals equal to or greater than those observed in the data.")
 
@@
 
+\newpage
+
+```{r table-psig-growth, tab.cap=tbl_psig_growth}
+
+psig.ft.g
+
+```
+
+\newpage
+
+```{r table-psig-mortality, tab.cap=tbl_psig_mort}
+
+psig.ft.m
+
+```
+
 <!---BLOCK_LANDSCAPE_STOP--->

### 2025-10-06 Analysis Code/Data_Prep_Analysis.R

@@ -352,31 +352,76 @@ if(RUN.SUMMARY == TRUE) {
 # Plotting the multi-species panels
 
 if(n_domain == 6) {
-  for(k in 1:2){ # 1 = growth, 2 = mortality
+  plt.dat <- readRDS(paste0(save.loc.fcn(k), "Processed_6Domain_Data.RDS"))
+  psig.dat <- read_csv(paste0(RESULTS.OTHER, "Permutation_results.csv"), show_col_types = FALSE)
+
+    for(k in 1:2){ # 1 = growth, 2 = mortality
+      font_to_use <- if("Times New Roman" %in% systemfonts::system_fonts()$family) {
+      "Times New Roman"
+    } else if("Times" %in% systemfonts::system_fonts()$family) {
+      "Times"
+    } else {
+      "serif"  # Fallback to default serif
+    }
     
-    plt.dat <- readRDS(paste0(save.loc.fcn(k), "Processed_6Domain_Data.RDS"))
     
-    AvB2 <- cm2.fcn(k, results.table = plt.dat$AvB2)
-    A_LMH2 <- cm2.fcn(k, results.table = plt.dat$A_LMH2)
-    B_LMH2 <- cm2.fcn(k, results.table = plt.dat$B_LMH2)
+    psig.join.fcn <- function(k, position) {
+      pos.adj <- 3*(k-1) + position
+      psig.cols <- grep(pos.adj, colnames(psig.dat))
+      psig.spp <- grep("Species", colnames(psig.dat))
+      join.table <- psig.dat[, c(psig.spp, psig.cols)] 
+      names(join.table) <- c("Species", "Domain", "sig.dist")
+      join.table <- join.table %>% mutate(
+        symb = case_when(
+          is.na(sig.dist) == TRUE ~ "",
+          sig.dist > 0.05 ~ "--",
+          sig.dist > 0.01 & sig.dist <= 0.05 ~ "*",
+          sig.dist > 0.001 & sig.dist <= 0.01 ~ "**",
+          sig.dist > 0 & sig.dist <= 0.001 ~ "***",
+          sig.dist == 0 ~ "****",
+          .default = "XX"
+        ),
+        yadj = ifelse(symb == "--", 0.075, -0.11)
+      )
+      return(join.table)
+    }
+    
[…]

### 2025-10-06 Analysis Code/Functions.R

@@ -782,6 +782,20 @@ sig.n.fcn <- function(t.contrast) {
 }
 
 
+# For the output: Sum of the permuation runs with significant findings exceeding the 
+# observed number of significant findings.
+nsig.comp.fcn <- function(sig.num, col.id, dat.nsig){
+  dat.nsig[, col.id] <- ifelse(dat.nsig[, col.id] >= sig.num, 1, 0)
+  return(sum(dat.nsig[, col.id]))
+}
+
+
+
+
+
+
+
+
 
 
 
@@ -801,8 +815,19 @@ cm2.fcn <- function(k, results.table){
 
 
 ## This plot provides the 
-diff.panel.fcn <- function(diff.dat, remove.y, fig.title, lab.right) {
+diff.panel.fcn <- function(diff.dat, remove.y, fig.title, lab.right, ft_to_use = font_to_use) {
   
+  # Position for astrices 
+  diff.dat <- diff.dat %>% mutate(
+    lab.x = case_when(
+      k == 1 & LCI.95 > 0 ~ -100,
+      k == 1 & LCI.95 < 0 ~ 100,
+      k == 2 & LCI.95 > 0 ~ -0.25,
+      k == 2 & LCI.95 < 0 ~ 0.25
+    )
+  )
+  
+
   # Kludgy way to get a conifer/deciduous break.  X312 = bigleaf maple.
   decid.break <- length(SEL.SPP) - which(SEL.SPP == "X312") + 1.5
   
@@ -817,36 +842,32 @@ diff.panel.fcn <- function(diff.dat, remove.y, fig.title, lab.right) {
                    fill = ifelse(significant, "black", "white"),
                    color = Domain),
                size = 2, shape = 21, stroke = 0.8) +
+    geom_text(aes(x = lab.x, 
+                  y = as.numeric(species_label) + y.offset + yadj,
+                  label = symb), size = 6) + 
     geom_hline(yintercept = decid.break, linetype = 2) +
     #scale_y_continuous(breaks = 1:length(levels(DvS2$species_label)),
     #                    labels = levels(DvS2$species_label)) +
     scale_y_continuous(breaks = 1:length(levels(diff.dat$species_label)),
                        labels = levels(diff.dat$species_label)) +
-    scale_color_viridis(discrete = TRUE, name = "Domain", option = "B", begin = 0.2, end = 0.6) +
+    scale_color_viridis(discrete = TRUE, name = NULL, option = "B", begin = 0.2, end = 0.6) +
[…]

### 2025-10-06 Analysis Code/Manuscript_information.R

@@ -53,6 +53,18 @@ diff.cwd.90range.min <- diff.range[which(diff.range == min(diff.range))]
 diff.cwd.90range.max <- diff.range[which(diff.range == max(diff.range))]
 diff.cwd.95max <- which(diff.cwd.stats$q95 == max(diff.cwd.stats$q95))
 
+# For supplemental, MC permutation p-values
+psig.dat <- read_csv(paste0(RESULTS.OTHER, "Permutation_results.csv"), show_col_types = FALSE)
+psig.dat2 <- psig.dat %>% dplyr::select(-order.c) %>%
+  left_join(gym.angio.dat %>% dplyr::select(SpeciesCode, sci_name, SFTWD_HRDWD), by = c("Species" = "SpeciesCode")) %>%
+  arrange(desc(SFTWD_HRDWD), sci_name) %>%
+  dplyr::select(-SFTWD_HRDWD, -Species) %>%
+  relocate(sci_name)
+
+num.cols <- grep("sig.dist", names(psig.dat2) )
+
+psig.dat2[, num.cols] <- apply(psig.dat2[, num.cols], 2, as.character)
+psig.dat2[is.na(psig.dat2)] <- ""
 
 analysis.stats <- list(tree.num = tree.num,
                        plot.num = plot.num,
@@ -64,6 +76,7 @@ analysis.stats <- list(tree.num = tree.num,
                        n.gym.spp = n.gym.spp,
                        init.cwd.stats = init.cwd.stats,
                        diff.cwd.stats = diff.cwd.stats,
+                       psig.dat2 = psig.dat2,
                        init.cwd.min.mean = init.cwd.min.mean, 
                        init.cwd.max.mean = init.cwd.max.mean, 
                        init.cwd.90range.min = init.cwd.90range.min,

### 2025-10-06 Analysis Code/Permutation_Test.R

@@ -241,19 +241,19 @@ perm.test.fcn <- function(arrays.grow, arrays.mort, PlotDat) {
 #plan(5, 4) BATCH.SIZE = 200 = 305 s, 5 outputs, 5.08 min, 61 s/iteration
 
 # plan(10, 2), batch size 500 = 1048.78 s, 20 outputs, 17.48 min, 52.45 sec/iteration. 14.6 hrs on my machine for 1000 outputs
-
+# I ran plan(10, 2), batch size = 500 for 500 outputs on my laptop. 7 hrs.  
 
 
 
 ## With 76 cores...
 #  I recommend trying BATCH.SIZE = 500, n.output = 38, n.process = 2:
-n.output <- 38   # Number of output per cycle
-n.process <- 76/n.output
+n.output <- 10   # Number of output per cycle
+n.process <- 20/n.output
 BATCH.SIZE <- 1000/n.process
 # I estimate your machine will complete the task in about 4 hours.
 
 
-total.output <- 1000  # Maybe try with 10 first?  There will be some warnings about unused cores, I think.
+total.output <- 500  # Maybe try with 10 first?  There will be some warnings about unused cores, I think.
 
 #n.output <- 4
 #n.process <- 20/n.output
@@ -278,45 +278,95 @@ x <- furrr::future_map(1:total.output, function(i) {
 }, .options = furrr_options(seed = TRUE))
 toc()
 
-
 plan(sequential)  # This closes parallel workers
 
 
-
-## Code for opening previously-run permutation tests
-x <- read_rds(paste0(DATA.LOC, "perm_test_out_092625.RDS"))
-
-gandy_table <- data.frame(do.call(rbind, lapply(x, function(xi) xi$gandy.out)))
-
-n.sto <- length(x)
-sto.rows <- nrow(x[[1]]$sig.tests.out)
-sto.cols <- ncol(x[[1]]$sig.tests.out)
-
-sig_tests_table <- map_dfr(seq_along(x), ~ {
-  as.data.frame(x[[.x]]$sig.tests.out) %>%
-    mutate(id = .x)
-})
+## Code for opening and combining previously-run permutation tests
+x1 <- read_rds(paste0(DATA.LOC, "perm_test_093025.RDS"))
+x2 <- read_rds(paste0(DATA.LOC, "perm_test_out500_20251001.RDS"))
[…]

### 2025-10-06 Global.R

@@ -11,13 +11,21 @@ library(tidyverse)
 library(here)         # File path management
 library(furrr)
 library(parallel)
+library(extrafont)
 # Generation of maps
 library(sf)
 library(ggspatial)
 library(maps)
 
 
-
+#library(extrafont)
+#font_import()
+#loadfonts(device = "win")
+#loadfonts(device = "postscript")
+#library(Cairo)
+#cairo_ps("test.eps", family = "Times")
+library(ragg) # Related to fonts
+library(systemfonts)
 ##### ---- Constants ---- #####
 
 # Code controls:

### 2025-10-13 Manuscript.Rmd
Updates to MS.  Added current Intro and Discussion material, cleaned up figure legends for Figs 3 & 4

@@
 Mean domain differences (circles) with 95% CIs (bars) that do not intersect zero are filled.
-Species above the dashed line are coniferous; those below are deciduous."))
+Species above the dashed line are coniferous; those below are deciduous.
+Empirical MC permutation p-values are provided for filled-circle contrasts, with \u2013\u2013 corresponding to p-values of > 0.05, while one to four asterisks correspond to p > 0.01, > 0.001, > 0.0001 and < 0.0001, respectively."))
 
@@
 Mean domain differences (circles) with 95% CIs (bars) that do not intersect zero are filled.
-Species above the dashed line are coniferous; those below are deciduous."))
+Species above the dashed line are coniferous; those below are deciduous.
+Empirical MC permutation p-values are provided for filled-circle contrasts, with \u2013\u2013 corresponding to p-values of > 0.05 while one to four asterisks correspond to p > 0.01, > 0.001, > 0.0001 and < 0.0001, respectively.
+"))
 
@@
 # Introduction
+
+There  are signs that tree species in western North America are in regional decline.
+Johnston et al.
+(2025) document a 29% 10-year mortality rate of 300+ yr old trees in Eastern Oregon.
+High-severity fires in California in 2020 and 2021 killed 13%-19% of giant sequoia (Sequoiadendron gianteum) with >1.07-m bole diameters (Stephenson and Brigham 2021, Shive et al.
+2021); recruitment in these burns was depressed (Soderberg et al.
+2022).
+Five of eight common species in western North America exhibit declines in relative density  as measured by relative stand density (Stanke et al.
+2021).
+Shriver et al.
+(2022) found that four of five juniper and piñon pine species were declining in portions of their range, comparing mortality to recruitment rates.
+
+Climate change as manifest through increases or intensification of drought and aridity can affect tree mortality directly through drought and indirectly by increasing wildfire frequency and severity and promoting insect and disease outbreak.
+Hammond et al.
+(2022) found globally that the intensity of drought and heat events are directly associated with widespread forest die-off events.
+Wildfire frequency and area burned in the western United States has increased, in concert with the longer fire seasons, hotter summer temperatures, and earlier snowmelt increasingly experienced in the western United States (Abatzoglou and Williams 2016, Westerling 2016, Juang et al.
+2022).
+There has also been a regional increase in the area of bark beetle outbreaks (Meddens et al.
+2012, Hicke et al.
+2016).
+Many of these climate-influence mortality causes may be attributable to the physical condition of trees following abnormal or prolonged exposure to water stress.
+
+Drought-induced mortality can be driven by carbon starvation and hydraulic failure (see reviews by Adams et al.
+2010 and McDowell et al.
+2008).
+Carbon starvation occurs when isohydric species (e.g., the piñon pine group), or those which mainly rely on stomatal closure to avoid excessive water loss through evapotranspiration (ET) receive insufficient atmospheric carbon dioxide for photosynthesis.
+Anisohydric plants, or those who do not strongly regulate water loss through stomatal closure, risk hydraulic failure if sufficient xylem cavitation occurs due to insufficient water uptake needed to meet ET demands (e.g., as seen in Populus tremuloides; Anderegg et al.
+2015).
+Adams et al.
+(2017) performed a review of drought-induced mortality across 23 species and concluded that hydraulic failure occurred broadly across taxa while non-structural carbohydrate reserves at mortality differed across taxa.
+Regardless of mechanism, drought (specifically, water deficit) appears to be an important direct driver of mortality for tree taxa (Hammond et al.
+2022.
+Repetitive citation.
+Others?).
+
+Drought and physiological stress from drought can indirectly increase mortality as well.
+Drought stress can increase tree susceptibility to wildfire, insects, and disease.
+Wildfire fire area (Abatzoglou and Williams, 2016; Parks et al.
+2025) and number (Westerling 2016) are positively associated with increasing aridity in the western US.
+Bark beetle outbreaks in semi-arid areas of the western US may be positively affected by or associated with acute intense drought and water-stressed trees (Negrón 2009, Gaylord et al.
+2015) but chronically stressed trees may not provide like bark beetle populations with the resources sufficient to promote population expansion (Huberty and Denno, 2004).
+Norway spruce (Picea abies) saplings inoculated with the necrotrophic pathogenic fungi Heterobasidion annosum sensu stricto and exposed to drought conditions were more likely to die than control saplings or saplings exposed to either drought conditions or inoculation, although Scots pine (Pinus sylvestris) seedlings did not show a similar relationship (Gomez-Gallego et al.
+2022).
+Other insects and disease may be promoted or inhibited by drought, depending on the circumstances and due to conflicting factors (Kolb et al.
+2016).
 
@@
 We obtained empirical means and 95% confidence intervals via bootstrap.
-For every species we sampled plots with replacement and obtained point estimates.
-For each domain we calculated a mean and upper and lower 95% CIs with the 2.5% and 97.5% quantiles from 1000 sampling iterations.
-This approach allows for nonsymmetrical CIs around a mean and prevents CIs from including negative values.
+For each of the 1000 bootstrap iterations we sampled FIA plots with replacement and obtained point estimates for each species' domain.
+For each domain we calculated a mean and upper and lower 95% CIs with the 2.5% and 97.5% quantiles from the  sampling iterations.
+This approach allows for nonsymmetrical CIs around means.
 We estimated contrasts between pairs of CWD domains by subtracting within-iteration domain point estimates.
@@
 Our analysis approach results in many comparisons.
-We conducted a Monte Carlo (MC) permutation analysis to provide species-level false positive rates.
-The MC permutation analysis had 1000 iterations; for each iteration the bootstrap sampling was conducted 1000 times.
+We conducted a Monte Carlo (MC) permutation analysis to provide species-level false positive rates (p-values).
+The MC permutation analysis had 1000 iterations; each iteration relied on the bootstrap confidence interval estimation described above.
 The domain assignments for each species was sampled without replacement.
-The distance of contrast CI bounds from zero were compared against test CI bounds: The permutation recorded a null significant value if the absolute values of the permutation iteration CI bounds were at least as far from zero as observed bounds were.
-The false detection rate for a
-
-[permutation analysis]
+The distance of MC permutation contrast CI bounds from zero were compared against test CI bounds.
+The MC permutation procedure recorded a null significant value if the absolute values of the permutation iteration CI bounds were at least as far from zero as observed bounds.
 
@@
 
-(Thoughts about permutation analysis): The number of significant findings per contrast was unexpected by chance except for Mortality, Above vs.
-Below.
-We examined the likelihood of each of the significant comparisons arising by chance, adjusted by the number of significant findings (120).
-Results were considered significant if their permutation p-value was below 0.00041 (0.05/120).
-We might need to run 10,000 permutations.
+The MC permutation analysis provided empirical p-values for observing contrast confidence intervals with bounds at least as far from zero as those observed (Supplemental figures S3 and S4), with values ranging from $p$ = 0.072 to $p$ < 0.0001 (41 instances).
 
-Mortality: 
-Picea engelmannii: Highest upper mortality = low. 
-Abies lasiocarpa: Lowest mortaity in Low. 
-- delta CWD lower for PIEN than for PILA.  Mean of High Above = 7.5 vs. 9ish.  
-- Similar patterns of delta CWD vs. CWD
-- ABLA = Least mortality in Low, in OLY penninsula, NW Cascades of WA.  High mortality overall (0.22)
-- PIEN = most mort in NE WA. High mortality overall (0.28)
-- PILA = most mort in mid/southern Sierras.  Seems like high mortality (0.20)
+# Discussion 
+We detected numerous species for whom growth and/or mortality rates differed by CWD category.
+We also found species for whom growth and mortality differed by state.
+Our observed changes in CWD were not large, but regionally, from southern California to the Olympic Peninsula in Washington, CWD increased in 9 out of 10 plots.
 
-ABLA and PIEN are conspecifics / codominants.
+Our regional probabilistically sampled design-based results provide a potentially valuable perspective on species status.
+Acer macrophyllum growth was negatively associated with initial CWD, and its growth was lower and mortality almost three times higher in California than Washington and Oregon.
+We cannot find primary literature that identifies an increase in mortality for this species in California; the degree of mortality observed seems severe.
+However, Betzen et al.
+(2021) found widespread evidence in Washington of A.
+macrophyllum decline (crown dieback, thinning, discoloration, reduced leaf size) and reduced growth and increased mortality in areas associated with higher temperatures and water limitations.
+Although these results align with our CWD findings for growth, they are not reflective (yet) of our state findings and are not apparent in our CWD findings for mortality.
+We encourage a more intensive assessment of A.
+macrophyllum’s status and causes for decline in California.
+
+A converse example is the mortality rate of Thuja plicata, where researchers and state agencies are concerned about die-off but our analysis did not find a strong mortality signal.
+Andrus et al.
+(2023) sampled T.
+plicata at nine locations in Washington, northern Oregon, and Idaho that were without canopy dieback, exhibited canopy dieback, or had died.
+They found that the mortality was associated with exceptionally warm and dry climate conditions.
+A collaboration of state and federal agencies is currently reaching out to the public to identify dieback sites for this species (website: Goodrich et al.
+2025).
+Our analysis found that T.
+plicata exhibited the lowest decadal mortality rate and third highest growth rate of any species we examined and the greatest growth in Medium domains and some evidence that mortality in the Above Low domain (Olympic peninsula, western Washington Cascades) was greater than Above High domain (NE Washington, southern Oregon Cascades).
+Our analysis may not be temporally sensitive enough to capture the same recent findings as Andrus et al., given that our FIA plot revisit data represented plot visits between 2011 to 2023.
+However, our analysis does indicate that numerous species are exhibiting mortality rates at a state level or overall that are higher and are potentially unsustainable (e.g., Abies grandis, Abies lasiocarpa, Picea engelmannii, Pinus contorta, Pinus monticola, and Alnus rubra have state or overall decadal mortality rates greater than 0.25).
+
+Abies lasiocarpa and Picea engelmannii are conspecifics and frequently codominant.
 Perret et al.
-2023 estimated rangewide mortality of ABLA of 25% +/-2.7, mort concentrated in eastern OR cascades (and ID and CO).
-83% of mort due to fire or bio disturbance.
-Conspecifics mostly affected by beetles, ABLA by fire.
-PIEN = higher survival and longevity (used to be; they estimate around a high of 1% annual mortality for fir, below 0.5% annually in spruce, PIEN), ABLA = faster growing, more prolific regeneration (Andrus 2018).
-We found that PIEN had higher mortality in the Above Low and Medium domains.
-We note that the Above domains represents a proportionally large increase for Low and Medium, with CWD means of approximately 50 and 80 mm, respectively.
-The same is true for ABLA, however, where the mean for Low is approximately 40 mm yet PIEN has less mortality in the Low domains and PIEN has more mortality in the Above Low domain.
-The difference may be geographical: ABLA low-mortality plots are primarily on the Olympic Peninsula of Washington and the northwestern Cascades, where rainfall is abundant.
-PIEN is generally absent from these locations.
-However, PIEN has the least mortality in the Above domain in the Blue Mountains of Oregon and Washington while ABLA does not show a similar finding for the same geographical region (their AH domains largely overlap and exist only in that mountain range).
-PIEN mortality in N&NE WA may have been driven by spruce beetles (WADNR 2023)
+(2023) estimated a rangewide mortality of A.
+lasiocarpa of 25% +/-2.7, with mortality concentrated in the eastern Oregon cascades as well as in Idaho and Colorado.
+83% of the mortality was due to fire or biogenic disturbance.
+These mortality rates are comparable to the state-level ones we found.
+Andrus et al.
+(2018) found that P.
+engelmannii exhibited higher survival and longevity in undisturbed plots in Colorado, with firs exhibiting mortalities of up to 1% annually while P.
+engelmannii had mortality rates below 0.5% annually.
+We found that P.
+engelmannii had higher mortality in the Above Low domain.
+We note that the Above domains represent a proportionally large increase for Low and Medium given initial CWD values.
+The same is true for A.
+lasiocarpa, yet P.
+engelmannii has greater mortality in the Low domains and A.
+lasiocarpa has less mortality in the Below Low domain although the MC permutation p-values were not very small.
+The difference may be geographical: A.
+lasiocarpa low-mortality plots are primarily on the Olympic Peninsula of Washington and the northwestern Cascades, where rainfall is abundant.
+P.
+engelmannii is generally absent from these locations.
+However, P.
+engelmannii has the least mortality in the Above domain in the Blue Mountains of Oregon and Washington while A.
+lasiocarpa does not show a similar finding for the same geographical region.
+P.
+engelmannii mortality in northern and northeastern Washington, WA may have been driven by spruce beetles (WADNR 2023).
+Our mortality estimate for Pinus contorta in Washington was twice that of Oregon’s and five times greater than Calfornia’s, while the growth confidence intervals for Washignton were lower than California’s and did not overlap.
+Mortality was highest in the Low CWD domains and growth lowest in the Above Low domain.
+The Above Low domain aligns well with plots for northern and northeastern Washington.
+Pine bark beetles (Dendroctonus ponderosae, Dendroctonus brevicomis, and Ips spp.) have affected this region for the past decade (WADNR 2023).
+This area corresponds with the portions of P.
+contorta’s range in the lowest forest stability index category for Stanke et al.
+(2021) while California’s Sierra Nevadas and the Oregon and Washington Cascade Range correspond to positive stability index values.
+Stanke et al.
+also found that insects (followed by fire) were a major driver of the percent change in relative tree density.
 
-Pinus contorta: Greatest mort in Low, least in High
-Pinus ponderosa: Greatest mort in High
-- In southern Sierra I see that they co-occur but have opposite mortality signals.
-Also a scattering in the Blue Mtns of opposite signals domains.
-Different fire strategies; Keely et al.
-2011 may be useful.
+Our analysis centered around initial CWD and delta CWD as a variable of interest relative to tree growth and mortality.
+There is ample evidence in the literature supporting the relationship or correlation of CWD or its cousin VPD to tree growth (Dyer 2004, Restaino et al.
+2016) and mortality (Anderegg et al.
+2015, Das et al.
+2013, Hammond et al.
+2022).
+However, there are some challenges in relying on it in a design-based regional analysis.
+We did not detect sizable changes in CWD between visits although CWD disproportionately increased between visits across sites.
+The changes we observed may have represented greater proportional change for some species than others, with species such as Abies amablis, Alnus rubra, and Juniperus occidentalis exhibiting a low initial CWD range.
 
-Growth: 
-PIEN: greater growth in High, above and below.  So, greater growth where least mortality.  
-ABLA: not really any difference in growth by domain.  Supposed to be faster growing than PIEN, but not seeing it.
-
-Lowest growth rate is for PIEN BL, at 15 in2.
-Highest growth for ABLA is 12.5 in2.
-ABLA is supposed to be a faster grower (height) but maybe not DIA?
-Diameter growth may be slow (trees 10-20" may be 150-200 yrs; Alexander et al.
-1990).
-On other hand, PIEN appears to have the capability of achieving greater diameters relatively sooner than ABLA (Parish et al.
-1999).
-PIEN growth in the High domains may be related to more growing days relative to other domains.
-
-Communities:
-Pinus lambertiana PILA
-Pinus monticola PIMO
-(Pinus ponderosa) PIPO
-(Pinus contorta) PICO
-
-These two are interesting because the lowest mortality domain for PIMO (Above, High) coincides geographically in the southern Sierra with the highest mortality domain for PILA (and for much of the highest-mortality domain for PIPO).
-PILA has the highest growth in AH whereas there isn't much of a growth story for PIMA.
-PIPO and PICO have great range overlap.
-PICO has the highest mortality in Low (NE WA), particularly AL and lowest mortality in BH (Sierras, OR Cascades).
-Note that the mortality rate for PICO in WA was the highest of any species in any state, and more than twice as high as the mortality rate for OR and 5 times higher than CA.
-PICO growth is lowest in AL.
-PIPO has the highest growth in Low.
-PIPO in NE WA has much less acreage of pine beetle damage in 2023 than PICO, not really a factor for PIMO.
-
-What is happening in the southern Sierras?
-Southern Sierras: Paper by Restaino et al.
-2019 examined mortality in Sierras.
-2012-2016 CA drought most extreme in past 1000 yrs (citations) - hot drought (high temps, low precip).
-Killed 129m trees in CA, primarily on west slope of Central & S Sierras (citation = USDA 2017).
-In the areas of the greatest mortality, Fettig et al.
-(2019) found tree mortality rates of around 50% with PIPO exhibiting almost 90% mortality rates.
-Mortality was hightest for PIPO at the lowest elevations and attributable to western pine beetles, and concentrated in the largest diameter trees.
-PILA was the second highest at 48%, and killed by Mt pine beetle.
-ABCO and CADE exhibited 26% and 23% mortality.
-Only one Quercus died.
-Restaino et al.
-2019 also found PIPO to experience high mortality followed by PILA, ABCO, and CADE.
-QUKE had 1% mortality (191 trees total, presumably 1 or 2 died?).
-Large PIPO were most likely to die, esp in dense stands and dry areas.
-For PILA mortality positively affected by stand BA size.
-CADE survived much better in wetter sites.
-ABCO mortality greater at dry sites, particularly for larger stands.
-In our study we had no evidence of low mortality rates for QUKE - about 20%.
-We had some evidence of high mortality for ABCO in the Sierras, although not overwhelming.
-CADE definitely had the highest mortality in the southern Sierras.
-PIMO appears to be doing great...why?
-
-N/NE wA?  
-McNellis et al.
-(2015) used FIA plot data to obtain mortality estimates but did not obtain high values for N/NE WA.
-Stanke (2021) shows a low forest stability index for PICO in that region.
-ABLA has a moderately negative index in that region.
-PIEN roughly aligns.
-PIPO has agreeably low mortality for that region.
-PSME is high relative to coastal trees, although the differences are not great (maybe why Stanke thinks they're doing fine).
-THPL is doing pretty well in NE WA; note that there is a call of alarm for the species (Andrus et al.
-2023), even in that region.
-
-Pseudotsuga menziesii PSME
-So much data.
-Over 4500 plots.
-Except for High, where mortality of Above and Below was approximately equivalent, we see that mortality was high in AL and AM.
-BL, BM had the lowest mortality.
-These coincided with moister plots.
-Growth was highest for Below except for BH.
-PSME mortality appears to be drought related - see Cline et al.
-2025.
--Tree mortality increased 2020-2023, prolongued drought and warming.  Soil moisture, temperature, and VPD key predictors of old tree (125-500yr) mortality. 
--strongest signal at mid-elevation west of Cascade crest, OR
--Strongest for suppressed/intermediate, non-dominant or codominant individuals, smaller diameter (competition exacerbated the effect of climate variables on smaller-tree mortality at drier sites). 
--insect mortality more prevalent following drought. 
-- lots of relevant citations
-- Western hemlock (Tsuga heterophylla) in the S/I cohort also died more frequently.  ACMA also did at one site.  
-- PNW heat dome of 2021 caused 11% mortality of C/D trees at Falls Creek site
-
-Acer macrophyllum ACMA
-We found that mortality was higher for BH than BL, BM.
-Maybe more interesting, growth was higher for Low, then Medium, then High irrespective of Above, Below.
-We cannot explain outright the high mortality for ACMA in California.
-However, Betzen et al.
-(2021) found widespread evidence in Washington of ACMA decline (crown dieback, thinning, discoloration, reduced leaf size) and reduced growth and mortality associated with higher temperatures and water limitations.
-These findings are technically in alignment with the overall pattern in growth and mortality we observed, but does not specifically address the very high mortality rate of CA relative to OR and WA.
+Interpreting CWD results at a regional scale becomes difficult because water stress alone can kill trees (Anderegg et al.
+2015, Maherali et al.
+2004, and van Nieuwstadt and Sheil 2005), but it can also increase the probability of insect (Hart et al.
+2014, Preisler et al.
+2012 , examples and counterexamples in Huberty and Denno 2004) and wildfire-caused mortality (Abatzoglou and Williams 2016).
+With reference to Manion’s (1991) “death spiral” (see also Bennett et al.
+2023 for a CWD-focused version), teasing apart the cause of mortality from predisposing, inciting, and contributing factors is difficult at best, as these factors can interact (Gaylord et al.
+2015).
+Interrelated predisposing conditions for individual tree mortality such as canopy openness (Lemmo et al.
+2024), structural complexity (Ma et al.
+2023), tree height (Stovall et al.
+2019), and within-stand competition (Ford et al.
+2017) would be missed by our regional evaluation.
 

### 2025-11-03 Manuscript.Rmd
Latest version of manuscript, added CWD analysis to supplemental materials.

@@
 
+# Correlation/fits for Initial CWD vs Delta CWD:
+CWD.rels <- read_rds(paste0(RESULTS.OTHER, "CWD_deltaCWD_relationship.rds"))
+CWD.lin.cor <- CWD.rels$lm1$r.squared^0.5 
+CWD.dAIC <- round(CWD.rels$aic.cwd$AIC[1] - CWD.rels$aic.cwd$AIC[2], 0)
+
 # Growth data for species overall, states, differences
@@
 
+grow.tot.sig <- grow.AvB.sig + grow.A.sig + grow.B.sig
+
 # Significant findings, Mort
@@
 mort.B.signeg <- length(mort.diff$B_LMH2$significant[mort.diff$B_LMH2$significant == TRUE & mort.diff$B_LMH2$Means < 0])
+
+mort.tot.sig <- mort.AvB.sig + mort.A.sig + mort.B.sig
 
@@
 fig.cwdmap_cap <- fig_nums(name = "fig_cwdmap", 
-                        caption = "Fuzzed FIA plot locations displayed with initial visit climatic water deficit (CWD) values (A) and CWD change between visits (C) for the U.S.
+                        caption = "Fuzzed FIA plot locations displayed with initial visit CWD values (A) and change in CWD between visits (C) for the U.S.
 states, north to south, of Washington, Oregon, and California (see inset map B for the location of these states within the USA).")
@@
 E.g., 'AL' is Above CWD change threshold, Low initial CWD value.
-The distribution of plot CWD change and initial values (C) is presented as well, with the blue horizontal line at a change in CWD of 5 mm representing the change threshold and the green vertical lines representing the quantile boundaries between Low, Medium, and High initial CWD values." )
+The distribution of plot CWD change and initial values (C) is presented as well, with the blue horizontal line indicating the CWD change threshold and the orange vertical lines representing the quantile boundaries between Low, Medium, and High initial CWD values." )
 
 fig.diff_grow <- fig_nums(name = "fig_diffgrow", 
-                          caption = paste0("Differences in mean growth rates (cm\u00B2/decade) between domains by species.  Domain abbreviations are described in the legend for ",
+                          caption = paste0("Differences in mean growth rates (cm\u00B2/decade) between domains by species.  Domain abbreviations are described in the legend for Figure ",
                                            f.ref("fig_domains"), 
@@
 fig.diff_mort <- fig_nums(name = "fig_diffmort", 
-                          caption = paste0("Differences in mean decadal mortality rates between domains by species.  Domain abbreviations are described in the legend for ",
+                          caption = paste0("Differences in mean decadal mortality rates between domains by species.  Domain abbreviations are described in the legend for Figure ",
                                            f.ref("fig_domains"), 
@@
 tab.summ_grow_mort <- tab_nums(name = "tbl_summ_grow_mort",
-                               caption = paste0("Summary of number of comparisons with 95% CIs that do not overlap with zero ([T]otal, [P]ositive-, and [N]egative-valued estimates) for Figures ", fig_nums("fig_diffgrow", display = "num"), " and ", fig_nums("fig_diffmort", display = "num"), " by analysis (T)ype (Growth or Mortality) and type of (C)omparison.
+                               caption = paste0("Summary of number of contrasts with 95% CIs that do not overlap with zero ([T]otal, [P]ositive-, and [N]egative-valued estimates) for Figures ", fig_nums("fig_diffgrow", display = "num"), " and ", fig_nums("fig_diffmort", display = "num"), " by analysis (T)ype (Growth or Mortality) and type of (C)omparison.
 The column Domains lists the domains compared; All is the sum of values within a Comparison.
@@
 (2025) document a 29% 10-year mortality rate of 300+ yr old trees in Eastern Oregon.
-High-severity fires in California in 2020 and 2021 killed 13%-19% of giant sequoia (Sequoiadendron gianteum) with >1.07-m bole diameters (Stephenson and Brigham 2021, Shive et al.
+Peret et al.
+(2022) estimated a decadal range-wide mortality rate of 25% for Subalpine fir (*Picea engelmannii*).
+High-severity fires in California in 2020 and 2021 killed 13%-19% of giant sequoia (*Sequoiadendron gianteum*) with >1.07-m bole diameters (Stephenson and Brigham 2021, Shive et al.
 2021); recruitment in these burns was depressed (Soderberg et al.
 2022).
-Five of eight common species in western North America exhibit declines in relative density  as measured by relative stand density (Stanke et al.
+Five of eight common species in western North America exhibit declines as measured by relative stand density (Stanke et al.
 2021).
 Shriver et al.
-(2022) found that four of five juniper and piñon pine species were declining in portions of their range, comparing mortality to recruitment rates.
+(2022) found that four of five juniper (*Juniperus* spp.) and piñon pine (*Pinus* spp.) species were declining in portions of their range, comparing mortality to recruitment rates.
 
-Climate change as manifest through increases or intensification of drought and aridity can affect tree mortality directly through drought and indirectly by increasing wildfire frequency and severity and promoting insect and disease outbreak.
+Climate change as manifest through increases or intensification in drought and aridity can affect tree mortality directly through physiological damage and indirectly by increasing wildfire frequency and severity and by promoting insect and disease outbreaks.
 Hammond et al.
 (2022) found globally that the intensity of drought and heat events are directly associated with widespread forest die-off events.
-Wildfire frequency and area burned in the western United States has increased, in concert with the longer fire seasons, hotter summer temperatures, and earlier snowmelt increasingly experienced in the western United States (Abatzoglou and Williams 2016, Westerling 2016, Juang et al.
+Wildfire frequency and area burned in the western United States has increased, in concert with the longer fire seasons, hotter summer temperatures, and earlier snowmelt (Abatzoglou and Williams 2016, Westerling 2016, Juang et al.
 2022).
-There has also been a regional increase in the area of bark beetle outbreaks (Meddens et al.
+There has also been a regional increase in the area of bark beetle (subfamily *Scolytinae* spp.) outbreaks (Meddens et al.
 2012, Hicke et al.
 2016).
-Many of these climate-influence mortality causes may be attributable to the physical condition of trees following abnormal or prolonged exposure to water stress.
+Many of these indirect climate-influenced mortality causes may be attributable to the physical condition of trees following abnormal or prolonged exposure to water stress.
 
@@
 2008).
-Carbon starvation occurs when isohydric species (e.g., the piñon pine group), or those which mainly rely on stomatal closure to avoid excessive water loss through evapotranspiration (ET) receive insufficient atmospheric carbon dioxide for photosynthesis.
-Anisohydric plants, or those who do not strongly regulate water loss through stomatal closure, risk hydraulic failure if sufficient xylem cavitation occurs due to insufficient water uptake needed to meet ET demands (e.g., as seen in Populus tremuloides; Anderegg et al.
+Carbon starvation occurs when isohydric species (e.g., piñon pine, *Pinus edulus*; Breshears et al.
+2009), or those which mainly rely on stomatal closure to avoid excessive water loss through evapotranspiration (ET), receive insufficient atmospheric carbon dioxide for photosynthesis.
+Anisohydric plants, or those who do not strongly regulate water loss through stomatal closure, risk hydraulic failure if sufficient xylem cavitation occurs due to insufficient water uptake needed to meet ET demands (e.g., as seen in *Populus tremuloides*; Anderegg et al.
 2015).
 Adams et al.
-(2017) performed a review of drought-induced mortality across 23 species and concluded that hydraulic failure occurred broadly across taxa while non-structural carbohydrate reserves at mortality differed across taxa.
+(2017) performed a review of drought-induced mortality across 26 species and concluded that hydraulic failure occurred broadly across taxa while non-structural carbohydrate reserves at mortality differed across taxa.
 Regardless of mechanism, drought (specifically, water deficit) appears to be an important direct driver of mortality for tree taxa (Hammond et al.
-2022.
-Repetitive citation.
-Others?).
+2022).
 
-Drought and physiological stress from drought can indirectly increase mortality as well.
 Drought stress can increase tree susceptibility to wildfire, insects, and disease.
-Wildfire fire area (Abatzoglou and Williams, 2016; Parks et al.
+Wildfire area (Abatzoglou and Williams, 2016; Parks et al.
 2025) and number (Westerling 2016) are positively associated with increasing aridity in the western US.
 Bark beetle outbreaks in semi-arid areas of the western US may be positively affected by or associated with acute intense drought and water-stressed trees (Negrón 2009, Gaylord et al.
-2015) but chronically stressed trees may not provide like bark beetle populations with the resources sufficient to promote population expansion (Huberty and Denno, 2004).
-Norway spruce (Picea abies) saplings inoculated with the necrotrophic pathogenic fungi Heterobasidion annosum sensu stricto and exposed to drought conditions were more likely to die than control saplings or saplings exposed to either drought conditions or inoculation, although Scots pine (Pinus sylvestris) seedlings did not show a similar relationship (Gomez-Gallego et al.
+2015) but chronically stressed trees may not provide bark beetle populations with the resources sufficient to promote population expansion (Huberty and Denno, 2004).
+Norway spruce (*Picea abies*) saplings inoculated with the necrotrophic pathogenic fungi *Heterobasidion annosum* sensu stricto and exposed to drought conditions were more likely to die than control saplings or saplings exposed to either drought conditions or inoculation, although Scots pine (*Pinus sylvestris*) seedlings did not show a similar relationship (Gomez-Gallego et al.
 2022).
@@
 2016).
+
+One regionally useful metric of water stress is Climatic Water Deficit (CWD; Stephenson 1998) which appears to be a reasonable covariate for examining tree growth and mortality.
+Stephenson (1990) defines CWD as evaporative demand not met by available water, or the difference between plant water demand (actual evapotranspiration, AET), and plant water supply (potential evapotranspiration, PET); these variables are driven by temperature and precipitation, respectively.
+CWD is geographically readily available from downscaled PRISM data (Daly et al.
+2008, Wang et al.
+2016).
+The Anderegg  et al.
+(2015) tree mortality study relied on CWD to capture drought occurrence.
+Das et al.
+(2013) found low-elevation California tree mortality best modeled using CWD.
+Restaino et al.
+(2016) found decreased Douglass fir growth with an increase in CWD.
+
+Data from the U.S.
+National Forest Inventory, collected by the US Forest Service Forest Inventory and Analysis Program (FIA), is well suited for evaluating regional species mortality and growth.
+Since 2001 the FIA program has gathered a variety of forestry data at plots along the west coast of the USA using a spatially balanced probability sample (McRoberts 2005, Reams et al.
+2005).
+The sample allows for a relatively unbaised design-based inference of population characteristics across a large geographical region.
+Data from plot revisits and individual tree data permit direct estimation of change features such as mortality and growth estimation.
+
+Monleon and Linz (2015) evaluated the evidence of tree species range shifts by subtracting the design-based mean estimates of annual FIA plot temperatures for seedlings from those of trees.
+FIA plot revisit data were mostly incomplete at the time of their analysis.
+Groom and Monleon (2023) used the revisit data to compare plot climate variable values (vapor pressure deficit [VPD], temperature, and precipitation) to estimate shifts in plot occupancy and density.
+Both studies used a single climate variable value associated with each plot for estimation; e.g., Groom and Monleon (2023) examined shifts in species density by estimating the difference in VPD means between plots with more trees of a given species and plots with fewer trees.
+They documented, however, that the initial climate variable values could (and often did) change between plot visits, with species often exhibiting positive or negative correlations between the initial visit and between-visits change value of the climate variable.
+They identified that their design-based analysis could potentially improve by including both initial visit and change values for the climate variable.
+
+The purpose of this analysis is to obtain design-based estimates of tree species’ growth and mortality across the western U.S.
+states of California, Oregon, and Washington as they relate to plot-level initial-visit water stress (Climatic Water Deficit, initial CWD) and the change in water stress (delta CWD) over time.
+We approached this bivariate analysis by dividing species’ plot data into domains based on initial and delta CWD values, finding growth and mortality estimates for each domain, and estimating domain contrasts to establish differences between domains.
+Our secondary goal includes providing decadal growth and mortality estimates for our analysis species overall and by state to establish species status and to inform our CWD analysis findings.
 
@@
 
-We used U.S.
-National Forest Inventory datasets for our analyses of forestland in California, Oregon, and Washington, USA.
+Our analysis relies on FIA data for forestland in California, Oregon, and Washington, USA.
 Forestland is defined as an area of 4,050 m$^2$ that is at least 10 percent stocked or potentially stocked with tree species, excluding urban and agricultural land uses (Reams et al., 2005).
-Additionally, as our study relied on plot revisit data, we only examined plots that were considered forestland during both visits.
+We constrained our analysis to only include remeasurements on areas considered forestland during both visits.
 
-FIA plots consist of four points where all trees with stem diameters greater than or equal to 12.7 cm (for most species measured at 1.37 m above the ground) were tallied in a 7.32 m radius circular subplot (total area 672.5 m$^2$) with additional trees sampled in an anulus around each subplot (macroplot) around each subplot, with a total radius of 18 m from the center point and a total area of 1012 m$^2$.
+FIA plots consist of four points where all trees with stem diameters greater than or equal to 12.7 cm (for most species measured at 1.37 m above the ground) were tallied in a 7.32 m radius circular subplot (total area 672.5 m$^2$) with additional trees sampled in an anulus around each subplot (macroplot), with a total radius 18 m from the center point and a total area of 1012 m$^2$.
 We did not include data from FIA microplots.
@@
 
-Since 2000, FIA plots have been selected via a spatially balanced probability sample (McRoberts 2005, Reams et al., 2005) with a target plot density of one ground plot every 24 km$^2$.
+Since 2000, FIA plots have been selected via spatially-balanced probability sampling (McRoberts 2005, Reams et al., 2005) with a target plot density of one ground plot every 24 km$^2$.
 Some regions and ownerships intensify the plot density; we excluded these additional plots to simplify stratification estimation, relying on FIA-defined strata and using the density of FIA Phase 1 sampling points within and across strata to determine strata weights.
 
-Plots in the three states are sampled every 10 years with a target of 1/10 of plots sampled in a panel every year (Thompson 2015).
+Plots in the three states are sampled every 10 years with a target of 1/10th of plots sampled in a panel every year (Thompson 2015).
 Our analysis depends on examining the fate and growth of trees from two sequential visits (referred to as the first and second plot visit), ideally from a complete sample panel remeasure.
 We obtained our FIA data from the USDA FIA DataMart web portal (https://research.fs.usda.gov/products/dataandtools/fia-datamart; downloaded May 5, 2025).
-They included plot data sampled from 2001 to 2023.
+The database included plot data sampled from 2001 to 2023.
 
@@
 
-We obtained plot-level CWD data from ClimateNA (Wang et al.
-2016) for June, July, and August (summer months) between 1980 and 2023 (Bryce: I just averaged the CWD values for these months.
-It wasn’t available seasonally, was it?
-Also, can you provide more info on how CWD is calculated?
-I know it is PET-AET, but I’m not sure how to interpret a monthly/seasonal value).
-Latitude and longitude were based on FIA-fuzzed coordinates (citation?
-Bryce – should we tighten this up by running the analysis on non-fuzzed coordinates?).
-We found the 20-year mean of summer CWD values for each plot prior to and including the first visit.
+We obtained plot-level CWD data from the model TerraClimate (Abatzoglou et al.
+2018) for June, July, and August (summer months) between 1980 and 2023.
+Latitude and longitude were based on FIA-fuzzed coordinates.
+We found the 20-year mean of summer CWD values for each plot prior to and including the first visit year.
 We refer this this value as the “initial CWD value”.
 We then found the mean of summer CWD values between visits and including the second visit year.
-The difference between this second value and initial CWD value is the “CWD change value” (maybe ‘difference’ really is a better, less loaded term).
-Both the initial CWD and CWD change values are mapped in Figure 1.
+The difference between this second value and initial CWD value is the “delta CWD value”.
+Both the initial CWD and delta CWD values for FIA plots are mapped in Figure 1.
 
 ## Analysis
-Estimation approach: Following Section 4.3.4 in Scott et al.
-(2005) we obtained point estimates of species growth and mortality using a ratio estimators (Eqn 4.16).
-For mortality, the numerator is the sum of stratum-weighted means of trees per plot of a given species that died in plots.
+Following Section 4.3.4 in Scott et al.
+(2005) we obtained point estimates of species growth and mortality using ratio estimators (Eqn 4.16).
+The numerator for species-specific mortality is the sum of stratum-weighted means of the number of trees that died between visits per plot.
 The number of sampled trees in each plot is expanded by the area sampled (subplot or macroplot area) adjusted by the proportion of the plot area available for sampling and the number of years between visits.
-The denominator is similar but is calculated as the sum of stratum-weighted means of all trees per plot of a given species.
+The denominator for mortality is similar but is calculated as the sum of stratum-weighted means of all trees per plot of a given species.
 
-The numerator of species-specific growth estimates is the sum of stratum-weighted means of the increase in basal area (cm$^2$) per plot while the denominator is the same as was used for the mortality estimate.
+The numerator of species-specific growth estimates is the sum of stratum-weighted means of the increase in basal area (cm$^2$) between visits per plot while the denominator is the same as was used for the mortality estimate.
 This produces a point estimate for a species’ mean growth.
-Mortality and growth values were standardized to a decadal period.
+Mortality and growth values were standardized to a decadal time period.
 
 The analysis finds species growth and mortality estimates for six domains based on CWD values (Figure 2C).
-The initial CWD values are divided into three quantiles: values below the first quartile (Low), between the first and third quartiles (Medium), and above the third quartile (High).
-These initial CWD value divisions allowed for species-specific assessments.
-For all species we defined an arbitrary boundary of 5 mm to define plots with a greater change in CWD (Above [the boundary]) and those with less CWD change (Below).
-This was done because species did not differ greatly in CWD change (Table S2).
-The category Below includes CWD changes that are negative (i.e., mean CWD was larger prior to the initial visit than between visits).
+The initial CWD values are divided into three species-specific quantiles: values below the first quartile (Low), between the first and third quartiles (Medium), and above the third quartile (High).
+Delta CWD plot values for all species were divided into these above a 5mm CWD change threshold (Above) and those below the threshold (Below).
+The 5mm threshold was selected because it is approximately the mean delta CWD value across species and because species' delta CWD value distributions did not differ greatly (Table S2).
+The category Below includes negative delta CWD values (i.e., mean CWD was larger prior to the initial visit than between visits).
+The domain designations AL, AM, and AH correspond to Above Low, Above Medium, and Above High.
+The Below domains are similarly labeled (BL, BM, BH).
 We also report species mortality and growth estimates for all plots across and within Oregon, Washington, and California irrespective of CWD values.
-We limited the number of species examined to those with more than 400 plots.
-We report domain estimates for all CWD-based domains with 10 or more plots.
+We limited the number of species examined to those with more than 400 occupied plots across the three states.
+We calculated domain estimates for all CWD-based domains with 10 or more plots.
+
+We examined contrasts in 10-year growth estimates (Figure `r fig_nums("fig_diffgrow", display = "num")`) and mortality estimates (Figure `r fig_nums("fig_diffmort", display = "num")`) among domains.
+The figure column "Above vs.
+Below Threshold" displays the mean point estimate and 95% CI for the contrast between corresponding Above and Below domains (e.g., for *Abies lasiocarpa*, subtracting the estimates for columns BL, BM, and BH from AL, AM, AH, respectively; Figure `r fig_nums("fig_domains", display = "num")`A).
+Negative estimates for the "Above vs.
+Below Threshold" in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")`  indicate that the 10-year growth or mortality rate was greater for a species in plots below the change estimates.
+`r t.ref("tbl_summ_grow_mort")` summarizes the number of domain differences in each of the columns in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")` where the 95% CI does not include zero (termed "significant" findings) as well as the number of contrasts per figure column.
 
@@
 For each of the 1000 bootstrap iterations we sampled FIA plots with replacement and obtained point estimates for each species' domain.
-For each domain we calculated a mean and upper and lower 95% CIs with the 2.5% and 97.5% quantiles from the  sampling iterations.
+For each domain we calculated a mean, upper, and lower 95% CI with the 2.5% and 97.5% quantiles from the  sampling iterations.
 This approach allows for nonsymmetrical CIs around means.
@@
 Our analysis approach results in many comparisons.
-We conducted a Monte Carlo (MC) permutation analysis to provide species-level false positive rates (p-values).
-The MC permutation analysis had 1000 iterations; each iteration relied on the bootstrap confidence interval estimation described above.
-The domain assignments for each species was sampled without replacement.
+We conducted a Monte Carlo (MC) permutation analysis to provide species-level false positive rates (p-values) for "significant" contrasts.
+The MC permutation analysis had 1000 iterations; each iteration estimated each species' bootstrap confidence interval as described above.
+The plot domain assignments for each species were sampled without replacement.
 The distance of MC permutation contrast CI bounds from zero were compared against test CI bounds.
 The MC permutation procedure recorded a null significant value if the absolute values of the permutation iteration CI bounds were at least as far from zero as observed bounds.
+
+## Dashboard
+We developed a data dashboard (https://tinyurl.com/FIAgrowmort) to allow readers to examine, for individual species, a species' domain estimates, the distribution of initial and delta CWD values, and the geographical distribution of plots and plot domains.
+All analysis files, manuscript markdowns, dashboard code, and ancillary data sources are available at the GitHub repository https://github.com/jeremygroom/Tree_Growth_Death.
 
@@
 Our study tracked the fate of `r comma(analysis.stats$tree.num)` trees from `r length(sel.spp)` species in `r comma(analysis.stats$plot.num)` plots that were visited twice by FIA crews.
-The two visits for `r pct.remeas[2]`% of plots were 10 years apart (`r pct.remeas[1]`% were 9 years, `r pct.remeas[3]`% were 11 years).
+Of the `r length(sel.spp)` species, `r analysis.stats$n.gym.spp` are gymnosperms.
+The two visits were 10 years apart for `r pct.remeas[2]`% of plots (`r pct.remeas[1]`% were 9 years, `r pct.remeas[3]`% were 11 years).
 Plots were distributed in `r analysis.stats$strat.num` strata.
-Of the `r length(sel.spp)` species, `r analysis.stats$n.gym.spp` are gymnosperms.
+
 Mean overall growth rates for species varied between `r min.grow.spp$mean` (*`r min.grow.spp$spp`*) and `r max.grow.spp$mean` (*`r max.grow.spp$spp`*, `r t.ref("tbl_growth")`) cm$^2$/decade.
+Six species had greater growth (non-overlapping CIs) in California than in Oregon and/or Washington: *Abies concolor*, *Abies grandis*, *Pinus ponderosa*, *Tsuga heterophylla*, *Arbutus menziesii*, and *Notolithocarpus densiflorus*.
+Three species had greater growth in Oregon (*Calocedrus decurrens*, *Picea engelmanii*, and *Pseudotsuga menziesii*) and two species had greater growth in Washington (*Abies amablis* and *Pinus monticola*).
 Decadal mean mortality rates varied between `r min.mort.spp$mean` (*`r min.mort.spp$spp`*) and `r max.mort.spp$mean` (*`r max.mort.spp$spp`*, `r t.ref("tbl_mort")`).
-Growth and mortality rates differed among states (Tables `r tab_nums("tbl_growth", display = "num")` and `r tab_nums("tbl_mort", display = "num")`).
+California had four species with greater mortality rates than Oregon and/or Washington: *Abies concolor*, *Pinus ponderosa*, *Acer macrophyllum*, and *Quercus kelloggii*.
+Oregon had one species with a greater mortality rate (*Tsuga mertensiana*) and Washington had four (*Abies grandis*, *Larix occidentalis*, *Picea engelmannii*, and *Pseudotsuga menziesii*).
 
-Forested plot 10-year differences in CWD do not visually coincide with initial CWD values (Figure `r fig_nums("fig_cwdmap", display = "num")`); however, `r analysis.stats$cwd.gt0.pct`% of plots increased in their CWD values to some degree during that time period.
-Tree species exhibited a variety of initial-visit CWD ranges, with means ranging from `r min.mean.init` (*`r min.mean.init.spp`*) to `r max.mean.init` (*`r max.mean.init.spp`*; Table S1).
-The CWD difference tended to be comparatively small, with a maximum 95th percentile value of `r max.q95.diff` (*`r max.q95.diff.spp`*, Table S2).
-The distribution of individual species plot domain values for CWD and change in CWD can be explored on the dashboard (https://tinyurl.com/FIAgrowmort).
+Delta CWD increased at `r analysis.stats$cwd.gt0.pct`% of plots.
+Initial and delta CWD values were positively linearly associated (correlation = `r CWD.lin.cor`); however, the two variables were better modeled by a 3rd order polynomial ($\Delta AIC = $ `r CWD.dAIC`, Figure S1, see model fits in Tables S1 and S2), and for most species delta CWD neither linearly related to nor monotonically increased with initial CWD values (Figure S2; also, see dashboard).
+Tree species exhibited a variety of initial-visit CWD ranges, with means ranging from `r min.mean.init` (*`r min.mean.init.spp`*) to `r max.mean.init` (*`r max.mean.init.spp`*; Table S3).
+Delta CWD values tended to be comparatively small, with a maximum 95th percentile value of `r max.q95.diff` (*`r max.q95.diff.spp`*, Table S4).
 
-The confidence intervals for 10-year mortality and growth rate domains did not necessarily follow the same patterns nor did all confidence intervals for a species' mortality or growth estimates overlap (e.g., Figures `r fig_nums("fig_domains", display = "num")`A and `r fig_nums("fig_domains", display = "num")`B).
-The relationship between initial CWD values and plot-level change in CWD were not reliably linear (e.g., Figure `r fig_nums("fig_domains", display = "num")`C) and varied by species (see dashboard).
-
-We examined differences in 10-year growth estimates (Figure `r fig_nums("fig_diffgrow", display = "num")`) and mortality estimates (Figure `r fig_nums("fig_diffmort", display = "num")`) among domains.
-The column "Above vs.
-Below Threshold" compares the domains above the CWD change threshold with those below it (e.g., for *Abies lasiocarpa*, subtracting the estimates for columns BL, BM, and BH from AL, AM, AH, respectively; Figure `r fig_nums("fig_domains", display = "num")`A).
-Negative estimates for the "Above vs.
-Below Threshold" in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")`  indicate that the 10-year growth or mortality rate was greater for a species in plots below the change estimates, respectively.
-`r t.ref("tbl_summ_grow_mort")` summarizes the number of domain differences in each of the columns in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")` where the 95% CI does not include zero as well as the number of comparisons per figure column.
-
-Growth domain comparisons (Figure `r fig_nums("fig_diffgrow", display = "num")`, `r t.ref("tbl_summ_grow_mort")`) did not appear to be produce substantially more "significant" negative vs.
-positive estimates, and all three columns of Figure `r fig_nums("fig_diffgrow", display = "num")` had similar numbers of "significant" comparisons.
+Individual species growth and mortality domain patterns were not always similar (e.g., Figures `r fig_nums("fig_domains", display = "num")`A and `r fig_nums("fig_domains", display = "num")`B).
+Growth domain contrasts (Figure `r fig_nums("fig_diffgrow", display = "num")`, `r t.ref("tbl_summ_grow_mort")`) did not appear to result in substantially more "significant" negative or positive estimates, and all three columns of Figure `r fig_nums("fig_diffgrow", display = "num")` had over 20 "significant" contrasts.
 Growth may have been susceptible (negatively or positively) to a change in CWD and differed by initial CWD values, and the responses are species-specific.
 *Abies amabilis*, *Abies grandis*, *Calocedrus decurrens*, *Pinus ponderosa*, *Pseudotsuga menziesii*, *Tsuga mertensiana*, *Acer macrophyllum*, and *Alnus rubra* exhibited greater growth in the Low domains than the Medium and/or High domains.
-Except for *Alnus rubra*, these results predominantly occurred in both the Below domain (2) or the Above and Below domains (5).
-*Thuja plicata* had higher growth in the Above and Below Medium domains than either High or Low domains.
+Except for *Alnus rubra*, these results predominantly occurred in both the Below domain (2 spp.) or the Above and Below domains (5 spp.).
+*Thuja plicata* had greater growth in the Above and Below Medium domains than either High or Low domains.
 *Abies concolor*, *Larix occidentalis*, *Picea engelmannii*, *Pinus contorta*, *Pinus lambertiana*, *Arbutus menziesii*, and *Quercus kelloggii* had greater growth in the more water stressed High domains than the Medium and/or Low domains.
-Except for *Larix occidentalis*, these results predominantly occurred in both the Above domain (3) or the Above and Below domains (3).
-Other species exhibited no 'significant' findings or displayed a mixture of results.
+Except for *Larix occidentalis*, these results predominantly occurred in both the Above domain (3 spp.) or the Above and Below domains (3 spp.).
+Other species exhibited no 'significant' findings (4 spp.) or displayed other result patterns (3 spp.).
 
-Mortality domain comparisons are similar in certain regards to the growth comparisons: results for the three types of domain comparisons did not appear to be predominantly positive or negative (Figure `r fig_nums("fig_diffmort", display = "num")`, `r t.ref("tbl_summ_grow_mort")`).
-There were fewer mortality estimates with "significant" findings than growth.
+Mortality domain contrasts are similar in certain regards to the growth contrasts: results for the three types of domain contrasts did not appear to be predominantly positive or negative (Figure `r fig_nums("fig_diffmort", display = "num")`, `r t.ref("tbl_summ_grow_mort")`).
+There were fewer mortality estimates with "significant" findings than growth (`r mort.tot.sig` vs `r grow.tot.sig`.
 There were fewer differences for "Above vs.
-Below Threshold" comparisons (`r t.ref("tbl_summ_grow_mort")`) than for the comparisons Above or Below the threshold, indicating that mortality may not be as strongly associated with the change in CWD as by the starting value.
+Below Threshold" contrasts (`r mort.n.AvB`, `r t.ref("tbl_summ_grow_mort")`) than for Above (`r mort.A.sig`) or Below (`r mort.B.sig`), indicating that mortality may not be as strongly associated with delta CWD as by the initial CWD value.
 *Larix occidentalis*, *Picea engelmannii*,	*Pinus contorta*,	*Pinus monticola*, *Thuja plicata*,	and *Arbutus menziesii* had greater mortality in the lower water deficit Low domains than Medium and/or High domains.
-These results occurred in Above (3) or both Above and Below (3) domains.
+These results occurred in Above (3 spp.) or both Above and Below (3 spp.) domains.
 Mortality rates for *Abies lasiocarpa*,	*Calocedrus decurrens*,	*Pinus lambertiana*, *Pinus ponderosa*,	*Pseudotsuga menziesii*,	*Tsuga heterophylla*,	*Tsuga mertensiana*,	*Acer macrophyllum*,	and *Notolithocarpus densiflorus* were greater in the High domains than the Medium and/or Low domains.
-These species were approximately evenly split between differences being present in the Above (4) or Below (5) domains, but none with 'significant' findings in both the Above and Below domains.
+These species were approximately evenly split between differences being present in the Above (4 spp.) or Below (5 spp.) domains, but none with 'significant' findings in both the Above and Below domains.
+Five species had no significant contrasts.
 
-(Section on geography?
-Growth highest for Abies grandis in BL, which coincides with coast range for OR, WA, CA plus Olympics.
-CA really only has trees in BL; these probably are the ones that grew the most.
-Then, mention some mortality patterns observed below, especially N/NE WA and Sierra Nevadas)
-
-The MC permutation analysis provided empirical p-values for observing contrast confidence intervals with bounds at least as far from zero as those observed (Supplemental figures S3 and S4), with values ranging from $p$ = 0.072 to $p$ < 0.0001 (41 instances).
+The MC permutation analysis provided empirical p-values for observing contrast confidence intervals with bounds at least as far from zero as those observed (Supplemental figures S5 and S6), with values ranging from a maximum of $p$ = 0.072 to a minimum $p$ < 0.0001 (41 instances).
+Except where explicitly stated, we only discuss individual species' domain findings if the MC permutation p-values are < 0.001 (indicated by three asterisks in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")`)
 
 # Discussion
-We detected numerous species for whom growth and/or mortality rates differed by CWD category.
-We also found species for whom growth and mortality differed by state.
-Our observed changes in CWD were not large, but regionally, from southern California to the Olympic Peninsula in Washington, CWD increased in 9 out of 10 plots.
 
-Our regional probabilistically sampled design-based results provide a potentially valuable perspective on species status.
-Acer macrophyllum growth was negatively associated with initial CWD, and its growth was lower and mortality almost three times higher in California than Washington and Oregon.
-We cannot find primary literature that identifies an increase in mortality for this species in California; the degree of mortality observed seems severe.
-However, Betzen et al.
-(2021) found widespread evidence in Washington of A.
-macrophyllum decline (crown dieback, thinning, discoloration, reduced leaf size) and reduced growth and increased mortality in areas associated with higher temperatures and water limitations.
-Although these results align with our CWD findings for growth, they are not reflective (yet) of our state findings and are not apparent in our CWD findings for mortality.
-We encourage a more intensive assessment of A.
-macrophyllum’s status and causes for decline in California.
+Across the 23 species examined we found no clear pattern of response to initial and/or delta CWD.
+Responses were species-specific, potentially driven in some cases by local phenomena and captured by a domain category or categories.
+We did find more “significant” CWD contrasts for species growth than mortality, possibly indicating that growth differences were more responsive as an indicator than mortality over a decade, although growth, although growth and mortality findings were not necessarily negatively related.
+Delta CWD increased almost everywhere between plot revisits although the magnitude of change was generally not large relative to initial CWD values.
+We additionally found a range of decadal mortality rates by state and region, some of which were concerningly high, such as those for *Picea engelmannii*, *Pinus contorta* in Washington, *Acer macrophyllum* in California, *Abies lasiocarpa*, and *Abies grandis* in Washington.
 
-A converse example is the mortality rate of Thuja plicata, where researchers and state agencies are concerned about die-off but our analysis did not find a strong mortality signal.
-Andrus et al.
-(2023) sampled T.
-plicata at nine locations in Washington, northern Oregon, and Idaho that were without canopy dieback, exhibited canopy dieback, or had died.
-They found that the mortality was associated with exceptionally warm and dry climate conditions.
-A collaboration of state and federal agencies is currently reaching out to the public to identify dieback sites for this species (website: Goodrich et al.
-2025).
-Our analysis found that T.
-plicata exhibited the lowest decadal mortality rate and third highest growth rate of any species we examined and the greatest growth in Medium domains and some evidence that mortality in the Above Low domain (Olympic peninsula, western Washington Cascades) was greater than Above High domain (NE Washington, southern Oregon Cascades).
-Our analysis may not be temporally sensitive enough to capture the same recent findings as Andrus et al., given that our FIA plot revisit data represented plot visits between 2011 to 2023.
-However, our analysis does indicate that numerous species are exhibiting mortality rates at a state level or overall that are higher and are potentially unsustainable (e.g., Abies grandis, Abies lasiocarpa, Picea engelmannii, Pinus contorta, Pinus monticola, and Alnus rubra have state or overall decadal mortality rates greater than 0.25).
+We believe the value of our analysis lies in its application to a relatively large number of tree species and its approach for investigating life history conditions such as growth and mortality that require an estimate of condition change.
+The availability of revisit data for FIA datasets is a relatively recent development.
+Prior to the availability of the revisit data, Monleon and Lintz (2015) found design-based estimates for the difference in mean sapling and mature tree plot variable values for (“initial”) plot temperature values to assess tree species’ range shifts.
+Groom and Monleon (2023), also using design-based estimation approaches, examined FIA revisit data and found evidence of tree species density shifts along initial values of climatic variables but documented evidence of interactions between those variables’ initial and change values.
+We are not aware of other design-based estimates of revisit data that examine more than one variable of interest simultaneously.
+Our approach allows us to capitalize on the amenability of the data to design-based estimation while accounting for the need, when examining revisit conditions, of capturing the initial and change value conditions.
+Examples of model-based analyses of mortality and growth using FIA revisit data include Shriver et al.
+(2022), Ogana et al.
+(2025), and Brandeis and Brandeis (2021).
+Examples of growth and mortality design-based estimation that rely on FIA revisit data, similar to our state and overall analyses, include Vickers and Goff (2025), Goeking and Windmuller-Campione (2021), and Perret et al.
+(2023).
+Stanke et al.
+(2021) found estimates of condition indices.
+Our bivariate examination of plot CWD shares some dimensional similarities with Dobrowski et al.
+(2013)  who examined the velocity of CWD (change in distance over time) along with actual evapotranspiration (AET) and minimum temperatures.
 
-Abies lasiocarpa and Picea engelmannii are conspecifics and frequently codominant.
-Perret et al.
-(2023) estimated a rangewide mortality of A.
-lasiocarpa of 25% +/-2.7, with mortality concentrated in the eastern Oregon cascades as well as in Idaho and Colorado.
-83% of the mortality was due to fire or biogenic disturbance.
-These mortality rates are comparable to the state-level ones we found.
-Andrus et al.
-(2018) found that P.
-engelmannii exhibited higher survival and longevity in undisturbed plots in Colorado, with firs exhibiting mortalities of up to 1% annually while P.
-engelmannii had mortality rates below 0.5% annually.
-We found that P.
-engelmannii had higher mortality in the Above Low domain.
-We note that the Above domains represent a proportionally large increase for Low and Medium given initial CWD values.
-The same is true for A.
-lasiocarpa, yet P.
-engelmannii has greater mortality in the Low domains and A.
-lasiocarpa has less mortality in the Below Low domain although the MC permutation p-values were not very small.
-The difference may be geographical: A.
-lasiocarpa low-mortality plots are primarily on the Olympic Peninsula of Washington and the northwestern Cascades, where rainfall is abundant.
-P.
-engelmannii is generally absent from these locations.
-However, P.
-engelmannii has the least mortality in the Above domain in the Blue Mountains of Oregon and Washington while A.
-lasiocarpa does not show a similar finding for the same geographical region.
-P.
-engelmannii mortality in northern and northeastern Washington, WA may have been driven by spruce beetles (WADNR 2023).
-Our mortality estimate for Pinus contorta in Washington was twice that of Oregon’s and five times greater than Calfornia’s, while the growth confidence intervals for Washignton were lower than California’s and did not overlap.
-Mortality was highest in the Low CWD domains and growth lowest in the Above Low domain.
-The Above Low domain aligns well with plots for northern and northeastern Washington.
-Pine bark beetles (Dendroctonus ponderosae, Dendroctonus brevicomis, and Ips spp.) have affected this region for the past decade (WADNR 2023).
-This area corresponds with the portions of P.
-contorta’s range in the lowest forest stability index category for Stanke et al.
-(2021) while California’s Sierra Nevadas and the Oregon and Washington Cascade Range correspond to positive stability index values.
-Stanke et al.
-also found that insects (followed by fire) were a major driver of the percent change in relative tree density.
-
-Our analysis centered around initial CWD and delta CWD as a variable of interest relative to tree growth and mortality.
-There is ample evidence in the literature supporting the relationship or correlation of CWD or its cousin VPD to tree growth (Dyer 2004, Restaino et al.
+Our analysis relied on plot-level initial CWD and delta CWD values as variables of interest for examining tree growth and mortality.
+CWD is a reasonable variable to use for examining regional water stress as it integrates climatic and soil water availability through considering precipitation and air temperatures (Stephenson 1990, 1998).
+There is ample evidence in the literature supporting the relationship or correlation between CWD and tree growth (Dyer 2004, Restaino et al.
 2016) and mortality (Anderegg et al.
@@
 2022).
-However, there are some challenges in relying on it in a design-based regional analysis.
-We did not detect sizable changes in CWD between visits although CWD disproportionately increased between visits across sites.
-The changes we observed may have represented greater proportional change for some species than others, with species such as Abies amablis, Alnus rubra, and Juniperus occidentalis exhibiting a low initial CWD range.
+However, there are some challenges in relying on CWD or any other climate variable in a design-based regional analysis.
+We did not detect sizable changes in CWD between visits, potentially signalling that 10 years is too soon to observe CWD-related species responses (perhaps moreso for mortality than growth).
+The changes we observed may have represented greater proportional change for some species than others, with species such as *Abies amablis*, *Alnus rubra*, and *Juniperus occidentalis* exhibiting narrow and low water stress initial CWD ranges.
+CWD effects can manifest in direct and indirect ways, potentially differing with species by geographic location and the biotic and abiotic impacts to those locations.
+It also cannot easily convey differences in impacts to trees within the same stands.
 
-Interpreting CWD results at a regional scale becomes difficult because water stress alone can kill trees (Anderegg et al.
+Interpreting CWD results at a regional scale is difficult because evidence of water stress can manifest in multiple overlapping ways.
+Water stress alone can kill trees (Anderegg et al.
 2015, Maherali et al.
@@
 With reference to Manion’s (1991) “death spiral” (see also Bennett et al.
-2023 for a CWD-focused version), teasing apart the cause of mortality from predisposing, inciting, and contributing factors is difficult at best, as these factors can interact (Gaylord et al.
+2023 for a CWD-related version), teasing apart the cause of mortality from predisposing, inciting, and contributing factors is difficult at best, as these factors can interact (Gaylord et al.
 2015).
@@
 
-# Discussion
+Although our current analysis offers a coarse examination of species status across a large area, our bivariate approach for assessing tree species growth and mortality by domains could be refined.
+Our climate domains could be divided further into sub-domains: trees could be assigned to tree size, stand competition, plot aspect, or plot elevation categories within the climate domain categories.
+This approach has its limits: only estimates for certain widespread tree species would be reasonable to report, and domains could not be divided too finely into sub-domains, limiting the number of variables that could be considered simultaneously.
+However, this approach would permit design-based estimation at a resolution better attuned to capturing factors affecting growth and mortality.
 
-The initial CWD values coincide fairly well with mean annual precipitation [NOAA Climate.gov](https://www.climate.gov/news-features/featured-images/new-maps-annual-average-temperature-and-precipitation-us-climate).
+We found some suggestion that growth patterns with MC permutation p-values less than < 0.001 were more frequently positive than negative for contrasts of Above or Below initial CWD values.
+This may indicate that more tree species are exhibiting greater growth at plots with less initial water stress (e.g., *Pseudotsuga menziesii*, *Pinus ponderosa*, and *Acer macrophyllum*) with counterexamples (e.g., *Pinus contorta*).
+We found fewer such contrasts for the decadal mortality analysis of CWD relative to decadal growth, with a fairly even number of significant contrasts for both Above and Below initial CWD values.
+Tree species could exhibit higher mortality at higher or lower initial water stress values.
+
+The number of "significant" Above vs.
+Below threshold contrasts for growth and mortality were approximately evenly distributed between positive and negative contrast estimates, with a positive difference indicating that some species exhibited higher growth or greater mortality in an Above domain (e.g., *Pinus lambertiana* for both growth and mortality) versus a Below domain (*Abies grandis* and *Pinus contorta*, respectively, for growth and mortality).
+This lack of a pattern across species may indicate that the distinction of Above vs.
+Below is generally not important, or that ‘significant’ findings are due to confounding factors such as geographical locations with different ecological pressures.
+
+The interpretation of contrast results may be more useful at an individual species level and when considering overall mortality and growth estimates.
+Regional analyses detected several species with relatively high mortality rates.
+*Acer macrophyllum* growth was negatively associated with initial CWD values, and its growth was lower and mortality almost three times higher in California than in Washington and Oregon.
+We cannot find primary literature that corroborates an increase in mortality for this species in California; the degree of mortality observed seems severe.
+McNellis et al.
+(2021) estimated a recent *Acer macrophyllum* overall mortality rate of 0.45%/yr (4.5%/decade).
+However, Betzen et al.
+(2021) found widespread evidence in Washington of *Acer macrophyllum* decline (crown dieback, thinning, discoloration, reduced leaf size) and reduced growth and increased mortality in locations with higher temperatures and water limitations.
+Although these results align with our regional CWD findings for growth, as displayed in our dashboard, they are not reflective (yet) of our state mortality findings for Washington.
+We encourage a more intensive assessment of *Acer macrophyllum*’s status and causes for decline in California.
+
+Our regional analyses found several other species with mortality rates that are relatively high and are potentially unsustainable (e.g., *Abies grandis*, *Abies lasiocarpa*, *Picea engelmannii*, *Pinus contorta*, *Pinus monticola*, and *Alnus rubra* have state or overall decadal mortality rates greater than 0.25).
+We do not believe our results are concerning for *Alnus rubra*.
+Our estimates for this short-lived tree species align with reported background mortality rates (e.g., Brown and Courtin 2018).
+
+Conspecifics *Abies lasiocarpa* and *Picea engelmannii* are frequently codominant and exhibit high mortality rates in our study, likely due to wildfire and insects.
+Andrus et al.
+(2018) found that *Picea engelmannii* exhibited higher survival and longevity compared to *Abies lasiocarpa* in undisturbed plots in Colorado, with firs exhibiting mortalities of up to 1% annually while *Picea engelmannii* had mortality rates below 0.5% annually, while our overall estimated mortality rates were substantially higher (although our mortality estimate for *Picea engelmannii* in Oregon was similar).
+Perret et al.
+(2023) estimated a rangewide mortality of *Abies lasiocarpa* of 25% +/-2.7, with mortality concentrated in the eastern Oregon Cascades as well as in Idaho and Colorado.
+They found that 83% of the mortality was due to fire or biogenic disturbance.
+These mortality rates are comparable to the state-level ones we found.
+We found that *Picea engelmannii* had lower mortality in the Above High domain, corresponding to stands in the Blue Mountains of Oregon and Washington, and higher mortality in the Above Low and Above High domains, largely coinciding with northern and northeastern Washington, where spruce beetle were elevated (WADNR 2023).
+
+Our mortality estimate for *Pinus contorta* in Washington was twice that of Oregon’s and five times greater than Calfornia’s, while the growth confidence intervals for Washignton were lower than California’s and did not overlap.
+Mortality was highest in the Low CWD domains and growth lowest in the Above Low domain.
+The Above Low domain aligns well with plots for northern and northeastern Washington.
+Pine bark beetles (*Dendroctonus ponderosae*, *Dendroctonus brevicomis*, and *Ips* spp.) have affected this region for the past decade (WADNR 2023).
+This area corresponds with the portions of *Pinus contorta*’s range in the lowest forest stability index category for Stanke et al.
+(2021) while California’s Sierra Nevadas and the Oregon and Washington Cascade Range correspond to positive stability index values.
+Stanke et al.
+also found, from FIA plot data, that insects (followed by fire) were related to the percent change in relative tree density.
+
+*Abies grandis* exhibited higher mortality in Washington than Oregon or California.
+We found no evidence of a difference in domain mortality geographically.
+This result may be due to an intersection in a fire suppression history promoting a densification of shade-tolerant but drought-intolerant firs in the arid climates of eastern Oregon and Washington, drought, and fir engraver beetle mortality (*Scolytus ventralis*; Washington DNR 2023, Merschel et al.
+2021).
+Johnson et al.
+(2025) report mortality rates of *Abies grandis* in protected areas in the Malheur National Forest of Oregon that exceed ours, primarily attributing fire as the cause.
+
+Our dashboard reveals some instances of domains representing geographical mortality differences.
+State estimates for *Pinus monticola* mortality do not appear to differ.
+Mortality for the Above High domain was low (see dashboard; MC permutation test results only found p < 0.01 when comparing Above High to Above Low) and predominantly represented by plots in the southern Sierra Nevada mountains.
+*Pinus lambertiana* and *Calocedrus decurrens* exhibited high mortality in approximately the same region.
+This region experienced an extreme drought that killed 170m trees between 2010 and 2021 (OEHHA 2022).
+Dudney et al.
+(2020) also estimated lower mortality rates for *Pinus monticola* than for *Pinus lambertiana* in the Sequoia and Kings Canyon National Parks, which are within this region.
+They attributed some of the difference in mortality to elevation, with mountain pine beetle (*Dendroctonus ponderosae*), fire, and white pine blister rust (*Cronartium ribicola*) more prevalent at lower elevations, corresponding to the range of *Pinus lambertiana*.
+Fettig et al.
+(2019) separately corroborated high tree mortality rates for *Pinus lambertiana* and *Calocedrus decurrens* in this region.
+
+An example species for which our results did not corroborate with other researchers is *Thuja plicata*.
+Researchers and state agencies are currently concerned about die-off but our analysis did not find a strong mortality signal.
+Andrus et al.
+(2023) sampled *Thuja plicata* at nine locations in Washington, northern Oregon, and Idaho that were without canopy dieback, exhibited canopy dieback, or had died.
+They found that the mortality was associated with exceptionally warm and dry climate conditions.
+A collaboration of state and federal agencies is currently reaching out to the public to identify dieback sites for this species (website: Goodrich et al.
+2025).
+Our analysis found that *Thuja plicata* exhibited the lowest decadal mortality rate and third highest growth rate of any species we examined.
+It also had its greatest growth in Medium domains.
+Our analysis may not be temporally sensitive enough to capture the same recent findings as Andrus et al., given that our FIA plot revisit data represented plot visits between 2011 to 2023.
+If species conditions change quickly (on the order of 3-5 years) the FIA plot data may not reflect the change until a full or nearly full panel of plots is complete following the change in conditions.
 
 # Conclusion
+
+We hope our approach, results, and dashboard assist in refining future species inquiries.
 
@@
 ::: {#refs}
+Abatzoglou, J.T., Dobrowski, S.Z., Parks, S.A.
+and Hegewisch, K.C., 2018.
+TerraClimate, a high-resolution global dataset of monthly climate and climatic water balance from 1958–2015.
+Scientific data, 5(1), pp.1-12.
+
+Alexander, R.R., Shearer, R.C. and Shepperd, W.D., 1990. Abies lasiocarpa (Hook.) Nutt. subalpine fir. Silvics of North America, 1, pp.60-70.
 
@@
 
-Wang, T., Hamann, A. Spittlehouse, D.L. and Carroll, C. 2016. Locally downscaled and spatially customizable climate data for historical and future periods for North America. PLoS One 11: e0156720
+Brandeis, T. and Brandeis, C., 2021. Assessing tree mortality probability in harvested hardwood stands using long-term forest inventory data. Forest Science, 67(2), pp.231-240.
 
-Alexander, R.R., Shearer, R.C. and Shepperd, W.D., 1990. Abies lasiocarpa (Hook.) Nutt. subalpine fir. Silvics of North America, 1, pp.60-70.
+Breshears, D.D., Myers, O.B., Meyer, C.W., Barnes, F.J., Zou, C.B., Allen, C.D., McDowell, N.G.
+and Pockman, W.T., 2009.
+Tree die‐off in response to global change‐type drought: mortality insights from a decade of plant water potential measurements.
+Frontiers in Ecology and the Environment, 7(4), pp.185-189.
 
-Washington State Department of Natural Resources (DNR) Forest Reliance Division.
+Brown, K.R.
+and Courtin, P.J., 2018.
+Can phosphorus additions increase long-term growth and survival of red alder (Alnus rubra Bong.) on periodically dry sites?.
+Forest Ecology and Management, 430, pp.545-557.
+
+Daly C, Halbleib M, Smith JI, Gibson WP, Doggett MK, Taylor GH, et al.
+Physiographically sensitive mapping of climatological temperature and precipitation across the conterminous United States.
+Int J Climatol.
+2008; 28(15):2031–64.
+doi: 10.1002/joc.1688 PMID: WOS:000261810700004.
+
+Goeking, S.A. and Windmuller-Campione, M.A., 2021. Comparative species assessments of five-needle pines throughout the western United States. Forest Ecology and Management, 496, p.119438.
+
+McRoberts, R.E.
+2005.
+The Enhanced Forest Inventory and Analysis Program, in: Bechtold, W.A., Patterson, P.L.
+(Eds.), The Enhanced Forest Inventory and Analysis Program – National Sampling Design and Estimation Procedures.
+Gen.
+Tech.
+Rep.
+U.S.
+Dep.
+Agric.
+For.
+Serv.
+SRS-80.
+
+Merschel, A.G., Beedlow, P.A., Shaw, D.C., Woodruff, D.R., Lee, E.H., Cline, S.P., Comeleo, R.L., Hagmann, R.K.
+and Reilly, M.J., 2021.
+An ecological perspective on living with fire in ponderosa pine forests of Oregon and Washington: Resistance, gone but not forgotten.
+Trees, Forests and People, 4, p.100074.
+
+Ogana, F.N., Green, P.C.
+and Radtke, P., 2025.
+Modeling the growth and yield of natural hardwood stands in the southern United States using the Forest Inventory and Analysis data.
+Forest Ecology and Management, 586, p.122722.
+
+Office of Environmental Health Hazard Assessment (OEHHA, 2022). Indicators of climate change in california, fourth edition, California Environmental Protection Agency, OEHHA.
+
+Reams, G.A, Smith, W.D., Hansen, M.H., Bechtold, W.A., Roesch, F.A., Moisen, G.G.
+2005.
+The forest inventory and analysis sampling frame, in: Bechtold, W.A., Patterson, P.L.
+(Eds.), The Enhanced Forest Inventory and Analysis Program – National Sampling Design and Estimation Procedures.
+Gen.
+Tech.
+Rep.
+U.S.
+Dep.
+Agric.
+For.
+Serv.
+SRS-80.
+
+Scott, C.T., W. A. Bechtold, G.A. Reams, W.D. Smith, J.A. Westfall, M.H. Hansen, Moisen, G.G. 2005. Sample-based estimators used by the Forest Inventory and Analysis
+National Information Management System, in: Bechtold, W.A., Patterson, P.L. (Eds.), The Enhanced Forest Inventory and Analysis Program – National Sampling Design
+and Estimation Procedures. Gen. Tech. Rep. U.S. Dep. Agric. For. Serv. SRS-80.
+
+Stephenson, N.L., 1990. Climatic control of vegetation distribution: the role of the water balance. The American Naturalist, 135(5), pp.649-670.
+
+Stephenson, N., 1998. Actual evapotranspiration and deficit: biologically meaningful correlates of vegetation distribution across spatial scales. Journal of biogeography, 25(5), pp.855-870.
+
+Vickers, L.A. and Goff, T.C., White Oak Sapling Dynamics Across the Eastern Us: Quantifying a Bottleneck to Sustainability. Available at SSRN 5398050.
+
+Washington State Department of Natural Resources (WADNR) Forest Reliance Division.
 2024.
 Forest health highlights in Washington: A summary of insect, disease, and other disturbance conditions affecting Washington’s forests.
-
-Washington State Department of Natural Resources (WADNR).
-2024.
-Forest Health Highlights in Washington – 2024.
 Washington State Department of Natural Resources, Forest Resilience Division, Olympia, WA.
@@
 <!---BLOCK_LANDSCAPE_STOP--->
+
+```{r tbl-summ-grow-mort, tab.cap = tab.summ_grow_mort}
+
+g.m.summ.ft
+
+```
 
@@
 ```
-
-\newpage
-
-```{r tbl-summ-grow-mort, tab.cap = tab.summ_grow_mort}
-
-g.m.summ.ft
-
-```

### 2025-11-03 Supplemental Material.Rmd
Latest version of manuscript, added CWD analysis to supplemental materials.

@@
 analysis.stats <- read_rds(paste0(RESULTS.OTHER, "analysis.stats.RDS"))
+# Correlation/fits for Initial CWD vs Delta CWD:
+CWD.rels <- read_rds(paste0(RESULTS.OTHER, "CWD_deltaCWD_relationship.rds"))
 
@@
 ```{r cwd_summary_tables}
+ft.lm1 <- CWD.rels$ft_lm1
+
 cwd.init <- analysis.stats$init.cwd.stats 
@@
 
-```{r table-legends}
+```{r legends}
+fig_nums <- captioner::captioner(prefix = " S")
 tab_nums <- captioner::captioner(prefix = "Table S")
+
+f.ref <- function(x) {
+ caption_text <- fig_nums(x)
+  # Remove any leading numbers and extract just "Figure X"
+  stringr::str_extract(caption_text, "\\d+")
+}
 
@@
 }
+
+tbl_cwd_lm1 <- tab_nums(name = "tbl_cwdlm1", 
+                         caption = "Linear regression model output for the linear fit of initial and delta CWD plot data (green line in Figure S1). Variable name CWD refers to initial CWD value.")
+
+tbl_cwd_lm2 <- tab_nums(name = "tbl_cwdlm2", 
+                         caption = "Linear regression model output for the third-order polynomial fit of initial and delta CWD plot data (orange line in Figure S1).
+Variable name CWD refers to initial CWD value.")
 
@@
 Values of zero represented no permutation results with confidence intervals equal to or greater than those observed in the data.")
+
+fig_cwd_all <- fig_nums(name = "fig_cwdall", 
+                        caption = "Initial CWD and Delta CWD values for all FIA plots occupied by analysis species.
+The green line is a linear fit of the values while the orange line is a third-order polynomial fit of the points.
+")
+
+fig_cwd_spp <- fig_nums(name = "fig_cwdspp", 
+                        caption = "Third-order polynomial fits for initial CWD and delta CWD values for FIA plots occupied by individual species.")
+
+```
+
+<br>
+
+<br>
+
+```{r cwdfig-all, fig.width=6, fig.height=6,  fig.dpi=300, fig.cap=fig_cwd_all, fig.cap.sep=""}
+
+# Output from Nplot_Climvar_analysis.R
+knitr::include_graphics(paste0(RESULTS.OTHER, "CWD_deltaCWD.png"))
+
+```
+
+\newpage
+
+```{r table-cwd-lm1, tab.cap=tbl_cwd_lm1}
+
+# Output from Nplot_Climvar_analysis.R
+#ft.lm1.1 <- ft.lm1 %>%
+#  compose(i = 2, j = 1, value = as_paragraph("CWD"), part = "body")
+
+ft.lm1.1 <- ft.lm1 %>%
+     mk_par(i = 2, j = 1, value = as_paragraph("CWD"), part = "body")
+
+ft.lm1.1
+
+```
+
+<br>
+
+<br>
+
+```{r table-cwd-lm2, tab.cap=tbl_cwd_lm2}
+
+# Output from Nplot_Climvar_analysis.R
+ft.lm2 <- CWD.rels$ft_lm2 %>%
+ set_formatter(
+      values = function(y) formatC(y, format = "e", digits = 2),
+      part = "body"
+    ) %>%
+  mk_par(i = 2:4, j = 1, value = c(as_paragraph("CWD"), as_paragraph("CWD^2"), as_paragraph("CWD^3")), part = "body")
+
+ft.lm2
+
+```
+
+\newpage
+
+```{r cwdfig-spp, fig.width = 6, fig.height= 6,  fig.dpi=300, fig.cap=fig_cwd_spp, fig.cap.sep=""}
+
+# Output from Nplot_Climvar_analysis.R
+knitr::include_graphics(paste0(RESULTS.OTHER, "CWD_spp_plt.png"))
 

### 2025-11-03 Analysis Code/Data_Prep_Analysis.R

@@ -356,14 +356,6 @@ if(n_domain == 6) {
   psig.dat <- read_csv(paste0(RESULTS.OTHER, "Permutation_results.csv"), show_col_types = FALSE)
 
     for(k in 1:2){ # 1 = growth, 2 = mortality
-      font_to_use <- if("Times New Roman" %in% systemfonts::system_fonts()$family) {
-      "Times New Roman"
-    } else if("Times" %in% systemfonts::system_fonts()$family) {
-      "Times"
-    } else {
-      "serif"  # Fallback to default serif
-    }
-    
     
     psig.join.fcn <- function(k, position) {
       pos.adj <- 3*(k-1) + position

### 2025-11-03 Analysis Code/Manuscript_information.R

@@ -86,7 +86,7 @@ analysis.stats <- list(tree.num = tree.num,
                        diff.cwd.90range.min = diff.cwd.90range.min,
                        diff.cwd.90range.max = diff.cwd.90range.max,
                        diff.cwd.95max = diff.cwd.95max
-                       ) 
+) 
 
 write_rds(analysis.stats, paste0(RESULTS.OTHER, "analysis.stats.RDS"))
 
@@ -261,7 +261,6 @@ inset_map <- ggplot() +
 cwd.map.fcn <- function(var.use, title.txt, lab.use, add_extras = FALSE) {
   p <- ggplot() +
     # Add state boundaries
-    geom_sf(data = west_states_proj, fill = "transparent", color = "black", size = 0.8) +
     # Add climate data points
     geom_point(data = climate_proj_df, 
                aes(x = X, y = Y, color = get(var.use)), 
@@ -276,6 +275,7 @@ cwd.map.fcn <- function(var.use, title.txt, lab.use, add_extras = FALSE) {
                             frame.colour = "black",
                             ticks.colour = "black"
                           )) +
+    geom_sf(data = west_states_proj, fill = "transparent", color = "black", size = 0.8) +
     coord_sf(xlim = x_range, ylim = y_range, expand = FALSE) +
     theme_void() +
     theme(
@@ -368,7 +368,7 @@ cwd.maps <- plot_grid(
 cwd.maps <- ggdraw() +
   draw_plot(cwd.plt.startcwd, x = 0, y = 0, width = 0.5, height = 1) +
   draw_plot(cwd.plt.deltcwd, x = 0.5, y = 0, width = 0.5, height = 1) 
-  
+
 
 # Print the final figure
 print(cwd.maps)
@@ -379,7 +379,7 @@ ggsave(paste0(RESULTS.OTHER, "cwd_maps_figure.png"), cwd.maps, device = 'png',
        bg = "white")
 
 ms.figures.other <- list(cwd.maps = cwd.maps)
-  
+
 write_rds(ms.figures.other, paste0(RESULTS.OTHER, "ms.figures.other.RDS"))
 
 

### 2025-11-03 Analysis Code/Nplot_Climvar_anaysis.R

@@ -1,10 +1,13 @@
 ## Species distributions, plot numbers, and climate variable distributions ##
 
-# --> Run Data_Prep_Analysis.R up through data.use (around line 91)
+# --> Run Data_Prep_Analysis.R up through data.use (around line 100)
 
 ### -- Number of plots per species -- ###
 ## This bit of code is for determining which species to investigate based on plot number.  
 #  In this case, 400 plots is used as a breakpoint.
+library(flextable)
+library(broom)
+
 n.spp.plts <- data.use %>% select(SPCD, puid) %>%
   distinct() %>%
   group_by(SPCD) %>%
@@ -66,18 +69,18 @@ maxlong <- max(map.dat.1$LON); minlong <- min(map.dat.1$LON)
 
 
 cwd.map.fcn <- function(var.use, title.txt, lab.use){
-ggplot(data = map.dat.1, aes(x = LON, y = LAT)) +
-  coord_fixed(xlim = c(minlong - 1, maxlong + 1),  ylim = c(minlat - 1, maxlat + 1), ratio = 1.3) +
-  geom_point(data = map.dat.1, aes(LON, LAT, color = get(var.use)), size = 0.5) +
-  #  geom_tile(data = map.dat.1, mapping = aes(x = LON, y = LAT, z = targ.spp), binwidth = 0.15, 
-  #            stat = "summary_2d", fun = mean, na.rm = TRUE, show.legend = FALSE) + 
-  scale_color_viridis(option = "H", begin = 0.1, end = 0.9, name = lab.use) +  
-  #                  labels = gsub(".", " ", QUANT.LEVELS[quants.used], fixed = TRUE)) +
-  #scale_fill_viridis(option = "H", begin = virid.min, end = virid.max) +
-  geom_polygon(data = west_df, mapping = aes(x = long, y = lat, group = group), color = "black", fill = "transparent") +
-  theme_void() +
-  theme(panel.border = element_rect(colour = "black", fill=NA, linewidth = 0.5)) + 
-  labs(title = title.txt)
+  ggplot(data = map.dat.1, aes(x = LON, y = LAT)) +
+    coord_fixed(xlim = c(minlong - 1, maxlong + 1),  ylim = c(minlat - 1, maxlat + 1), ratio = 1.3) +
[…]

### 2025-11-03 Global.R

@@ -115,6 +115,17 @@ spp.id <- names(orig)[(ncol(orig) - (length(grep("X", names(orig))) - 1)):ncol(o
 spp.names <- readxl::read_xlsx(paste0(DATA.LOC, "FullSppNames.xlsx")) %>% dplyr::filter(SPCD %in% spp.list)
 
 
+## Plotting
+font_to_use <- if("Times New Roman" %in% systemfonts::system_fonts()$family) {
+  "Times New Roman"
+} else if("Times" %in% systemfonts::system_fonts()$family) {
+  "Times"
+} else {
+  "serif"  # Fallback to default serif
+}
+
+
+
 ## Items for parallel computing
 #n.cores <- round(detectCores() * 0.75, 0) # Number of cores, package "parallel".  Using 75% of the available cores, rounded.
 n.cores <- N.WORKERS

### 2025-11-12 Manuscript.Rmd
Updated the manuscript text and added mountain range names to the manuscript figure and to the maps in the app

@@
 CWD.rels <- read_rds(paste0(RESULTS.OTHER, "CWD_deltaCWD_relationship.rds"))
-CWD.lin.cor <- CWD.rels$lm1$r.squared^0.5 
+CWD.lin.cor <- format(round(CWD.rels$lm1$r.squared^0.5, 2), nsmall = 2) # Last digit = 0, want to show.
 CWD.dAIC <- round(CWD.rels$aic.cwd$AIC[1] - CWD.rels$aic.cwd$AIC[2], 0)
@@
 
-There  are signs that tree species in western North America are in regional decline.
+Numerous tree species in western North America exhibit increased mortality or decline.
+van Mantgem et al.
+(2009) modeled mortality rates for the genera *Abies* and *Pinus* in the western USA as increasing from less than 0.05% annually in 1960 to almost 0.2% by 2010.
 Johnston et al.
-(2025) document a 29% 10-year mortality rate of 300+ yr old trees in Eastern Oregon.
-Peret et al.
-(2022) estimated a decadal range-wide mortality rate of 25% for Subalpine fir (*Picea engelmannii*).
-High-severity fires in California in 2020 and 2021 killed 13%-19% of giant sequoia (*Sequoiadendron gianteum*) with >1.07-m bole diameters (Stephenson and Brigham 2021, Shive et al.
-2021); recruitment in these burns was depressed (Soderberg et al.
+(2025) documented a 29% 10-year mortality rate of 300+ yr old trees in Eastern Oregon.
+Perret et al.
+(2023) estimated a decadal range-wide mortality rate of 25% for Subalpine fir (*Picea engelmannii*).
+High-severity fires in California in 2020 and 2021 killed 13%-19% of extant giant sequoia (*Sequoiadendron gianteum*) with >1.07-m bole diameters (Stephenson and Brigham 2021, Shive et al.
+2022); recruitment in these burns was depressed (Soderberg et al.
 2022).
@@
 (2022) found globally that the intensity of drought and heat events are directly associated with widespread forest die-off events.
-Wildfire frequency and area burned in the western United States has increased, in concert with the longer fire seasons, hotter summer temperatures, and earlier snowmelt (Abatzoglou and Williams 2016, Westerling 2016, Juang et al.
+Wildfire frequency and area burned in the western United States has increased, in concert with the longer fire seasons, hotter summer temperatures, and earlier snowmelt (Abatzoglou and Williams 2016, Westerlin 2016, Juang et al.
 2022).
@@
 
-Drought-induced mortality can be driven by carbon starvation and hydraulic failure (see reviews by Adams et al.
+Carbon starvation and hydraulic failure can drive drought-induced mortality (see reviews by Adams et al.
 2010 and McDowell et al.
@@
 
-One regionally useful metric of water stress is Climatic Water Deficit (CWD; Stephenson 1998) which appears to be a reasonable covariate for examining tree growth and mortality.
-Stephenson (1990) defines CWD as evaporative demand not met by available water, or the difference between plant water demand (actual evapotranspiration, AET), and plant water supply (potential evapotranspiration, PET); these variables are driven by temperature and precipitation, respectively.
-CWD is geographically readily available from downscaled PRISM data (Daly et al.
-2008, Wang et al.
-2016).
+One regionally useful metric of water stress is Climatic Water Deficit (CWD; Stephenson 1998).
+Stephenson (1990) defines CWD as evaporative demand not met by available water, or the difference between plant water demand (actual evapotranspiration, AET), and plant water supply (potential evapotranspiration); these variables are driven by temperature and precipitation, respectively.
 The Anderegg  et al.
@@
 National Forest Inventory, collected by the US Forest Service Forest Inventory and Analysis Program (FIA), is well suited for evaluating regional species mortality and growth.
-Since 2001 the FIA program has gathered a variety of forestry data at plots along the west coast of the USA using a spatially balanced probability sample (McRoberts 2005, Reams et al.
-2005).
+Since 1999 the FIA program has gathered a variety of forestry data at plots nationwide using a spatially balanced probability sample (McRoberts 2005, Reams et al.
+2005, Woudenberg et al.
+2010).
 The sample allows for a relatively unbaised design-based inference of population characteristics across a large geographical region.
@@
 
+Climatic data such as mean annual temperature and CWD can be combined with FIA data to obtain climate-related design-based estimates.
 Monleon and Linz (2015) evaluated the evidence of tree species range shifts by subtracting the design-based mean estimates of annual FIA plot temperatures for seedlings from those of trees.
 FIA plot revisit data were mostly incomplete at the time of their analysis.
-Groom and Monleon (2023) used the revisit data to compare plot climate variable values (vapor pressure deficit [VPD], temperature, and precipitation) to estimate shifts in plot occupancy and density.
+Groom and Monleon (2023) used a mostly-completed revisit data set to compare plot climate variable values (vapor pressure deficit [VPD], temperature, and precipitation) to estimate shifts in plot occupancy and density.
 Both studies used a single climate variable value associated with each plot for estimation; e.g., Groom and Monleon (2023) examined shifts in species density by estimating the difference in VPD means between plots with more trees of a given species and plots with fewer trees.
@@
 The purpose of this analysis is to obtain design-based estimates of tree species’ growth and mortality across the western U.S.
-states of California, Oregon, and Washington as they relate to plot-level initial-visit water stress (Climatic Water Deficit, initial CWD) and the change in water stress (delta CWD) over time.
+states of California, Oregon, and Washington as they relate to plot-level initial-visit water stress (initial CWD) and the change in water stress between plot visits (delta CWD).
 We approached this bivariate analysis by dividing species’ plot data into domains based on initial and delta CWD values, finding growth and mortality estimates for each domain, and estimating domain contrasts to establish differences between domains.
@@
 Our analysis relies on FIA data for forestland in California, Oregon, and Washington, USA.
-Forestland is defined as an area of 4,050 m$^2$ that is at least 10 percent stocked or potentially stocked with tree species, excluding urban and agricultural land uses (Reams et al., 2005).
-We constrained our analysis to only include remeasurements on areas considered forestland during both visits.
+Forestland is defined as an area of 4,050 m$^2$ that is at least 10 percent stocked or potentially stocked with tree species, excluding urban and agricultural land uses (Reams et al.
+2005).
+We constrained our analysis to only include remeasurements for areas considered forestland during both visits.
 
 FIA plots consist of four points where all trees with stem diameters greater than or equal to 12.7 cm (for most species measured at 1.37 m above the ground) were tallied in a 7.32 m radius circular subplot (total area 672.5 m$^2$) with additional trees sampled in an anulus around each subplot (macroplot), with a total radius 18 m from the center point and a total area of 1012 m$^2$.
-We did not include data from FIA microplots.
+We did not include data from microplots, where seedlings and saplings are measures, as the analysis was not intended to estimate species recruitment but rather to follow the trajectory of established trees.
 Bechtold and Scott (2005) provide plot design details and measurement protocols.
 
-Since 2000, FIA plots have been selected via spatially-balanced probability sampling (McRoberts 2005, Reams et al., 2005) with a target plot density of one ground plot every 24 km$^2$.
-Some regions and ownerships intensify the plot density; we excluded these additional plots to simplify stratification estimation, relying on FIA-defined strata and using the density of FIA Phase 1 sampling points within and across strata to determine strata weights.
+The FIA program selected plots with a spatially-balanced probability sampling design (McRoberts 2005, Reams et al., 2005) that used a target plot density of one ground plot every 24 km$^2$.
+Some regions and ownerships intensify the plot density; we excluded these additional plots to simplify stratification estimation, relying on FIA-defined strata and using the density of FIA Phase 1 sampling points within and across strata to determine stratum weights.
 
-Plots in the three states are sampled every 10 years with a target of 1/10th of plots sampled in a panel every year (Thompson 2015).
-Our analysis depends on examining the fate and growth of trees from two sequential visits (referred to as the first and second plot visit), ideally from a complete sample panel remeasure.
+Plots in the three states are sampled every 10 years with a target of annually sampling a panel of 10% of plots every year (Thompson 2015).
+Our analysis depends on examining the fate and growth of trees measured during two sequential visits (referred to as the first and second plot visit), ideally from a complete sample panel remeasure.
 We obtained our FIA data from the USDA FIA DataMart web portal (https://research.fs.usda.gov/products/dataandtools/fia-datamart; downloaded May 5, 2025).
@@
 Latitude and longitude were based on FIA-fuzzed coordinates.
-We found the 20-year mean of summer CWD values for each plot prior to and including the first visit year.
-We refer this this value as the “initial CWD value”.
+We defined the 'initial CWD' values as the 20-year mean of summer CWD values for each plot prior to and including the first visit year.
 We then found the mean of summer CWD values between visits and including the second visit year.
-The difference between this second value and initial CWD value is the “delta CWD value”.
-Both the initial CWD and delta CWD values for FIA plots are mapped in Figure 1.
+We then defined 'delta CWD' values as the difference between this second value and initial CWD value.
+Figure 1 presents maps of both the initial CWD and delta CWD values for forested and nonforested FIA plots.
 
 ## Analysis
-Following Section 4.3.4 in Scott et al.
-(2005) we obtained point estimates of species growth and mortality using ratio estimators (Eqn 4.16).
+We obtained point estimates of species growth and mortality using ratio estimators (Eqn 4.16 in Scott et al.
+2005).
+We expanded the number of sampled trees in each plot by the area sampled (subplot or macroplot area) adjusted by the proportion of the plot area available for sampling and the number of years between visits.
 The numerator for species-specific mortality is the sum of stratum-weighted means of the number of trees that died between visits per plot.
-The number of sampled trees in each plot is expanded by the area sampled (subplot or macroplot area) adjusted by the proportion of the plot area available for sampling and the number of years between visits.
 The denominator for mortality is similar but is calculated as the sum of stratum-weighted means of all trees per plot of a given species.
@@
 This produces a point estimate for a species’ mean growth.
-Mortality and growth values were standardized to a decadal time period.
+We standardized mortality and growth values to a decadal time period.
 
-The analysis finds species growth and mortality estimates for six domains based on CWD values (Figure 2C).
-The initial CWD values are divided into three species-specific quantiles: values below the first quartile (Low), between the first and third quartiles (Medium), and above the third quartile (High).
+We estimated, for each species examined, six domains based on CWD values (Figure 2C).
+The initial CWD plot values are divided into three species-specific quantiles: values below the first quartile (Low), between the first and third quartiles (Medium), and above the third quartile (High).
 Delta CWD plot values for all species were divided into these above a 5mm CWD change threshold (Above) and those below the threshold (Below).
@@
 
-We examined contrasts in 10-year growth estimates (Figure `r fig_nums("fig_diffgrow", display = "num")`) and mortality estimates (Figure `r fig_nums("fig_diffmort", display = "num")`) among domains.
+For each species we examined contrasts among growth (Figure `r fig_nums("fig_diffgrow", display = "num")`) and mortality (Figure `r fig_nums("fig_diffmort", display = "num")`) CWD domain estimates.
 The figure column "Above vs.
@@
 The plot domain assignments for each species were sampled without replacement.
+Given this sampling approach, and since species distributions were not independent, the p-values are strictly interperetable at the species level (e.g., we cannot speak to the likelihood of observing a specific number of significant findings for a contrast type).
 The distance of MC permutation contrast CI bounds from zero were compared against test CI bounds.
@@
 Our study tracked the fate of `r comma(analysis.stats$tree.num)` trees from `r length(sel.spp)` species in `r comma(analysis.stats$plot.num)` plots that were visited twice by FIA crews.
-Of the `r length(sel.spp)` species, `r analysis.stats$n.gym.spp` are gymnosperms.
+Of the `r length(sel.spp)` species, `r analysis.stats$n.gym.spp` were gymnosperms.
 The two visits were 10 years apart for `r pct.remeas[2]`% of plots (`r pct.remeas[1]`% were 9 years, `r pct.remeas[3]`% were 11 years).
@@
 Oregon had one species with a greater mortality rate (*Tsuga mertensiana*) and Washington had four (*Abies grandis*, *Larix occidentalis*, *Picea engelmannii*, and *Pseudotsuga menziesii*).
+Seven species had overall mortality rates $\geq$ 0.20 (*Abies concolor*, *Abies lasiocarpa*, *Picea engelmannii*, *Pinus contorta*, *Pinus lambertiana*, *Alnus rubra*, and *Quercus chryselopis*) while at a state level seven species had mortality rates $\geq$ 0.25 (California: *Acer macrophyllum* and *Alnus rubra*; Oregon: *Abies lasiocarpa*; Washington: *Abies grandis*, *Picea engelmannii*, *Pinus contorta*, and *Pinus monticola*).
 
 Delta CWD increased at `r analysis.stats$cwd.gt0.pct`% of plots.
-Initial and delta CWD values were positively linearly associated (correlation = `r CWD.lin.cor`); however, the two variables were better modeled by a 3rd order polynomial ($\Delta AIC = $ `r CWD.dAIC`, Figure S1, see model fits in Tables S1 and S2), and for most species delta CWD neither linearly related to nor monotonically increased with initial CWD values (Figure S2; also, see dashboard).
+Initial and delta CWD values were positively linearly associated (correlation = `r CWD.lin.cor`); however, the two variables were better modeled by a 3rd order polynomial ($\Delta AIC$ =  `r CWD.dAIC`, Figure S1, see model fits in Tables S1 and S2), and for most species delta CWD neither linearly related to nor monotonically increased with initial CWD values (Figure S2; also, see dashboard).
 Tree species exhibited a variety of initial-visit CWD ranges, with means ranging from `r min.mean.init` (*`r min.mean.init.spp`*) to `r max.mean.init` (*`r max.mean.init.spp`*; Table S3).
@@
 
-Individual species growth and mortality domain patterns were not always similar (e.g., Figures `r fig_nums("fig_domains", display = "num")`A and `r fig_nums("fig_domains", display = "num")`B).
-Growth domain contrasts (Figure `r fig_nums("fig_diffgrow", display = "num")`, `r t.ref("tbl_summ_grow_mort")`) did not appear to result in substantially more "significant" negative or positive estimates, and all three columns of Figure `r fig_nums("fig_diffgrow", display = "num")` had over 20 "significant" contrasts.
-Growth may have been susceptible (negatively or positively) to a change in CWD and differed by initial CWD values, and the responses are species-specific.
+Individual species growth and mortality domain patterns could differ (e.g., Figures `r fig_nums("fig_domains", display = "num")`A and `r fig_nums("fig_domains", display = "num")`B).
+Growth domain contrasts (Figure `r fig_nums("fig_diffgrow", display = "num")`, `r t.ref("tbl_summ_grow_mort")`) did not appear to result in substantially more "significant" negative or positive estimates, indicating that species overall did not exhibit more or less growth in the Above, Low, or High domains.
+All three columns of Figure `r fig_nums("fig_diffgrow", display = "num")` had over 20 "significant" contrasts.
+Depending on the species, growth may have been susceptible (negatively or positively) to a change in CWD and/or differed by initial CWD values.
 *Abies amabilis*, *Abies grandis*, *Calocedrus decurrens*, *Pinus ponderosa*, *Pseudotsuga menziesii*, *Tsuga mertensiana*, *Acer macrophyllum*, and *Alnus rubra* exhibited greater growth in the Low domains than the Medium and/or High domains.
@@
 Mortality domain contrasts are similar in certain regards to the growth contrasts: results for the three types of domain contrasts did not appear to be predominantly positive or negative (Figure `r fig_nums("fig_diffmort", display = "num")`, `r t.ref("tbl_summ_grow_mort")`).
-There were fewer mortality estimates with "significant" findings than growth (`r mort.tot.sig` vs `r grow.tot.sig`.
+There were fewer mortality estimates with "significant" findings than growth (`r mort.tot.sig` vs `r grow.tot.sig`).
 There were fewer differences for "Above vs.
-Below Threshold" contrasts (`r mort.n.AvB`, `r t.ref("tbl_summ_grow_mort")`) than for Above (`r mort.A.sig`) or Below (`r mort.B.sig`), indicating that mortality may not be as strongly associated with delta CWD as by the initial CWD value.
+Below Threshold" contrasts (`r mort.AvB.sig`, `r t.ref("tbl_summ_grow_mort")`) than for Above (`r mort.A.sig`) or Below (`r mort.B.sig`), indicating that mortality may not be as strongly associated with delta CWD as by the initial CWD value.
 *Larix occidentalis*, *Picea engelmannii*,	*Pinus contorta*,	*Pinus monticola*, *Thuja plicata*,	and *Arbutus menziesii* had greater mortality in the lower water deficit Low domains than Medium and/or High domains.
@@
 Mortality rates for *Abies lasiocarpa*,	*Calocedrus decurrens*,	*Pinus lambertiana*, *Pinus ponderosa*,	*Pseudotsuga menziesii*,	*Tsuga heterophylla*,	*Tsuga mertensiana*,	*Acer macrophyllum*,	and *Notolithocarpus densiflorus* were greater in the High domains than the Medium and/or Low domains.
-These species were approximately evenly split between differences being present in the Above (4 spp.) or Below (5 spp.) domains, but none with 'significant' findings in both the Above and Below domains.
+These species were approximately evenly split between differences being present in the Above (4 spp.) or Below (5 spp.) domains, but none with "significant" findings in both the Above and Below domains.
 Five species had no significant contrasts.
@@
 The MC permutation analysis provided empirical p-values for observing contrast confidence intervals with bounds at least as far from zero as those observed (Supplemental figures S5 and S6), with values ranging from a maximum of $p$ = 0.072 to a minimum $p$ < 0.0001 (41 instances).
-Except where explicitly stated, we only discuss individual species' domain findings if the MC permutation p-values are < 0.001 (indicated by three asterisks in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")`)
+Except where explicitly stated, we only discuss individual species' domain findings if the species has one or more relevant MC permutation p-values < 0.001 (indicated by three or four asterisks in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")`)
 
@@
 
-Across the 23 species examined we found no clear pattern of response to initial and/or delta CWD.
+Across the 23 species examined we found no clear pattern of response to initial and/or delta CWD values.
 Responses were species-specific, potentially driven in some cases by local phenomena and captured by a domain category or categories.
-We did find more “significant” CWD contrasts for species growth than mortality, possibly indicating that growth differences were more responsive as an indicator than mortality over a decade, although growth, although growth and mortality findings were not necessarily negatively related.
+We did find more “significant” CWD contrasts for species growth than mortality, possibly indicating that growth differences by CWD domain were more readily detectable than mortality over a decade, although growth and mortality findings were not necessarily negatively related.
 Delta CWD increased almost everywhere between plot revisits although the magnitude of change was generally not large relative to initial CWD values.
-We additionally found a range of decadal mortality rates by state and region, some of which were concerningly high, such as those for *Picea engelmannii*, *Pinus contorta* in Washington, *Acer macrophyllum* in California, *Abies lasiocarpa*, and *Abies grandis* in Washington.
+We additionally found for some species that decadal mortality rates by state and/or region were concerningly high.
 
-We believe the value of our analysis lies in its application to a relatively large number of tree species and its approach for investigating life history conditions such as growth and mortality that require an estimate of condition change.
-The availability of revisit data for FIA datasets is a relatively recent development.
-Prior to the availability of the revisit data, Monleon and Lintz (2015) found design-based estimates for the difference in mean sapling and mature tree plot variable values for (“initial”) plot temperature values to assess tree species’ range shifts.
-Groom and Monleon (2023), also using design-based estimation approaches, examined FIA revisit data and found evidence of tree species density shifts along initial values of climatic variables but documented evidence of interactions between those variables’ initial and change values.
-We are not aware of other design-based estimates of revisit data that examine more than one variable of interest simultaneously.
-Our approach allows us to capitalize on the amenability of the data to design-based estimation while accounting for the need, when examining revisit conditions, of capturing the initial and change value conditions.
+The number of "significant" Above vs.
+Below threshold contrasts for growth and mortality were approximately evenly distributed between positive and negative contrast estimates, with a positive difference indicating that some species exhibited higher growth or greater mortality in an Above domain (e.g., *Pinus lambertiana* for both growth and mortality) versus a Below domain (*Abies grandis* and *Pinus contorta*, respectively, for growth and mortality).
+This lack of a pattern across species may indicate that the distinction of Above vs.
+Below is not generalizable across species, or that "significant" findings are due to confounding factors such as geographical locations with different ecological pressures.
+
+Contrast results may be more informative at an individual species level and when considering state-level mortality and growth estimates.
+Regional analyses detected several species with relatively high mortality rates.
+*Acer macrophyllum* growth was negatively associated with initial CWD values, and its growth was lower and mortality almost three times higher in California than in Washington and Oregon.
+We cannot find primary literature that corroborates an increase in mortality for this species in California.
+McNellis et al.
+(2021) estimated a recent *Acer macrophyllum* overall mortality rate of 0.45%/yr (4.5%/decade) while our estimates for Washington and Oregon were twice this level, and California's estimate was six times as large.
+However, Betzen et al.
+(2021) found widespread evidence in Washington of *Acer macrophyllum* decline (crown dieback, thinning, discoloration, reduced leaf size) and reduced growth and increased mortality in locations with higher temperatures and water limitations.
+Although these results align with our contrast findings for growth, as displayed in our dashboard, they are not reflective (yet) of our state mortality findings for Washington.
+We encourage a more intensive assessment of *Acer macrophyllum*’s status in California.
+
+We do not believe our results are concerning for *Alnus rubra*.  Our estimates for this short-lived tree species align with reported background mortality rates (e.g., Brown and Courtin 2018).
+
+Conspecifics *Abies lasiocarpa* and *Picea engelmannii* are frequently codominant and exhibit high mortality rates in our study.
+Their respective mortality rates in our analysis appear to be elevated; Andrus et al.
+(2018) found that *Picea engelmannii* exhibited higher survival and longevity compared to *Abies lasiocarpa* in undisturbed plots in Colorado, with firs exhibiting a mortality rates of up to 1% annually while *Picea engelmannii* had mortality rates below 0.5% annually.
+Mortality estimates for *Abies lasiocarpa* from Perret et al.
+(2023) approximated our state and overall estimates at 25% +/-2.7, with mortality concentrated in the eastern Oregon Cascades as well as in Idaho and Colorado.
+They found that 83% of the mortality was due to fire or biogenic disturbance.
+The pressures on them appear to differ by CWD values and location.
+We found that *Picea engelmannii* had lower mortality in the Above High domain, corresponding to stands in the Blue Mountains of Oregon and Washington, and higher mortality in the Above Low and Above High domains, largely coinciding with northern and northeastern Washington, where spruce beetle were elevated (WADNR 2023).
+Its mortality rate in Washington was almost 5 times higher than in Oregon.
+*Abies lasiocarpa* did not have substantial mortality contrast findings (some evidence of BH - BL difference, but the MC permutation p-value was greater than 0.001) and its mortality was high but similar across all three states.
+
+Our findings for *Pinus contorta* in Washington likely reflected insect pressures on the species.
+Our Washington mortality estimate was twice that of Oregon’s and five times greater than Calfornia’s, while the growth confidence intervals for Washington were lower than California’s and did not overlap.
+Mortality was highest in the Low CWD domains and growth lowest in the Above Low domain and also high in the Below Low and Below Medium domains.
+These domains aligns well with plots for Washington, in regions that have experienced water stress (Raymond et al.
+2023).
+Pine bark beetles (*Dendroctonus ponderosae*, *Dendroctonus brevicomis*, and *Ips* spp.) have affected *Pinus contorta* for the past decade in this region (WADNR 2023).
+This area also corresponds, at the level of ecoregion subsections, with the portions of *Pinus contorta*’s range in the lowest forest stability index category for Stanke et al.
+(2021) while California’s Sierra Nevadas and the Oregon and Washington Cascade Range correspond to positive stability index values.
+Stanke et al.
+also found, from FIA plot data, that insects (followed by fire) were related to the percent change in relative tree density.
+
+*Abies grandis* exhibited higher mortality in Washington than Oregon or California.
+We found no evidence of a difference in domain mortality geographically, although the greatest growth domain was Below Low, which corresponded to the more mesic plots west of the Cascade Crest of Washington and Oregon.
+The mortality result may be due to an intersection in a fire suppression history promoting a densification of shade-tolerant but drought-intolerant firs in the arid climates of eastern Oregon and Washington, drought, and fir engraver beetle mortality (*Scolytus ventralis*; Washington DNR 2023, Merschel et al.
+2021).
+Johnson et al.
+(2025) report mortality rates of *Abies grandis* in protected areas in the Malheur National Forest of Oregon that exceed ours, primarily attributing fire as the cause.
+
+Our dashboard reveals some instances of domains representing geographical mortality differences.
+State estimates for *Pinus monticola* mortality do not appear to differ.
+Mortality for the Above High domain was low (see dashboard; MC permutation test results only found p < 0.01 when comparing Above High to Above Low) and predominantly represented by plots in the southern Sierra Nevada mountains of California.
+*Pinus lambertiana* and *Calocedrus decurrens* exhibited high mortality in approximately the same region, a result supported by Fettig et al.
+(2019).
+This region experienced an extreme drought that killed 170m trees between 2010 and 2021 (OEHHA 2022).
+Dudney et al.
+(2020) also estimated lower mortality rates for *Pinus monticola* than for *Pinus lambertiana* in the Sequoia and Kings Canyon National Parks, which are within this region.
+They attributed some of the difference in mortality to elevation, with mountain pine beetle (*Dendroctonus ponderosae*), fire, and white pine blister rust (*Cronartium ribicola*) more prevalent at lower elevations, corresponding to the range of *Pinus lambertiana*.
+
+Our results for *Thuja plicata* did not corroborate with those from other researchers who are currently concerned about mortality in this species.
+Andrus et al.
+(2023) sampled *Thuja plicata* at nine locations in Washington, northern Oregon, and Idaho that were without canopy dieback, exhibited canopy dieback, or had died.
+They found that the mortality was associated with exceptionally warm and dry climate conditions.
+A collaboration of state and federal agencies is currently reaching out to the public to identify dieback sites for this species (website: Goodrich et al.
+2025).
+Our analysis found that *Thuja plicata* exhibited the lowest decadal mortality rate and third highest growth rate of any species we examined.
+It also had its greatest growth in Medium domains.
+Our analysis may not be temporally sensitive enough to capture the same recent findings as Andrus et al., given that our FIA plot revisit data represented plot visits between 2011 to 2023.
+If species conditions change quickly (on the order of 3-5 years) the FIA plot data may not reflect the change until a full or nearly full panel of plots is complete following the change in conditions.
+Another non-exclusive possibility is that our study serves as one example of many broad-scale design-based analyses of spatially-balanced samples that offer a relatively unbiased assessment of species condition that can assist managers in identifying species of concern and prioritizing restoration actions.
+*Thuja plicata* may indeed be senescing in many regions, but other species may be faring worse.
+
+We believe the value of our analysis lies in its application to a relatively large number of tree species and its approach for investigating life history conditions such as growth and mortality that require the documentation of tree condition change.
+The availability of full-panel FIA revisit data is a relatively recent development.
+Prior to the availability of the revisit data, Monleon and Lintz (2015) assessed tree species’ range shifts using the single-visit data set, comparing mean plot temperature values for sapling and mature trees.
+Groom and Monleon (2023) accessed to the more complete FIA revisit data set and found evidence of tree species density shifts along initial values of climatic variables but documented evidence of interactions between those variables’ initial and change values.
+The literature of model- and design-based estimation of FIA revisit data is growing.
 Examples of model-based analyses of mortality and growth using FIA revisit data include Shriver et al.
@@
 Our bivariate examination of plot CWD shares some dimensional similarities with Dobrowski et al.
-(2013)  who examined the velocity of CWD (change in distance over time) along with actual evapotranspiration (AET) and minimum temperatures.
+(2013)  who examined the velocity of CWD (change in distance over time) along with AET and minimum temperatures.
+However, we are not aware of other design-based estimates of revisit data that examine more than one variable of interest simultaneously.
+Our approach allows us to capitalize on the amenability of the data to design-based estimation while accounting for the need, when examining revisit conditions, of capturing the initial and change value conditions.
 
-Our analysis relied on plot-level initial CWD and delta CWD values as variables of interest for examining tree growth and mortality.
-CWD is a reasonable variable to use for examining regional water stress as it integrates climatic and soil water availability through considering precipitation and air temperatures (Stephenson 1990, 1998).
-There is ample evidence in the literature supporting the relationship or correlation between CWD and tree growth (Dyer 2004, Restaino et al.
+CWD is a defensible variable to use for examining regional water stress as it integrates climatic and soil water availability through considering precipitation and air temperatures (Stephenson 1990, 1998).
+CWD correlates well with tree growth (Dyer 2004, Restaino et al.
 2016) and mortality (Anderegg et al.
@@
 2022).
-However, there are some challenges in relying on CWD or any other climate variable in a design-based regional analysis.
+However, there are some challenges in relying on CWD or any other climate variable in a design-based regional analysis such as this one.
+The challenges stem from the remeasurement timeframe, using a fixed delta CWD domain threshold, the broad, fixed scope of our analysis, and the lack of incorporation of local-scale conditions.
+
 We did not detect sizable changes in CWD between visits, potentially signalling that 10 years is too soon to observe CWD-related species responses (perhaps moreso for mortality than growth).
-The changes we observed may have represented greater proportional change for some species than others, with species such as *Abies amablis*, *Alnus rubra*, and *Juniperus occidentalis* exhibiting narrow and low water stress initial CWD ranges.
+The CWD changes we observed may have represented greater proportional change for some species than others, with species such as *Abies amablis*, *Alnus rubra*, and *Juniperus occidentalis* exhibiting narrow and low water stress initial CWD ranges.
 CWD effects can manifest in direct and indirect ways, potentially differing with species by geographic location and the biotic and abiotic impacts to those locations.
@@
 
-Interpreting CWD results at a regional scale is difficult because evidence of water stress can manifest in multiple overlapping ways.
+The scale of our analysis and the lack of incorporation of other variables likely limited our analysis because the evidence of water stress can manifest in multiple overlapping ways.
 Water stress alone can kill trees (Anderegg et al.
@@
 
-Although our current analysis offers a coarse examination of species status across a large area, our bivariate approach for assessing tree species growth and mortality by domains could be refined.
+Although our current analysis offers a coarse examination of species status across a large area, future analyses could refine our bivariate approach for assessing tree species growth and mortality by domains.
 Our climate domains could be divided further into sub-domains: trees could be assigned to tree size, stand competition, plot aspect, or plot elevation categories within the climate domain categories.
@@
 
-We found some suggestion that growth patterns with MC permutation p-values less than < 0.001 were more frequently positive than negative for contrasts of Above or Below initial CWD values.
-This may indicate that more tree species are exhibiting greater growth at plots with less initial water stress (e.g., *Pseudotsuga menziesii*, *Pinus ponderosa*, and *Acer macrophyllum*) with counterexamples (e.g., *Pinus contorta*).
-We found fewer such contrasts for the decadal mortality analysis of CWD relative to decadal growth, with a fairly even number of significant contrasts for both Above and Below initial CWD values.
-Tree species could exhibit higher mortality at higher or lower initial water stress values.
-
-The number of "significant" Above vs.
-Below threshold contrasts for growth and mortality were approximately evenly distributed between positive and negative contrast estimates, with a positive difference indicating that some species exhibited higher growth or greater mortality in an Above domain (e.g., *Pinus lambertiana* for both growth and mortality) versus a Below domain (*Abies grandis* and *Pinus contorta*, respectively, for growth and mortality).
-This lack of a pattern across species may indicate that the distinction of Above vs.
-Below is generally not important, or that ‘significant’ findings are due to confounding factors such as geographical locations with different ecological pressures.
-
-The interpretation of contrast results may be more useful at an individual species level and when considering overall mortality and growth estimates.
-Regional analyses detected several species with relatively high mortality rates.
-*Acer macrophyllum* growth was negatively associated with initial CWD values, and its growth was lower and mortality almost three times higher in California than in Washington and Oregon.
-We cannot find primary literature that corroborates an increase in mortality for this species in California; the degree of mortality observed seems severe.
-McNellis et al.
-(2021) estimated a recent *Acer macrophyllum* overall mortality rate of 0.45%/yr (4.5%/decade).
-However, Betzen et al.
-(2021) found widespread evidence in Washington of *Acer macrophyllum* decline (crown dieback, thinning, discoloration, reduced leaf size) and reduced growth and increased mortality in locations with higher temperatures and water limitations.
-Although these results align with our regional CWD findings for growth, as displayed in our dashboard, they are not reflective (yet) of our state mortality findings for Washington.
-We encourage a more intensive assessment of *Acer macrophyllum*’s status and causes for decline in California.
-
-Our regional analyses found several other species with mortality rates that are relatively high and are potentially unsustainable (e.g., *Abies grandis*, *Abies lasiocarpa*, *Picea engelmannii*, *Pinus contorta*, *Pinus monticola*, and *Alnus rubra* have state or overall decadal mortality rates greater than 0.25).
-We do not believe our results are concerning for *Alnus rubra*.
-Our estimates for this short-lived tree species align with reported background mortality rates (e.g., Brown and Courtin 2018).
-
-Conspecifics *Abies lasiocarpa* and *Picea engelmannii* are frequently codominant and exhibit high mortality rates in our study, likely due to wildfire and insects.
-Andrus et al.
-(2018) found that *Picea engelmannii* exhibited higher survival and longevity compared to *Abies lasiocarpa* in undisturbed plots in Colorado, with firs exhibiting mortalities of up to 1% annually while *Picea engelmannii* had mortality rates below 0.5% annually, while our overall estimated mortality rates were substantially higher (although our mortality estimate for *Picea engelmannii* in Oregon was similar).
-Perret et al.
-(2023) estimated a rangewide mortality of *Abies lasiocarpa* of 25% +/-2.7, with mortality concentrated in the eastern Oregon Cascades as well as in Idaho and Colorado.
-They found that 83% of the mortality was due to fire or biogenic disturbance.
-These mortality rates are comparable to the state-level ones we found.
-We found that *Picea engelmannii* had lower mortality in the Above High domain, corresponding to stands in the Blue Mountains of Oregon and Washington, and higher mortality in the Above Low and Above High domains, largely coinciding with northern and northeastern Washington, where spruce beetle were elevated (WADNR 2023).
-
-Our mortality estimate for *Pinus contorta* in Washington was twice that of Oregon’s and five times greater than Calfornia’s, while the growth confidence intervals for Washignton were lower than California’s and did not overlap.
-Mortality was highest in the Low CWD domains and growth lowest in the Above Low domain.
-The Above Low domain aligns well with plots for northern and northeastern Washington.
-Pine bark beetles (*Dendroctonus ponderosae*, *Dendroctonus brevicomis*, and *Ips* spp.) have affected this region for the past decade (WADNR 2023).
-This area corresponds with the portions of *Pinus contorta*’s range in the lowest forest stability index category for Stanke et al.
-(2021) while California’s Sierra Nevadas and the Oregon and Washington Cascade Range correspond to positive stability index values.
-Stanke et al.
-also found, from FIA plot data, that insects (followed by fire) were related to the percent change in relative tree density.
-
-*Abies grandis* exhibited higher mortality in Washington than Oregon or California.
-We found no evidence of a difference in domain mortality geographically.
-This result may be due to an intersection in a fire suppression history promoting a densification of shade-tolerant but drought-intolerant firs in the arid climates of eastern Oregon and Washington, drought, and fir engraver beetle mortality (*Scolytus ventralis*; Washington DNR 2023, Merschel et al.
-2021).
-Johnson et al.
-(2025) report mortality rates of *Abies grandis* in protected areas in the Malheur National Forest of Oregon that exceed ours, primarily attributing fire as the cause.
-
-Our dashboard reveals some instances of domains representing geographical mortality differences.
-State estimates for *Pinus monticola* mortality do not appear to differ.
-Mortality for the Above High domain was low (see dashboard; MC permutation test results only found p < 0.01 when comparing Above High to Above Low) and predominantly represented by plots in the southern Sierra Nevada mountains.
-*Pinus lambertiana* and *Calocedrus decurrens* exhibited high mortality in approximately the same region.
-This region experienced an extreme drought that killed 170m trees between 2010 and 2021 (OEHHA 2022).
-Dudney et al.
-(2020) also estimated lower mortality rates for *Pinus monticola* than for *Pinus lambertiana* in the Sequoia and Kings Canyon National Parks, which are within this region.
-They attributed some of the difference in mortality to elevation, with mountain pine beetle (*Dendroctonus ponderosae*), fire, and white pine blister rust (*Cronartium ribicola*) more prevalent at lower elevations, corresponding to the range of *Pinus lambertiana*.
-Fettig et al.
-(2019) separately corroborated high tree mortality rates for *Pinus lambertiana* and *Calocedrus decurrens* in this region.
-
-An example species for which our results did not corroborate with other researchers is *Thuja plicata*.
-Researchers and state agencies are currently concerned about die-off but our analysis did not find a strong mortality signal.
-Andrus et al.
-(2023) sampled *Thuja plicata* at nine locations in Washington, northern Oregon, and Idaho that were without canopy dieback, exhibited canopy dieback, or had died.
-They found that the mortality was associated with exceptionally warm and dry climate conditions.
-A collaboration of state and federal agencies is currently reaching out to the public to identify dieback sites for this species (website: Goodrich et al.
-2025).
-Our analysis found that *Thuja plicata* exhibited the lowest decadal mortality rate and third highest growth rate of any species we examined.
-It also had its greatest growth in Medium domains.
-Our analysis may not be temporally sensitive enough to capture the same recent findings as Andrus et al., given that our FIA plot revisit data represented plot visits between 2011 to 2023.
-If species conditions change quickly (on the order of 3-5 years) the FIA plot data may not reflect the change until a full or nearly full panel of plots is complete following the change in conditions.
+Our analysis may overall capture significant changes in mortality and growth for the tree species examined, with some trees exhibiting increases in mortality and decreases in growth either in portions of their ranges or across their entire range.
+For some species in semi-arid environments such as *Pinus pondersosa*, *Pseudotsuga menziesii*, *Pinus contorta*, *Picea engelmannii*, and *Abies lasiocarpa*, we may be observing ecosystem shifts with a change in water stress combined with disturbances such as wildfire and insect mortality (Halofsky et al.
+2020, Parks et al.
+2019, Raymond et al.
+2023).
+Our hope is that our bivariate domain-based estimation approach is useful for others, while our results for individual species serve bolster or direct others' inquiries into the status of the species we examined.
+For managers, the results may be used to bolster action for resiliency.
+With the shift in ecosystems, landowners can assist with species migration when planting or replanting (Raymond et al.
+2023), and forest thinning can help reduce water stress and the impacts of fire and insects (Bernal et al.
+2023).
 
@@
 
-We hope our approach, results, and dashboard assist in refining future species inquiries.
+Our analysis serves as an example of bivariate design-based estimation, where we use estimation domains based on initial CWD and delta CWD values of plots.
+For the 23 species we examined, we found no general patterns in domains with higher or lower tree growth or mortality; the results appeared species specific.
+We believe our analysis approach can be refined for individual species analysis, with the potential for integrating a third variable that could introduce information besides a condition variable and change variable.
+Our analysis and the analyses of others provide valuable and reliable insights into the condition of several tree species that exhibit abnormally high mortality (and low growth).
+Our hope si that these findings will help drive management and further monitoring of these species in the rapidly changing landscape of western North America.
 
@@
 ::: {#refs}
+Abatzoglou, J.T. and Williams, A.P., 2016. Impact of anthropogenic climate change on wildfire across western US forests. Proceedings of the National Academy of sciences, 113(42), pp.11770-11775.
+
 Abatzoglou, J.T., Dobrowski, S.Z., Parks, S.A.
@@
 
+Adams, H.D., Macalady, A.K., Breshears, D.D., Allen, C.D., Stephenson, N.L., Saleska, S.R., Huxman, T.E.
+and McDowell, N.G., 2010.
+Climate‐induced tree mortality: Earth system consequences.
+Eos, Transactions American Geophysical Union, 91(17), pp.153-154.
+
+Adams, H.D., Zeppel, M.J., Anderegg, W.R., Hartmann, H., Landhäusser, S.M., Tissue, D.T., Huxman, T.E., Hudson, P.J., Franz, T.E., Allen, C.D.
+and Anderegg, L.D., 2017.
+A multi-species synthesis of physiological mechanisms in drought-induced tree mortality.
+Nature ecology & evolution, 1(9), pp.1285-1291.
+
 Alexander, R.R., Shearer, R.C. and Shepperd, W.D., 1990. Abies lasiocarpa (Hook.) Nutt. subalpine fir. Silvics of North America, 1, pp.60-70.
+
+Anderegg, W.R., Flint, A., Huang, C.Y., Flint, L., Berry, J.A., Davis, F.W., Sperry, J.S.
+and Field, C.B., 2015.
+Tree mortality predicted from drought-induced vascular damage.
+Nature geoscience, 8(5), pp.367-371.
+
+Andrus, R.A., Harvey, B.J., Chai, R.K.
+and Veblen, T.T., 2018.
+Different vital rates of Engelmann spruce and subalpine fir explain discordance in understory and overstory dominance.
+Canadian Journal of Forest Research, 48(12), pp.1554-1562.
+
+Bechtold, W.A., Scott, C.T.
+2005.
+The Forest Inventory and Analysis plot design, in: Bechtold, W.A., Patterson, P.L.
+(Eds.), The Enhanced Forest Inventory and Analysis Program – National Sampling Design and Estimation Procedures.
+Gen.
+Tech.
+Rep.
+U.S.
+Dep.
+Agric.
+For.
+Serv.
+SRS-80.
+
+Bennett, M., Shaw, D.C. and Lowrey, L., 2023. Recent Douglas-fir mortality in the Klamath Mountains ecoregion of Oregon: Evidence for a decline spiral. Journal of Forestry, 121(3), pp.246-261.
 
@@
 
+Das, A.J., Stephenson, N.L., Flint, A., Das, T. and van Mantgem, P.J., 2013. Climatic correlates of tree mortality in water-and energy-limited forests. PLoS One, 8(7), p.e69917.
+
+Dobrowski, S.Z., Abatzoglou, J., Swanson, A.K., Greenberg, J.A., Mynsberge, A.R., Holden, Z.A.
+and Schwartz, M.K., 2013.
+The climate velocity of the contiguous United States during the 20th century.
+Global change biology, 19(1), pp.241-251.
+
+Dudney, J.C., Nesmith, J.C., Cahill, M.C., Cribbs, J.E., Duriscoe, D.M., Das, A.J., Stephenson, N.L.
+and Battles, J.J., 2020.
+Compounding effects of white pine blister rust, mountain pine beetle, and fire threaten four white pine species.
+Ecosphere, 11(10), p.e03263.
+
+Dyer, J.M., 2004. A water budget approach to predicting tree species growth and abundance, utilizing paleoclimatology sources. Climate Research, 28(1), pp.1-10.
+
+Fettig, C.J., Mortenson, L.A., Bulaon, B.M.
+and Foulk, P.B., 2019.
+Tree mortality following drought in the central and southern Sierra Nevada, California, US.
+Forest Ecology and Management, 432, pp.164-178.
+
+Ford, K.R., Breckheimer, I.K., Franklin, J.F., Freund, J.A., Kroiss, S.J., Larson, A.J., Theobald, E.J.
+and HilleRisLambers, J., 2017.
+Competition alters tree growth responses to climate at individual and stand scales.
+Canadian Journal of Forest Research, 47(1), pp.53-62.
+
+Gaylord, M.L., Kolb, T.E. and McDowell, N.G., 2015. Mechanisms of piñon pine mortality after severe drought: a retrospective study of mature trees. Tree physiology, 35(8), pp.806-816.
+
 Goeking, S.A. and Windmuller-Campione, M.A., 2021. Comparative species assessments of five-needle pines throughout the western United States. Forest Ecology and Management, 496, p.119438.
+
+Gomez‐Gallego, M., Galiano, L., Martínez‐Vilalta, J., Stenlid, J., Capador‐Barreto, H.D., Elfstrand, M., Camarero, J.J.
+& Oliva, J.
+(2022) Interaction of drought‐ and pathogen‐induced mortality in Norway spruce and Scots pine.
+Plant, Cell & Environment, 45, 2292–2305.
+https://doi.org/10.1111/pce.14360
+
+Goodrich, B., Fischer, M., Buhl, C.  2025. Western redcedar dieback.  https://storymaps.arcgis.com/stories/1405dab5f59246aa83849ec43f72b15a. Accessed 11 November 2025.
+
+Groom, J.D. and Monleon, V.J., 2023. West coast US tree species show stronger evidence for shifts in density than range. Forest Ecology and Management, 545, p.121262. 
+Hammond, W.M., Williams, A.P., Abatzoglou, J.T., Adams, H.D., Klein, T., López, R., Sáenz-Romero, C., Hartmann, H., Breshears, D.D.
+and Allen, C.D., 2022.
+Global field observations of tree die-off reveal hotter-drought fingerprint for Earth’s forests.
+Nature communications, 13(1), p.1761.
+https://doi.org/10.1038/s41467-022-29289-2
+
+Hart, S.J., Veblen, T.T., Eisenhart, K.S., Jarvis, D.
+and Kulakowski, D., 2014.
+Drought induces spruce beetle (Dendroctonus rufipennis) outbreaks across northwestern Colorado.
+Ecology, 95(4), pp.930-939.
+
+Hicke, J.A., Meddens, A.J. and Kolden, C.A., 2016. Recent tree mortality in the western United States from bark beetles and forest fires. Forest Science, 62(2), pp.141-153.
+
+Huberty, A.F. and Denno, R.F., 2004. Plant water stress and its consequences for herbivorous insects: a new synthesis. Ecology, 85(5), pp.1383-1398.
+
+Johnston, J.D., Stephens, P., Schmidt, M., Jeronimo, S., Brackett, A., Wuenschel, A., Ulrich, D.E., Paine, R., Duloisy, L., Greenler, S., Larson, A.J., and Churchill, D.J.
+2025.
+Significant Mortality of Old Trees Across a Dry Forest Landscape, Oregon, USA.
+Forest Ecology and Management, Volume 593, 122907, ISSN 0378-1127, https://doi.org/10.1016/j.foreco.2025.122907.
+
+Juang, C.S., Williams, A.P., Abatzoglou, J.T., Balch, J.K., Hurteau, M.D.
+and Moritz, M.A., 2022.
+Rapid growth of large forest fires drives the exponential response of annual forest‐fire area to aridity in the western United States.
+Geophysical Research Letters, 49(5), p.e2021GL097131.
+
+Kolb, T.E., Fettig, C.J., Ayres, M.P., Bentz, B.J., Hicke, J.A., Mathiasen, R., Stewart, J.E.
+and Weed, A.S., 2016.
+Observed and anticipated impacts of drought on forest insects and diseases in the United States.
+Forest Ecology and Management, 380, pp.321-334.
+
+Lemmo, S.L., Kerhoulas, L.P., Sherriff, R.L. and Beckmann, J.J., 2024. Drought effects on tree mortality and regeneration in northern California. Forest Ecology and Management, 563, p.121969.
+
+Ma, Q., Su, Y., Niu, C., Ma, Q., Hu, T., Luo, X., Tai, X., Qiu, T., Zhang, Y., Bales, R.C.
+and Liu, L., 2023.
+Tree mortality during long-term droughts is lower in structurally complex forest stands.
+Nature communications, 14(1), p.7467.
+
+Maherali, H., Pockman, W.T. and Jackson, R.B., 2004. Adaptive variation in the vulnerability of woody plants to xylem cavitation. Ecology, 85(8), pp.2184-2199.
+
+Manion, P.D. 1991. Tree Disease Concepts, 2nd ed, 402. Englewood Cliffs, NJ: Prentice-Hall.
+
+McDowell, N., Pockman, W.T., Allen, C.D., Breshears, D.D., Cobb, N., Kolb, T., Plaut, J., Sperry, J., West, A., Williams, D.G.
+and Yepez, E.A., 2008.
+Mechanisms of plant survival and mortality during drought: why do some plants survive while others succumb to drought?.
+New phytologist, 178(4), pp.719-739.
+
+McNellis, B.E., Smith, A.M., Hudak, A.T. and Strand, E.K., 2021. Tree mortality in western US forests forecasted using forest inventory and Random Forest classification.  Ecosphere, 12(3), p.e03419.
 
@@
 
+Meddens, A.J., Hicke, J.A.
+and Ferguson, C.A., 2012.
+Spatiotemporal patterns of observed bark beetle‐caused tree mortality in British Columbia and the western United States.
+Ecological Applications, 22(7), pp.1876-1891.
+
 Merschel, A.G., Beedlow, P.A., Shaw, D.C., Woodruff, D.R., Lee, E.H., Cline, S.P., Comeleo, R.L., Hagmann, R.K.
@@
 Trees, Forests and People, 4, p.100074.
+
+Monleon, V.J., Lintz, H.E., Delzon, S., 2015. Evidence of tree species’ range shifts in a complex landscape. PLoS One 10 (1), e0118069.
+
+Negrón, J.F., McMillin, J.D., Anhold, J.A.
+and Coulson, D., 2009.
+Bark beetle-caused mortality in a drought-affected ponderosa pine landscape in Arizona, USA.
+Forest Ecology and Management, 257(4), pp.1353-1362.
+
+Office of Environmental Health Hazard Assessment (OEHHA, 2022). Indicators of climate change in California, fourth edition, California Environmental Protection Agency, OEHHA.
 
@@
 
-Office of Environmental Health Hazard Assessment (OEHHA, 2022). Indicators of climate change in california, fourth edition, California Environmental Protection Agency, OEHHA.
+Parks, S.A., Coop, J.D. and Davis, K.T., 2025. Intensifying fire season aridity portends ongoing expansion of severe wildfire in western US forests. Global Change Biology, 31(8), p.e70429.
+
+Perret, D.L., Bell, D.M., Gray, A.N., Shaw, J.D.
+and Zald, H.S., 2023.
+Range-wide population assessments for subalpine fir indicate widespread disturbance-driven decline.
+Forest Ecology and Management, 542, p.121128.
+
+Preisler, H.K., Hicke, J.A., Ager, A.A.
+and Hayes, J.L., 2012.
+Climate and weather influences on spatial temporal patterns of mountain pine beetle populations in Washington and Oregon.
+Ecology, 93(11), pp.2421-2434.
 
@@
 
-Scott, C.T., W. A. Bechtold, G.A. Reams, W.D. Smith, J.A. Westfall, M.H. Hansen, Moisen, G.G. 2005. Sample-based estimators used by the Forest Inventory and Analysis
-National Information Management System, in: Bechtold, W.A., Patterson, P.L. (Eds.), The Enhanced Forest Inventory and Analysis Program – National Sampling Design
-and Estimation Procedures. Gen. Tech. Rep. U.S. Dep. Agric. For. Serv. SRS-80.
+Restaino, C.M., Peterson, D.L.
+and Littell, J., 2016.
+Increased water deficit decreases Douglas fir growth throughout western US forests.
+Proceedings of the National academy of Sciences, 113(34), pp.9557-9562.
+
+Scott, C.T., W.
+A.
+Bechtold, G.A.
+Reams, W.D.
+Smith, J.A.
+Westfall, M.H.
+Hansen, Moisen, G.G.
+2005.
+Sample-based estimators used by the Forest Inventory and Analysis National Information Management System, in: Bechtold, W.A., Patterson, P.L.
+(Eds.), The Enhanced Forest Inventory and Analysis Program – National Sampling Design and Estimation Procedures.
+Gen.
+Tech.
+Rep.
+U.S.
+Dep.
+Agric.
+For.
+Serv.
+SRS-80.
+
+Shive, K.L., Wuenschel, A., Hardlund, L.J., Morris, S., Meyer, M.D.
+and Hood, S.M., 2022.
+Ancient trees and modern wildfires: Declining resilience to wildfire in the highly fire-adapted giant sequoia.
+Forest Ecology and Management, 511, p.120110.
+
+Shriver, R.K., Yackulic, C.B., Bell, D.M., & Bradford, J.B.
+(2022).
+Dry forest decline is driven by both declining recruitment and increasing mortality in response to warm, dry conditions.
+Global Ecology and Biogeography, 31, 2259–2269.
+https://doi.org/10.1111/geb.13582
+
+Soderberg, D.N., A.J.
+Das, N.
+L.
+Stephenson, M.D.
+Meyer, C.A.
+Brigham, and J.
+Flickinger.
+2024.
+“Assessing giant sequoia Mortality and regeneration following high-severity wildfire.” Ecosphere 15(3): e4789.
+https://doi.org/10.1002/ecs2.4789
+
+Stanke, H., Finley, A.O., Domke, G.M., Weed, A.S., MacFarlane, D.W., 2021.
+Over half of western United States’ most abundant tree species in decline.
+Nat.
+Commun.
+12, 451.
+https://doi.org/10.1038/s41467-020-20678-z.
+
+Stephenson, N.L., and C. Brigham. 2021. Preliminary estimates of sequoia mortality in the 2020 Castle Fire. National Park Service, Sequoia and Kings Canyon National Parks.
+
+Stephenson, N.L., 1998. Actual evapotranspiration and deficit: biologically meaningful correlates of vegetation distribution across spatial scales. Journal of biogeography, 25(5), pp.855-870.
 
@@
 
-Stephenson, N., 1998. Actual evapotranspiration and deficit: biologically meaningful correlates of vegetation distribution across spatial scales. Journal of biogeography, 25(5), pp.855-870.
+Stovall, A.E., Shugart, H. and Yang, X., 2019. Tree height explains mortality risk during an intense drought. Nature communications, 10(1), p.4385.
+Thompson, J. 2015. PNW-FIADB Users Manual: A data dictionary and user guide for the PNW-FIADB database. Forest Inventory and Analysis, Pacific Northwest Research Station, Portland, Oregon.
 
-Vickers, L.A. and Goff, T.C., White Oak Sapling Dynamics Across the Eastern Us: Quantifying a Bottleneck to Sustainability. Available at SSRN 5398050.
+Van Mantgem, P.J., Stephenson, N.L., Byrne, J.C., Daniels, L.D., Franklin, J.F., Fulé, P.Z., Harmon, M.E., Larson, A.J., Smith, J.M., Taylor, A.H.
+and Veblen, T.T., 2009.
+Widespread increase of tree mortality rates in the western United States.
+Science, 323(5913), pp.521-524.
+DOI: 10.1126/science.1165000
+
+Van Nieuwstadt, M.G. and Sheil, D., 2005. Drought, fire and tree survival in a Borneo rain forest, East Kalimantan, Indonesia. Journal of ecology, 93(1), pp.191-201.
+Vickers, L.A. and Goff, T.C., 2025. White oak sapling dynamics across the eastern US: Quantifying a bottleneck to sustainability. Forest Ecology and Management, 598, p.123232.
 
@@
 https://dnr.wa.gov/sites/default/files/2025-03/rp_2023_fhh_report_final_wr.pdf, accessed 9/18/2025
+
+Westerling ALR. 2016 Increasing western US forest wildfire activity: sensitivity to changes in the timing of spring. Phil. Trans. R. Soc. B 371: 20150178. http://dx.doi.org/10.1098/rstb.2015.0178
+
+Woudenberg, S.W., Conkling, B.L., O'Connell, B.M., LaPoint, E.B., Turner, J.A.
+and Waddell, K.L., 2010.
+The Forest Inventory and Analysis Database: Database description and users manual version 4.0 for Phase 2.
+Gen.
+Tech.
+Rep.
+RMRS-GTR-245.
+Fort Collins, CO: US Department of Agriculture, Forest Service, Rocky Mountain Research Station.
+336 p.
 

### 2025-11-12 Analysis Code/Functions.R

@@ -978,15 +978,25 @@ domain.map.fcn <- function(domain.matrix, spp.num, n.plots.used, virid.use, poin
   #virid.max <- max(n.plots.used$loc[n.plots.used$n > 0]) / 10
   map.virid.use <- virid.use[n.plots.used$loc[n.plots.used$n > 0]]
   
-  ggplot(data = map.dat.1, aes(x = LON, y = LAT)) +
-    coord_fixed(xlim = c(minlong - 1, maxlong + 1),  ylim = c(minlat - 1, maxlat + 1), ratio = 1.3) +
-    geom_point(data = map.dat.1, aes(LON, LAT, color = factor(targ.spp)), size = point.size) +
+  
+  #browser()
+  map.dat.2 <- st_as_sf(map.dat.1, coords = c("LON", "LAT"), crs = 4326) #%>%
+#    st_transform(crs = 3310)  # Match your projection
+  
+  
+  ggplot() +# + data = map.dat.2, aes(x = LON, y = LAT)) +
+    #coord_fixed(xlim = c(minlong - 1, maxlong + 1),  ylim = c(minlat - 1, maxlat + 1), ratio = 1.3) +
+    geom_sf(data = map.dat.2, aes(color = factor(targ.spp)), size = point.size) +
     #  geom_tile(data = map.dat.1, mapping = aes(x = LON, y = LAT, z = targ.spp), binwidth = 0.15, 
     #            stat = "summary_2d", fun = mean, na.rm = TRUE, show.legend = FALSE) + 
     scale_color_manual(values = map.virid.use) +  
     #                  labels = gsub(".", " ", DOMAIN.LEVELS[domains.used], fixed = TRUE)) +
     #scale_fill_viridis(option = "H", begin = virid.min, end = virid.max) +
     geom_polygon(data = west_df, mapping = aes(x = long, y = lat, group = group), color = "black", fill = "transparent") +
+    geom_text(data = mountain_ranges_df,
+              aes(x = X, y = Y, label = name, angle = ang),
+              size = 3, fontface = "bold.italic", color = mountain_ranges_df$colr,
+              check_overlap = TRUE) +
     theme_void() +
     theme(panel.border = element_rect(colour = "black", fill=NA, linewidth = 0.5),
           legend.position = "none")

### 2025-11-12 Analysis Code/Manuscript_information.R

@@ -257,6 +257,28 @@ inset_map <- ggplot() +
   theme(panel.background = element_rect(fill = "lightblue", color = "black", linewidth = 0.5),
         plot.background = element_rect(fill = "white", color = "black", linewidth = 0.5))
 
+
+# Create a data frame with mountain range names and approximate coordinates
+mountain_ranges <- data.frame(
+  name = c("Cascade Range", "Sierra Nevada", "Coast Ranges", 
+           "Blue Mountains", "Klamath\nMountains"),
+  #lon = c(-121.5, -119.5, -123.5, -123.5, -118.5, -123.0),
+  lon = c(-122, -119.5, -123.8,  -118.5, -123.0),
+  lat = c(45.5, 37.5, 44.0, 45.2, 42)
+) %>%
+  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
+  st_transform(crs = 3310)  # Match your projection
+
+# Extract coordinates for plotting
+coords <- st_coordinates(mountain_ranges)
+mountain_ranges_df <- cbind(mountain_ranges, coords) %>%
+  st_drop_geometry() %>%
+  mutate(ang = c(82, -55, 82, 65, 0),
+         #colr = c("white", "black", "white", "white", "black", "black"))
+         colr = rep("black", 5))
+
+
+
 # Enhanced mapping function
 cwd.map.fcn <- function(var.use, title.txt, lab.use, add_extras = FALSE) {
   p <- ggplot() +
@@ -277,6 +299,10 @@ cwd.map.fcn <- function(var.use, title.txt, lab.use, add_extras = FALSE) {
                           )) +
     geom_sf(data = west_states_proj, fill = "transparent", color = "black", size = 0.8) +
     coord_sf(xlim = x_range, ylim = y_range, expand = FALSE) +
+    geom_text(data = mountain_ranges_df,
+              aes(x = X, y = Y, label = name, angle = ang),
+              size = 3, fontface = "bold.italic", color = mountain_ranges_df$colr,
+              check_overlap = TRUE) +
     theme_void() +
     theme(
       panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.8),

### 2025-11-12 Global.R

@@ -126,6 +126,28 @@ font_to_use <- if("Times New Roman" %in% systemfonts::system_fonts()$family) {
 
 
 
+# Shinyapp: Create a data frame with mountain range names and approximate coordinates
+mountain_ranges <- data.frame(
+  name = c("Cascade Range", "Sierra Nevada", "Coast Ranges", 
+           "Blue Mountains", "Klamath\nMountains"),
+  #lon = c(-121.5, -119.5, -123.5, -123.5, -118.5, -123.0),
+  lon = c(-122, -119.5, -123.8,  -118.5, -123.0),
+  lat = c(45.5, 37.5, 44.0, 45.2, 42)
+) %>%
+  st_as_sf(coords = c("lon", "lat"), crs = 4326) #%>%
+#st_transform(crs = 3310)  # Match your projection
+
+# Extract coordinates for plotting
+mtn_coords <- st_coordinates(mountain_ranges)
+mountain_ranges_df <- cbind(mountain_ranges, mtn_coords) %>%
+  st_drop_geometry() %>%
+  mutate(ang = c(82, -55, 82, 65, 0),
+         #colr = c("white", "black", "white", "white", "black", "black"))
+         colr = rep("black", 5))
+
+
+
+
 ## Items for parallel computing
 #n.cores <- round(detectCores() * 0.75, 0) # Number of cores, package "parallel".  Using 75% of the available cores, rounded.
 n.cores <- N.WORKERS

### 2025-11-12 app.R

@@ -168,7 +168,7 @@ server <- function(input, output, session) {
     if (!data_loaded() || is.null(input$species) || input$species == "") {
       return(NULL)
     }
-    
+
     tryCatch({
       # Get analysis type
       k <- as.numeric(input$analysis_type)
@@ -184,7 +184,7 @@ server <- function(input, output, session) {
         plt_dat <- analysis_data$mort_output
       }
       
-#      browser()
+
       plt_dat2 <- cm2.fcn(k, plt_dat$domain.summaries) # Transforming growth data to cm2
       domain.matrix <- mort_grow_dat$domain.matrix
       domain.n <- mort_grow_dat$domain.n
@@ -194,7 +194,7 @@ server <- function(input, output, session) {
       virid.use <<- viridis_pal(option = "H", begin = 0.1, end = 0.9)(n_domain)  # Get colors for plotting
       
       
-     #browser()
+#     browser()
       # Generate the plot using pair.plts.fcn with SHINYAPP.IN.USE = TRUE
       plot_result <- pair.plts.fcn(
         sppnum.to.plot = input$species, 

### 2025-11-14 Manuscript.Rmd
Fourth version of the manuscript; edited the app slightly to make species/type of data shown easier to switch between and added mountain range names.

@@
 
-Numerous tree species in western North America exhibit increased mortality or decline.
+Numerous tree species in western North America exhibit increased mortality rates or decline.
 van Mantgem et al.
@@
 2022).
-Five of eight common species in western North America exhibit declines as measured by relative stand density (Stanke et al.
+Five of eight common tree species in western North America exhibit declines as measured by relative stand density (Stanke et al.
 2021).
@@
 
-Climate change as manifest through increases or intensification in drought and aridity can affect tree mortality directly through physiological damage and indirectly by increasing wildfire frequency and severity and by promoting insect and disease outbreaks.
+Climate change as manifest through increases or intensification in drought and aridity can affect tree mortality directly through physiological damage as well as indirectly by increasing wildfire frequency and severity and by promoting insect and disease outbreaks.
 Hammond et al.
 (2022) found globally that the intensity of drought and heat events are directly associated with widespread forest die-off events.
-Wildfire frequency and area burned in the western United States has increased, in concert with the longer fire seasons, hotter summer temperatures, and earlier snowmelt (Abatzoglou and Williams 2016, Westerlin 2016, Juang et al.
+Wildfire frequency and area burned in the western United States has increased in concert with the longer fire seasons, hotter summer temperatures, and earlier snowmelt (Abatzoglou and Williams 2016, Westerlin 2016, Juang et al.
 2022).
@@
 Das et al.
-(2013) found low-elevation California tree mortality best modeled using CWD.
+(2013) determined that low-elevation California tree mortality was best modeled using CWD.
 Restaino et al.
@@
 2010).
-The sample allows for a relatively unbaised design-based inference of population characteristics across a large geographical region.
+The sample allows for relatively unbiased design-based inference of population characteristics across a large geographical region.
 Data from plot revisits and individual tree data permit direct estimation of change features such as mortality and growth estimation.
@@
 Climatic data such as mean annual temperature and CWD can be combined with FIA data to obtain climate-related design-based estimates.
-Monleon and Linz (2015) evaluated the evidence of tree species range shifts by subtracting the design-based mean estimates of annual FIA plot temperatures for seedlings from those of trees.
+Monleon and Linz (2015) evaluated the evidence of tree species range shifts by subtracting the design-based mean estimates of FIA plot temperatures for seedlings from those of trees.
 FIA plot revisit data were mostly incomplete at the time of their analysis.
 Groom and Monleon (2023) used a mostly-completed revisit data set to compare plot climate variable values (vapor pressure deficit [VPD], temperature, and precipitation) to estimate shifts in plot occupancy and density.
-Both studies used a single climate variable value associated with each plot for estimation; e.g., Groom and Monleon (2023) examined shifts in species density by estimating the difference in VPD means between plots with more trees of a given species and plots with fewer trees.
-They documented, however, that the initial climate variable values could (and often did) change between plot visits, with species often exhibiting positive or negative correlations between the initial visit and between-visits change value of the climate variable.
-They identified that their design-based analysis could potentially improve by including both initial visit and change values for the climate variable.
-
-The purpose of this analysis is to obtain design-based estimates of tree species’ growth and mortality across the western U.S.
+Both studies used a single climate variable value associated with each plot for estimation; e.g., Groom and Monleon (2023) examined shifts in species density by estimating the difference in VPD means between plots that gained trees of a given species and plots that lost trees.
+They documented, however, that the initial climate variable values could (and often did) change between plot visits, with species' plots often exhibiting positive or negative correlations between the initial visit values and between-visits change value of the climate variable.
+They identified that their design-based analysis would potentially improve if they could simultaneously include both initial visit and change values for the climate variable.
+
+The purpose of this analysis is to demonstrate an approach for a bivariate design-based estimation procedure.
+We estimate tree species’ growth and mortality across the western U.S.
 states of California, Oregon, and Washington as they relate to plot-level initial-visit water stress (initial CWD) and the change in water stress between plot visits (delta CWD).
 We approached this bivariate analysis by dividing species’ plot data into domains based on initial and delta CWD values, finding growth and mortality estimates for each domain, and estimating domain contrasts to establish differences between domains.
-Our secondary goal includes providing decadal growth and mortality estimates for our analysis species overall and by state to establish species status and to inform our CWD analysis findings.
+Secondarily we provide decadal growth and mortality estimates for our analysis species overall and by state to establish species status and to inform our CWD analysis findings.
 
@@
 
-FIA plots consist of four points where all trees with stem diameters greater than or equal to 12.7 cm (for most species measured at 1.37 m above the ground) were tallied in a 7.32 m radius circular subplot (total area 672.5 m$^2$) with additional trees sampled in an anulus around each subplot (macroplot), with a total radius 18 m from the center point and a total area of 1012 m$^2$.
-We did not include data from microplots, where seedlings and saplings are measures, as the analysis was not intended to estimate species recruitment but rather to follow the trajectory of established trees.
-Bechtold and Scott (2005) provide plot design details and measurement protocols.
+FIA relies on three related sampling procedures.
+Phase 1 involves the classification of distributed points of remote sensing data into (at least) the categories of forest and non-forest.
+Phase 1 plots are useful for estimation stratification.
+The phase 2 permanent ground plots are a subset of Phase 1 points.
+They consist of four points where all trees with stem diameters greater than or equal to 12.7 cm (for most species measured at 1.37 m above the ground) were tallied in a 7.32 m radius circular subplot (total area 672.5 m$^2$) with additional trees sampled in an annulus around each subplot (macroplot), with a total radius 18 m from the center point and a total area of 1012 m$^2$.
+We did not include data from microplots, where seedlings and saplings are measured, as the analysis was not intended to estimate species recruitment but rather to follow the trajectory of established trees.
+Phase 3 plots are an intensification of Phase 2 plots; we do not rely on their data here.
+Reams et al.
+(2005) and Bechtold and Scott (2005) provide plot design details and measurement protocols.
 
 The FIA program selected plots with a spatially-balanced probability sampling design (McRoberts 2005, Reams et al., 2005) that used a target plot density of one ground plot every 24 km$^2$.
-Some regions and ownerships intensify the plot density; we excluded these additional plots to simplify stratification estimation, relying on FIA-defined strata and using the density of FIA Phase 1 sampling points within and across strata to determine stratum weights.
+Some regions and ownerships intensify their plot density; we excluded these additional plots to simplify stratification estimation, relying on FIA-defined strata and using the density of FIA Phase 1 sampling points within and across strata to determine stratum weights.
 
@@
 The initial CWD plot values are divided into three species-specific quantiles: values below the first quartile (Low), between the first and third quartiles (Medium), and above the third quartile (High).
-Delta CWD plot values for all species were divided into these above a 5mm CWD change threshold (Above) and those below the threshold (Below).
-The 5mm threshold was selected because it is approximately the mean delta CWD value across species and because species' delta CWD value distributions did not differ greatly (Table S2).
-The category Below includes negative delta CWD values (i.e., mean CWD was larger prior to the initial visit than between visits).
+Delta CWD plot values for all species were divided into those above a 5mm CWD change threshold (Above) and those below the threshold (Below).
+The 5mm threshold was selected because species' delta CWD value distributions did not differ greatly and because 5mm is approximately the mean delta CWD value across species (Table S2).
+The Below domains may includes negative delta CWD values (i.e., mean CWD was larger prior to the initial visit than between visits).
 The domain designations AL, AM, and AH correspond to Above Low, Above Medium, and Above High.
@@
 We limited the number of species examined to those with more than 400 occupied plots across the three states.
-We calculated domain estimates for all CWD-based domains with 10 or more plots.
+We calculated estimates for all CWD-based domains with 10 or more plots.
 
@@
 The figure column "Above vs.
-Below Threshold" displays the mean point estimate and 95% CI for the contrast between corresponding Above and Below domains (e.g., for *Abies lasiocarpa*, subtracting the estimates for columns BL, BM, and BH from AL, AM, AH, respectively; Figure `r fig_nums("fig_domains", display = "num")`A).
+Below Threshold" displays the mean point estimate and 95% CI for contrast between corresponding Above and Below domains (e.g., for *Abies lasiocarpa*, subtracting the estimates for columns BL, BM, and BH from AL, AM, AH, respectively; Figure `r fig_nums("fig_domains", display = "num")`A).
 Negative estimates for the "Above vs.
@@
 We conducted a Monte Carlo (MC) permutation analysis to provide species-level false positive rates (p-values) for "significant" contrasts.
-The MC permutation analysis had 1000 iterations; each iteration estimated each species' bootstrap confidence interval as described above.
-The plot domain assignments for each species were sampled without replacement.
-Given this sampling approach, and since species distributions were not independent, the p-values are strictly interperetable at the species level (e.g., we cannot speak to the likelihood of observing a specific number of significant findings for a contrast type).
+For each of 1000 MC permutation analysis iterations the code estimated each all species' bootstrap confidence intervals as described above.
+The MC permutation analysis sampled each species' plot domain assignments without replacement.
+Given this sampling approach, and since species distributions were not independent, the p-values are strictly interpretable at the species level (e.g., we cannot speak to the likelihood of observing a specific number of significant findings for a contrast type).
 The distance of MC permutation contrast CI bounds from zero were compared against test CI bounds.
@@
 ## Dashboard
-We developed a data dashboard (https://tinyurl.com/FIAgrowmort) to allow readers to examine, for individual species, a species' domain estimates, the distribution of initial and delta CWD values, and the geographical distribution of plots and plot domains.
+We developed a data dashboard (https://tinyurl.com/FIAgrowmort) to enable ourselves and readers to examine individual species domain estimates, the distribution of initial and delta CWD values, and the geographical distribution of plots and plot domains.
 All analysis files, manuscript markdowns, dashboard code, and ancillary data sources are available at the GitHub repository https://github.com/jeremygroom/Tree_Growth_Death.
@@
 Decadal mean mortality rates varied between `r min.mort.spp$mean` (*`r min.mort.spp$spp`*) and `r max.mort.spp$mean` (*`r max.mort.spp$spp`*, `r t.ref("tbl_mort")`).
-California had four species with greater mortality rates than Oregon and/or Washington: *Abies concolor*, *Pinus ponderosa*, *Acer macrophyllum*, and *Quercus kelloggii*.
-Oregon had one species with a greater mortality rate (*Tsuga mertensiana*) and Washington had four (*Abies grandis*, *Larix occidentalis*, *Picea engelmannii*, and *Pseudotsuga menziesii*).
+Four species in California exhibited greater mortality rates than in Oregon and/or Washington: *Abies concolor*, *Pinus ponderosa*, *Acer macrophyllum*, and *Quercus kelloggii*.
+One species had a greater mortality rate in Oregon (*Tsuga mertensiana*) and Washington had five (*Abies grandis*, *Larix occidentalis*, *Picea engelmannii*, *Pinus contorta*, and *Pseudotsuga menziesii*).
 Seven species had overall mortality rates $\geq$ 0.20 (*Abies concolor*, *Abies lasiocarpa*, *Picea engelmannii*, *Pinus contorta*, *Pinus lambertiana*, *Alnus rubra*, and *Quercus chryselopis*) while at a state level seven species had mortality rates $\geq$ 0.25 (California: *Acer macrophyllum* and *Alnus rubra*; Oregon: *Abies lasiocarpa*; Washington: *Abies grandis*, *Picea engelmannii*, *Pinus contorta*, and *Pinus monticola*).
@@
 
+The dashboard revealed domains of interest ("significant") for certain species that aligned with geographical regions.
+For the Above and Below Low domains *Abies lasiocarpa* had less mortality in northeastern Washington and *Tsuga mertensiana* exhibited greater growth in the same region.
+*Pinus contorta* exhibited less growth and greater mortality in northern Washington (Above Low).
+In the southern Sierra Nevada *Pinus lambertiana* had greater growth (Above Medium and High) and greater mortality (Above High), *Calocedrus decurrens* had greater mortality (Above High), and *Pinus monticola* had less mortality (Above High).
+*Acer macrophyllum* exhibited less growth in the Sierra Nevada and southern Klamath Mountains (Above High) and greater growth in northwestern Washington in the Coast and Cascade ranges (Above and Below Low).
+
 The MC permutation analysis provided empirical p-values for observing contrast confidence intervals with bounds at least as far from zero as those observed (Supplemental figures S5 and S6), with values ranging from a maximum of $p$ = 0.072 to a minimum $p$ < 0.0001 (41 instances).
-Except where explicitly stated, we only discuss individual species' domain findings if the species has one or more relevant MC permutation p-values < 0.001 (indicated by three or four asterisks in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")`)
+Except where explicitly stated, we only discuss individual species' domain findings if the species has one or more relevant MC permutation p-values < 0.001 (indicated by three or four asterisks in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")`).
 
@@
 
-Conspecifics *Abies lasiocarpa* and *Picea engelmannii* are frequently codominant and exhibit high mortality rates in our study.
-Their respective mortality rates in our analysis appear to be elevated; Andrus et al.
+Conspecifics *Abies lasiocarpa* and *Picea engelmannii* are frequently codominant and exhibit high mortality rates.
+Andrus et al.
 (2018) found that *Picea engelmannii* exhibited higher survival and longevity compared to *Abies lasiocarpa* in undisturbed plots in Colorado, with firs exhibiting a mortality rates of up to 1% annually while *Picea engelmannii* had mortality rates below 0.5% annually.
 Mortality estimates for *Abies lasiocarpa* from Perret et al.
-(2023) approximated our state and overall estimates at 25% +/-2.7, with mortality concentrated in the eastern Oregon Cascades as well as in Idaho and Colorado.
-They found that 83% of the mortality was due to fire or biogenic disturbance.
-The pressures on them appear to differ by CWD values and location.
+(2023) approximated our state and overall estimates at 25% +/-2.7; they found the highest level of mortality in the eastern Oregon Cascades as well as in Idaho and Colorado.
+They attributed 83% of the mortality to fire or biogenic disturbance.
+We found that the pressures on the two species appear to differ by CWD values and location.
 We found that *Picea engelmannii* had lower mortality in the Above High domain, corresponding to stands in the Blue Mountains of Oregon and Washington, and higher mortality in the Above Low and Above High domains, largely coinciding with northern and northeastern Washington, where spruce beetle were elevated (WADNR 2023).
@@
 Our Washington mortality estimate was twice that of Oregon’s and five times greater than Calfornia’s, while the growth confidence intervals for Washington were lower than California’s and did not overlap.
-Mortality was highest in the Low CWD domains and growth lowest in the Above Low domain and also high in the Below Low and Below Medium domains.
-These domains aligns well with plots for Washington, in regions that have experienced water stress (Raymond et al.
+The domains with the highest mortality and lowest growth align well with Washington regions that have experienced water stress (Raymond et al.
 2023).
@@
 This area also corresponds, at the level of ecoregion subsections, with the portions of *Pinus contorta*’s range in the lowest forest stability index category for Stanke et al.
-(2021) while California’s Sierra Nevadas and the Oregon and Washington Cascade Range correspond to positive stability index values.
+(2021) while California’s Sierra Nevada and the Oregon and Washington Cascade Range correspond to positive stability index values.
 Stanke et al.
@@
 Johnson et al.
-(2025) report mortality rates of *Abies grandis* in protected areas in the Malheur National Forest of Oregon that exceed ours, primarily attributing fire as the cause.
+(2025) report mortality rates of *Abies grandis* in protected areas in the Malheur National Forest of Oregon that exceed ours, primarily attributing the cause to fire.
 
@@
 
-We did not detect sizable changes in CWD between visits, potentially signalling that 10 years is too soon to observe CWD-related species responses (perhaps moreso for mortality than growth).
+We did not detect sizable changes in CWD between visits, potentially signalling that 10 years is too soon to observe CWD-related species responses (perhaps more true for mortality than growth).
 The CWD changes we observed may have represented greater proportional change for some species than others, with species such as *Abies amablis*, *Alnus rubra*, and *Juniperus occidentalis* exhibiting narrow and low water stress initial CWD ranges.
@@
 Our analysis may overall capture significant changes in mortality and growth for the tree species examined, with some trees exhibiting increases in mortality and decreases in growth either in portions of their ranges or across their entire range.
-For some species in semi-arid environments such as *Pinus pondersosa*, *Pseudotsuga menziesii*, *Pinus contorta*, *Picea engelmannii*, and *Abies lasiocarpa*, we may be observing ecosystem shifts with a change in water stress combined with disturbances such as wildfire and insect mortality (Halofsky et al.
+For some species in semi-arid environments such as *Pinus ponderosa*, *Pseudotsuga menziesii*, *Pinus contorta*, *Picea engelmannii*, and *Abies lasiocarpa*, we may be observing ecosystem shifts with a change in water stress combined with disturbances such as wildfire and insect mortality (Halofsky et al.
 2020, Parks et al.
@@
 Our hope is that our bivariate domain-based estimation approach is useful for others, while our results for individual species serve bolster or direct others' inquiries into the status of the species we examined.
-For managers, the results may be used to bolster action for resiliency.
+For managers, the results may be used to inform action for resiliency.
 With the shift in ecosystems, landowners can assist with species migration when planting or replanting (Raymond et al.
@@
 Our analysis and the analyses of others provide valuable and reliable insights into the condition of several tree species that exhibit abnormally high mortality (and low growth).
-Our hope si that these findings will help drive management and further monitoring of these species in the rapidly changing landscape of western North America.
+Our hope is that these findings will help drive management and further monitoring of these species in the rapidly changing landscape of western North America.
 

### 2025-11-14 Analysis Code/Functions.R

@@ -985,7 +985,6 @@ domain.map.fcn <- function(domain.matrix, spp.num, n.plots.used, virid.use, poin
   
   
   ggplot() +# + data = map.dat.2, aes(x = LON, y = LAT)) +
-    #coord_fixed(xlim = c(minlong - 1, maxlong + 1),  ylim = c(minlat - 1, maxlat + 1), ratio = 1.3) +
     geom_sf(data = map.dat.2, aes(color = factor(targ.spp)), size = point.size) +
     #  geom_tile(data = map.dat.1, mapping = aes(x = LON, y = LAT, z = targ.spp), binwidth = 0.15, 
     #            stat = "summary_2d", fun = mean, na.rm = TRUE, show.legend = FALSE) + 
@@ -995,8 +994,10 @@ domain.map.fcn <- function(domain.matrix, spp.num, n.plots.used, virid.use, poin
     geom_polygon(data = west_df, mapping = aes(x = long, y = lat, group = group), color = "black", fill = "transparent") +
     geom_text(data = mountain_ranges_df,
               aes(x = X, y = Y, label = name, angle = ang),
-              size = 3, fontface = "bold.italic", color = mountain_ranges_df$colr,
+              size = 4, fontface = "bold.italic", color = mountain_ranges_df$colr,
               check_overlap = TRUE) +
+
+    coord_sf(xlim = c(minlong - 1, maxlong + 1),  ylim = c(minlat - 1, maxlat + 1)) +
     theme_void() +
     theme(panel.border = element_rect(colour = "black", fill=NA, linewidth = 0.5),
           legend.position = "none")

### 2025-11-14 app.R

@@ -45,19 +45,21 @@ ui <- fluidPage(
   
   # Control panel at top
   fluidRow(
-    column(6,
-           selectInput("analysis_type", 
-                       "Analysis Type:",
-                       choices = list("Growth" = 1, "Mortality" = 2),
-                       selected = 1)
-    ),
-    column(6,
+    column(3,
            selectInput("species", 
                        "Select Species:",
                        choices = NULL,  # Will be populated in server
                        selected = NULL)
+           ),    
+    column(3,
+           selectInput("analysis_type", 
+                       "Analysis Type:",
+                       choices = list("Growth" = 1, "Mortality" = 2),
+                       selected = 1)
     )
   ),
+  
+
   
   # Main plot area
   fluidRow(

### 2025-11-19 Manuscript.Rmd
Latest manuscript changes to RMD file

@@
 grow.tot.sig <- grow.AvB.sig + grow.A.sig + grow.B.sig
+grow.tot <- grow.n.AvB + grow.n.A + grow.n.B
 
@@
 mort.tot.sig <- mort.AvB.sig + mort.A.sig + mort.B.sig
+mort.tot <- mort.n.AvB + mort.n.A + mort.n.B
 
@@
 fig.domains <- fig_nums(name = "fig_domains",
-                        caption = "Decadal (A) growth and (B) mortality domain estimates for *Calocedrus decurrens*.
+                        caption = "This figure illustrates our bivariate domain approach using *Calocedrus decurrens* as an example.
+Here we depict decadal (A) growth and (B) mortality domain estimates.
 The six domains are defined as FIA plots that fall [A]bove or [B]elow the CWD change threshold with [L]ow, [M]edium, or [H]igh initial CWD values (see text for details).
@@
 
+Western North American forests are experiencing accelerating tree mortality rates in response to intensifying drought and water stress driven by climate change.
+Tree mortality and growth reflect cumulative effects of direct physiological stress (xylem cavitation, carbon starvation), and indirect climate-mediated impacts (wildfire, insects, pathogens).
+However, obtaining design-based estimates for these demographic rates requires addressing a methodological challenge: revisit data are necessary to measure tree fate and growth, but plots differ in their initial water stress conditions and the magnitude of change between visits, necessitating a bivariate rather than univariate estimation approach.
+We developed a bivariate design-based estimation framework that stratifies plots into domains defined by both baseline initial-visit climatic water deficit (CWD) and change in CWD between visits.
+We applied this approach to `r comma(analysis.stats$plot.num)` National Forest Inventory plots spanning California, Oregon, and Washington, tracking `r comma(analysis.stats$tree.num)` trees from `r length(sel.spp)` species over decade-long intervals.
+We estimated decadal growth and mortality overall, by state, and across six CWD domains per species, using bootstrap confidence intervals and Monte Carlo permutation tests to assess statistical significance.
+Species responses were highly variable, with no universal pattern of increased mortality or reduced growth in domains experiencing elevated water stress.
+Several species exhibited overall mortality rates exceeding 0.20 per decade (20% decadal mortality), with state estimates of 0.34 for *Picea engelmannii* in Washington, 0.27 for *Acer macrophyllum* in California, and 0.37 for *Pinus contorta* in Washington.
+Domain-specific patterns aligned with known regional disturbances (bark beetles, drought, fire) for multiple species.
+Our bivariate framework provides a template for monitoring forest vital rates as ecosystem-level indicators of climate change response, while our species-specific findings identify or validate populations warranting immediate conservation assessment and inform forest management strategies.
+
 # Introduction
 
-Numerous tree species in western North America exhibit increased mortality rates or decline.
-van Mantgem et al.
-(2009) modeled mortality rates for the genera *Abies* and *Pinus* in the western USA as increasing from less than 0.05% annually in 1960 to almost 0.2% by 2010.
+Western North American forests are experiencing increases in tree mortality across multiple species in concert with changing climate conditions.
+Species and tree groups have incurred extraordinary losses in specific regions.
 Johnston et al.
 (2025) documented a 29% 10-year mortality rate of 300+ yr old trees in Eastern Oregon.
-Perret et al.
-(2023) estimated a decadal range-wide mortality rate of 25% for Subalpine fir (*Picea engelmannii*).
 High-severity fires in California in 2020 and 2021 killed 13%-19% of extant giant sequoia (*Sequoiadendron gianteum*) with >1.07-m bole diameters (Stephenson and Brigham 2021, Shive et al.
@@
 2022).
+Species and genera have also seen their mortality rates increase overall.
+van Mantgem et al.
+(2009) modeled mortality rates for the genera *Abies* and *Pinus* in the western USA as increasing from less than 0.05% annually in 1960 to almost 0.2% by 2010.
+Perret et al.
+(2023) estimated a decadal range-wide mortality rate of 25% for Subalpine fir (*Picea engelmannii*).
 Five of eight common tree species in western North America exhibit declines as measured by relative stand density (Stanke et al.
@@
 
-Climate change as manifest through increases or intensification in drought and aridity can affect tree mortality directly through physiological damage as well as indirectly by increasing wildfire frequency and severity and by promoting insect and disease outbreaks.
+Climate change as manifest through increases or intensification in drought and aridity can affect tree mortality directly and indirectly.
+Direct impacts include physiological damage while indirect causes include increasing wildfire, insect, and disease outbreak frequency and severity.
 Hammond et al.
@@
 Carbon starvation and hydraulic failure can drive drought-induced mortality (see reviews by Adams et al.
-2010 and McDowell et al.
+2010, Sala et al.
+2010, 2012, and McDowell et al.
 2008).
@@
 
+Given the importance of water stress on tree mortality, selecting an appropriate metric is critical for rangewide assessments.
 One regionally useful metric of water stress is Climatic Water Deficit (CWD; Stephenson 1998).
@@
 Both studies used a single climate variable value associated with each plot for estimation; e.g., Groom and Monleon (2023) examined shifts in species density by estimating the difference in VPD means between plots that gained trees of a given species and plots that lost trees.
-They documented, however, that the initial climate variable values could (and often did) change between plot visits, with species' plots often exhibiting positive or negative correlations between the initial visit values and between-visits change value of the climate variable.
+
+Groom et al.
+(2023) documented, however, that the initial climate variable values could (and often did) change between plot visits, with species' plots often exhibiting positive or negative correlations between the initial visit values and between-visits change value of the climate variable.
 They identified that their design-based analysis would potentially improve if they could simultaneously include both initial visit and change values for the climate variable.
 
-The purpose of this analysis is to demonstrate an approach for a bivariate design-based estimation procedure.
-We estimate tree species’ growth and mortality across the western U.S.
-states of California, Oregon, and Washington as they relate to plot-level initial-visit water stress (initial CWD) and the change in water stress between plot visits (delta CWD).
-We approached this bivariate analysis by dividing species’ plot data into domains based on initial and delta CWD values, finding growth and mortality estimates for each domain, and estimating domain contrasts to establish differences between domains.
-Secondarily we provide decadal growth and mortality estimates for our analysis species overall and by state to establish species status and to inform our CWD analysis findings.
+Our study has two objectives.
+The first is to demonstrate, across 23 species, a bivariate design-based estimation approach for tree growth and mortality that simultaneously accounts for initial climatic conditions and change between measurements.
+This addresses a gap in design-based analyses of forest change.
+Second, we provide current growth and mortality estimates for our analysis species across our study area and by state to establish species status and to inform our climate analysis findings.
 
@@
 Our analysis depends on examining the fate and growth of trees measured during two sequential visits (referred to as the first and second plot visit), ideally from a complete sample panel remeasure.
+Tree growth is the observed increase in marked tree bole diameter between visits (cm$^2$) while mortality is captured from the change in tree condition.
 We obtained our FIA data from the USDA FIA DataMart web portal (https://research.fs.usda.gov/products/dataandtools/fia-datamart; downloaded May 5, 2025).
@@
 Latitude and longitude were based on FIA-fuzzed coordinates.
-We defined the 'initial CWD' values as the 20-year mean of summer CWD values for each plot prior to and including the first visit year.
-We then found the mean of summer CWD values between visits and including the second visit year.
-We then defined 'delta CWD' values as the difference between this second value and initial CWD value.
+We defined the 'initial CWD' values as the 20-year mean of summer CWD values for each plot prior to and including the first visit year (initial CWD = mean(summer CWD from visit year$_{-19}$ to year$_0$).
+We then found the mean of summer CWD values between visits and including the second visit year (second visit CWD = mean(summer CWD from year$_1$ to year$_10$) .
+We then defined delta CWD = second visit CWD - initial CWD.
 Figure 1 presents maps of both the initial CWD and delta CWD values for forested and nonforested FIA plots.
@@
 
-We estimated, for each species examined, six domains based on CWD values (Figure 2C).
-The initial CWD plot values are divided into three species-specific quantiles: values below the first quartile (Low), between the first and third quartiles (Medium), and above the third quartile (High).
-Delta CWD plot values for all species were divided into those above a 5mm CWD change threshold (Above) and those below the threshold (Below).
+We estimated, for each species examined, six domains based on CWD values.
+The domains allow us to parse the plot CWD values into three levels for initial CWD values and two levels for delta CWD values.
+We created the domains based on initial CWD plot values by dividing the values into three species-specific quantiles: values below the first quartile (Low), between the first and third quartiles (Medium), and above the third quartile (High; Figure 2C).
+We defined the two domains for delta CWD plot values for all species as those above a 5mm CWD change threshold (Above) and those below the threshold (Below).
 The 5mm threshold was selected because species' delta CWD value distributions did not differ greatly and because 5mm is approximately the mean delta CWD value across species (Table S2).
@@
 The Below domains are similarly labeled (BL, BM, BH).
-We also report species mortality and growth estimates for all plots across and within Oregon, Washington, and California irrespective of CWD values.
+Parsing the plot data into six domains allow for bivariate examinations of each species' data.
+This bivariate approach allows us to ask, for example, whether trees in historically dry sites (High initial CWD) respond differently to increased water stress (Above threshold delta CWD) compared to trees in mesic sites (Low initial CWD) experiencing similar increases.
+We calculated estimates for all CWD-based domains with 10 or more plots.
+
+We additionally report species mortality and growth estimates for all plots across and within Oregon, Washington, and California irrespective of CWD values.
 We limited the number of species examined to those with more than 400 occupied plots across the three states.
-We calculated estimates for all CWD-based domains with 10 or more plots.
 
@@
 
-Our analysis approach results in many comparisons.
+Our analysis approach results in hundreds of comparisons.
 We conducted a Monte Carlo (MC) permutation analysis to provide species-level false positive rates (p-values) for "significant" contrasts.
@@
 
+Given the large number of contrasts conducted (`r mort.tot` each for growth and mortality), we adopted a conservative approach to interpretation.
+Aside from discussing patterns of findings across species, we focus discussion on contrasts where (1) the 95% CI does not include zero, and (2) the MC permutation test yields p < 0.001, indicating that the observed pattern is unlikely under random domain assignment.
+Contrasts with less stringent p-values are shown in figures but are generally not discussed individually as they may represent Type I errors.
+
 ## Dashboard
@@
 The two visits were 10 years apart for `r pct.remeas[2]`% of plots (`r pct.remeas[1]`% were 9 years, `r pct.remeas[3]`% were 11 years).
-Plots were distributed in `r analysis.stats$strat.num` strata.
+Plots were distributed across `r analysis.stats$strat.num` strata spanning forested regions in all three states, from coastal temperate rainforest to semi-arid interior woodlands.
 
@@
 Three species had greater growth in Oregon (*Calocedrus decurrens*, *Picea engelmanii*, and *Pseudotsuga menziesii*) and two species had greater growth in Washington (*Abies amablis* and *Pinus monticola*).
+
 Decadal mean mortality rates varied between `r min.mort.spp$mean` (*`r min.mort.spp$spp`*) and `r max.mort.spp$mean` (*`r max.mort.spp$spp`*, `r t.ref("tbl_mort")`).
@@
 Growth domain contrasts (Figure `r fig_nums("fig_diffgrow", display = "num")`, `r t.ref("tbl_summ_grow_mort")`) did not appear to result in substantially more "significant" negative or positive estimates, indicating that species overall did not exhibit more or less growth in the Above, Low, or High domains.
+
 All three columns of Figure `r fig_nums("fig_diffgrow", display = "num")` had over 20 "significant" contrasts.
@@
 Below Threshold" contrasts (`r mort.AvB.sig`, `r t.ref("tbl_summ_grow_mort")`) than for Above (`r mort.A.sig`) or Below (`r mort.B.sig`), indicating that mortality may not be as strongly associated with delta CWD as by the initial CWD value.
-*Larix occidentalis*, *Picea engelmannii*,	*Pinus contorta*,	*Pinus monticola*, *Thuja plicata*,	and *Arbutus menziesii* had greater mortality in the lower water deficit Low domains than Medium and/or High domains.
-These results occurred in Above (3 spp.) or both Above and Below (3 spp.) domains.
+
+*Larix occidentalis*, *Picea engelmannii*,	*Pinus contorta*,	*Pinus monticola*, and *Thuja plicata* had greater mortality in the lower water deficit Low domains than Medium and/or High domains.
+These results occurred in Above (3 spp.) or both Above and Below (2 spp.) domains.
 Mortality rates for *Abies lasiocarpa*,	*Calocedrus decurrens*,	*Pinus lambertiana*, *Pinus ponderosa*,	*Pseudotsuga menziesii*,	*Tsuga heterophylla*,	*Tsuga mertensiana*,	*Acer macrophyllum*,	and *Notolithocarpus densiflorus* were greater in the High domains than the Medium and/or Low domains.
 These species were approximately evenly split between differences being present in the Above (4 spp.) or Below (5 spp.) domains, but none with "significant" findings in both the Above and Below domains.
+*Arbutus menziesii* had greater mortality in Medium domains than the High domains.
 Five species had no significant contrasts.
 
-The dashboard revealed domains of interest ("significant") for certain species that aligned with geographical regions.
+The interactive dashboard revealed geographic patterns that were associated with specific domains while not necessarily captured by state-level estimates.
 For the Above and Below Low domains *Abies lasiocarpa* had less mortality in northeastern Washington and *Tsuga mertensiana* exhibited greater growth in the same region.
@@
 
-Across the 23 species examined we found no clear pattern of response to initial and/or delta CWD values.
-Responses were species-specific, potentially driven in some cases by local phenomena and captured by a domain category or categories.
-We did find more “significant” CWD contrasts for species growth than mortality, possibly indicating that growth differences by CWD domain were more readily detectable than mortality over a decade, although growth and mortality findings were not necessarily negatively related.
+Analysis of 23 tree species revealed diverse, species-specific associations with climatic water deficit, with different species exhibiting contrasting relationships among growth, mortality, and water stress.
+Responses were potentially driven by differences in drought tolerance strategies, local disturbance regimes, and biotic interactions.
+We found more “significant” CWD contrasts for species growth than mortality, possibly indicating that growth differences by CWD domain were more readily detectable than mortality over a decade, although growth and mortality findings were not necessarily negatively related.
 Delta CWD increased almost everywhere between plot revisits although the magnitude of change was generally not large relative to initial CWD values.
-We additionally found for some species that decadal mortality rates by state and/or region were concerningly high.
-
-The number of "significant" Above vs.
-Below threshold contrasts for growth and mortality were approximately evenly distributed between positive and negative contrast estimates, with a positive difference indicating that some species exhibited higher growth or greater mortality in an Above domain (e.g., *Pinus lambertiana* for both growth and mortality) versus a Below domain (*Abies grandis* and *Pinus contorta*, respectively, for growth and mortality).
-This lack of a pattern across species may indicate that the distinction of Above vs.
-Below is not generalizable across species, or that "significant" findings are due to confounding factors such as geographical locations with different ecological pressures.
+We additionally found concerningly high state and/or overall decadal mortality rates for some species.
 
 Contrast results may be more informative at an individual species level and when considering state-level mortality and growth estimates.
-Regional analyses detected several species with relatively high mortality rates.
+Regional analyses detected several species with relatively high mortality rates including *Acer macrophyllum*, *Picea engelmannii*, *Abies lasiocarpa*, *Pinus contorta*, and *Alnus rubra*.
+We do not believe our results are concerning for *Alnus rubra*.
+Our estimates for this short-lived tree species align with reported background mortality rates (e.g., Brown and Courtin 2018).
+However, the mortality rate for *Acer macrophyllum* in California appears to be both unreported elsewhere and substantially higher in California than in Oregon or Washington.
+
 *Acer macrophyllum* growth was negatively associated with initial CWD values, and its growth was lower and mortality almost three times higher in California than in Washington and Oregon.
@@
 
-We do not believe our results are concerning for *Alnus rubra*.  Our estimates for this short-lived tree species align with reported background mortality rates (e.g., Brown and Courtin 2018).
-
 Conspecifics *Abies lasiocarpa* and *Picea engelmannii* are frequently codominant and exhibit high mortality rates.
@@
 We found that the pressures on the two species appear to differ by CWD values and location.
-We found that *Picea engelmannii* had lower mortality in the Above High domain, corresponding to stands in the Blue Mountains of Oregon and Washington, and higher mortality in the Above Low and Above High domains, largely coinciding with northern and northeastern Washington, where spruce beetle were elevated (WADNR 2023).
+We found that *Picea engelmannii* had lower mortality in the Above High domain, corresponding to stands in the Blue Mountains of Oregon and Washington, and higher mortality in the Above Low and Above High domains, largely coinciding with northern and northeastern Washington, where spruce beetle infestations were elevated (WADNR 2023).
 Its mortality rate in Washington was almost 5 times higher than in Oregon.
@@
 also found, from FIA plot data, that insects (followed by fire) were related to the percent change in relative tree density.
+
+The absence of consistent contrast patterns across species is itself informative, indicating that tree responses to water stress conditions are potentially mediated by local adaptation to conditions, species-specific physiological traits, and local environmental conditions rather than following a general response curve.
+The number of "significant" Above vs.
+Below threshold contrasts for growth and mortality were approximately evenly distributed between positive and negative contrast estimates, with a positive difference indicating that some species exhibited higher growth or greater mortality in an Above domain (e.g., *Pinus lambertiana* for both growth and mortality) versus a Below domain (*Abies grandis* and *Pinus contorta*, respectively, for growth and mortality).
+This lack of a pattern across species may indicate that the distinction of Above vs.
+Below is not generalizable across species, or that "significant" findings are due to confounding factors such as geographical locations with different ecological pressures.
 
@@
 2015).
-Interrelated predisposing conditions for individual tree mortality such as canopy openness (Lemmo et al.
+Our regional analysis necessarily aggregates across local-scale variation in interrelated predisposing conditions for individual tree mortality such as canopy openness (Lemmo et al.
 2024), structural complexity (Ma et al.
@@
 2019), and within-stand competition (Ford et al.
-2017) would be missed by our regional evaluation.
+2017).
+For management applications, particularly those for individual stands, our regional and domain estimates should be viewed as general results that may be modified by these local factors.
 
@@
 2023).
-Our hope is that our bivariate domain-based estimation approach is useful for others, while our results for individual species serve bolster or direct others' inquiries into the status of the species we examined.
+Our hope is that our bivariate domain-based estimation approach is useful for others, while our results for individual species serve to bolster or direct others' inquiries into the status of the species we examined.
 For managers, the results may be used to inform action for resiliency.
@@
 Our analysis and the analyses of others provide valuable and reliable insights into the condition of several tree species that exhibit abnormally high mortality (and low growth).
-Our hope is that these findings will help drive management and further monitoring of these species in the rapidly changing landscape of western North America.
+Our hope is that these findings will inform management and encourage focused monitoring efforts for these species in the rapidly changing landscape of western North America.
 
@@
 Stephenson, N.L., 1990. Climatic control of vegetation distribution: the role of the water balance. The American Naturalist, 135(5), pp.649-670.     
+Sala, A., Piper, F. and Hoch, G., 2010. Physiological mechanisms of drought-induced tree mortality are far from being resolved. The New Phytologist, 186(2), pp.274-281.
+
+Sala, A., Woodruff, D.R. and Meinzer, F.C., 2012. Carbon dynamics in trees: feast or famine?. Tree physiology, 32(6), pp.764-775.
 

### 2025-12-12 Manuscript.Rmd
Incorporated edits from B. Frank.  This added a figure, changed the supplimental and main RMarkdown files, and altered many of the scripts involved in generating plots and tables.

@@
 ---
-title: "Tree growth, mortality, and climatic stress in west-coast states, USA"
+title: "Tree growth, mortality, and climatic water deficit in west-coast states, USA"
 author:
@@
 ```{r load-data}
+n.spp <- length(sel.spp)
+
 # Some general analysis info + plots
@@
 
+# Function to extract a value out of the contrast figures by species and contrast
+fig.val.fcn <- function(dat, spp.x, cont, rnd){
+dat.source <- if(dat == "g") grow.diff else mort.diff
+k <- if(dat == "g") 1 else 2
+IvS2 <- cm2.fcn(k, results.table = dat.source$IvS2)
+I_LMH2 <- cm2.fcn(k, results.table = dat.source$I_LMH2)
+S_LMH2 <- cm2.fcn(k, results.table = dat.source$S_LMH2)
+tab.comb <- bind_rows(IvS2, I_LMH2, S_LMH2)
+
+val.out <- round(tab.comb$Means[tab.comb$Species == spp.x & tab.comb$Domain == cont], rnd)
+return(val.out)
+}
+
 # Significant findings, Growth
-grow.n.AvB <- nrow(grow.diff$AvB2)
-grow.n.A <- nrow(grow.diff$A_LMH2)
-grow.n.B <- nrow(grow.diff$B_LMH2)
-
-grow.AvB.sig <- length(grow.diff$AvB2$significant[grow.diff$AvB2$significant == TRUE])
-grow.AvB.signeg <- length(grow.diff$AvB2$significant[grow.diff$AvB2$significant == TRUE & grow.diff$AvB2$Means < 0])
-grow.A.sig <- length(grow.diff$A_LMH2$significant[grow.diff$A_LMH2$significant == TRUE])
-grow.A.signeg <- length(grow.diff$A_LMH2$significant[grow.diff$A_LMH2$significant == TRUE & grow.diff$A_LMH2$Means < 0])
-grow.B.sig <- length(grow.diff$B_LMH2$significant[grow.diff$B_LMH2$significant == TRUE])
-grow.B.signeg <- length(grow.diff$B_LMH2$significant[grow.diff$B_LMH2$significant == TRUE & grow.diff$B_LMH2$Means < 0])
-
-grow.tot.sig <- grow.AvB.sig + grow.A.sig + grow.B.sig
-grow.tot <- grow.n.AvB + grow.n.A + grow.n.B
+grow.n.IvS <- nrow(grow.diff$IvS2)
+grow.n.I <- nrow(grow.diff$I_LMH2)
+grow.n.S <- nrow(grow.diff$S_LMH2)
+
+grow.IvS.sig <- length(grow.diff$IvS2$significant[grow.diff$IvS2$significant == TRUE])
+grow.IvS.signeg <- length(grow.diff$IvS2$significant[grow.diff$IvS2$significant == TRUE & grow.diff$IvS2$Means < 0])
+grow.I.sig <- length(grow.diff$I_LMH2$significant[grow.diff$I_LMH2$significant == TRUE])
+grow.I.signeg <- length(grow.diff$I_LMH2$significant[grow.diff$I_LMH2$significant == TRUE & grow.diff$I_LMH2$Means < 0])
+grow.S.sig <- length(grow.diff$S_LMH2$significant[grow.diff$S_LMH2$significant == TRUE])
+grow.S.signeg <- length(grow.diff$S_LMH2$significant[grow.diff$S_LMH2$significant == TRUE & grow.diff$S_LMH2$Means < 0])
+
+grow.tot.sig <- grow.IvS.sig + grow.I.sig + grow.S.sig
+grow.tot <- grow.n.IvS + grow.n.I + grow.n.S
 
 # Significant findings, Mort
-mort.n.AvB <- nrow(mort.diff$AvB2)
-mort.n.A <- nrow(mort.diff$A_LMH2)
-mort.n.B <- nrow(mort.diff$B_LMH2)
-
-mort.AvB.sig <- length(mort.diff$AvB2$significant[mort.diff$AvB2$significant == TRUE])
-mort.AvB.signeg <- length(mort.diff$AvB2$significant[mort.diff$AvB2$significant == TRUE & mort.diff$AvB2$Means < 0])
-mort.A.sig <- length(mort.diff$A_LMH2$significant[mort.diff$A_LMH2$significant == TRUE])
-mort.A.signeg <- length(mort.diff$A_LMH2$significant[mort.diff$A_LMH2$significant == TRUE & mort.diff$A_LMH2$Means < 0])
-mort.B.sig <- length(mort.diff$B_LMH2$significant[mort.diff$B_LMH2$significant == TRUE])
-mort.B.signeg <- length(mort.diff$B_LMH2$significant[mort.diff$B_LMH2$significant == TRUE & mort.diff$B_LMH2$Means < 0])
-
-mort.tot.sig <- mort.AvB.sig + mort.A.sig + mort.B.sig
-mort.tot <- mort.n.AvB + mort.n.A + mort.n.B
+mort.n.IvS <- nrow(mort.diff$IvS2)
+mort.n.I <- nrow(mort.diff$I_LMH2)
+mort.n.S <- nrow(mort.diff$S_LMH2)
+
+mort.IvS.sig <- length(mort.diff$IvS2$significant[mort.diff$IvS2$significant == TRUE])
+mort.IvS.signeg <- length(mort.diff$IvS2$significant[mort.diff$IvS2$significant == TRUE & mort.diff$IvS2$Means < 0])
+mort.I.sig <- length(mort.diff$I_LMH2$significant[mort.diff$I_LMH2$significant == TRUE])
+mort.I.signeg <- length(mort.diff$I_LMH2$significant[mort.diff$I_LMH2$significant == TRUE & mort.diff$I_LMH2$Means < 0])
+mort.S.sig <- length(mort.diff$S_LMH2$significant[mort.diff$S_LMH2$significant == TRUE])
+mort.S.signeg <- length(mort.diff$S_LMH2$significant[mort.diff$S_LMH2$significant == TRUE & mort.diff$S_LMH2$Means < 0])
+
+mort.tot.sig <- mort.IvS.sig + mort.I.sig + mort.S.sig
+mort.tot <- mort.n.IvS + mort.n.I + mort.n.S
 
 # Summary of significant Growth and Mortality findings for table creation.
-datasets.grow <- list(grow.diff$AvB2, grow.diff$A_LMH2, grow.diff$B_LMH2)
-datasets.mort <- list(mort.diff$AvB2, mort.diff$A_LMH2, mort.diff$B_LMH2)
-datatypes <- list("Above vs. Below", "Above", "Below")
-n.comp.grow <- list(grow.n.AvB, grow.n.A, grow.n.B)
-n.comp.mort <- list(mort.n.AvB, mort.n.A, mort.n.B)
+datasets.grow <- list(grow.diff$IvS2, grow.diff$I_LMH2, grow.diff$S_LMH2)
+datasets.mort <- list(mort.diff$IvS2, mort.diff$I_LMH2, mort.diff$S_LMH2)
+datatypes <- list("Increasing vs. Stable", "Increasing", "Stable")
+n.comp.grow <- list(grow.n.IvS, grow.n.I, grow.n.S)
+n.comp.mort <- list(mort.n.IvS, mort.n.I, mort.n.S)
 
@@
 grow.all.cm <- cm2.fcn(1, grow.all)
-s.g.dat <- state.table.fcn(grow.state.cm, grow.all.cm, 1)
+s.g.dat <- state.table.fcn(grow.state.cm, grow.all.cm, 1) %>%
+  left_join(spp.cat %>% dplyr::select(-spp_cat), by = c(`Species Name` = "Species")) %>%
+  rename("Category" = "cat_abbr") %>%
+  relocate(Category, .after = `Species Name`)
 
@@
 # State mortality
-s.m.dat <- state.table.fcn(mort.state, mort.all, 2)
+s.m.dat <- state.table.fcn(mort.state, mort.all, 2) %>%
+  left_join(spp.cat %>% dplyr::select(-spp_cat), by = c(`Species Name` = "Species")) %>%
+  rename("Category" = "cat_abbr") %>%
+  relocate(Category, .after = `Species Name`)
 
@@
 
-# CWD amounts
+# CWD amounts, tapping values from the initial & delta CWD tables in the suppliment
+init.cwd.stats <- analysis.stats$init.cwd.stats %>%
+  mutate(range05.95 = q95 - q05) %>%
+  arrange(Mean)
+
+delta.cwd.stats <- analysis.stats$diff.cwd.stats %>%
+  mutate(range05.95 = q95 - q05) %>%
+  arrange(Mean)
+
+cwd.summ.fcn <- function(data.2use, col.interest, place, stat.use) {
+end.start <- switch(stat.use, "min" = 1, "max" = n.spp) # Which row end (top or bottom) to start on?
+place2 <- if(stat.use == "min") place else n.spp - place + 1
+
+x <- data.2use %>% arrange(get(col.interest))
+x2 <- x[place2, ]
+x2.range <- paste0("(5th percentile = ", x2$q05, "mm, 95th percentile = ", x2$q95, "mm)")
+x2.range.full <- paste0("(5th-95th percentile range = ", x2$range05.95, "mm, 5th percentile = ", x2$q05, "mm, 95th percentile = ", x2$q95, "mm)")
+x2.mean.full <- paste0("(mean = ", x2$Mean, "mm, 5th percentile = ", x2$q05, "mm, 95th percentile = ", x2$q95, "mm)")
+
+return(list(spp.out = x2$Species, range.alone = x2$range05.95, mean.alone = x2$Mean,
+            spp.range = x2.range, 
+            spp.range.full = x2.range.full, spp.mean.full = x2.mean.full
+            ))
+}
+
+#Initial CWD: Tree species with the min/max means and ranges.
+init.mean.sm <- cwd.summ.fcn(init.cwd.stats, col.interest = "Mean", place = 1, stat.use = "min") #smallest mean
+init.mean.sm2 <-   cwd.summ.fcn(init.cwd.stats, col.interest = "Mean", place = 2, stat.use = "min") # 2nd smallest mean
+init.range.sm <- cwd.summ.fcn(init.cwd.stats, col.interest = "range05.95", place = 1, stat.use = "min") # smallest range
+init.range.sm2 <- cwd.summ.fcn(init.cwd.stats, col.interest = "range05.95", place = 2, stat.use = "min")
+init.mean.lg <- cwd.summ.fcn(init.cwd.stats, col.interest = "Mean", place = 1, stat.use = "max") # largest mean
+init.mean.lg2 <- cwd.summ.fcn(init.cwd.stats, col.interest = "Mean", place = 2, stat.use = "max")
+init.range.lg <- cwd.summ.fcn(init.cwd.stats, col.interest = "range05.95", place = 1, stat.use = "max")
+init.range.lg2 <- cwd.summ.fcn(init.cwd.stats, col.interest = "range05.95", place = 2, stat.use = "max")
+
+#Initial CWD: Tree species with the min/max means and ranges.
+delta.mean.sm <- cwd.summ.fcn(delta.cwd.stats, col.interest = "Mean", place = 1, stat.use = "min") #smallest mean
+delta.mean.sm2 <-   cwd.summ.fcn(delta.cwd.stats, col.interest = "Mean", place = 2, stat.use = "min") # 2nd smallest mean
+delta.range.sm <- cwd.summ.fcn(delta.cwd.stats, col.interest = "range05.95", place = 1, stat.use = "min") # smallest range
+delta.range.sm2 <- cwd.summ.fcn(delta.cwd.stats, col.interest = "range05.95", place = 2, stat.use = "min")
+delta.mean.lg <- cwd.summ.fcn(delta.cwd.stats, col.interest = "Mean", place = 1, stat.use = "max") # largest mean
+delta.mean.lg2 <- cwd.summ.fcn(delta.cwd.stats, col.interest = "Mean", place = 2, stat.use = "max")
+delta.range.lg <- cwd.summ.fcn(delta.cwd.stats, col.interest = "range05.95", place = 1, stat.use = "max")
+delta.range.lg2 <- cwd.summ.fcn(delta.cwd.stats, col.interest = "range05.95", place = 2, stat.use = "max")
+
 min.mean.init <- analysis.stats$init.cwd.stats$Mean[analysis.stats$init.cwd.min.mean]
@@
 Mean estimates with confidence intervals and plot number are provided.
-Estimates are not made for states with fewer than 10 plots.")
+Estimates are not made for states with fewer than 10 plots.
+Ecological category abbreviations (Category) are provided in the text.")
 
@@
 fig.domains <- fig_nums(name = "fig_domains",
-                        caption = "This figure illustrates our bivariate domain approach using *Calocedrus decurrens* as an example.
+                        caption = "This figure illustrates our domain analysis approach using *Calocedrus decurrens* as an example.
 Here we depict decadal (A) growth and (B) mortality domain estimates.
-The six domains are defined as FIA plots that fall [A]bove or [B]elow the CWD change threshold with [L]ow, [M]edium, or [H]igh initial CWD values (see text for details).
-E.g., 'AL' is Above CWD change threshold, Low initial CWD value.
-The distribution of plot CWD change and initial values (C) is presented as well, with the blue horizontal line indicating the CWD change threshold and the orange vertical lines representing the quantile boundaries between Low, Medium, and High initial CWD values." )
+The six domains are defined as FIA plots that fall into the [I]ncreasing or [S]table CWD change threshold with [L]ow, [M]edium, or [H]igh initial CWD values (see text for details).
+E.g., 'IL' is in the Increasing CWD change category, Low initial CWD value.
+The distribution of plot CWD change and initial values (C) is presented as well, with the blue horizontal line indicating the CWD change category and the orange vertical lines representing the quantile boundaries between Low, Medium, and High initial CWD values." )
+
+fig.cwd_initdelt <- fig_nums(name = "fig_cwdmeans", 
+                          caption = paste0("Mean species plot values for the climatic water deficit initial and change values."))
 
@@
 Mean domain differences (circles) with 95% CIs (bars) that do not intersect zero are filled.
-Species above the dashed line are coniferous; those below are deciduous.
-Empirical MC permutation p-values are provided for filled-circle contrasts, with \u2013\u2013 corresponding to p-values of > 0.05, while one to four asterisks correspond to p > 0.01, > 0.001, > 0.0001 and < 0.0001, respectively."))
+Species above the dashed line are conifers; those below are hardwoods."))
 
@@
 Mean domain differences (circles) with 95% CIs (bars) that do not intersect zero are filled.
-Species above the dashed line are coniferous; those below are deciduous.
-Empirical MC permutation p-values are provided for filled-circle contrasts, with \u2013\u2013 corresponding to p-values of > 0.05 while one to four asterisks correspond to p > 0.01, > 0.001, > 0.0001 and < 0.0001, respectively.
-"))
+Species above the dashed line are conifers; those below are hardwoods."))
 
@@
 Tree mortality and growth reflect cumulative effects of direct physiological stress (xylem cavitation, carbon starvation), and indirect climate-mediated impacts (wildfire, insects, pathogens).
-However, obtaining design-based estimates for these demographic rates requires addressing a methodological challenge: revisit data are necessary to measure tree fate and growth, but plots differ in their initial water stress conditions and the magnitude of change between visits, necessitating a bivariate rather than univariate estimation approach.
-We developed a bivariate design-based estimation framework that stratifies plots into domains defined by both baseline initial-visit climatic water deficit (CWD) and change in CWD between visits.
-We applied this approach to `r comma(analysis.stats$plot.num)` National Forest Inventory plots spanning California, Oregon, and Washington, tracking `r comma(analysis.stats$tree.num)` trees from `r length(sel.spp)` species over decade-long intervals.
-We estimated decadal growth and mortality overall, by state, and across six CWD domains per species, using bootstrap confidence intervals and Monte Carlo permutation tests to assess statistical significance.
+Relating design-based estimates for mortality and growth to a climate variable can be challenging, as directly assessing mortality and growth requires the visitation of plots at two different points in time.
+Plots can differ both in their initial climate variable value and the change in that value between visits, necessitating estimation domains that jointly consider initial conditions and temporal change.
+We developed an analysis framework that partitions plots into domains defined by their initial-visit climatic water deficit (CWD) value and the change in the CWD value between visits.
+We applied this approach to `r comma(analysis.stats$plot.num)` National Forest Inventory plots spanning California, Oregon, and Washington, tracking `r comma(analysis.stats$tree.num)` trees from `r n.spp` species over decade-long intervals.
+We estimated decadal growth and mortality overall, by state, and across six CWD domains per species, along with associated measures of uncertainty.
 Species responses were highly variable, with no universal pattern of increased mortality or reduced growth in domains experiencing elevated water stress.
-Several species exhibited overall mortality rates exceeding 0.20 per decade (20% decadal mortality), with state estimates of 0.34 for *Picea engelmannii* in Washington, 0.27 for *Acer macrophyllum* in California, and 0.37 for *Pinus contorta* in Washington.
+Several species exhibited overall mortality rates exceeding 0.20 (i.e., 20% decadal mortality), with estimates of 0.34 for *Picea engelmannii* in Washington, 0.27 for *Acer macrophyllum* in California, and 0.37 for *Pinus contorta* in Washington.
 Domain-specific patterns aligned with known regional disturbances (bark beetles, drought, fire) for multiple species.
-Our bivariate framework provides a template for monitoring forest vital rates as ecosystem-level indicators of climate change response, while our species-specific findings identify or validate populations warranting immediate conservation assessment and inform forest management strategies.
+Our study contributes a template for monitoring forest vital rates as ecosystem-level indicators of climate change response, and our species-specific findings identify or validate populations warranting immediate conservation assessment that can inform forest management strategies.
 
@@
 2009), or those which mainly rely on stomatal closure to avoid excessive water loss through evapotranspiration (ET), receive insufficient atmospheric carbon dioxide for photosynthesis.
-Anisohydric plants, or those who do not strongly regulate water loss through stomatal closure, risk hydraulic failure if sufficient xylem cavitation occurs due to insufficient water uptake needed to meet ET demands (e.g., as seen in *Populus tremuloides*; Anderegg et al.
+Anisohydric plants, or those that do not strongly regulate water loss through stomatal closure, risk hydraulic failure if sufficient xylem cavitation occurs due to insufficient water uptake needed to meet ET demands (e.g., as seen in *Populus tremuloides*; Anderegg et al.
 2015).
@@
 2022).
-
-Drought stress can increase tree susceptibility to wildfire, insects, and disease.
-Wildfire area (Abatzoglou and Williams, 2016; Parks et al.
-2025) and number (Westerling 2016) are positively associated with increasing aridity in the western US.
-Bark beetle outbreaks in semi-arid areas of the western US may be positively affected by or associated with acute intense drought and water-stressed trees (Negrón 2009, Gaylord et al.
-2015) but chronically stressed trees may not provide bark beetle populations with the resources sufficient to promote population expansion (Huberty and Denno, 2004).
-Norway spruce (*Picea abies*) saplings inoculated with the necrotrophic pathogenic fungi *Heterobasidion annosum* sensu stricto and exposed to drought conditions were more likely to die than control saplings or saplings exposed to either drought conditions or inoculation, although Scots pine (*Pinus sylvestris*) seedlings did not show a similar relationship (Gomez-Gallego et al.
-2022).
-Other insects and disease may be promoted or inhibited by drought, depending on the circumstances and due to conflicting factors (Kolb et al.
-2016).
 
@@
 National Forest Inventory, collected by the US Forest Service Forest Inventory and Analysis Program (FIA), is well suited for evaluating regional species mortality and growth.
-Since 1999 the FIA program has gathered a variety of forestry data at plots nationwide using a spatially balanced probability sample (McRoberts 2005, Reams et al.
+Since 1999 the FIA program has gathered a variety of forestry data at plots nationwide using a spatially balanced sampling design (McRoberts 2005, Reams et al.
 2005, Woudenberg et al.
 2010).
-The sample allows for relatively unbiased design-based inference of population characteristics across a large geographical region.
+The sampling design allows for approximately unbiased design-based inference of population characteristics across large geographical regions.
 Data from plot revisits and individual tree data permit direct estimation of change features such as mortality and growth estimation.
@@
 Monleon and Linz (2015) evaluated the evidence of tree species range shifts by subtracting the design-based mean estimates of FIA plot temperatures for seedlings from those of trees.
-FIA plot revisit data were mostly incomplete at the time of their analysis.
-Groom and Monleon (2023) used a mostly-completed revisit data set to compare plot climate variable values (vapor pressure deficit [VPD], temperature, and precipitation) to estimate shifts in plot occupancy and density.
+However, FIA plot revisit data were mostly incomplete at the time of their analysis.
+In contrast, Groom and Monleon (2023) used a mostly-completed revisit data set to compare plot climate variable values (vapor pressure deficit [VPD], temperature, and precipitation) to estimate shifts in plot occupancy and density.
 Both studies used a single climate variable value associated with each plot for estimation; e.g., Groom and Monleon (2023) examined shifts in species density by estimating the difference in VPD means between plots that gained trees of a given species and plots that lost trees.
@@
 Our study has two objectives.
-The first is to demonstrate, across 23 species, a bivariate design-based estimation approach for tree growth and mortality that simultaneously accounts for initial climatic conditions and change between measurements.
+The first is to demonstrate, across 23 species, a design-based estimation approach for tree growth and mortality that simultaneously accounts for initial climatic conditions and change between measurements.
 This addresses a gap in design-based analyses of forest change.
@@
 Our analysis relies on FIA data for forestland in California, Oregon, and Washington, USA.
-Forestland is defined as an area of 4,050 m$^2$ that is at least 10 percent stocked or potentially stocked with tree species, excluding urban and agricultural land uses (Reams et al.
+Forestland is defined as an area of at least 4,050 m$^2$ (i.e., one acre) that is at least 10 percent stocked or potentially stocked with tree species, excluding urban and agricultural land uses (Reams et al.
 2005).
@@
 
-FIA relies on three related sampling procedures.
-Phase 1 involves the classification of distributed points of remote sensing data into (at least) the categories of forest and non-forest.
-Phase 1 plots are useful for estimation stratification.
-The phase 2 permanent ground plots are a subset of Phase 1 points.
-They consist of four points where all trees with stem diameters greater than or equal to 12.7 cm (for most species measured at 1.37 m above the ground) were tallied in a 7.32 m radius circular subplot (total area 672.5 m$^2$) with additional trees sampled in an annulus around each subplot (macroplot), with a total radius 18 m from the center point and a total area of 1012 m$^2$.
+In the coterminous United States, the FIA program uses a spatially balanced, quasi-systematic sampling design.
+Plots are randomly located and produce sampling intensities of approximately 2,400 hectares per plot, with some denser areas such as those located on national forest system lands.
+In the western United States, each plot is assigned to one of ten panels, and approximately one panel is measured each year in perpetuity, yielding ten year remeasurement intervals on each plot, barring exceptions owing to optional temporal intensifications (Thompson 2015).
+In this study, we use only those plots that are part of the standard sampling intensity, and have been remeasured at an interval of ten years.
+Each FIA plot consists, in part, of four 7.32 m radius subplots where all trees with stem diameters greater than or equal to 12.7 cm (for most species measured at 1.37 m above the ground) were tallied with additional trees sampled in an annulus around each subplot (macroplot), with a total radius 18 m from the center point and a total area of 1012 m$^2$.
 We did not include data from microplots, where seedlings and saplings are measured, as the analysis was not intended to estimate species recruitment but rather to follow the trajectory of established trees.
-Phase 3 plots are an intensification of Phase 2 plots; we do not rely on their data here.
 Reams et al.
@@
 
-The FIA program selected plots with a spatially-balanced probability sampling design (McRoberts 2005, Reams et al., 2005) that used a target plot density of one ground plot every 24 km$^2$.
-Some regions and ownerships intensify their plot density; we excluded these additional plots to simplify stratification estimation, relying on FIA-defined strata and using the density of FIA Phase 1 sampling points within and across strata to determine stratum weights.
-
-Plots in the three states are sampled every 10 years with a target of annually sampling a panel of 10% of plots every year (Thompson 2015).
 Our analysis depends on examining the fate and growth of trees measured during two sequential visits (referred to as the first and second plot visit), ideally from a complete sample panel remeasure.
@@
 
-Individual trees within plots were retained in the analysis if they were alive at the initial visit and met the stem diameter criterion during that visit (i.e., plot ingrowth was not included).
+Remeasured trees within plots were retained in the analysis if they were alive at the initial visit and met the stem diameter criterion during that visit (i.e., plot ingrowth was not included).
 The growth analysis additionally restricted trees considered to those that were alive at both visits.
 The mortality analysis excluded trees killed through silvicultural practices.
-Both analyses excluded trees that shrank in diameter, physically moved due to landslides or other event, were erroneously tallied at either visit, were missed due to data collection procedural changes, or existed at any time in a nonsampled area.
-We used only FIA plots sampled specifically for tree mortality and growth.
+Both analyses excluded trees that shrank in diameter, physically moved due to landslides or other events, were erroneously tallied at either visit, were missed due to data collection procedural changes, or existed at either visit in a nonsampled area.
 
@@
 
-We obtained plot-level CWD data from the model TerraClimate (Abatzoglou et al.
+We obtained plot-level CWD data from the TerraClimate dataset (Abatzoglou et al.
 2018) for June, July, and August (summer months) between 1980 and 2023.
-Latitude and longitude were based on FIA-fuzzed coordinates.
-We defined the 'initial CWD' values as the 20-year mean of summer CWD values for each plot prior to and including the first visit year (initial CWD = mean(summer CWD from visit year$_{-19}$ to year$_0$).
-We then found the mean of summer CWD values between visits and including the second visit year (second visit CWD = mean(summer CWD from year$_1$ to year$_10$) .
-We then defined delta CWD = second visit CWD - initial CWD.
-Figure 1 presents maps of both the initial CWD and delta CWD values for forested and nonforested FIA plots.
+B.
+Frank determined that the Pearson's correlation between the fuzzed and unfuzzed CWD plot values was > 0.99, so for simplicity we used the fuzzed CWD values and spatial coordinates.
+We defined 'initial CWD' values as the 20-year mean of summer CWD values for each plot prior to and including the first visit year (initial CWD = mean(summer CWD from visit year$_{-19}$ to year$_0$)).
+We then found the mean of summer CWD values between visits and including the second visit year (second visit CWD = mean(summer CWD from year$_1$ to year$_10$)).
+We then defined 'delta CWD' as the second visit CWD values minus initial CWD values.
+Figure `r fig_nums("fig_cwdmap", display = "num")` presents maps of both the initial CWD and delta CWD values for forested and nonforested FIA plots.
+
+## Species  Classification
+
+We coarsely categorized tree species by several taxonomic and ecological characteristics, such as tree type (angiosperm [hardwood] vs.
+conifer), drought tolerance, fire adaption, elevation, shade tolerance, and mesic and riparian specialization.
+We obtained the category information from https://plants.usda.gov/ (accessed 12/12/2025) and Burns and Honkala (1990a, 1990b).
+Categorization was done to aid the discussion of results; some species could fall under more than one category (e.g., we classified *Picea engelmannii* as a high elevation conifer [HEC] while it could have been classified as a shade-tolerant conifer [STC]).
+Our categories with abbreviations also include fire-adapted conifers (FAC), drought-tolerant conifers (DTC), drought-tolerant hardwoods (DTH) and riparian and mesic hardwoods (RMH).
+These abbreviations are used in Tables `r tab_nums("tbl_growth", display = "num")` and `r tab_nums("tbl_mort", display = "num")`.
 
 ## Analysis
+We limited the number of species examined to those with more than 400 occupied plots across the states of Oregon, Washington, and California.
 We obtained point estimates of species growth and mortality using ratio estimators (Eqn 4.16 in Scott et al.
 2005).
-We expanded the number of sampled trees in each plot by the area sampled (subplot or macroplot area) adjusted by the proportion of the plot area available for sampling and the number of years between visits.
+We expanded the number of sampled trees in each plot by the area measured (subplot or macroplot area) adjusted by the proportion of the plot area available for measurement and the number of years between visits.
 The numerator for species-specific mortality is the sum of stratum-weighted means of the number of trees that died between visits per plot.
-The denominator for mortality is similar but is calculated as the sum of stratum-weighted means of all trees per plot of a given species.
+The denominator for mortality is the sum of stratum-weighted means of all live trees per plot of a given species at the initial visit.
+Thus, the estimator describes the proportion of trees that transitioned from live to dead between remeasurements.
 
 The numerator of species-specific growth estimates is the sum of stratum-weighted means of the increase in basal area (cm$^2$) between visits per plot while the denominator is the same as was used for the mortality estimate.
-This produces a point estimate for a species’ mean growth.
+This produces a point estimate for a species’ mean growth on an individual tree basis.
 We standardized mortality and growth values to a decadal time period.
 
-We estimated, for each species examined, six domains based on CWD values.
-The domains allow us to parse the plot CWD values into three levels for initial CWD values and two levels for delta CWD values.
-We created the domains based on initial CWD plot values by dividing the values into three species-specific quantiles: values below the first quartile (Low), between the first and third quartiles (Medium), and above the third quartile (High; Figure 2C).
-We defined the two domains for delta CWD plot values for all species as those above a 5mm CWD change threshold (Above) and those below the threshold (Below).
+We report species mortality and growth estimates for all plots across and within Oregon, Washington, and California.
+Separately, we investigate climatic effects on tree mortality and growth by partitioning trees into subpopulations (domains).
+For individual species we estimate domain means and measures of uncertainty for both mortality and growth.
+In this study we use six domains based on CWD values.
+We created the domains based on initial CWD plot values by partitioning the values into three species-specific quantiles: values below the first quartile (Low), between the first and third quartiles (Medium), and above the third quartile (High; Figure `r fig_nums("fig_domains", display = "num")`C).
+We defined the two domains for delta CWD plot values for all species as those above a 5mm CWD change threshold, which we interpret as increasing in CWD (Increasing), and those below the threshold (Stable).
 The 5mm threshold was selected because species' delta CWD value distributions did not differ greatly and because 5mm is approximately the mean delta CWD value across species (Table S2).
-The Below domains may includes negative delta CWD values (i.e., mean CWD was larger prior to the initial visit than between visits).
-The domain designations AL, AM, and AH correspond to Above Low, Above Medium, and Above High.
-The Below domains are similarly labeled (BL, BM, BH).
-Parsing the plot data into six domains allow for bivariate examinations of each species' data.
-This bivariate approach allows us to ask, for example, whether trees in historically dry sites (High initial CWD) respond differently to increased water stress (Above threshold delta CWD) compared to trees in mesic sites (Low initial CWD) experiencing similar increases.
+The Stable domains may include some negative delta CWD values (i.e., mean CWD was larger prior to the initial visit than between visits), however these instances were rare and of small magnitude.
+The domain designations IL, IM, and IH correspond to Increasing Low, Increasing Medium, and Increasing High.
+The Stable domains are similarly labeled (SL, SM, SH).
+These six within-domain estimates allowed us to establish mortality and growth rates for trees that experienced similar climatic conditions for a given species.
 We calculated estimates for all CWD-based domains with 10 or more plots.
 
-We additionally report species mortality and growth estimates for all plots across and within Oregon, Washington, and California irrespective of CWD values.
-We limited the number of species examined to those with more than 400 occupied plots across the three states.
-
-For each species we examined contrasts among growth (Figure `r fig_nums("fig_diffgrow", display = "num")`) and mortality (Figure `r fig_nums("fig_diffmort", display = "num")`) CWD domain estimates.
-The figure column "Above vs.
-Below Threshold" displays the mean point estimate and 95% CI for contrast between corresponding Above and Below domains (e.g., for *Abies lasiocarpa*, subtracting the estimates for columns BL, BM, and BH from AL, AM, AH, respectively; Figure `r fig_nums("fig_domains", display = "num")`A).
-Negative estimates for the "Above vs.
-Below Threshold" in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")`  indicate that the 10-year growth or mortality rate was greater for a species in plots below the change estimates.
-`r t.ref("tbl_summ_grow_mort")` summarizes the number of domain differences in each of the columns in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")` where the 95% CI does not include zero (termed "significant" findings) as well as the number of contrasts per figure column.
-
-We obtained empirical means and 95% confidence intervals via bootstrap.
-For each of the 1000 bootstrap iterations we sampled FIA plots with replacement and obtained point estimates for each species' domain.
-For each domain we calculated a mean, upper, and lower 95% CI with the 2.5% and 97.5% quantiles from the  sampling iterations.
-This approach allows for nonsymmetrical CIs around means.
+In addition to within-domain estimates, we examined contrasts among growth (Figure `r fig_nums("fig_diffgrow", display = "num")`) and mortality (Figure `r fig_nums("fig_diffmort", display = "num")`) across CWD domains.
+This allows us to establish if trees between two CWD domains are experiencing meaningful differences in growth and mortality.
+For example, this allows us to ask whether trees in historically dry sites (High initial CWD) respond differently to increased water stress (Increasing, or above threshold delta CWD) compared to trees in mesic sites (Low initial CWD) experiencing similar increases, among other comparisons.
+
+We obtained empirical means and 95% confidence intervals via a nonparametric bootstrap.
+For each of the 1000 bootstrap iterations we sampled FIA plots with replacement and obtained point estimates for each domain, yielding a maximum of `r n.spp * 6 ` in-domain point estimates for each iteration.
+For each domain we calculated a mean, upper, and lower 95% CI with the 2.5% and 97.5% quantiles from the  bootstrap iterations.
+This approach allows for asymmetrical CIs around means.
 We estimated contrasts between pairs of CWD domains by subtracting within-iteration domain point estimates.
 
-Our analysis approach results in hundreds of comparisons.
-We conducted a Monte Carlo (MC) permutation analysis to provide species-level false positive rates (p-values) for "significant" contrasts.
-For each of 1000 MC permutation analysis iterations the code estimated each all species' bootstrap confidence intervals as described above.
-The MC permutation analysis sampled each species' plot domain assignments without replacement.
-Given this sampling approach, and since species distributions were not independent, the p-values are strictly interpretable at the species level (e.g., we cannot speak to the likelihood of observing a specific number of significant findings for a contrast type).
-The distance of MC permutation contrast CI bounds from zero were compared against test CI bounds.
-The MC permutation procedure recorded a null significant value if the absolute values of the permutation iteration CI bounds were at least as far from zero as observed bounds.
-
-Given the large number of contrasts conducted (`r mort.tot` each for growth and mortality), we adopted a conservative approach to interpretation.
-Aside from discussing patterns of findings across species, we focus discussion on contrasts where (1) the 95% CI does not include zero, and (2) the MC permutation test yields p < 0.001, indicating that the observed pattern is unlikely under random domain assignment.
-Contrasts with less stringent p-values are shown in figures but are generally not discussed individually as they may represent Type I errors.
+Our  contrast analysis approach resulted in hundreds of comparisons.
+Given this large number of tests, we caution readers to interpret individual significant results (i.e., contrast 95% CIs that do not including zero) cautiously, as some proportion likely represent Type-I errors and occurred by chance.
+We did not apply a multiple comparisons correction as our analysis is inherently exploratory, aimed at identifying patterns within and across species rather than testing a specific *a priori* hypothesis.
+We did not want to apply corrections because we did not want to inflate the Type-II error rate and obscure real relationships in favor of reducing false positives.
 
@@
 
-Our study tracked the fate of `r comma(analysis.stats$tree.num)` trees from `r length(sel.spp)` species in `r comma(analysis.stats$plot.num)` plots that were visited twice by FIA crews.
-Of the `r length(sel.spp)` species, `r analysis.stats$n.gym.spp` were gymnosperms.
+Our study tracked the fate of `r comma(analysis.stats$tree.num)` trees from `r n.spp` species in `r comma(analysis.stats$plot.num)` plots that were visited twice by FIA crews.
+Of the `r n.spp` species, `r analysis.stats$n.gym.spp` were gymnosperms.
 The two visits were 10 years apart for `r pct.remeas[2]`% of plots (`r pct.remeas[1]`% were 9 years, `r pct.remeas[3]`% were 11 years).
@@
 
-Mean overall growth rates for species varied between `r min.grow.spp$mean` (*`r min.grow.spp$spp`*) and `r max.grow.spp$mean` (*`r max.grow.spp$spp`*, `r t.ref("tbl_growth")`) cm$^2$/decade.
-Six species had greater growth (non-overlapping CIs) in California than in Oregon and/or Washington: *Abies concolor*, *Abies grandis*, *Pinus ponderosa*, *Tsuga heterophylla*, *Arbutus menziesii*, and *Notolithocarpus densiflorus*.
-Three species had greater growth in Oregon (*Calocedrus decurrens*, *Picea engelmanii*, and *Pseudotsuga menziesii*) and two species had greater growth in Washington (*Abies amablis* and *Pinus monticola*).
-
-Decadal mean mortality rates varied between `r min.mort.spp$mean` (*`r min.mort.spp$spp`*) and `r max.mort.spp$mean` (*`r max.mort.spp$spp`*, `r t.ref("tbl_mort")`).
-Four species in California exhibited greater mortality rates than in Oregon and/or Washington: *Abies concolor*, *Pinus ponderosa*, *Acer macrophyllum*, and *Quercus kelloggii*.
-One species had a greater mortality rate in Oregon (*Tsuga mertensiana*) and Washington had five (*Abies grandis*, *Larix occidentalis*, *Picea engelmannii*, *Pinus contorta*, and *Pseudotsuga menziesii*).
-Seven species had overall mortality rates $\geq$ 0.20 (*Abies concolor*, *Abies lasiocarpa*, *Picea engelmannii*, *Pinus contorta*, *Pinus lambertiana*, *Alnus rubra*, and *Quercus chryselopis*) while at a state level seven species had mortality rates $\geq$ 0.25 (California: *Acer macrophyllum* and *Alnus rubra*; Oregon: *Abies lasiocarpa*; Washington: *Abies grandis*, *Picea engelmannii*, *Pinus contorta*, and *Pinus monticola*).
+## Climatic Water Deficit
 
 Delta CWD increased at `r analysis.stats$cwd.gt0.pct`% of plots.
-Initial and delta CWD values were positively linearly associated (correlation = `r CWD.lin.cor`); however, the two variables were better modeled by a 3rd order polynomial ($\Delta AIC$ =  `r CWD.dAIC`, Figure S1, see model fits in Tables S1 and S2), and for most species delta CWD neither linearly related to nor monotonically increased with initial CWD values (Figure S2; also, see dashboard).
-Tree species exhibited a variety of initial-visit CWD ranges, with means ranging from `r min.mean.init` (*`r min.mean.init.spp`*) to `r max.mean.init` (*`r max.mean.init.spp`*; Table S3).
-Delta CWD values tended to be comparatively small, with a maximum 95th percentile value of `r max.q95.diff` (*`r max.q95.diff.spp`*, Table S4).
+The increases occurred in all forested regions of the three states (Figure `r fig_nums("fig_cwdmap", display = "num")`C).
+Plots where CWD decreased (indicated by negative delta CWD values) are presented alone in Figure S1.
+For Washington and Oregon negative CWD values were almost entirely found in the Coast Ranges and the mesic west slope of the Cascades Range crest.
+The Blue Mountains and Northern and northeastern Washington lacked negative delta CWD plots entirely.
+In California negative delta CWD plots with values < 1mm were primarily found in the northern Sierra Nevada and in the southern portions of the state.
+Overall, initial and delta CWD values were weakly and positively linearly associated (correlation = `r CWD.lin.cor`).The two variables were better modeled with a 3rd order polynomial although the R^2^ value still indicated that the variance was not well explained by the model ($\Delta AIC$ =  `r CWD.dAIC`, R^2^ = `r round(CWD.rels$lm2$r.squared, 2)`, Figure S2, see model fits in Tables S1 and S2), implying that other variables likely affect the relationship between initial and delta CWD.
+For most species delta CWD neither linearly related to nor monotonically increased with initial CWD values (Figure S3; see dashboard).
+For instance, the relationship between the initial and delta CWD values for *Calocedrus decurrens* may be described overall with a concave curve, with some of the most negative delta CWD values occurring in the High domain (along with some of largest delta CWD values; Figures `r fig_nums("fig_domains", display = "num")`C).
+
+The distribution of initial CWD values across the three states (Figure `r fig_nums("fig_cwdmap", display = "num")`A) reflected total annual precipitation normals for 1991-2000 (check against https://www.climate.gov/media/13728).
+The most mesic areas, the Coast Ranges and the west slope of the Cascade Range, coincided with the areas with the lowest initial CWD values.
+
+Initial CWD means and ranges differed among tree species (Table S3).
+*`r init.mean.sm$spp.out`* had the smallest mean value for initial CWD (`r init.mean.sm$mean.alone` mm) while *`r init.range.sm$spp.out`* had both the largest mean value (`r init.mean.lg$mean.alone` mm) and smallest range `r init.range.sm$spp.range`.
+*`r init.range.lg$spp.out`* occupied the plots with the largest range of initial CWD values `r init.range.lg$spp.range` and was the most widely distributed of our species (dashboard).
+
+Species did not differ greatly by delta CWD (Table S4).
+The riparian species *`r delta.mean.sm$spp.out`* had the smallest mean delta CWD value of `r delta.mean.sm$mean.alone` mm `r delta.mean.sm$spp.range`.
+The species with the second smallest mean (`r delta.mean.sm2$mean.alone`mm) and smallest range of delta CWD values `r delta.mean.sm2$spp.range` was the largely coastal *`r delta.mean.sm2$spp.out`*.
+The mostly high-elevation *`r delta.mean.lg$spp.out`* had the greatest mean increase in CWD values of `r delta.mean.lg$mean.alone` mm `r delta.mean.lg$spp.range`.
+*`r delta.range.lg$spp.out`*, another high-elevation species, had the greatest range in delta CWD values `r delta.range.lg$spp.range`.
+
+## Growth
+
+Overall growth rates differed by species but in some instances coincided with species’ ecological traits.
+For instance, the drought-tolerant hardwoods (*Arbutus menziesii*, *Notholithocarpus densiflorus*, *Quercus chrysolepis*, and *Quercus kelloggii*) exhibited the lowest growth rate, with all mean decadal basal area increases < 75 cm^2^ (Table 1).
+*Quercus chrysolepis* had the least mean growth at 40.5 cm^2^/decade.
+Riparian and mesic hardwoods like *Acer macrophyllum* and *Alnus rubra* on average grew by 145 cm^2^/decade, although *Acer macrophyllum* growth varied by state; its mean growth in California was 52.3 cm^2^/decade, less than half of its growth rate in Oregon and almost four times less than its growth rate in Washington.
+High-elevation conifers like *Abies concolor* and *Picea engelmannii* would both have grown at similar rates of approximately 170-190 cm^2^/decade except that the Washington rate for *Picea engelmannii* was only 125.4 cm/decade.
+The other high-elevation conifers, *Abies lasiocarpa*,  and *Tsuga mertensiana*, grew more slowly at almost 70 cm^2^/decade.
+
+Shade tolerant conifers (*Abies amabilis*, *Abies grandis*, *Pinus lambertiana*, *Thuja plicata*, and *Tsuga heterophylla*) grew at moderate (109.9 cm^2^/decade for *Abies amablis*) to high rates (*Pinus lambertiana* had the highest overall growth rate of 253.6 cm^2^/decade).
+The fire-adapted conifers exhibited a range of mean growth rates, with *Pseudotsuga menziesii* having the greatest mean growth (229.9 cm^2^/decade), *Pinus monticola* and *Larix occidentalis* exhibiting moderate growth (124.8 and 106.4 cm^2^/decade, respectively), and *Pinus contorta* with the lowest mean growth rate (69.3 cm^2^/decade) The three state confidence intervals for *Pinus contorta* did not overlap; growth was greatest in California (83.8 cm^2^/decade) and least in Washington (57.0 cm^2^/decade).
+*Pinus monticola* exhibited a somewhat reversed pattern, with the greatest growth in Washington (172.4 cm^2^/decade) and less growth in Oregon and California (approximately 116 cm^2^/decade for both).
+The means for drought-tolerant conifers (*Calocedrus decurrens*, *Juniperus occidentalis*, *Pinus jeffreyi*, and *Pinus ponderosa*) were above 140 cm^2^/decade except for *Juniperus occidentalis* with a rate of 68.7 cm^2^/decade.
 
 Individual species growth and mortality domain patterns could differ (e.g., Figures `r fig_nums("fig_domains", display = "num")`A and `r fig_nums("fig_domains", display = "num")`B).
-Growth domain contrasts (Figure `r fig_nums("fig_diffgrow", display = "num")`, `r t.ref("tbl_summ_grow_mort")`) did not appear to result in substantially more "significant" negative or positive estimates, indicating that species overall did not exhibit more or less growth in the Above, Low, or High domains.
-
-All three columns of Figure `r fig_nums("fig_diffgrow", display = "num")` had over 20 "significant" contrasts.
+Growth domain contrasts (Figure `r fig_nums("fig_diffgrow", display = "num")`, `r t.ref("tbl_summ_grow_mort")`) did not appear to result in substantially more significant  negative or positive estimates, indicating that species overall did not exhibit more or less growth in the Increasing, Low, or High domains.
+The figure column “Increasing vs.
+Stable” displays the mean point estimate and 95% CI for contrast between corresponding Increasing and Stable domains (e.g., for *Abies lasiocarpa*, subtracting the estimates for columns SL, SM, and SH from IL, IM, IH, respectively; Figure `r fig_nums("fig_domains", display = "num")`A).
+Negative estimates for the “Increasing vs.
+Stable” in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")` indicate that the 10-year growth or mortality rate was greater for a species in plots below the change estimates.
+`r t.ref("tbl_summ_grow_mort")` summarizes the number of domain differences in each of the columns in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")` where the 95% CI does not include zero as well as the number of contrasts per figure column.
+
+We observed varying growth responses across the CWD domains (Figure `r fig_nums("fig_diffgrow", display = "num")`).
 Depending on the species, growth may have been susceptible (negatively or positively) to a change in CWD and/or differed by initial CWD values.
-*Abies amabilis*, *Abies grandis*, *Calocedrus decurrens*, *Pinus ponderosa*, *Pseudotsuga menziesii*, *Tsuga mertensiana*, *Acer macrophyllum*, and *Alnus rubra* exhibited greater growth in the Low domains than the Medium and/or High domains.
-Except for *Alnus rubra*, these results predominantly occurred in both the Below domain (2 spp.) or the Above and Below domains (5 spp.).
-*Thuja plicata* had greater growth in the Above and Below Medium domains than either High or Low domains.
-*Abies concolor*, *Larix occidentalis*, *Picea engelmannii*, *Pinus contorta*, *Pinus lambertiana*, *Arbutus menziesii*, and *Quercus kelloggii* had greater growth in the more water stressed High domains than the Medium and/or Low domains.
-Except for *Larix occidentalis*, these results predominantly occurred in both the Above domain (3 spp.) or the Above and Below domains (3 spp.).
-Other species exhibited no 'significant' findings (4 spp.) or displayed other result patterns (3 spp.).
+We consider a lower initial CWD value to indicate a more favorable condition for plant growth, and indeed *Abies amabilis*, *Abies grandis*, *Calocedrus decurrens*, *Pinus ponderosa*, *Pseudotsuga menziesii*, *Tsuga mertensiana*, *Acer macrophyllum*, and *Alnus rubra* exhibited greater growth in the Low domains than the Medium and/or High domains.
+This was exemplified by *Acer macrophyllum* and *Abies grandis* with a mean difference of `r fig.val.fcn("g", "X312", "IH - IL", 0)` cm^2^/decade, IH-IL and `r fig.val.fcn("g", "X312", "SH - SL", 0)` cm^2^/decade, SH-SL, respectively while contrasts for species like *Tsuga mertensiana* and *Quercus chrysolepis* were less (-0.20 for IM-IL and SM-SL, and `r fig.val.fcn("g", "X805", "SH - SM", 0)` for SH-SM, respectively).
+Except for *Alnus rubra*, these results predominantly occurred in both the Stable domain (2 spp.) or the Increasing and Stable domains (5 spp.).
+However, this finding is not uniform for all species, and we found that *Thuja plicata* had greater growth in the Increasing (`r fig.val.fcn("g", "X242", "IH - IM", 0)` cm^2^/decade for IH-IM and `r fig.val.fcn("g", "X242", "IM - IL", 0)` cm^2^/decade for IM-IL) and Stable (`r fig.val.fcn("g", "X242", "SH - SM", 0)` cm^2^/decade for SH-SM and `r fig.val.fcn("g", "X242", "SM - SL", 0)` cm^2^/decade for SM-SL) Medium domains than either High or Low domains.
+*Abies concolor*, *Larix occidentalis*, *Picea engelmannii*, *Pinus contorta*, *Pinus lambertiana*, *Arbutus menziesii*, and *Quercus kelloggii* had greater growth in the more water stressed High domains than the Medium and/or Low domains, with some of the greatest differences for *Pinus lambertiana* (`r fig.val.fcn("g", "X117", "IH - IL", 0)` cm^2^/decade for IH-IL) and *Picea engelmannii* (`r fig.val.fcn("g", "X93", "SH - SL", 0)` cm^2^/decade for SH-SL).
+Except for *Larix occidentalis*, these results predominantly occurred in both the Increasing domain (3 spp.) or the Increasing and Stable domains (3 spp.).
+Other species exhibited no significant findings (4 spp.) or displayed other result patterns (3 spp.).
+
+## Mortality
+
+Overall decadal mean mortality was highest among high elevation conifers.
+Three out of the four species had mortality rates of 0.20/decade or higher, including *Picea engelmannii* with the highest mean mortality rate of 0.28/decade.
+The same species exhibited a mean mortality rate that was almost five times greater in Washington (0.34) than in Oregon (0.07).
+Drought-tolerant conifers had overall mortality rates that were relatively low (*Calocedrus decurrens* had the greatest mortality of the four species at 0.15/decade), although *Calocedrus decurrens* and *Pinus ponderosa* both exhibited higher mortality and lower growth rates in California than Oregon.
+The mean mortality rates for drought-tolerant hardwoods ranged from 0.14 to 0.20, with *Quercus chrysolepis* and *Quercus kelloggii* having the highest mortality rates (0.20 and 0.18, respectively).
+For the fire-adapted conifers, *Pinus monticola* and *Pinus contorta* had the greatest mortality rates (0.19 and 0.20, respectively), with *Pinus contorta* exhibiting a low decadal mortality rate in California (0.07), more than twice that rate in Oregon (0.17), and five times the California rate in Washington (0.37).
+For the riparian and mesic hardwoods, *Alnus rubra* had an overall mortality rate of 0.21, compared 0.13 for *Acer macrophyllum*.
+However, *Acer macrophyllum* exhibited a mortality rate of 0.27 in California, almost three times as high as rates for Washington and Oregon.
+Mortality rates for shade-tolerant conifers varied greatly, from 0.04 (*Thuja plicata*) to 0.20 (*Pinus lambertiana*).
+The Washington mortality rate for *Abies grandis* was approximately twice as high as mortality rates for Oregon and California.
+
+A  suite of species had greater mortality in the Low domains than the Medium and/or High domains including *Larix occidentalis*, *Picea engelmannii*, *Pinus contorta*, *Pinus monticola*, and *Thuja plicata*.
+Some of the larger mean mortality contrasts included *Picea engelmannii* and *Pinus contorta* (`r fig.val.fcn("m", "X93", "IH - IL", 2)` and `r fig.val.fcn("m", "X108", "IH - IL", 2)`/decade, respectively, for IH-IL) while *Thuja plicata* represented a smaller estimate (`r fig.val.fcn("m", "X242", "IH - IL", 2)`/decade for IH-IL).
+These results occurred in Increasing (3 spp.) or both Increasing and Stable (2 spp.) domains.
+Mortality rates for *Abies lasiocarpa*, *Calocedrus decurrens*, *Pinus lambertiana*, *Pinus ponderosa*, *Pseudotsuga menziesii*, *Tsuga heterophylla*, *Tsuga mertensiana*, *Acer macrophyllum*, and *Notolithocarpus densiflorus* were greater in the High domains than the Medium and/or Low domains.
+Larger estimates include `r fig.val.fcn("m", "X19", "SH - SL", 2)`/decade for the *Abies lasiocarpa* SH-SL contrast, and `r fig.val.fcn("m", "X117", "IH - IM", 2)`/decade for the *Pinus lambertiana* IH-IM contrast.
+*Pseudotsuga mensiesii* exhibited smaller contrast estimates of around `r fig.val.fcn("m", "X202", "SH - SM", 2)`/decade for SH-SL and SH-SM.
+These species were approximately evenly split between differences being present in the Increasing (4 spp.) or Stable (5 spp.) domains, but none with significant findings in both the Increasing and Stable domains.
+*Arbutus menziesii* had greater mortality in Medium domains than the High domains (approximately `r fig.val.fcn("m", "X361", "IH - IM", 2)`/decade for IH-IM and SH-SM).
+Five species had no significant mortality contrasts.
 
 Mortality domain contrasts are  similar in certain regards to the growth contrasts: results for the three types of domain contrasts did not appear to be predominantly positive or negative (Figure `r fig_nums("fig_diffmort", display = "num")`, `r t.ref("tbl_summ_grow_mort")`).
-There were fewer mortality estimates with "significant" findings than growth (`r mort.tot.sig` vs `r grow.tot.sig`).
-There were fewer differences for "Above vs.
-Below Threshold" contrasts (`r mort.AvB.sig`, `r t.ref("tbl_summ_grow_mort")`) than for Above (`r mort.A.sig`) or Below (`r mort.B.sig`), indicating that mortality may not be as strongly associated with delta CWD as by the initial CWD value.
-
-*Larix occidentalis*, *Picea engelmannii*,	*Pinus contorta*,	*Pinus monticola*, and *Thuja plicata* had greater mortality in the lower water deficit Low domains than Medium and/or High domains.
-These results occurred in Above (3 spp.) or both Above and Below (2 spp.) domains.
-Mortality rates for *Abies lasiocarpa*,	*Calocedrus decurrens*,	*Pinus lambertiana*, *Pinus ponderosa*,	*Pseudotsuga menziesii*,	*Tsuga heterophylla*,	*Tsuga mertensiana*,	*Acer macrophyllum*,	and *Notolithocarpus densiflorus* were greater in the High domains than the Medium and/or Low domains.
-These species were approximately evenly split between differences being present in the Above (4 spp.) or Below (5 spp.) domains, but none with "significant" findings in both the Above and Below domains.
-*Arbutus menziesii* had greater mortality in Medium domains than the High domains.
-Five species had no significant contrasts.
-
+There were fewer mortality estimates with significant findings than growth (46 vs 74).
+There were fewer differences for “Increasing vs.
+Stable” contrasts (9) than for Increasing (21) or Stable (16), indicating that mortality may not be as strongly associated with delta CWD as by the initial CWD value.
+
+## Geography  
 The interactive dashboard revealed geographic patterns that were associated with specific domains while not necessarily captured by state-level estimates.
-For the Above and Below Low domains *Abies lasiocarpa* had less mortality in northeastern Washington and *Tsuga mertensiana* exhibited greater growth in the same region.
-*Pinus contorta* exhibited less growth and greater mortality in northern Washington (Above Low).
-In the southern Sierra Nevada *Pinus lambertiana* had greater growth (Above Medium and High) and greater mortality (Above High), *Calocedrus decurrens* had greater mortality (Above High), and *Pinus monticola* had less mortality (Above High).
-*Acer macrophyllum* exhibited less growth in the Sierra Nevada and southern Klamath Mountains (Above High) and greater growth in northwestern Washington in the Coast and Cascade ranges (Above and Below Low).
-
-The MC permutation analysis provided empirical p-values for observing contrast confidence intervals with bounds at least as far from zero as those observed (Supplemental figures S5 and S6), with values ranging from a maximum of $p$ = 0.072 to a minimum $p$ < 0.0001 (41 instances).
-Except where explicitly stated, we only discuss individual species' domain findings if the species has one or more relevant MC permutation p-values < 0.001 (indicated by three or four asterisks in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")`).
+For the Increasing and Stable Low domains *Abies lasiocarpa* had less mortality in northeastern Washington and *Tsuga mertensiana* exhibited greater growth in the same region.
+*Pinus contorta* exhibited less growth and greater mortality in northern Washington (Increasing Low).
+In the southern Sierra Nevada *Pinus lambertiana* had greater growth (Increasing Medium and High) and greater mortality (Increasing High), *Calocedrus decurrens* had greater mortality (Increasing High), and *Pinus monticola* had less mortality (Increasing High).
+*Acer macrophyllum* exhibited less growth in the Sierra Nevada and southern Klamath Mountains (Increasing High) and greater growth in northwestern Washington in the Coast and Cascade ranges (Increasing and Stable Low).
 
@@
 Responses were potentially driven by differences in drought tolerance strategies, local disturbance regimes, and biotic interactions.
-We found more “significant” CWD contrasts for species growth than mortality, possibly indicating that growth differences by CWD domain were more readily detectable than mortality over a decade, although growth and mortality findings were not necessarily negatively related.
+We found more significant CWD contrasts for species growth than mortality, possibly indicating that growth differences by CWD domain were more readily detectable than mortality over a decade, although growth and mortality findings were not necessarily negatively related.
 Delta CWD increased almost everywhere between plot revisits although the magnitude of change was generally not large relative to initial CWD values.
-We additionally found concerningly high state and/or overall decadal mortality rates for some species.
-
-Contrast results may be more informative at an individual species level and when considering state-level mortality and growth estimates.
+We additionally found concerningly high decadal mortality rates for some species within and across states.
+
+Contrast results serve to highlight differences in mortality and growth rates for different subpopulations of trees undergoing different climatic experiences.
 Regional analyses detected several species with relatively high mortality rates including *Acer macrophyllum*, *Picea engelmannii*, *Abies lasiocarpa*, *Pinus contorta*, and *Alnus rubra*.
@@
 Our estimates for this short-lived tree species align with reported background mortality rates (e.g., Brown and Courtin 2018).
-However, the mortality rate for *Acer macrophyllum* in California appears to be both unreported elsewhere and substantially higher in California than in Oregon or Washington.
-
+
+The high mortality rate for *Acer macrophyllum* in California appears to be both unreported elsewhere and substantially greater than in Oregon or Washington.
 *Acer macrophyllum* growth was negatively associated with initial CWD values, and its growth was lower and mortality almost three times higher in California than in Washington and Oregon.
@@
 We found that the pressures on the two species appear to differ by CWD values and location.
-We found that *Picea engelmannii* had lower mortality in the Above High domain, corresponding to stands in the Blue Mountains of Oregon and Washington, and higher mortality in the Above Low and Above High domains, largely coinciding with northern and northeastern Washington, where spruce beetle infestations were elevated (WADNR 2023).
+We found that *Picea engelmannii* had lower mortality in the Increasing High domain, corresponding to stands in the Blue Mountains of Oregon and Washington, and higher mortality in the Increasing Low and Increasing High domains, largely coinciding with northern and northeastern Washington, where spruce beetle infestations were elevated (Washington Department of Natural Resources [WADNR] 2023).
 Its mortality rate in Washington was almost 5 times higher than in Oregon.
@@
 Pine bark beetles (*Dendroctonus ponderosae*, *Dendroctonus brevicomis*, and *Ips* spp.) have affected *Pinus contorta* for the past decade in this region (WADNR 2023).
-This area also corresponds, at the level of ecoregion subsections, with the portions of *Pinus contorta*’s range in the lowest forest stability index category for Stanke et al.
+This area also corresponds, at the level of ecoregion subsections, with the portions of *Pinus contorta*’s range in the lowest forest stability index category of Stanke et al.
 (2021) while California’s Sierra Nevada and the Oregon and Washington Cascade Range correspond to positive stability index values.
@@
 The absence of consistent contrast patterns across species is itself informative, indicating that tree responses to water stress conditions are potentially mediated by local adaptation to conditions, species-specific physiological traits, and local environmental conditions rather than following a general response curve.
-The number of "significant" Above vs.
-Below threshold contrasts for growth and mortality were approximately evenly distributed between positive and negative contrast estimates, with a positive difference indicating that some species exhibited higher growth or greater mortality in an Above domain (e.g., *Pinus lambertiana* for both growth and mortality) versus a Below domain (*Abies grandis* and *Pinus contorta*, respectively, for growth and mortality).
-This lack of a pattern across species may indicate that the distinction of Above vs.
-Below is not generalizable across species, or that "significant" findings are due to confounding factors such as geographical locations with different ecological pressures.
+The number of significant Increasing vs.
+Stable threshold contrasts for growth and mortality were approximately evenly distributed between positive and negative contrast estimates, with a positive difference indicating that some species exhibited higher growth or greater mortality in an Increasing domain (e.g., *Pinus lambertiana* for both growth and mortality) versus a Stable domain (*Abies grandis* and *Pinus contorta*, respectively, for growth and mortality).
+This lack of a pattern across species may indicate that the distinction of Increasing vs.
+Stable is not generalizable across species, or that significant findings are due to confounding factors such as geographical locations with different ecological pressures.
 
 *Abies grandis* exhibited higher mortality in Washington than Oregon or California.
-We found no evidence of a difference in domain mortality geographically, although the greatest growth domain was Below Low, which corresponded to the more mesic plots west of the Cascade Crest of Washington and Oregon.
+We found no evidence of a difference in domain mortality geographically, although the greatest growth domain was Stable Low, which corresponded to the more mesic plots west of the Cascade Crest of Washington and Oregon.
 The mortality result may be due to an intersection in a fire suppression history promoting a densification of shade-tolerant but drought-intolerant firs in the arid climates of eastern Oregon and Washington, drought, and fir engraver beetle mortality (*Scolytus ventralis*; Washington DNR 2023, Merschel et al.
@@
 State estimates for *Pinus monticola* mortality do not appear to differ.
-Mortality for the Above High domain was low (see dashboard; MC permutation test results only found p < 0.01 when comparing Above High to Above Low) and predominantly represented by plots in the southern Sierra Nevada mountains of California.
+Mortality for the Increasing High domain was low (see dashboard; MC permutation test results only found p < 0.01 when comparing Increasing High to Increasing Low) and predominantly represented by plots in the southern Sierra Nevada mountains of California.
 *Pinus lambertiana* and *Calocedrus decurrens* exhibited high mortality in approximately the same region, a result supported by Fettig et al.
 (2019).
-This region experienced an extreme drought that killed 170m trees between 2010 and 2021 (OEHHA 2022).
+This region experienced an extreme drought that killed, as estimated from aerial detection surveys, 170 million trees between 2010 and 2021 (Office of Environmental Health Hazard Assessment 2022).
 Dudney et al.
@@
 2025).
-Our analysis found that *Thuja plicata* exhibited the lowest decadal mortality rate and third highest growth rate of any species we examined.
+Our analysis found that *Thuja plicata* exhibited the lowest decadal mortality rate and third highest growth rate of any species we examined, and relatively strong sampling precision.
 It also had its greatest growth in Medium domains.
@@
 If species conditions change quickly (on the order of 3-5 years) the FIA plot data may not reflect the change until a full or nearly full panel of plots is complete following the change in conditions.
-Another non-exclusive possibility is that our study serves as one example of many broad-scale design-based analyses of spatially-balanced samples that offer a relatively unbiased assessment of species condition that can assist managers in identifying species of concern and prioritizing restoration actions.
-*Thuja plicata* may indeed be senescing in many regions, but other species may be faring worse.
+Another non-exclusive possibility is that our results, based on an extensive, unbiased network of plots, offer a robust and unbiased mortality estimate for this species and others.
+*Thuja plicata* may indeed be senescing in specific regions, but other species may be overall faring worse.
 
 We believe the value of our analysis lies in its application to a relatively large number of tree species and its approach for investigating life history conditions such as growth and mortality that require the documentation of tree condition change.
-The availability of full-panel FIA revisit data is a relatively recent development.
+The availability of full-panel FIA revisit data in the states we investigated is a relatively recent development.
 Prior to the availability of the revisit data, Monleon and Lintz (2015) assessed tree species’ range shifts using the single-visit data set, comparing mean plot temperature values for sapling and mature trees.
@@
 (2023).
-Stanke et al.
-(2021) found estimates of condition indices.
-Our bivariate examination of plot CWD shares some dimensional similarities with Dobrowski et al.
+Our examination approach for plot initial and delta CWD values shares some dimensional similarities with Dobrowski et al.
 (2013)  who examined the velocity of CWD (change in distance over time) along with AET and minimum temperatures.
-However, we are not aware of other design-based estimates of revisit data that examine more than one variable of interest simultaneously.
 Our approach allows us to capitalize on the amenability of the data to design-based estimation while accounting for the need, when examining revisit conditions, of capturing the initial and change value conditions.
@@
 
-We did not detect sizable changes in CWD between visits, potentially signalling that 10 years is too soon to observe CWD-related species responses (perhaps more true for mortality than growth).
+We did not detect sizable changes in CWD between visits, potentially signaling that 10 years is too soon to observe CWD-related species responses (perhaps more true for mortality than growth).
+Another possibility is that the TerraClimate dataset is relatively spatially coarse (~4 km, Abatzoglou et al.
+2018), potentially dampening the range of possible results from a finer resolution model.
 The CWD changes we observed may have represented greater proportional change for some species than others, with species such as *Abies amablis*, *Alnus rubra*, and *Juniperus occidentalis* exhibiting narrow and low water stress initial CWD ranges.
@@
 
-Although our current analysis offers a coarse examination of species status across a large area, future analyses could refine our bivariate approach for assessing tree species growth and mortality by domains.
+Although our current analysis offers a coarse examination of species status across a large area, future analyses could refine our approach for assessing tree species growth and mortality by domains.
 Our climate domains could be divided further into sub-domains: trees could be assigned to tree size, stand competition, plot aspect, or plot elevation categories within the climate domain categories.
-This approach has its limits: only estimates for certain widespread tree species would be reasonable to report, and domains could not be divided too finely into sub-domains, limiting the number of variables that could be considered simultaneously.
+This approach has its limits: only estimates for certain widespread tree species would be reasonable to report, and domains could not be divided too finely into sub-domains due to sample size constraints, limiting the number of domain variables that could be considered simultaneously.
 However, this approach would permit design-based estimation at a resolution better attuned to capturing factors affecting growth and mortality.
 
-Our analysis may overall capture significant changes in mortality and growth for the tree species examined, with some trees exhibiting increases in mortality and decreases in growth either in portions of their ranges or across their entire range.
+Our analysis captures significant changes in mortality and growth for the tree species we examined, with some trees exhibiting increases in mortality and decreases in growth either in portions of their ranges or across their entire range.
 For some species in semi-arid environments such as *Pinus ponderosa*, *Pseudotsuga menziesii*, *Pinus contorta*, *Picea engelmannii*, and *Abies lasiocarpa*, we may be observing ecosystem shifts with a change in water stress combined with disturbances such as wildfire and insect mortality (Halofsky et al.
@@
 2023).
-Our hope is that our bivariate domain-based estimation approach is useful for others, while our results for individual species serve to bolster or direct others' inquiries into the status of the species we examined.
+Our hope is that our results for individual species serve to bolster or direct others' inquiries into the status of the species we examined.
 For managers, the results may be used to inform action for resiliency.
@@
 
-Our analysis serves as an example of bivariate design-based estimation, where we use estimation domains based on initial CWD and delta CWD values of plots.
+Our analysis serves as an example of design-based estimation, where we use estimation domains based on initial CWD and delta CWD values of plots.
 For the 23 species we examined, we found no general patterns in domains with higher or lower tree growth or mortality; the results appeared species specific.
@@
 
+Burns, R.M., and B.H. Honkala, tech coords. 1990a. Silvics of North America. Volume 1. Conifers. Agricultural Handbook 654. U.S. Department of Agriculture, Forest Service, Washington DC. 675p.
+
+Burns, R.M., and B.H. Honkala, tech coords. 1990b. Silvics of North America. Volume 2. Hardwoods. Agricultural Handbook 654. U.S. Department of Agriculture, Forest Service, Washington DC. 877p.
+
 Daly C, Halbleib M, Smith JI, Gibson WP, Doggett MK, Taylor GH, et al.
@@
 
-Office of Environmental Health Hazard Assessment (OEHHA, 2022). Indicators of climate change in California, fourth edition, California Environmental Protection Agency, OEHHA.
+Office of Environmental Health Hazard Assessment. 2022. Indicators of climate change in California, fourth edition, California Environmental Protection Agency.
 
@@
 
+```{r cwdmeansfig, fig.width = 7, fig.height=6, fig.dpi=300, fig.cap=fig.cwd_initdelt, fig.cap.sep=""}
+
+# Output from Manuscript_information.R
+knitr::include_graphics(paste0(RESULTS.OTHER, "Spp_InitDeltCWD.png"))
+
+```
+
 ```{r diff_grow_fig, fig.height = 6.5, width = 6.5, fig.dpi=300, fig.cap=fig.diff_grow, fig.cap.sep=""}

### 2025-12-12 Supplemental Material.Rmd
Incorporated edits from B. Frank.  This added a figure, changed the supplimental and main RMarkdown files, and altered many of the scripts involved in generating plots and tables.

@@
 analysis.stats <- read_rds(paste0(RESULTS.OTHER, "analysis.stats.RDS"))
+cwd.init <- analysis.stats$init.cwd.stats  
+cwd.diff <- analysis.stats$diff.cwd.stats
+
 # Correlation/fits for Initial CWD vs Delta CWD:
@@
 
-cwd.init <- analysis.stats$init.cwd.stats 
-cwd.diff <- analysis.stats$diff.cwd.stats
+cwd.init2 <- cwd.init %>% dplyr::select(-UCI, -LCI)
+cwd.diff2 <- cwd.diff %>% dplyr::select(-UCI, -LCI)
 
@@
 
+fig.negcwdmap_cap <- fig_nums(name = "fig_cwdnegmap", 
+                        caption = "Fuzzed FIA plot locations for plots with negative CWD values.
+The initial CWD values for those plots are displayed in A while the negative delta CWD values are displayed in B.
+The inset map (B) shows the locations of the study states within the USA: From north to south, Washington, Oregon, and California.")
+
 fig_cwd_all <- fig_nums(name = "fig_cwdall", 
@@
 
-<br>
+```{r cwdfig-negcwdmap, fig.height = 7,  fig.dpi=300, fig.cap=fig.negcwdmap_cap, fig.cap.sep=""}
 
-<br>
+# Output from Manuscript_information.R
+knitr::include_graphics(paste0(RESULTS.OTHER, "cwdneg_maps_figure.png"))
+
+```
 

### 2025-12-12 Analysis Code/Data_Prep_Analysis.R

@@ -38,6 +38,8 @@ library(tictoc)  # For development, timing routines.
 library(data.table)
 library(matrixStats) # For fast iteration processing
 library(abind) # for combining matrices into arrays
+library(ggrepel) # For plotting, avoiding overlapping labels
+
 #### 2) Base data prep --------------------------------------------------------
 # These are the base data sets that will be transformed by the analysis types.
 
@@ -291,6 +293,16 @@ for(k in 1:length(ANALYSIS.TYPE)) {  # 1 = grow, 2 = mortality
     saveRDS(list(bootstrap_results = bootstrap_results, domain.summaries = domain.summaries, all_dat = all_dat), 
             file = paste0(save.loc.fcn(k), "Domain_Analysis_Output.RDS"))
     
+ 
+    ## TO ADAPT THE INPUT WITHOUT RERUNNING THE BOOTSTRAP/PERMUTATION TEST SKIP THE    
+    #   BOOTSTRAP AND saveRDS() CALL AND RUN THE FOLLOWING. THIS WAS USED TO RENAME THE
+    #   CHANGE DOMAINS FROM ABOVE/BELOW TO INCREASING/STABLE.
+#    dmd <- readRDS(paste0(save.loc.fcn(k), "Domain_Analysis_Output.RDS"))
+#    bootstrap_results = dmd$bootstrap_results
+#    domain.summaries = dmd$domain.summaries
+#   all_dat = dmd$all_dat
+    
+    
     ## The code in this section is dedicated to preparing data files for plotting the differences in six-domain results.
     ## This process has two steps: first, the domain matrix pairs are subtracted from one another 
     #  using the domain.diff.fcn function.  Then, using the differenced matrices, the matrices
@@ -298,25 +310,25 @@ for(k in 1:length(ANALYSIS.TYPE)) {  # 1 = grow, 2 = mortality
     #  prepped for inclusion.  
     if(n_domain == 6) {
       
-    # Domains compared: Above vs. Below threshold 
-    A.vec <- c("AL", "AM", "AH")
-    B.vec <- c("BL", "BM", "BH")
[…]

### 2025-12-12 Analysis Code/Functions.R

@@ -58,7 +58,7 @@ clim.mort.resp.fcn <- function(spcd, clim.var, treedat.sel, clim.dat) {
       mutate(., var.quants = ifelse(!!sym(var1) > quant.lims[2], 1, ifelse(!!sym(var1) < quant.lims[1], -1, 0)),
              deltvar.quants = ifelse(!!sym(var.delt) > quant.lims.delta, 1, 0),
              vq = case_match(var.quants, -1 ~ "L", 0 ~ "M", 1 ~ "H"),
-             dvq = case_match(deltvar.quants, 0 ~ "B", 1 ~ "A"),
+             dvq = case_match(deltvar.quants, 0 ~ "S", 1 ~ "I"),
              var.deltvar = factor(paste0(dvq, vq), levels = DOMAIN.LEVELS))
     }}
   
@@ -153,7 +153,7 @@ clim.growth.resp.fcn <- function(spcd, clim.var, treedat.sel, clim.dat) {
         mutate(., var.quants = ifelse(!!sym(var1) > quant.lims[2], 1, ifelse(!!sym(var1) < quant.lims[1], -1, 0)),
                deltvar.quants = ifelse(!!sym(var.delt) > quant.lims.delta, 1, 0),
                vq = case_match(var.quants, -1 ~ "L", 0 ~ "M", 1 ~ "H"),
-               dvq = case_match(deltvar.quants, 0 ~ "B", 1 ~ "A"),
+               dvq = case_match(deltvar.quants, 0 ~ "S", 1 ~ "I"),
                var.deltvar = factor(paste0(dvq, vq), levels = DOMAIN.LEVELS))
       } else {
         values$errorMessage <- "Error: Expecting 6 or 9 estimation domains. Check Global.R:DOMAIN.LEVELS and Functions.R:clim.mort.resp.fcn."
@@ -815,7 +815,7 @@ cm2.fcn <- function(k, results.table){
 
 
 ## This plot provides the 
-diff.panel.fcn <- function(diff.dat, remove.y, fig.title, lab.right, ft_to_use = font_to_use) {
+diff.panel.fcn <- function(diff.dat, remove.y, fig.title, lab.right, ft_to_use = font_to_use, plt.mcperm) {
   
   # Position for astrices 
   diff.dat <- diff.dat %>% mutate(
@@ -842,9 +842,9 @@ diff.panel.fcn <- function(diff.dat, remove.y, fig.title, lab.right, ft_to_use =
                    fill = ifelse(significant, "black", "white"),
[…]

### 2025-12-12 Analysis Code/Manuscript_information.R

@@ -54,7 +54,11 @@ diff.cwd.90range.max <- diff.range[which(diff.range == max(diff.range))]
 diff.cwd.95max <- which(diff.cwd.stats$q95 == max(diff.cwd.stats$q95))
 
 # For supplemental, MC permutation p-values
-psig.dat <- read_csv(paste0(RESULTS.OTHER, "Permutation_results.csv"), show_col_types = FALSE)
+psig.dat <- read_csv(paste0(RESULTS.OTHER, "Permutation_results.csv"), show_col_types = FALSE) %>%
+  ## Changing domain designations from A/B to I/S ##
+  mutate(across(everything(), ~str_replace_all(., "A", "I")),
+         mutate(across(everything(), ~str_replace_all(., "B", "S"))))
+
 psig.dat2 <- psig.dat %>% dplyr::select(-order.c) %>%
   left_join(gym.angio.dat %>% dplyr::select(SpeciesCode, sci_name, SFTWD_HRDWD), by = c("Species" = "SpeciesCode")) %>%
   arrange(desc(SFTWD_HRDWD), sci_name) %>%
@@ -89,6 +93,37 @@ analysis.stats <- list(tree.num = tree.num,
 ) 
 
 write_rds(analysis.stats, paste0(RESULTS.OTHER, "analysis.stats.RDS"))
+
+
+##### Figure: Mean initial/delta CWD by species (scatterplo) -----------------------
+cwd.init.ci <- init.cwd.stats %>% dplyr::select(Species, Mean, UCI, LCI) %>% rename("Mean_Init" = Mean, "UCI_Init" = UCI, "LCI_Init" = LCI)
+cwd.diff.ci <- diff.cwd.stats %>% dplyr::select(Species, Mean, UCI, LCI) %>% rename("Mean_Diff" = Mean, "UCI_Diff" = UCI, "LCI_Diff" = LCI)
+cwd.ci <- left_join(cwd.init.ci, cwd.diff.ci, by = "Species")
+
+# CI bars don't add much. Sticking with points
+cwd.ci.plt <- ggplot(cwd.ci, aes(Mean_Init, Mean_Diff, group = Species, label = Species)) + 
+  #geom_errorbar(aes(ymin = LCI_Diff, ymax = UCI_Diff)) +  
+  #geom_errorbar(aes(xmin = LCI_Init, xmax = UCI_Init)) +
+  geom_point() + 
+  geom_text_repel(
[…]

### 2025-12-12 Analysis Code/Nplot_Climvar_anaysis.R

@@ -97,6 +97,41 @@ ggplot(map.dat.1, aes(ELEV, difference)) +
   theme_bw()
 
 
+# Examining elevation and delta CWD, highlighting negative plots
+map.dat.2 <- map.dat.1 %>% 
+  mutate(neg.delt.cwd = ifelse(difference < 0, 1, 0)) %>%
+#  separate_wider_regex(var = puid, c(var1 = ".*", "_", var2 = ".*", "_", var3 = ".*"))
+  separate_wider_delim(cols = puid, delim = "_", names = c("var1", "state", "var3"), too_many = "merge") %>%
+  dplyr::select(-var1, -var3)
+
+  # All plots, negative CWD = black points.
+ggplot(map.dat.2, aes(pre_mean, ELEV, color = state)) + geom_point(alpha = 0.15) +
+  geom_point(data = map.dat.2 %>% filter(neg.delt.cwd == 1, pre_mean < 250), aes(pre_mean, ELEV)) +
+  theme_bw() + 
+  labs(x = "Initial CWD", y = "Elevation (m)") + 
+  theme(text = element_text(family = font_to_use))
+
+  # Only negative CWD plots, Initial CWD x Elevation, color = negative delta CWD value
+ggplot(data = map.dat.2 %>% filter(neg.delt.cwd == 1, pre_mean < 250), aes(pre_mean, ELEV, color = difference)) + geom_point(alpha = 0.5) +
+  theme_bw() +
+  scale_color_viridis(discrete = FALSE, option = "H", name = "Delta\nCWD", begin = 0.1, end = 0.9) +
+  labs(x = "Initial CWD", y = "Elevation (m)") + 
+  theme(text = element_text(family = font_to_use))
+
+ # Only negative CWD plots, Delta CWD x Elevation.  Seeing if any relationship exists here.  
+ggplot(data = map.dat.2 %>% filter(neg.delt.cwd == 1), aes(ELEV, difference, color = state)) + geom_point(alpha = 0.5) + geom_smooth() +
+  theme_bw() + 
+  labs(y = "Delta CWD", x = "Elevation (m)") + 
+  theme(text = element_text(family = font_to_use))
+
+  # Only negative CWD plots, Initial CWD vs. Delta CWD, no elevation.  Just checking.
+ggplot(data = map.dat.2 %>% filter(neg.delt.cwd == 1, pre_mean < 250), aes(pre_mean, difference, color = state)) + geom_point(alpha = 0.5) + geom_smooth() +
[…]

### 2025-12-12 Global.R

@@ -77,7 +77,7 @@ INIT.CLIM.BREAKS <- c(75, 125) # if USE.QUANT.PROBS are FALSE, these are the dom
 if(THREE.CHANGE.CATEGORIES) {
 DOMAIN.LEVELS <- c("LiHc", "MiHc", "HiHc", "LiMc", "MiMc", "HiMc", "LiLc", "MiLc", "HiLc")
 } else {
-DOMAIN.LEVELS <- c("AL", "AM", "AH", "BL", "BM", "BH") # Above/Below threshold and Low/Medium/High
+DOMAIN.LEVELS <- c("IL", "IM", "IH", "SL", "SM", "SH") # Increasing/Stable threshold and Low/Medium/High
 }
 N.PLOT.LIM <- 10  # Limit to the minimum number of plots for which an estimate will be calculated
 BS.N <- 1000    # Bootstrap iteration number
@@ -114,6 +114,7 @@ spp.id <- names(orig)[(ncol(orig) - (length(grep("X", names(orig))) - 1)):ncol(o
 
 spp.names <- readxl::read_xlsx(paste0(DATA.LOC, "FullSppNames.xlsx")) %>% dplyr::filter(SPCD %in% spp.list)
 
+spp.cat <- read_csv(paste0(DATA.LOC, "Species_Categories.csv"))
 
 ## Plotting
 font_to_use <- if("Times New Roman" %in% systemfonts::system_fonts()$family) {

### 2025-12-16 Manuscript.Rmd
Next revision version (first complete revision after B.Frank's edits).  The permutation analysis has been removed and a new figure of mean species initial/delta CWD values and a new map of negative initial/delta CWD values have been added.

@@
 x2 <- x[place2, ]
-x2.range <- paste0("(5th percentile = ", x2$q05, "mm, 95th percentile = ", x2$q95, "mm)")
-x2.range.full <- paste0("(5th-95th percentile range = ", x2$range05.95, "mm, 5th percentile = ", x2$q05, "mm, 95th percentile = ", x2$q95, "mm)")
-x2.mean.full <- paste0("(mean = ", x2$Mean, "mm, 5th percentile = ", x2$q05, "mm, 95th percentile = ", x2$q95, "mm)")
+x2.range <- paste0("(5th percentile = ", x2$q05, " mm, 95th percentile = ", x2$q95, " mm)")
+x2.range.full <- paste0("(5th-95th percentile range = ", x2$range05.95, " mm, 5th percentile = ", x2$q05, " mm, 95th percentile = ", x2$q95, " mm)")
+x2.mean.full <- paste0("(mean = ", x2$Mean, " mm, 5th percentile = ", x2$q05, " mm, 95th percentile = ", x2$q95, " mm)")
 
@@
 # 
-fig_nums <- captioner::captioner(prefix = "")
+fig_nums <- captioner::captioner()
 tab_nums <- captioner::captioner(prefix = "Table")
@@
 Western North American forests are experiencing accelerating tree mortality rates in response to intensifying drought and water stress driven by climate change.
-Tree mortality and growth reflect cumulative effects of direct physiological stress (xylem cavitation, carbon starvation), and indirect climate-mediated impacts (wildfire, insects, pathogens).
+Tree growth and mortality reflect cumulative effects of direct physiological stress (xylem cavitation, carbon starvation), and indirect climate-mediated impacts (wildfire, insects, pathogens).
 Relating design-based estimates for mortality and growth to a climate variable can be challenging, as directly assessing mortality and growth requires the visitation of plots at two different points in time.
@@
 We applied this approach to `r comma(analysis.stats$plot.num)` National Forest Inventory plots spanning California, Oregon, and Washington, tracking `r comma(analysis.stats$tree.num)` trees from `r n.spp` species over decade-long intervals.
-We estimated decadal growth and mortality overall, by state, and across six CWD domains per species, along with associated measures of uncertainty.
+We estimated decadal growth and mortality across three states, by state, and for six CWD domains per species, along with associated measures of uncertainty.
 Species responses were highly variable, with no universal pattern of increased mortality or reduced growth in domains experiencing elevated water stress.
@@
 2022).
-There has also been a regional increase in the area of bark beetle (subfamily *Scolytinae* spp.) outbreaks (Meddens et al.
+There have also been regional increases in the area of bark beetle (subfamily *Scolytinae* spp.) outbreaks (Meddens et al.
 2012, Hicke et al.
@@
 Data from the U.S.
-National Forest Inventory, collected by the US Forest Service Forest Inventory and Analysis Program (FIA), is well suited for evaluating regional species mortality and growth.
+National Forest Inventory, collected by the US Forest Service Forest Inventory and Analysis Program (FIA), is well suited for evaluating regional species growth and mortality.
 Since 1999 the FIA program has gathered a variety of forestry data at plots nationwide using a spatially balanced sampling design (McRoberts 2005, Reams et al.
@@
 Our analysis relies on FIA data for forestland in California, Oregon, and Washington, USA.
-Forestland is defined as an area of at least 4,050 m$^2$ (i.e., one acre) that is at least 10 percent stocked or potentially stocked with tree species, excluding urban and agricultural land uses (Reams et al.
+Forestland is defined as an area of at least 4,050 m^2^ (i.e., one acre) that is at least 10 percent stocked or potentially stocked with tree species, excluding urban and agricultural land uses (Reams et al.
 2005).
@@
 
-In the coterminous United States, the FIA program uses a spatially balanced, quasi-systematic sampling design.
+In the coterminous United States the FIA program uses a spatially balanced, quasi-systematic sampling design.
 Plots are randomly located and produce sampling intensities of approximately 2,400 hectares per plot, with some denser areas such as those located on national forest system lands.
 In the western United States, each plot is assigned to one of ten panels, and approximately one panel is measured each year in perpetuity, yielding ten year remeasurement intervals on each plot, barring exceptions owing to optional temporal intensifications (Thompson 2015).
-In this study, we use only those plots that are part of the standard sampling intensity, and have been remeasured at an interval of ten years.
-Each FIA plot consists, in part, of four 7.32 m radius subplots where all trees with stem diameters greater than or equal to 12.7 cm (for most species measured at 1.37 m above the ground) were tallied with additional trees sampled in an annulus around each subplot (macroplot), with a total radius 18 m from the center point and a total area of 1012 m$^2$.
+In this study, we use only those plots that are part of the standard sampling intensity and have been remeasured at an interval of ten years.
+Each FIA plot consists, in part, of four 7.32 m radius subplots where all trees with stem diameters greater than or equal to 12.7 cm (for most species measured at 1.37 m above the ground) were tallied with additional trees sampled in an annulus around each subplot (macroplot), with a total radius 18 m from the center point and a total area of 1012 m^2^.
 We did not include data from microplots, where seedlings and saplings are measured, as the analysis was not intended to estimate species recruitment but rather to follow the trajectory of established trees.
@@
 Our analysis depends on examining the fate and growth of trees measured during two sequential visits (referred to as the first and second plot visit), ideally from a complete sample panel remeasure.
-Tree growth is the observed increase in marked tree bole diameter between visits (cm$^2$) while mortality is captured from the change in tree condition.
+Tree growth is the observed increase in marked tree bole diameter between visits (cm^2^) while mortality is captured from the change in tree condition.
 We obtained our FIA data from the USDA FIA DataMart web portal (https://research.fs.usda.gov/products/dataandtools/fia-datamart; downloaded May 5, 2025).
@@
 We obtained the category information from https://plants.usda.gov/ (accessed 12/12/2025) and Burns and Honkala (1990a, 1990b).
-Categorization was done to aid the discussion of results; some species could fall under more than one category (e.g., we classified *Picea engelmannii* as a high elevation conifer [HEC] while it could have been classified as a shade-tolerant conifer [STC]).
+Categorization was done to aid the presentation of results; some species could fall under more than one category (e.g., we classified *Picea engelmannii* as a high elevation conifer [HEC] while it could have been classified as a shade-tolerant conifer [STC]).
 Our categories with abbreviations also include fire-adapted conifers (FAC), drought-tolerant conifers (DTC), drought-tolerant hardwoods (DTH) and riparian and mesic hardwoods (RMH).
@@
 We expanded the number of sampled trees in each plot by the area measured (subplot or macroplot area) adjusted by the proportion of the plot area available for measurement and the number of years between visits.
-The numerator for species-specific mortality is the sum of stratum-weighted means of the number of trees that died between visits per plot.
-The denominator for mortality is the sum of stratum-weighted means of all live trees per plot of a given species at the initial visit.
-Thus, the estimator describes the proportion of trees that transitioned from live to dead between remeasurements.
-
-The numerator of species-specific growth estimates is the sum of stratum-weighted means of the increase in basal area (cm$^2$) between visits per plot while the denominator is the same as was used for the mortality estimate.
+The mortality estimator describes the proportion of trees that transitioned from live to dead between remeasurements.
+For a given species the numerator is the sum of stratum-weighted means, where each mean is the sum of trees that died between visits within each stratum divided by the number of plots within the stratum.
+The denominator for mortality is also a sum of stratum-weighted means, except each mean is the sum of all live trees within a stratum at the initial visit divided by the number of plots within the stratum.
+
+The numerator of species-specific growth estimates is a sum of stratum-weighted means, where each mean is the sum of tree basal area increase (cm^2^) between visits within the stratum divided by the number of plots in the stratum while the denominator is the same as was used for the mortality estimate.
 This produces a point estimate for a species’ mean growth on an individual tree basis.
@@
 
-We report species mortality and growth estimates for all plots across and within Oregon, Washington, and California.
+We report species mortality and growth estimates for all plots across ("overall") and within Oregon, Washington, and California.
 Separately, we investigate climatic effects on tree mortality and growth by partitioning trees into subpopulations (domains).
@@
 We created the domains based on initial CWD plot values by partitioning the values into three species-specific quantiles: values below the first quartile (Low), between the first and third quartiles (Medium), and above the third quartile (High; Figure `r fig_nums("fig_domains", display = "num")`C).
-We defined the two domains for delta CWD plot values for all species as those above a 5mm CWD change threshold, which we interpret as increasing in CWD (Increasing), and those below the threshold (Stable).
-The 5mm threshold was selected because species' delta CWD value distributions did not differ greatly and because 5mm is approximately the mean delta CWD value across species (Table S2).
-The Stable domains may include some negative delta CWD values (i.e., mean CWD was larger prior to the initial visit than between visits), however these instances were rare and of small magnitude.
+We defined the two domains for delta CWD plot values for all species as those above a 5 mm CWD change threshold, which we interpret as increasing in CWD (Increasing), and those below the threshold (Stable).
+The 5 mm threshold was applied across species because species' delta CWD value distributions did not differ greatly and because 5 mm is approximately the grand mean delta CWD value (Table S2).
+The Stable domains may include some negative delta CWD values (i.e., the mean CWD value was larger prior to the initial visit than between visits); however, these instances were rare and of small magnitude.
 The domain designations IL, IM, and IH correspond to Increasing Low, Increasing Medium, and Increasing High.
 The Stable domains are similarly labeled (SL, SM, SH).
-These six within-domain estimates allowed us to establish mortality and growth rates for trees that experienced similar climatic conditions for a given species.
-We calculated estimates for all CWD-based domains with 10 or more plots.
-
-In addition to within-domain estimates, we examined contrasts among growth (Figure `r fig_nums("fig_diffgrow", display = "num")`) and mortality (Figure `r fig_nums("fig_diffmort", display = "num")`) across CWD domains.
+These six within-domain estimates allowed us to establish mortality and growth rates for trees of a given species that experienced similar climatic conditions.
+We calculated growth and mortality estimates for all CWD-based domains with 10 or more plots.
+
+In addition to within-domain estimates, we examined growth and mortality estimate contrasts among CWD domains.
 This allows us to establish if trees between two CWD domains are experiencing meaningful differences in growth and mortality.
@@
 We obtained empirical means and 95% confidence intervals via a nonparametric bootstrap.
-For each of the 1000 bootstrap iterations we sampled FIA plots with replacement and obtained point estimates for each domain, yielding a maximum of `r n.spp * 6 ` in-domain point estimates for each iteration.
+For each of the 1000 bootstrap iterations we sampled FIA plots with replacement and obtained species point estimates for each domain, yielding a maximum of `r n.spp * 6 ` domain point estimates for each iteration.
 For each domain we calculated a mean, upper, and lower 95% CI with the 2.5% and 97.5% quantiles from the  bootstrap iterations.
@@
 Our  contrast analysis approach resulted in hundreds of comparisons.
-Given this large number of tests, we caution readers to interpret individual significant results (i.e., contrast 95% CIs that do not including zero) cautiously, as some proportion likely represent Type-I errors and occurred by chance.
-We did not apply a multiple comparisons correction as our analysis is inherently exploratory, aimed at identifying patterns within and across species rather than testing a specific *a priori* hypothesis.
-We did not want to apply corrections because we did not want to inflate the Type-II error rate and obscure real relationships in favor of reducing false positives.
+Given the large number of tests, we caution readers to interpret individual significant results (i.e., contrast 95% CIs that do not including zero) cautiously, as some proportion likely represent Type-I errors and occurred by chance.
+We did not apply a multiple comparisons correction as (1) our analysis is inherently exploratory, aimed at identifying patterns within and across species rather than testing a specific *a priori* hypothesis, and (2) because we did not want to inflate the Type-II error rate and obscure real relationships in favor of reducing false positive findings.
 
 ## Dashboard
-We developed a data dashboard (https://tinyurl.com/FIAgrowmort) to enable ourselves and readers to examine individual species domain estimates, the distribution of initial and delta CWD values, and the geographical distribution of plots and plot domains.
+We developed a data dashboard (https://tinyurl.com/FIAgrowmort) to enable ourselves and readers to examine individual species domain estimates, the distribution of initial and delta CWD values, and the fuzzed geographic locations of plots and plot domains.
 All analysis files, manuscript markdowns, dashboard code, and ancillary data sources are available at the GitHub repository https://github.com/jeremygroom/Tree_Growth_Death.
@@
 The Blue Mountains and Northern and northeastern Washington lacked negative delta CWD plots entirely.
-In California negative delta CWD plots with values < 1mm were primarily found in the northern Sierra Nevada and in the southern portions of the state.
-Overall, initial and delta CWD values were weakly and positively linearly associated (correlation = `r CWD.lin.cor`).The two variables were better modeled with a 3rd order polynomial although the R^2^ value still indicated that the variance was not well explained by the model ($\Delta AIC$ =  `r CWD.dAIC`, R^2^ = `r round(CWD.rels$lm2$r.squared, 2)`, Figure S2, see model fits in Tables S1 and S2), implying that other variables likely affect the relationship between initial and delta CWD.
+In California negative delta CWD plots with values < -1.0 mm were primarily found in the northern Sierra Nevada and in the southern portions of the state.
+Overall, initial and delta CWD values were weakly and positively linearly associated (correlation = `r CWD.lin.cor`).
+The two variables were better modeled with a 3rd order polynomial although the resulting R^2^ value still indicated that the variance was not well explained by the model ($\Delta AIC$ =  `r CWD.dAIC`, R^2^ = `r round(CWD.rels$lm2$r.squared, 2)`, Figure S2, see model fits in Tables S1 and S2), implying that other processes likely affected the relationship between initial and delta CWD.
 For most species delta CWD neither linearly related to nor monotonically increased with initial CWD values (Figure S3; see dashboard).
-For instance, the relationship between the initial and delta CWD values for *Calocedrus decurrens* may be described overall with a concave curve, with some of the most negative delta CWD values occurring in the High domain (along with some of largest delta CWD values; Figures `r fig_nums("fig_domains", display = "num")`C).
+For instance, the relationship between the initial and delta CWD values for *Calocedrus decurrens* may be described overall with a concave curve, with some of the most negative delta CWD values occurring in the High domain (along with some of largest delta CWD values; Figure `r fig_nums("fig_domains", display = "num")`C).
 
@@
 Initial CWD means and ranges differed among tree species (Table S3).
+The `r n.spp` appeared to form two different groups by initial CWd and delta CWD (Figure `r fig_nums("fig_cwdmeans", display = "num")`).
+Eight species had mean delta CWD values < 4.5 mm and mean initial CWD values < 100 mm.
 *`r init.mean.sm$spp.out`* had the smallest mean value for initial CWD (`r init.mean.sm$mean.alone` mm) while *`r init.range.sm$spp.out`* had both the largest mean value (`r init.mean.lg$mean.alone` mm) and smallest range `r init.range.sm$spp.range`.
@@
 *Quercus chrysolepis* had the least mean growth at 40.5 cm^2^/decade.
-Riparian and mesic hardwoods like *Acer macrophyllum* and *Alnus rubra* on average grew by 145 cm^2^/decade, although *Acer macrophyllum* growth varied by state; its mean growth in California was 52.3 cm^2^/decade, less than half of its growth rate in Oregon and almost four times less than its growth rate in Washington.
+Riparian and mesic hardwoods like *Acer macrophyllum* and *Alnus rubra* on average grew by 145 cm^2^/decade, although growth for *Acer macrophyllum* varied by state; its mean growth in California was 52.3 cm^2^/decade, less than half of its growth rate in Oregon and almost four times less than its growth rate in Washington.
 High-elevation conifers like *Abies concolor* and *Picea engelmannii* would both have grown at similar rates of approximately 170-190 cm^2^/decade except that the Washington rate for *Picea engelmannii* was only 125.4 cm/decade.
-The other high-elevation conifers, *Abies lasiocarpa*,  and *Tsuga mertensiana*, grew more slowly at almost 70 cm^2^/decade.
+The other high-elevation conifers, *Abies lasiocarpa*,  and *Tsuga mertensiana* grew more slowly at almost 70 cm^2^/decade.
 
@@
 Negative estimates for the “Increasing vs.
-Stable” in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")` indicate that the 10-year growth or mortality rate was greater for a species in plots below the change estimates.
+Stable” in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")` indicate that the 10-year growth or mortality rate was greater for a species in Stable domains.
 `r t.ref("tbl_summ_grow_mort")` summarizes the number of domain differences in each of the columns in Figures `r fig_nums("fig_diffgrow", display = "num")` and `r fig_nums("fig_diffmort", display = "num")` where the 95% CI does not include zero as well as the number of contrasts per figure column.
@@
 Depending on the species, growth may have been susceptible (negatively or positively) to a change in CWD and/or differed by initial CWD values.
-We consider a lower initial CWD value to indicate a more favorable condition for plant growth, and indeed *Abies amabilis*, *Abies grandis*, *Calocedrus decurrens*, *Pinus ponderosa*, *Pseudotsuga menziesii*, *Tsuga mertensiana*, *Acer macrophyllum*, and *Alnus rubra* exhibited greater growth in the Low domains than the Medium and/or High domains.
-This was exemplified by *Acer macrophyllum* and *Abies grandis* with a mean difference of `r fig.val.fcn("g", "X312", "IH - IL", 0)` cm^2^/decade, IH-IL and `r fig.val.fcn("g", "X312", "SH - SL", 0)` cm^2^/decade, SH-SL, respectively while contrasts for species like *Tsuga mertensiana* and *Quercus chrysolepis* were less (-0.20 for IM-IL and SM-SL, and `r fig.val.fcn("g", "X805", "SH - SM", 0)` for SH-SM, respectively).
+We consider a lower initial CWD value to indicate a more favorable condition for plant growth, and *Abies amabilis*, *Abies grandis*, *Calocedrus decurrens*, *Pinus ponderosa*, *Pseudotsuga menziesii*, *Tsuga mertensiana*, *Acer macrophyllum*, and *Alnus rubra* exhibited greater growth in the Low domains than the Medium and/or High domains.
+This was exemplified by *Acer macrophyllum* and *Abies grandis* with a mean difference of `r fig.val.fcn("g", "X312", "IH - IL", 0)` cm^2^/decade, IH-IL and `r fig.val.fcn("g", "X312", "SH - SL", 0)` cm^2^/decade, SH-SL, respectively while less for contrasts for species like *Tsuga mertensiana* (approximately -0.20 for IM-IL and SM-SL) and *Quercus chrysolepis* (`r fig.val.fcn("g", "X805", "SH - SM", 0)` for SH-SM).
 Except for *Alnus rubra*, these results predominantly occurred in both the Stable domain (2 spp.) or the Increasing and Stable domains (5 spp.).
-However, this finding is not uniform for all species, and we found that *Thuja plicata* had greater growth in the Increasing (`r fig.val.fcn("g", "X242", "IH - IM", 0)` cm^2^/decade for IH-IM and `r fig.val.fcn("g", "X242", "IM - IL", 0)` cm^2^/decade for IM-IL) and Stable (`r fig.val.fcn("g", "X242", "SH - SM", 0)` cm^2^/decade for SH-SM and `r fig.val.fcn("g", "X242", "SM - SL", 0)` cm^2^/decade for SM-SL) Medium domains than either High or Low domains.
+However, this finding is not uniform for all species, and we found that *Thuja plicata* had greater growth in the Medium domains than either High or Low domains (Increasing: `r fig.val.fcn("g", "X242", "IH - IM", 0)` cm^2^/decade for IH-IM and `r fig.val.fcn("g", "X242", "IM - IL", 0)` cm^2^/decade for IM-IL; Stable: `r fig.val.fcn("g", "X242", "SH - SM", 0)` cm^2^/decade for SH-SM and `r fig.val.fcn("g", "X242", "SM - SL", 0)` cm^2^/decade for SM-SL).
 *Abies concolor*, *Larix occidentalis*, *Picea engelmannii*, *Pinus contorta*, *Pinus lambertiana*, *Arbutus menziesii*, and *Quercus kelloggii* had greater growth in the more water stressed High domains than the Medium and/or Low domains, with some of the greatest differences for *Pinus lambertiana* (`r fig.val.fcn("g", "X117", "IH - IL", 0)` cm^2^/decade for IH-IL) and *Picea engelmannii* (`r fig.val.fcn("g", "X93", "SH - SL", 0)` cm^2^/decade for SH-SL).
@@
 The same species exhibited a mean mortality rate that was almost five times greater in Washington (0.34) than in Oregon (0.07).
-Drought-tolerant conifers had overall mortality rates that were relatively low (*Calocedrus decurrens* had the greatest mortality of the four species at 0.15/decade), although *Calocedrus decurrens* and *Pinus ponderosa* both exhibited higher mortality and lower growth rates in California than Oregon.
+Drought-tolerant conifers had overall mortality rates that were relatively low (*Calocedrus decurrens* had the greatest mortality of the four species at 0.15/decade).
+*Calocedrus decurrens* and *Pinus ponderosa* both exhibited higher mortality and lower growth rates in California than Oregon.
 The mean mortality rates for drought-tolerant hardwoods ranged from 0.14 to 0.20, with *Quercus chrysolepis* and *Quercus kelloggii* having the highest mortality rates (0.20 and 0.18, respectively).
@@
 A suite of species had greater mortality in the Low domains than the Medium and/or High domains including *Larix occidentalis*, *Picea engelmannii*, *Pinus contorta*, *Pinus monticola*, and *Thuja plicata*.
-Some of the larger mean mortality contrasts included *Picea engelmannii* and *Pinus contorta* (`r fig.val.fcn("m", "X93", "IH - IL", 2)` and `r fig.val.fcn("m", "X108", "IH - IL", 2)`/decade, respectively, for IH-IL) while *Thuja plicata* represented a smaller estimate (`r fig.val.fcn("m", "X242", "IH - IL", 2)`/decade for IH-IL).
+Some of the most negative mean mortality contrasts included *Picea engelmannii* and *Pinus contorta* (`r fig.val.fcn("m", "X93", "IH - IL", 2)` and `r fig.val.fcn("m", "X108", "IH - IL", 2)`/decade, respectively, for IH-IL) while the estimate for *Thuja plicata* was smaller (`r fig.val.fcn("m", "X242", "IH - IL", 2)`/decade for IH-IL).
 These results occurred in Increasing (3 spp.) or both Increasing and Stable (2 spp.) domains.
 Mortality rates for *Abies lasiocarpa*, *Calocedrus decurrens*, *Pinus lambertiana*, *Pinus ponderosa*, *Pseudotsuga menziesii*, *Tsuga heterophylla*, *Tsuga mertensiana*, *Acer macrophyllum*, and *Notolithocarpus densiflorus* were greater in the High domains than the Medium and/or Low domains.
-Larger estimates include `r fig.val.fcn("m", "X19", "SH - SL", 2)`/decade for the *Abies lasiocarpa* SH-SL contrast, and `r fig.val.fcn("m", "X117", "IH - IM", 2)`/decade for the *Pinus lambertiana* IH-IM contrast.
+The greatest positive estimates include `r fig.val.fcn("m", "X19", "SH - SL", 2)`/decade for the *Abies lasiocarpa* SH-SL contrast, and `r fig.val.fcn("m", "X117", "IH - IM", 2)`/decade for the *Pinus lambertiana* IH-IM contrast.
 *Pseudotsuga mensiesii* exhibited smaller contrast estimates of around `r fig.val.fcn("m", "X202", "SH - SM", 2)`/decade for SH-SL and SH-SM.
@@
 
-Contrast results serve to highlight differences in mortality and growth rates for different subpopulations of trees undergoing different climatic experiences.
 Regional analyses detected several species with relatively high mortality rates including *Acer macrophyllum*, *Picea engelmannii*, *Abies lasiocarpa*, *Pinus contorta*, and *Alnus rubra*.
@@
 The high mortality rate for *Acer macrophyllum* in California appears to be both unreported elsewhere and substantially greater than in Oregon or Washington.
-*Acer macrophyllum* growth was negatively associated with initial CWD values, and its growth was lower and mortality almost three times higher in California than in Washington and Oregon.
+*Acer macrophyllum* growth was negatively associated with initial CWD values (i.e., the least growth occurred in the High domains), and its growth was lower and mortality almost three times higher in California than in Washington and Oregon.
 We cannot find primary literature that corroborates an increase in mortality for this species in California.
@@
 (2021) found widespread evidence in Washington of *Acer macrophyllum* decline (crown dieback, thinning, discoloration, reduced leaf size) and reduced growth and increased mortality in locations with higher temperatures and water limitations.
-Although these results align with our contrast findings for growth, as displayed in our dashboard, they are not reflective (yet) of our state mortality findings for Washington.
+Although these results align with our contrast findings for growth, they are not reflective (yet) of our state mortality findings for Washington.
 We encourage a more intensive assessment of *Acer macrophyllum*’s status in California.
 
-Conspecifics *Abies lasiocarpa* and *Picea engelmannii* are frequently codominant and exhibit high mortality rates.
+The high-elevation conspecifics *Abies lasiocarpa* and *Picea engelmannii* are frequently codominant and exhibit high mortality rates.
 Andrus et al.
@@
 Its mortality rate in Washington was almost 5 times higher than in Oregon.
-*Abies lasiocarpa* did not have substantial mortality contrast findings (some evidence of BH - BL difference, but the MC permutation p-value was greater than 0.001) and its mortality was high but similar across all three states.
-
-Our findings for *Pinus contorta* in Washington likely reflected insect pressures on the species.
-Our Washington mortality estimate was twice that of Oregon’s and five times greater than Calfornia’s, while the growth confidence intervals for Washington were lower than California’s and did not overlap.
+*Abies lasiocarpa* did have lower mortality in the Stable Low domain than the other Stable domains which corresponds to trees found in the more mesic portion of Washington State west of the Cascade Range crest.
+
+The lower growth and higher mortality findings for *Pinus contorta* in Washington likely reflected insect pressures on the species.
 The domains with the highest mortality and lowest growth align well with Washington regions that have experienced water stress (Raymond et al.
@@
 
-The absence of consistent contrast patterns across species is itself informative, indicating that tree responses to water stress conditions are potentially mediated by local adaptation to conditions, species-specific physiological traits, and local environmental conditions rather than following a general response curve.
-The number of significant Increasing vs.
-Stable threshold contrasts for growth and mortality were approximately evenly distributed between positive and negative contrast estimates, with a positive difference indicating that some species exhibited higher growth or greater mortality in an Increasing domain (e.g., *Pinus lambertiana* for both growth and mortality) versus a Stable domain (*Abies grandis* and *Pinus contorta*, respectively, for growth and mortality).
-This lack of a pattern across species may indicate that the distinction of Increasing vs.
-Stable is not generalizable across species, or that significant findings are due to confounding factors such as geographical locations with different ecological pressures.
-
 *Abies grandis* exhibited higher mortality in Washington than Oregon or California.
@@
 
+The absence of consistent contrast patterns across species is itself informative, indicating that tree responses to water stress conditions are potentially mediated by local adaptation to conditions, species-specific physiological traits, and local environmental conditions rather than following a general response curve.
+The number of significant Increasing vs.
+Stable threshold contrasts for growth and mortality were approximately evenly distributed between positive and negative contrast estimates, with a positive difference indicating that some species exhibited higher growth or greater mortality in an Increasing domain (e.g., *Pinus lambertiana* for both growth and mortality) versus a Stable domain (*Abies grandis* and *Pinus contorta*, respectively, for growth and mortality).
+This lack of a pattern across species may indicate that the distinction of Increasing vs.
+Stable is not generalizable across species, or that significant findings are due to confounding factors such as geographical locations with different ecological pressures.
+
 Our dashboard reveals some instances of domains representing geographical mortality differences.
 State estimates for *Pinus monticola* mortality do not appear to differ.
-Mortality for the Increasing High domain was low (see dashboard; MC permutation test results only found p < 0.01 when comparing Increasing High to Increasing Low) and predominantly represented by plots in the southern Sierra Nevada mountains of California.
+Mortality for the Increasing High domain was low (see dashboard) and was predominantly represented by plots in the southern Sierra Nevada mountains of California.
 *Pinus lambertiana* and *Calocedrus decurrens* exhibited high mortality in approximately the same region, a result supported by Fettig et al.
@@
 (2023) sampled *Thuja plicata* at nine locations in Washington, northern Oregon, and Idaho that were without canopy dieback, exhibited canopy dieback, or had died.
-They found that the mortality was associated with exceptionally warm and dry climate conditions.
+They found that mortality was associated with exceptionally warm and dry climate conditions.
 A collaboration of state and federal agencies is currently reaching out to the public to identify dieback sites for this species (website: Goodrich et al.
 2025).
-Our analysis found that *Thuja plicata* exhibited the lowest decadal mortality rate and third highest growth rate of any species we examined, and relatively strong sampling precision.
+Our analysis found that *Thuja plicata* exhibited the lowest decadal mortality rate and third highest growth rate of any species we examined, and demonstrated relatively strong sampling precision.
 It also had its greatest growth in Medium domains.
@@
 We did not detect sizable changes in CWD between visits, potentially signaling that 10 years is too soon to observe CWD-related species responses (perhaps more true for mortality than growth).
-Another possibility is that the TerraClimate dataset is relatively spatially coarse (~4 km, Abatzoglou et al.
-2018), potentially dampening the range of possible results from a finer resolution model.
-The CWD changes we observed may have represented greater proportional change for some species than others, with species such as *Abies amablis*, *Alnus rubra*, and *Juniperus occidentalis* exhibiting narrow and low water stress initial CWD ranges.
-CWD effects can manifest in direct and indirect ways, potentially differing with species by geographic location and the biotic and abiotic impacts to those locations.
-It also cannot easily convey differences in impacts to trees within the same stands.
-
-The scale of our analysis and the lack of incorporation of other variables likely limited our analysis because the evidence of water stress can manifest in multiple overlapping ways.
+Another possibility is that the spatial coarseness of the TerraClimate dataset (~4 km, Abatzoglou et al.
+2018) potentially dampens the range of results we might see with a finer resolution climate model.
+The CWD changes we observed may have represented greater proportional change for some species than others, with species such as *Abies amablis*, *Alnus rubra*, and *Juniperus occidentalis* exhibiting narrow initial CWD ranges.
+
+The scale of our analysis and the small number of variables we considered (two) likely limited our analysis because the evidence of water stress can manifest in multiple overlapping ways.
 Water stress alone can kill trees (Anderegg et al.
@@
 Although our current analysis offers a coarse examination of species status across a large area, future analyses could refine our approach for assessing tree species growth and mortality by domains.
-Our climate domains could be divided further into sub-domains: trees could be assigned to tree size, stand competition, plot aspect, or plot elevation categories within the climate domain categories.
+Our climate domains could be divided further into sub-domains: trees could be assigned to categories by size, stand competition, plot aspect, or plot elevation within the climate domain categories.
 This approach has its limits: only estimates for certain widespread tree species would be reasonable to report, and domains could not be divided too finely into sub-domains due to sample size constraints, limiting the number of domain variables that could be considered simultaneously.
-However, this approach would permit design-based estimation at a resolution better attuned to capturing factors affecting growth and mortality.
+However, this approach would likely permit design-based estimation at a resolution better attuned to capturing factors affecting growth and mortality.
 
@@
 Our analysis serves as an example of design-based estimation, where we use estimation domains based on initial CWD and delta CWD values of plots.
-For the 23 species we examined, we found no general patterns in domains with higher or lower tree growth or mortality; the results appeared species specific.
+For the 23 species we examined, we found no general patterns in domains related to tree growth or mortality; the results appeared species specific.
 We believe our analysis approach can be refined for individual species analysis, with the potential for integrating a third variable that could introduce information besides a condition variable and change variable.
-Our analysis and the analyses of others provide valuable and reliable insights into the condition of several tree species that exhibit abnormally high mortality (and low growth).
+Our analysis and the analyses of others provide valuable and reliable insights into the condition of several tree species that exhibit abnormally high mortality and low growth.
 Our hope is that these findings will inform management and encourage focused monitoring efforts for these species in the rapidly changing landscape of western North America.

### 2025-12-16 Analysis Code/Data_Prep_Analysis.R

@@ -406,6 +406,7 @@ if(n_domain == 6) {
     legend.side <- switch(k, "1" = c(TRUE, TRUE, FALSE),
            "2" = c(FALSE, FALSE, TRUE))
     
+    IvS2 <- IvS2 %>% mutate(Domain = factor(Domain, levels = c("IL - SL", "IM - SM", "IH - SH")))
     
     p1 <- diff.panel.fcn(IvS2, remove.y = FALSE, fig.title = "Increasing vs. Stable", lab.right = legend.side[1], plt.mcperm = PLT.MCPERM)
     p2 <- diff.panel.fcn(I_LMH2, remove.y = TRUE, fig.title = "Increasing, High/Med/Low", lab.right = legend.side[2], plt.mcperm = PLT.MCPERM)

### 2025-12-16 Analysis Code/Manuscript_information.R

@@ -112,6 +112,7 @@ cwd.ci.plt <- ggplot(cwd.ci, aes(Mean_Init, Mean_Diff, group = Species, label = 
     nudge_x = 1,
     box.padding = 0.5,
     nudge_y = 0.05,
+    fontface = "italic", 
     # segment.curvature = -1e-10,
     #segment.ncp = 3,
     segment.angle = 20) +
